### 2021-10-29 products/manuscript/Manuscript_Burn.Rmd
Part 3 - Update

---
title: "Evaluating Machine Learning Methodologies for Predicting Thermal Burn Patient Outcomes"
subtitle: ""
author: Morgan Taylor
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

# Summary/Abstract
_Write a summary of your project._

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 
Detonation of a nuclear device over an urban area would generate an unprecedented volume of patients, with injuries resulting from the blast itself, from the thermal energy and fires, and from the radiation emitted from the device. Of all of the injury etiologies, thermal burns may be the least understood, and treatment of such injuries has the greatest resource constraints. A single patient with an extensive burn injury requires a greater commitment of supplies, personnel, and time to achieve optimal outcomes [Saffle 2005]. In particular, thermal burn patients lose heat, plasma, and fluids that, if left uncorrected, can cause hypothermia, hypovolemic shock, and renal insufficiency [Evans 1952]. Moreover, severely burned patients are particularly susceptible to infection [Church 2006], as skin is the primary line of defense in the immune response [Goldman & Schafer, 2011]. The volume of resources required to treat burn patients can be exemplified by treatment cost. Total charges average \$268,435 for surviving patients and \$354,560 for non-surviving patients [NBR 2019 report]. For comparison, the average cost per stay across all healthcare etiologies is $9,700 [Pftuntner et al. 2013]. The accumulation of these factors on a individual patient level suggests the burn care community is poorly positioned to transition to mass casualty care.

To further complicate matters, the broader systemic infrastructure for thermal burn patients is severely limited [Kearns 2020]. Specialized burn centers are designed to care for patients with burns as well as other skin and soft tissue injuries and disorders [NBR Data Dictionary]. Highly specialized teams comprised of burn surgeon(s), nurses, anesthesiologist(s), respiratory therapists, occupational therapists, physical therapists, dietitians, and psychosocial experts staff these burn centers, but there is a severe shortage of trained personnel to supplement these locations [Al-Mousawi et al 2010]. Across the entire United States (US), there are 133 burn centers, only 72 of which are verified by the American Burn Association (ABA). Similarly, there are approximately 300 burn surgeons whose teams cover approximately 2,000 specialty burn beds [Kearns 2020]. Moreover, the sparse nature of these burn centers dictate there are several locations in the US that are hundreds of miles from the closest facility [DHS Burn source]. Kearns et al. [2020] writes, "nationally, the burn care community could manage approximately 2,000 patients if time permits (up to 120 hours), and transportation resources exist to treat and redistribute the patients to centers across the nation." The two assumptions listed are unlikely to be met in a real-world scenario, thus emphasizing the limited capability of the burn healthcare infrastructure to withstand a catastrophic event, such as a nuclear device detonation.

In such circumstances, it is imperative to transition to population patient care by employing triage practices. Coming from the french word "trier," which means "to sort," a healthcare provider triages potential patients by using an established system or plan to determine a treatment priority for each patient [Iserson 2006]. Historically, triage systems are tailored to trauma etiologies and the most commonly used ones are not evidenced-based [Jenkins et al 2008]. The existing triage system for burn patients sorts patients to healthcare destination (outpatient or non-burn center hospital, burn center, and expectant care) based on age and total percent body surface area burned (%TBSA). These criteria are leveraged from prediction scores, such as the Modified Baux Score or the Abbreviated Burn Severity Index (ABSI), which are developed from logistic regressions with patient mortality as the outcome of interest [Halgas et al., 2018]. However, other healthcare disciplines have increasingly transitioned to ML models, which are automatically generated and continually improved, potentially increasing predictive accuracy [Rojas et al., 2018; Taylor et al., 2016]. Preliminary research suggests ML models using artificial neural network (ANN) models can predict burn patient mortality more accurately than commonly used prediction scores [Liu & Salinas, 2015], but there is evidence to suggest random forest models may be more advantageous in emergent applications [Taylor et al., 2016]. The purpose of this study is to examine the accuracy and applicability of ML methods in assessing thermal burn patient mortality and length of stay in burn centers.

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
The ABA created the National Burn Repository (NBR), which is a near-comprehensive record of patients treated at 101 participating burn centers in the US from the years 2009 to 2018. The data set contains 221,519 patient entries, 3,712 of which are reported from international facilities. The data is most frequently collected using the ABA Burn Care Quality Platform and other similar software [NBR Data Dictionary]. Key variables captured in this repository include: distance from patient residence to burn center (based on zip code), age, race, ethnicity, sex, patient occupation, marital status, living situation, injury date, if injury was work related, location of injury, alcohol/drug use, circumstances of injury, ICD-9 / ICD-10 E-code, injury severity, initial clinical presentation, burn center admission information, date of procedure, procedures completed, total length of stay, total ICU length of stay, total ventilator stay, hospital discharge disposition, care directives, and patient financial data.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._
The long-term goal of this analysis is to better prepare healthcare providers and emergency managers for catastrophic disasters with a significant proportion of burn patients, as would be the case in a nuclear detonation. The primary objective of this analysis is to identify a more sophisticated, machine-learning based model that more accurately predicts burn patient length of stay in hospitals and mortality. Particularly, this study has the following specific objectives:
1.	Create two RF models that respectively predict thermal burn patient length of stay in hospitals and mortality under normal care circumstances
2.	Create two ANN models that respectively predict thermal burn patient length of stay and mortality under normal care circumstances
3.	Compare the accuracy of the ML models with pre-existing, logistic regression-based prediction models

Following identification of a more appropraite outcome prediction model, future studies will apply resource limitations to the model to identify significant predicors of patient mortality and length of stay under catastrophic circumstances.

# Methods and Results

_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
Access to the ABA NBR was achieved via application to the ABA Quality and Burn Registry Committee. Due to Health Insurance Portability and Accountability Act (HIPAA) regulations, the data cannot be provided for the purposes of reproducibility.

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
The initial

Inclusion and exclusion criteria were applied based on the guidance from the ABA NBR Report [2019]. Records were excluded if "Admit Type" or "Admit Status" were readmission, admission for reconstruction/rehabilitation, outpatient encounter, same patient, scheduled/elective admission, or acute admission, not burn-related.

## Exploratory analysis

_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

Table \@ref(tab:summarytable) shows a table summarizing the data.

```{r summarytable,  echo=FALSE}
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable, caption = 'Data summary table.')
```

Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.

```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
knitr::include_graphics("../../results/resultfigure.png")
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.

```{r resulttable,  echo=FALSE}
resulttable=readRDS("../../results/resulttable.rds")
knitr::kable(resulttable, caption = 'Linear model fit table.')
```

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@Leek2015a] discusses types of analyses.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

# References

### 2021-10-29 products/manuscript/Manuscript_Fallout.Rmd
Part 3 - Update

---
title: "Examining the Fallout Radioactivity Across Europe After the Chernobyl Meltdown"
subtitle: ""
author: Morgan Taylor
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

# Summary/Abstract
_Write a summary of your project._

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 
On April 26, 1989, a power failure at the number 4 reactor at the Chernobyl Nuclear Power Plant in the Soviet Union caused the worst nuclear disaster in history, resulting in an estimated $68 billion (adjusted for inflation) of damages. Due to the unprecedented nature of the event and the delay in notification of the general public, many questions still remain regarding the concentration and distribution of fallout radiation across several countries. As the controversy surrounding the true impact of the event remains even more than thirty years later, this study aims to better understand the distribution of the different radioisotopes from a nuclear reactor meltdown.

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
A publicly available dataset identifying sampling locations from across several European countries can be accessed [here](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data).

There are nine variables included: country abbreviation, country numerical code, sampling location (city or state), latitude of sampling location, longitude of sampling location, date of sampling, and concentration of three radioisotopes (Iodine 131, Caesium 134, and Ceasium 137). There are 2051 observations included, but not all observations have complete data.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._
There are three primary questions this study will endeavor to answer:

1. Is there a significant reduction in radioisotope concentration captured within the study period? If so, do the results represent known half-life decomposition for each radioisotope type?

2. What is the geospatial dispersion of fallout radiation captured through this publicly available sampling dataset?

3. Does the geospatial dispersion reported in the dataset match commonly accepted models for nuclear reactor meltdown? (Pending approval)

# Methods and Results

_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
The raw data was obtained from [this online repository](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data), but it is unclear from where the original poster obtained this data. The data were then imported into [RStudio 4.1](https://www.rstudio.com/), and the [tidyverse](https://www.tidyverse.org/) package was used to adjust variable formats and create subsets for future analyses. In lieu of using the calendar date for the sample collection, a new variable representing the number of days since the Chernobyl meltdown was created. This decision was motivated by the fact that radioactive decay is a relative measure of elapsed time rather than a calendar date. The [geosphere](https://cran.r-project.org/web/packages/geosphere/geosphere.pdf) package was used in a different feature engineering application to determine the distance between the Chernobyl nuclear power plant and each of the sampling locations. The full R code for data import and cleaning is available in the supplementary information.

## Exploratory analysis

_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

A total 2,051 observations were included in the original dataset, but 20 were removed for having no concentration recorded for any of the radioisotopes in question.

Table \@ref(tab:isotope_table) describes the three radioisotopes captured in the data.

```{r isotopetable,  echo=FALSE}
#load data for entire document
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")
data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")

#load data
mydata <- readRDS(data_location)
mydata_long <- readRDS(data_location2)

#define row labels
table1::label(mydata$Iodine131) <- "Iodine 131 (Bq/m^3)"
table1::label(mydata$Caesium134) <- "Caesium 134 (Bq/m^3)"
table1::label(mydata$Caesium137) <- "Caesium 137 (Bq/m^3)"

#now combine into table
table1::table1(~ Iodine131 + Caesium134 + Caesium137, data = mydata)
```

Figure \@ref(fig:radioisotopesdays_50) depicts the relationship between the first 50 days after the incident and the measured concentration of each of the three radioisotopes. Radioactive decay is often used as an example of exponential decay; however, this pattern is not represented with this data. This phenomenon can be explained by the fact that fallout radiation instead follows a Poisson distribution pattern, due to the geographical dispersion of the particles.

```{r radioisotopesdays_50,  fig.cap='Radioisotope concentrations in European countries by days after Chernobyl..', echo=FALSE}
knitr::include_graphics("../../results/radioisotopesdays_50.png")
```
Fallout radioisotope concentrations were collected from 96 different locations across 96 countries.

Figure \@ref(fig:i131samplelocations) depicts the relationship between time elapsed and the measured concentration of I-131 across the different countries. Figures \@ref(fig:cs134samplelocations) and \@ref(fig:cs137samplelocations) are identical representations but for Cs-134 and Cs-137 respectively.

```{r i131locations,  fig.cap='Concentration of I-131 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/i131samplelocations.png")
```

```{r cs134locations,  fig.cap='Concentration of Cs-134 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/cs134samplelocations.png")
```

```{r cs137locations,  fig.cap='Concentration of Cs-137 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/cs137samplelocations.png")
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@Leek2015a] discusses types of analyses.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

# References

### 2021-10-29 code/analysis_code/edascript.R

@@ -1,0 +1,340 @@
+######################################
+#Exploratory Data Analysis Script
+######################################
+
+#this script loads the processed data and conducts the exploratory data analysis
+#and saves the results to the results folder
+
+######################################
+#Libraries and Options
+######################################
+
+#load needed packages
+library(here) #for data loading/saving
+library(summarytools) #for overall df summary
+library(ggplot2) #for plotting
+library(table1) #to create summary tables for summary statistics
+library(broom) #for cleaning up output from lm()
+
+#global environment options
+# formatting for script to avoid scientific notation output
+options(scipen=999)
+
+# globally set theme for ggplots
+ggplot2::theme_set(theme_classic())
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed_data","processeddata.rds")
+data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")
+
+#load data
+mydata <- readRDS(data_location)
+mydata_long <- readRDS(data_location2)
+
+######################################
+#Data Overview
+######################################
+
+# summarize data to see list of variables and types
+summarytools::dfSummary(mydata)
+
+summarytools::dfSummary(mydata_long)
+
+######################################
+# Outcome Variable Exploration
+######################################
+#start by examining each of the radioisotopes
+#units of each are Bq/m^3
+
+#using the wide format df (most complete df)
+
+#Iodine 131
+#start with base summary statistics
+summary(mydata$Iodine131)
+
+#create a density plot
+ggplot2::ggplot(data = mydata, aes(x = Iodine131)) +
+  geom_density()
+#this is expected as increased rates will be less frequent
+#more likely to be in USSR than in EU countries
+
+#create a box plot
+ggplot2::ggplot(data = mydata, aes(y = Iodine131)) +
+  geom_boxplot()
+#also to be expected for same descriptions as listed above
+#clearly does not follow a normal distribution
+
+#Caesium 134
+#start with base summary statistics
+summary(mydata$Caesium134)
+#more NA values
+#makes sense given Caesium 134 is less likely than I-131 to have extended fallout
+
+#create a density plot
+ggplot2::ggplot(data = mydata, aes(x = Caesium134)) +
+  geom_density()
+#this is expected as increased rates will be less frequent
+#more likely to be in USSR than in EU countries
+
+#create a box plot
+ggplot2::ggplot(data = mydata, aes(y = Caesium134)) +
+  geom_boxplot()
[…]

### 2021-10-29 code/analysis_code/halflifescript.R

@@ -1,0 +1,90 @@
+######################################
+#Map Creation Script
+######################################
+
+#this script loads the processed data and creates a map of sample locations
+#as well as radioisotope concentrations
+
+######################################
+#Libraries and Options
+######################################
+
+#load needed packages
+library(here) #for data loading/saving
+library(summarytools) #for overall df summary
+library(ggplot2) #for plotting
+library(table1) #to create summary tables for summary statistics
+library(radsafer) #to create theoretical half-life decay values for each isotope
+library(data.table) #to find minimums and maximums by group
+
+#global environment options
+# formatting for script to avoid scientific notation output
+options(scipen=999)
+
+# globally set theme for ggplots
+ggplot2::theme_set(theme_classic())
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed_data","processeddata.rds")
+data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")
+
+#load data
+mydata <- readRDS(data_location)
+mydata_long <- readRDS(data_location2)
+
+######################################
+#Data Overview
+######################################
+
+# summarize data to see list of variables and types
+summarytools::dfSummary(mydata)
+
+summarytools::dfSummary(mydata_long)
+
+######################################
+#Calculate Expected Decay
+######################################
+#first create separate df for each isotope from mydata
+#iodine 131
+i131 <- dplyr::filter(mydata_long, Radioisotope == "Iodine131")
+
+#caesium 134
+cs134 <- dplyr::filter(mydata_long, Radioisotope == "Caesium134")
+
+#caesium 137
+cs137 <- dplyr::filter(mydata_long, Radioisotope == "Caesium137")
+
+#use the radsafer package
+#calculate the correct activity-dependent value based on radioactive decay
+#units: percent of dose on days elapsed since meltdown
+#iodine 131
+mydata$I131.Decay <- radsafer::dk_correct(RN_select = "I-131", 
+                                   time_unit = "days", 
+                                   time_lapse = mydata$Day,
+                                   A1 = 1,
+                                   num = TRUE)
+
+#caesium 134
+mydata$Cs134.Decay <- radsafer::dk_correct(RN_select = "Cs-134", 
+                                          time_unit = "days", 
+                                          time_lapse = mydata$Day,
+                                          A1 = 1,
+                                          num = TRUE)
+
+#caesium 137
[…]

### 2021-10-29 code/analysis_code/mapscript.R

@@ -1,0 +1,65 @@
+######################################
+#Map Creation Script
+######################################
+
+#this script loads the processed data and creates a map of sample locations
+#as well as radioisotope concentrations
+
+######################################
+#Libraries and Options
+######################################
+
+#load needed packages
+library(here) #for data loading/saving
+library(summarytools) #for overall df summary
+library(ggplot2) #for plotting
+library(table1) #to create summary tables for summary statistics
+library(mapproj) #to create map
+library(grid) #to create map
+library(rworldmap) #to create world map
+
+library(ggmap)
+library(ggthemes)
+
+#global environment options
+# formatting for script to avoid scientific notation output
+options(scipen=999)
+
+# globally set theme for ggplots
+ggplot2::theme_set(theme_classic())
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed_data","processeddata.rds")
+data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")
+
+#load data
+mydata <- readRDS(data_location)
+mydata_long <- readRDS(data_location2)
+
+######################################
+#Data Overview
+######################################
+
+# summarize data to see list of variables and types
+summarytools::dfSummary(mydata)
+
+summarytools::dfSummary(mydata_long)
+
+######################################
+#Create European Map
+######################################
+#get world map
+worldmap <- getMap()
+
+#subset to the Eurasia continent
+country <- worldmap[which(worldmap$continent == "Eurasia"), ]
+
+#extract longitude and latitude border coordinates
+coordinates <- base::lapply(country, function(i){
+                    df <- data.frame(worldmap@polygons[[i]]@Polygons[[1]]@coords)
+                    df$region = as.character(worldmap$NAME[i])
+                    colnames(df) <- list("long", "lat", "region")
+                    return(df)
+})
+#stuck on error: no method for coercing this S4 class to a vector

### 2021-10-29 code/processing_code/processingscript.R

@@ -1,66 +1,146 @@
-###############################
-# processing script
-#
+######################################
+#Processing Script
+######################################
+
 #this script loads the raw data, processes and cleans it 
 #and saves it as Rds file in the processed_data folder
 
-#load needed packages. make sure they are installed.
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing
+######################################
+#Libraries and Options
+######################################
+
+#load needed packages
 library(here) #to set paths
+library(tidyverse) #for data processing
+library(geosphere) #for distance calculations
 
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","exampledata.xlsx")
+##Import data
+#define path to data
+data_location <- here::here("data","raw_data","Chernobyl_ Chemical_Radiation.csv")
 
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
+#load data
+rawdata <- utils::read.csv(data_location)
 
-#take a look at the data
-dplyr::glimpse(rawdata)
+######################################
+#Cleaning and Feature Engineering
+######################################
 
-#dataset is so small, we can print it to the screen.
-#that is often not possible.
-print(rawdata)
+#overview
+utils::str(rawdata)
 
-# looks like we have measurements for height (in centimeters) and weight (in kilogram)
+#rename variable columns
+processeddata <- rawdata
+colnames(processeddata) <- c("Country-Abbrev", "Country-Code", "Sample.Location", "Longitude", "Latitude", "Date", "Iodine131", "Caesium134", "Caesium137")
 
-# there are some problems with the data: 
-# There is an entry which says "sixty" instead of a number. 
-# Does that mean it should be a numeric 60? It somehow doesn't make
-# sense since the weight is 60kg, which can't happen for a 60cm person (a baby)
-# Since we don't know how to fix this, we need to remove the person.
-# This "sixty" entry also turned all Height entries into characters instead of numeric.
-# We need to fix that too.
-# Then there is one person with a height of 6. 
-# that could be a typo, or someone mistakenly entered their height in feet.
-# Since we unfortunately don't know, we'll have to remove this person.
-# similarly, there is a person with weight of 7000, which is impossible,
-# and one person with missing weight.
[…]

### 2021-11-10 code/processing_code/processing-declarations.Rmd
Disasters Revision
Giving up on the time series data

---
title: "Data Processing: Disaster Declaration Summaries"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the raw data and performs some cleaning as well as feature engineering. The raw data comes from [the OpenFEMA Dataset website](https://www.fema.gov/openfema-data-page/disaster-declarations-summaries-v2).

<br>

Overall, processing this dataset involves the following elements:

* Raw data summarization and overview
* Removal of irrelevant variables
* Summarization of key variables for each disaster declaration and state
* Feature engineering to create duration of incident and response
* Saving the processed dataset as an .rds

<br>

To keep track of processing changes, each iteration of processed data will be named `missions_#`.

<br>

Ultimately, this processed dataset will be joined to the Disaster Declaration Summaries processed dataset for the analysis.

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
```

---

## Load Raw Data
Load the raw data downloaded from provided hyerlink.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location_dss <- here::here("data", "raw_data", "DisasterDeclarationsSummaries.csv")

#load data
declarations_raw <- utils::read.csv(data_location_dss, na.strings = c("", "NA"))
```

---

## Raw Data Overview
Examine the summary and structure of the raw data.
```{r raw data overview}
#summary using base R
summary(declarations_raw)

#summary and structure using the skimr package
skimr::skim(declarations_raw)
```

### 2021-11-10 code/processing_code/processing-missions.Rmd
Disasters Revision
Giving up on the time series data

---
title: "Data Processing: Mission Assignments"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the raw data and performs some cleaning as well as feature engineering. The raw data comes from [the OpenFEMA Dataset website](https://www.fema.gov/openfema-data-page/mission-assignments-v1).

<br>

Overall, processing this dataset involves the following elements:

* Raw data summarization and overview
* Removal of irrelevant variables
* Summarization of mission costs for each state and disaster declaration
* Feature engineering to create cost share proportions
* Saving the processed dataset as an .rds

<br>

To keep track of processing changes, each iteration of processed data will be named `missions_#`.

<br>

Ultimately, this processed dataset will be joined to the Disaster Declaration Summaries processed dataset for the analysis.

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
```

---

## Load Raw Data
Load the raw data downloaded from provided hyerlink.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location_ma <- here::here("data", "raw_data", "MissionAssignments.csv")

#load data
missions_raw <- utils::read.csv(data_location_ma, na.strings = c("", "NA"))
```

---

## Raw Data Overview
Examine the summary and structure of the raw data.
```{r raw data overview}
#summary using base R
summary(missions_raw)

#summary and structure using the skimr package
skimr::skim(missions_raw)
```
There are 22 variables in the raw dataset and 21,472 observations. However, there are multiple observations for each disaster declaration and state, as they are listed by each request for assistance or ammendment for funding. Time to clean!

---

## Data Cleaning and Wrangling
For the purposes of this project, we only need detail at the disaster declaration number and state level. So we will need to summarize, but first let's remove the variables that are too granular for this analysis. 
```{r variable list}
#print list of variables
colnames(missions_raw)
```

---

Before we get rid of any variables, we need to confirm the date range of this dataset to be able to filter the `DisasterDeclarationsSummaries.csv` dataset.
```{r date range}
#first need to hide na values of `dateRquested` variable
dates <- na.omit(missions_raw$dateRequested)

#first date of requested assistance
min(dates)

#last date of requested assistance
max(dates)
```
So, we know this dataset ranges from February 2, 2012 to November 10, 2021.

---

Now we can remove variables. We only need to keep the following:

* stateorTribe
* disasterNumber
* agency
* stateCostSharePct
* federalCostSharePct
* requestedAmount
* obligationAmount

```{r}
#select target variables
missions_1 <- missions_raw %>%
                dplyr::select(stateorTribe, 
                              disasterNumber, 
                              agency,
                              stateCostSharePct,
                              federalCostSharePct,
                              requestedAmount, 
                              obligationAmount)
```
We now have a dataset with 7 variables and 21,472 variables

---

Before we can sum the rows, we need to address the missing values for the `stateorTribe` variable. Based on the structure of the data, we can use the `fill` function from `tidyr` to write the value based on the previous row.
```{r state name fill}
#fill in missing values for state or tribe
missions_2 <- missions_1 %>%
                tidyr::fill(stateorTribe)
```

---

Now, we can sum the rows to determine the total cost by disaster declaration number, state, and agency that provided the assistance.
```{r row sums}
#sum rows
missions_3 <- missions_2 %>%
                dplyr::group_by(disasterNumber, stateorTribe, agency) %>%
                dplyr::summarize(StateCSProp = sum(stateCostSharePct),
                                 FEMACSProp = sum(federalCostSharePct),
                                 RequestedAmt = sum(requestedAmount),
                                 ObligatedAmt = sum(obligationAmount))
```

---

The next step is to determine the requested and obligated cost share for state governments and FEMA. We will do this first by calculating the cost share percentages for state government and FEMA by agency and disaster declaration, and then we will apply the percentages to the requested and obligated funding amounts.
```{r cost share}
#create variable for state government and FEMA cost share percentages
missions_4 <- missions_3 %>%
                dplyr::mutate(StateCSPct = StateCSProp / (FEMACSProp + StateCSProp),
                              FEMACSPct = FEMACSProp / (FEMACSProp + StateCSProp)) %>%
                subset(select = -c(StateCSProp, FEMACSProp))

#apply cost share percentages to requested and obligated funding
#also create a variable that represents the total percent for total funding requested and/or obligated (e.g. 1)
missions_5 <- missions_4 %>%
                dplyr::mutate(StateReqAmt = RequestedAmt * StateCSPct,
                              FEMAReqAmt = RequestedAmt * FEMACSPct,
                              StateOblAmt = ObligatedAmt * StateCSPct,
                              FEMAOblAmt = ObligatedAmt * FEMACSPct,
                              TotalCSPct = StateCSPct + FEMACSPct)
```

---

Currently, the dataset is a mixture of wide and long formatting. To integrate effectively with the Disaster Declaration Summaries dataframe, we need to convert it completely to long formatting.
```{r long format}
#the gather function works positionally, so first reorder the columns to be in the right order
#essentially grouping columns by type of value (e.g. cost share percent, requested funding, obligated funding)
missions_6 <- missions_5[, c(1:3, 12, 6, 7, 4, 8, 9, 5, 10, 11)]

#trying piping several gather functions
missions_7 <- missions_6 %>%
                tidyr::gather(PayOrg, CostSharePct, TotalCSPct:FEMACSPct) %>%
                tidyr::gather(PayOrg2, ReqAmt, RequestedAmt:FEMAReqAmt) %>%
                tidyr::gather(PayOrg3, OblAmt, ObligatedAmt:FEMAOblAmt) %>%
                subset(select = -c(PayOrg2, PayOrg3))

#relabel values of `PayOrg`
missions_7$PayOrg[missions_7$PayOrg == "TotalCSPct"] <- "Total"
missions_7$PayOrg[missions_7$PayOrg == "StateCSPct"] <- "State"
missions_7$PayOrg[missions_7$PayOrg == "FEMACSPct"] <- "FEMA"
```

---

We now have our processed dataset for mission assignments! Let's save it into the `processed_data` folder.
```{r save processed data}
#create a mirror set for documentation purposes
missions_processed <- missions_7

#location to save processed file
missions_data_location <- here::here("data","processed_data","missionsprocessed.rds")

#save as an RDS
saveRDS(missions_processed, file = missions_data_location)
```

### 2021-11-11 code/analysis_code/analysis-eda.Rmd
Finished processing disasters, started EDA

---
title: "Analysis: Exploratory Data Analysis"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the processed data and conducts the exploratory data analysis. Follow the processing markdowns in the `processing_code` folder to generate the processed data.

<br>

This analysis examines the predictors of allocation of FEMA funds during disaster declarations. In terms of the datasets, the two outcomes of interest are Requested Amount (`ReqAmt`) and Obligated Amount (`OblAmt`).

<br>

For each variable, the following steps will be completed:

1. Produce and print some numerical output (e.g. table, summary statistics)
2. Create a histogram or density plot (continuous variables only)
3. Scatterplot, boxplot, or other similar plots against the main outcome of interest
4. Any other exploration steps that may be useful.

The steps will be numbered for each variable as specified above.

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
* summarytools: for overall dataframe summary
* ggplot2: for plotting data
* car: for creating QQ plots
* gt: for creating tables for summary statistics / other numerical outputs
* scales: for percent calculation

```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
library(summarytools) #for overall dataframe summary
library(ggplot2) #for plotting data
library(car) #for creating QQ plots
library(gt) #for creating tables
library(scales) #for percent calculation

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)

# globally set theme for ggplots
ggplot2::theme_set(theme_classic())
```

---

## Load Processed Data
Load the data generated from the markdowns in the `processing_code` folder.
```{r load data}
#define file path
data_location <- here::here("data", "processed_data", "combinedprocessed.rds")

#load data
EDAdata <- readRDS(data_location)
```

---

## Data Overview
To better understand the data, let's use `summarytools` to better visualize the data.
```{r data overview}
summarytools::dfSummary(EDAdata)
```

---

## Outcome of Interest: Requested FEMA Funds
The first of two primary outcomes of interest is the total amount of funds the state requested from FEMA for a disaster declaration.
```{r}
#summary statistics
base::summary(EDAdata$ReqAmt)

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = ReqAmt)) +
  geom_density()

#boxplot
ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
  geom_boxplot()

#crop to $1M
ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
  geom_boxplot() +
  ylim(0, 1000000)

##QQ plot
car::qqPlot(EDAdata$ReqAmt)
```

## Outcome of Interest: Obligated FEMA Funds

### 2021-11-11 code/processing_code/processing-combine.Rmd
Finished processing disasters, started EDA

---
title: "Data Processing: Creating the Combined Dataframe"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the processed data from disaster declarations and mission assignments and combines them into one dataframe for analysis.

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
```

---

## Load Data
Load the processed data from the `processed_data` folder.
```{r load data}
#define path to processed population data
data_location1 <- here::here("data", "processed_data", "declarationsprocessed.rds")
data_location2 <- here::here("data", "processed_data", "missionsprocessed.rds")
data_location3 <- here::here("data", "raw_data", "DisasterDeclarationsSummaries.csv")

#load data
declarations <- readRDS(data_location1)
missions <- readRDS(data_location2)
raw_decs <- utils::read.csv(data_location3, na.strings = c("", "NA"))
```

---

## Individual Dataset Overview
Examine the summary and structure of the individual datasets.
```{r dataset overview}
#summary using base R
summary(declarations)
summary(missions)
```

---

## Joining the Datasets
The two datasets could be joined by state or declaration number. 
```{r}
# #first change name of states column in missions df
# names(missions)[names(missions) == 'stateorTribe'] <- 'state'
# 
# #create df with states and declaration numbers from each df
# missions_nums <- as.data.frame(table(missions$state, missions$disasterNumber))
# declarations_nums <- as.data.frame(table(declarations$state, declarations$disasterNumber))

#we want a full join to keep all of the data
combined <- dplyr::full_join(declarations, missions, by = 'disasterNumber')
```

---

It's not a flawless join, so let's do some investigation to see where there might be gaps.
```{r}
summary(combined)

#filter the rows that aren't in declarations data but are in missions data
missing_nums <- combined %>% 
                  filter(is.na(state)) %>%
                  select(disasterNumber)

#unique list of missing declarations
missing_declarations <- unique(missing_nums)

#check missing declarations in raw declarations
decs_check <- raw_decs %>%
                dplyr::filter(disasterNumber %in% missing_declarations$disasterNumber)

#the missing values represent declarations in territories, not states
#for now, remove - we can always come back and adjust as needed
combined_processed <- combined %>%
                        dplyr::filter(!is.na(IncidentYear))
```

---

## Save Processed Data

Let's save it into the `processed_data` folder.
```{r save processed data}

#location to save processed file
combined_data_location <- here::here("data","processed_data","combinedprocessed.rds")

#save as an RDS
saveRDS(combined_processed, file = combined_data_location)
```

### 2021-11-11 code/processing_code/processing-declarations.Rmd
Finished processing disasters, started EDA

@@
 * Removal of irrelevant variables
+* Feature engineering to create duration of incident and response
 * Summarization of key variables for each disaster declaration and state
-* Feature engineering to create duration of incident and response
 * Saving the processed dataset as an .rds
@@
 
-To keep track of processing changes, each iteration of processed data will be named `missions_#`.
+To keep track of processing changes, each iteration of processed data will be named `declarations_#`. The final processed dataset will be `declarations_processed`.
 
@@
 
-Ultimately, this processed dataset will be joined to the Disaster Declaration Summaries processed dataset for the analysis.
+Ultimately, this processed dataset will be joined to the Mission Assignments processed dataset for the analysis.
 
@@
 * skimr: for data summarization
+* lubridate: for date wrangling
 ```{r libraries, include = FALSE}
@@
 library(skimr) #for data summarization
+library(lubridate) #for date wrangling
 ```
@@
 ```
+There are 23 variables and 62,589 observations.
+However, there are multiple lines per county that receives a disaster declaration, and the dataset includes every federally declared disaster since the 1950s.
+So there's plenty of cleaning to do!
+
+---
+
+## Data Cleaning and Wrangling
+For the purposes of this project, we only need detail at the disaster declaration number and state level.
+So we will need to summarize, but first let's remove the variables that are too granular for this analysis.
+```{r variable list}
+#print list of variables
+colnames(declarations_raw)
+```
+
+---
+
+The `MissionAssignments.csv` data captures data from requested federal assistance from February 2, 2012 to November 11, 2021 (see `processing-missions.Rmd` for more detail).
+We need to filter the disaster declaration data to capture the same range, but we can do this by disaster declaration number.
+```{r time filter}
+#define path to processed missions data
+data_location_miss <- here::here("data", "processed_data", "missionsprocessed.rds")
+
+#load data
+missions_processed <- readRDS(data_location_miss)
+
+#filter based on disaster number in missions dataset
+declarations_1 <- declarations_raw %>%
+                dplyr::filter(disasterNumber %in% missions_processed$disasterNumber)
+```
+Luckily, the first disaster represented in this subset matches the first disaster captured in the mission assignments dataset.
+Since the data is real-time updated on date of download (November 10, 2021), we don't have to worry about lining up the end points of these datasets.
+We can now move forward with the cleaning.
+
+---
+
+To make the dataset a little easier to manage, let's get rid of the variables we know are junk or aren't relevant to this analysis. The following variables can be removed now:
+
+* femaDeclarationString
+* fipsStateCode
+* fipsCountyCode
+* placeCode
+* designatedArea
+* declarationRequestNumber
+* hash
+* lastRefresh
+* id
+```{r variable removal 1}
+#remove variables
+declarations_2 <- declarations_1 %>%
+                    dplyr::select(-c(femaDeclarationString,
+                                     fipsStateCode,
+                                     fipsCountyCode,
+                                     placeCode,
+                                     designatedArea,
+                                     declarationRequestNumber,
+                                     hash,
+                                     lastRefresh,
+                                     id))
+```
+We now have a dataset with 15 variables and 20,961 observations.
+
+---
+
+The next step is to conduct some feature engineering to create variables that represent the time elapsed between `incidentBeginDate` and `incidentEndDate` as well as `incidentBeginDate` and `disasterCloseoutDate`.
+These correspond to incident duration and response duration, respectively.
+```{r incident and response days}
+#use `difftime` function to find the number of days 
+#response days will be absurdly long, but keeping the same to maintain units
+declarations_3 <- declarations_2 %>%
+                    dplyr::mutate(IncidentDuration = difftime(incidentEndDate,
+                                                              incidentBeginDate,
+                                                              units = c("days")),
+                                  ResponseDays = difftime(disasterCloseoutDate,
+                                                          incidentBeginDate,
+                                                          units = c("days")))
+```
+
+---
+
+The last date-related item is to create a variable that represents the year and another that represents the month of the disaster.
+Month will be particularly relevant for seasonal patterns (e.g.
+hurricane season), but year is also important to consider given administration changes and climate change.
+```{r month and year variable}
+#create year and month variable
+#drop all original time variables too
+declarations_4 <- declarations_3 %>%
+                    dplyr::mutate(IncidentYear = lubridate::year(incidentBeginDate),
+                                  IncidentMonth = lubridate::month(incidentBeginDate)) %>%
+                    subset(select = -c(incidentEndDate, 
+                                       incidentBeginDate,
+                                       disasterCloseoutDate,
+                                       declarationDate,
+                                       fyDeclared))
+```
+
+---
+
+We now have a dataset with 13 variables and 20, 961 observations.
+However, now that we've removed the county-level data, we need to combine rows to represent disaster declarations per state.
+The program declarations vary at a county-level, so the first step is to sum the rows for the program declaration variables.
+We don't need to worry about IA, as it was replaced by IH before our study period.
+```{r program sums}
+#group by variables not being summed (equivalent of finding the duplicates)
+#then sum the 0/1 values for the programs
+declarations_5 <- declarations_4 %>%
+                    dplyr::group_by(disasterNumber, 
+                                    state, 
+                                    declarationType,
+                                    incidentType,
+                                    declarationTitle,
+                                    IncidentDuration,
+                                    ResponseDays,
+                                    IncidentYear,
+                                    IncidentMonth) %>%
+                    dplyr::summarize(IH = sum(ihProgramDeclared),
+                                     PA = sum(paProgramDeclared),
+                                     HM = sum(hmProgramDeclared))
+```
+
+---
+
+Now we have a curious dilemma.
+The sums of each program represent the number of counties that had enough damage to receive federal assistance.
+However, not every state has the same number of counties, so the current variables aren't meaningful.
+There are two options: have a variable that reflects at least one program declaration anywhere in the state or have a variable that reflects the percent of counties receiving a program declaration.
+For now, let's do both.
+First, we'll start by loading and joining the processed population dataset.
+```{r join population data}
+#define path to processed population data
+data_location_pop <- here::here("data", "processed_data", "popsprocessed.rds")
+
+#load data
+pops <- readRDS(data_location_pop)
+
+#rename columns
+names(pops) <- c("state", "Population", "Counties")
+
+#combine with most recent declarations dataframe
+declarations_6 <- merge(declarations_5, pops, by="state")
+```
+
+---
+
+Next, we can create the two sets of programs variables.
+```{r programs}
+#create percents (Really proportions) of program declarations per number of counties in each state for each disaster
+declarations_7 <- declarations_6 %>%
+                    dplyr::mutate(IH_pct = IH / Counties,
+                                  PA_pct = PA / Counties,
+                                  HM_pct = HM / Counties)
+
+#recode existing program values to be 0 or 1
+declarations_8 <- declarations_7 %>%
+                    dplyr::mutate(IH = ifelse(IH > 0, 1, 0),
+                                  PA = ifelse(PA > 0, 1, 0),
+                                  HM = ifelse(HM > 0, 1, 0))
+
+#reorganize columns to clean things up a bit
+declarations_processed <- declarations_8[, c(1, 2, 8, 9, 3:7, 10, 15, 11, 16, 12, 17, 13, 14)]
+```
+
+---
+
+## Processed Data Overview
+We have our processed dataset! Let's take a look at the overview to see how it changed.
+```{r processed data overview}
+#summary using base R
+summary(declarations_processed)
+
+#summary and structure using the skimr package
+skimr::skim(declarations_processed)
+```
+
+---
+
+## Save Processed Data
+
+Time to save it to the `processed_data` folder and move onto the next step!
+```{r save processed data}
+
+#location to save processed file
+declarations_data_location <- here::here("data","processed_data","declarationsprocessed.rds")
+
+#save as an RDS
+saveRDS(declarations_processed, file = declarations_data_location)
+```

### 2021-11-11 code/processing_code/processing-missions.Rmd
Finished processing disasters, started EDA

@@
 
-To keep track of processing changes, each iteration of processed data will be named `missions_#`.
+To keep track of processing changes, each iteration of processed data will be named `missions_#`. The final processed dataset will be `missions_processed`.
 
@@
 There are 22 variables in the raw dataset and 21,472 observations.
-However, there are multiple observations for each disaster declaration and state, as they are listed by each request for assistance or ammendment for funding.
+However, there are multiple observations for each disaster declaration and state, as they are listed by each request for assistance or amendment for funding.
 Time to clean!
@@
 
-We now have our processed dataset for mission assignments! Let's save it into the `processed_data` folder.
+The last step is to make the formatting of state names consistent.
+```{r state names}
+unique(missions_7$stateorTribe)
+
+#there's probably a more efficient way to do this, but fix all of the non-two letter abbreviations
+missions_7$stateorTribe[missions_7$stateorTribe == "Virgin Islands"] <- "VI"
+missions_7$stateorTribe[missions_7$stateorTribe == "Seminole Tribe of Florida"] <- "FL"
+missions_7$stateorTribe[missions_7$stateorTribe == "Ponca Tribe of Nebraska"] <- "NE"
+missions_7$stateorTribe[missions_7$stateorTribe == "MUSCOGEE (CREEK) NATION"] <- "OK"
+missions_7$stateorTribe[missions_7$stateorTribe == "Standing Rock Sioux Tribe"] <- "ND"
+missions_7$stateorTribe[missions_7$stateorTribe == "Santa Clara Pueblo"] <- "NM"
+missions_7$stateorTribe[missions_7$stateorTribe == "Pueblo of Acoma"] <- "NM"
+missions_7$stateorTribe[missions_7$stateorTribe == "District of Columbia (DC)"] <- "DC"
+missions_7$stateorTribe[missions_7$stateorTribe == "Puerto Rico"] <- "PR"
+missions_7$stateorTribe[missions_7$stateorTribe == "American Samoa"] <- "AS"
+missions_7$stateorTribe[missions_7$stateorTribe == "Northern Mariana Islands"] <- "MP"
+missions_7$stateorTribe[missions_7$stateorTribe == "Choctaw Nation"] <- "OK"
+missions_7$stateorTribe[missions_7$stateorTribe == "POTAWATOMI NATION"] <- "OK"
+missions_7$stateorTribe[missions_7$stateorTribe == "Oglala Sioux Tribe of the Pine Ridge Reservation"] <- "SD"
+```
+
+---
+
+After integrating the data, I have realized I really only care about the FEMA funds. So, just subset `PayOrg` to represent FEMA and then drop the `PayOrg` variable.
+```{r FEMA filter}
+missions_8 <- missions_7 %>%
+                dplyr::filter(PayOrg == "FEMA") %>%
+                subset(select = -c(PayOrg))
+
+```
+
+---
+
+## Processed Data Overview
+We have our processed dataset! Let's take a look at the overview to see how it changed.
+```{r processed data overview}
+#create a mirror set for documentation purposes
+missions_processed <- missions_8
+
+#summary using base R
+summary(missions_processed)
+
+#not using skimr because it does not do well with long format data
+```
+
+---
+
+## Save Processed Data
+
+Let's save it into the `processed_data` folder.
 ```{r save processed data}
-#create a mirror set for documentation purposes
-missions_processed <- missions_7
 

### 2021-11-11 code/processing_code/processing-populations.Rmd
Finished processing disasters, started EDA

---
title: "Data Processing: Population Estimates"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the raw data and performs some cleaning as well as feature engineering. The raw data comes from  [the U.S> Department of Agriculture Economic Research Service](https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/).

<br>

Overall, processing this dataset involves the following elements:

* Raw data summarization and overview
* Removal of irrelevant variables
* Feature engineering to determine number of counties per state
* Saving the processed dataset as an .rds

<br>

To keep track of processing changes, each iteration of processed data will be named `pops_#`. The final processed dataset will be `populations_processed`.

<br>

Ultimately, this processed dataset will be joined to the Disaster Declaration Summaries processed dataset for the analysis.

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
library(lubridate) #for date wrangling
```

---

## Load Raw Data
Load the raw data downloaded from provided hyerlink.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location_pops <- here::here("data", "raw_data", "PopulationEstimates.csv")

#load data
populations_raw <- utils::read.csv(data_location_pops, na.strings = c("", "NA"))
```

---

## Raw Data Overview
Examine the summary and structure of the raw data.
```{r raw data overview}
#summary using base R
summary(populations_raw)

#summary and structure using the skimr package
skimr::skim(populations_raw)
```
There are 14 variables and 3,283 observations. The data is structured in such a way that it includes totals for the states as well as for each county - time to clean!

---

## Data Cleaning and Wrangling
First, let's remove the unimportant variables.
```{r variable list}
#print list of variables
colnames(populations_raw)
```

We only want to keep the following variables:

* FIPStxt
* State
* Area.name
* Population.2020
```{r variable selection}
#select key variables
pops_1 <- populations_raw %>%
                dplyr::select(FIPStxt,
                              State,
                              Area.name,
                              Population.2020)
```
We will eventually get rid of FIPStxt, but we need it to separate state and county level data.

---

To create the state-level subset, filter the rows based on values of FIPStxt.
```{r state filter}
#create a vector of FIPStxt values
state_fipstxt  <- c("01000",
                    "02000",
                    "03000",
                    "04000",
                    "05000",
                    "06000",
                    "07000",
                    "08000",
                    "09000",
                    "10000",
                    "11000",
                    "12000",
                    "13000",
                    "14000",
                    "15000",
                    "16000",
                    "17000",
                    "18000",
                    "19000",
                    "20000",
                    "21000",
                    "22000",
                    "23000",
                    "25000",
                    "26000",
                    "27000",
                    "28000",
                    "29000",
                    "30000",
                    "31000",
                    "32000",
                    "33000",
                    "34000",
                    "35000",
                    "36000",
                    "37000",
                    "38000",
                    "39000",
                    "40000",
                    "41000",
                    "42000",
                    "43000",
                    "44000",
                    "45000",
                    "46000",
                    "47000",
                    "48000",
                    "49000",
                    "50000",
                    "51000",
                    "52000",
                    "53000",
                    "54000",
                    "55000",
                    "56000",
                    "60000",
                    "66000",
                    "69000",
                    "72000",
                    "78000")

#subset based on which ones are in this vector
#then drop FIPStxt and Area.name
pops_states <- pops_1 %>%
                dplyr::filter(FIPStxt %in% state_fipstxt) %>%
                subset(select = -c(FIPStxt,
                                   Area.name))
```
Now we have a dataframe with state abbreviations and population estimates.

---

The last step is to count the number of counties per state and add it to the state dataframe.
```{r counties count}
#subset pops_1 data that doesn't include states
#remove first row too since it's the entire US
pops_2 <- pops_1 %>%
            dplyr::filter(! FIPStxt %in% state_fipstxt) %>%
            dplyr::slice(-1)

#count the number of occurrences of each state (which represents the number of counties)
pops_3 <- as.data.frame(table(pops_2$State))

#relabel columns
names(pops_3) <- c("State", "Counties")

#combine with states dataframe
pops_processed <- merge(pops_states, pops_3, by="State")
```

---

## Processed Data Overview
We have our processed dataset! Let's take a look at the overview to see how it changed.
```{r processed data overview}
#summary using base R
summary(pops_processed)

#summary and structure using the skimr package
skimr::skim(pops_processed)
```

---

## Save Processed Data

We now have our processed dataset for state populations and counties! Let's save it into the `processed_data` folder.
```{r save processed data}

#location to save processed file
pops_data_location <- here::here("data","processed_data","popsprocessed.rds")

#save as an RDS
saveRDS(pops_processed, file = pops_data_location)
```

### 2021-11-11 code/analysis_code/analysis-eda.Rmd
EDA complete, linear analysis started

@@
 * car: for creating QQ plots
-* gt: for creating tables for summary statistics / other numerical outputs
-* scales: for percent calculation
+* gtsummary: for creating tables for summary statistics / other numerical outputs
+* scales: for percent calculation and axis formatting
+* cowplot: for placing plots side by side
+* table1: for making summary tables
 
@@
 library(car) #for creating QQ plots
-library(gt) #for creating tables
+library(gtsummary) #for creating tables
 library(scales) #for percent calculation
+library(cowplot) #for placing plots side by side
+library(table1) #for making summary tables
 
@@
 ```
+Looking at the output of `summarytools::dfsummary()`
+
+* There are 445 distinct disaster numbers across 2012 - 2021. 
+* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297.
+* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486.
+* `ResponseDays` are missing 72% of the time, which suggests it should not be included in the analysis.
+* There are 16 possible incident types
+* There is a clear effect of both incident month and incident year.
+
+Let's dive a little deeper into each variable.
 
@@
 ## Outcome of Interest: Requested FEMA Funds
-The first of two primary outcomes of interest is the total amount of funds the state requested from FEMA for a disaster declaration.
+The first of two primary outcomes of interest is the total amount of funds the state requested from the federal government for a disaster declaration.
 ```{r}
@@
 
-#boxplot
+#try a log transformation out of curiosity 
+#this does ignore all negative values
+ggplot2::ggplot(data = EDAdata, aes(x = log(ReqAmt))) +
+  geom_density()
+
+#normal boxplot
 ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
@@
 
-#crop to $1M
+#crop to $1M to get a better idea
 ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
@@
 
+#log transformed boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = log(ReqAmt))) +
+  geom_boxplot()
+
 ##QQ plot
@@
 ```
+These outputs show the requested amounts do not follow a normal distribution, which is to be expected given the broad range of disasters.
+We may need to remove outliers in the analysis, but for now, we'll leave them to capture the true nature of disaster declarations.
+The largest outlier is likely the COVID-19 Pandemic.
+The log transformation does a better job of showing the distribution of the data, so we may need to consider an exponential model in the analysis; however, log(negative number) is undefined, so we would be ignoring all reimbursement data.
+The transformed and untransformed figures suggest there are significant tails in the data.
+
+---
 
 ## Outcome of Interest: Obligated FEMA Funds
+The alternative outcome of interest is the funding obligated from FEMA.
+While this may be the same as `ReqAmt` (especially for large-scale disasters), FEMA doesn't always pay the amount the states request.
+```{r obligated funds}
+#summary statistics
+base::summary(EDAdata$OblAmt)
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = OblAmt)) +
+  geom_density()
+
+#try a log transformation out of curiosity 
+ggplot2::ggplot(data = EDAdata, aes(x = log(OblAmt))) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
+  geom_boxplot()
+
+#crop to $1M to get a better idea
+ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
+  geom_boxplot() +
+  ylim(0, 1000000)
+
+#log transformed boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = log(OblAmt))) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$OblAmt)
+```
+It appears that obligated funds and requested funds have similar distributions and characteristics. The same commentary about the requested funds analysis applies here as well.
+
+---
+
+## Predictor: State
+```{r}
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$state, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#filter top 5 states
+#start with obligated amount
+obl_fig_state <- EDAdata %>%
+                    dplyr::add_count(state) %>%
+                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
+                    ggplot2::ggplot(aes(x = state, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      labs(x = "State or Territory",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated to five most disaster-prone states")
+obl_fig_state
+
+#save file
+obl_state_fig_file = here("results","obl-state.png")
+ggsave(filename = obl_state_fig_file, plot = obl_fig_state)
+
+#repeat for requested funds
+req_fig_state <- EDAdata %>%
+                    dplyr::add_count(state) %>%
+                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
+                    ggplot2::ggplot(aes(x = state, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      labs(x = "State or Territory",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested by five most disaster-prone states")
+req_fig_state
+
+#save file
+req_state_fig_file = here("results","req-state.png")
+ggsave(filename = req_state_fig_file, plot = req_fig_state)
+
+#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
+decs_state_5 <- EDAdata %>%
+                    dplyr::add_count(state) %>%
+                    dplyr::filter(dense_rank(-n) < 6) %>%
+                    gtsummary::select(state, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = state,
+                      label = list(state ~ "State or Territory",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
+                    gtsummary::modify_caption("**Table 1. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_state_5
+
+#save file
+decs_state_file = here("results", "decs-state-5.Rds")
+saveRDS(decs_state_5, file = decs_state_file)
+```
+
+---
+
+## Predictor: Incident Year
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$IncidentYear)
+
+#examine graphical relationship with outcomes
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = IncidentYear)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = IncidentYear)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$IncidentYear)
+
+#histogram distribution
+year_hist <- EDAdata %>% 
+               ggplot2::ggplot(aes(x=IncidentYear)) + 
+                        geom_bar(fill="steelblue4") + 
+                        scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
+                        scale_y_continuous(expand = c(0, 0), limits = c(0, 125)) +
+                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
+                        labs(x = "Incident Year",
+                             y = "Number of Declarations",
+                             title = "Disaster Declarations per Year (2012 - 2021)")
+year_hist
+
+#save file
+year_fig_file = here("results","incidentyear.png")
+ggsave(filename = year_fig_file, plot = year_hist)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident year
+req_fig_year <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentYear)) +
+                      geom_bar(aes(y = ReqAmt),
+                               stat = "identity",
+                               fill = "tomato") +
+                      scale_y_continuous(expand = c(0, 0), 
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
+                      labs(x = "Incident Year",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_year
+
+#repeat for obligated funds
+obl_fig_year <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentYear)) +
+                      geom_bar(aes(y = OblAmt),
+                               stat = "identity",
+                               fill = "turquoise4") +
+                      scale_y_continuous(expand = c(0, 0),
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
+                      labs(x = "Incident Year",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_year
+
+#combine into one plot
+plot_col <- cowplot::plot_grid(req_fig_year, 
+                                     obl_fig_year,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col
+
+title1 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds per year",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_year <- cowplot::plot_grid(
+                              title1, plot_col,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_year
+
+#save file
+funds_year_fig_file = here("results","funds-years.png")
+ggsave(filename = funds_year_fig_file, plot = funds_fig_year)
+
+#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
+decs_year_3 <- EDAdata %>%
+                    dplyr::add_count(IncidentYear) %>%
+                    dplyr::filter(dense_rank(-n) < 4) %>%
+                    gtsummary::select(IncidentYear, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = IncidentYear,
+                      label = list(IncidentYear ~ "Year",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Year**") %>%
+                    gtsummary::modify_caption("**Table 2. FEMA Funding for the 3 Costliest Years (2012 - 2021)**") %>%
+                    gtsummary::bold_labels()
+decs_year_3
+
+#save file
+decs_year_file = here("results", "decs-year-3.Rds")
+saveRDS(decs_year_3, file = decs_year_file)
+```
+
+---
+
+## Incident Month
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$IncidentMonth)
+
+#examine graphical relationship with outcomes
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = IncidentMonth)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = IncidentMonth)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$IncidentMonth)
+
+#histogram distribution
+month_hist <- EDAdata %>% 
+               ggplot2::ggplot(aes(x=IncidentMonth)) + 
+                        geom_bar(fill="darkseagreen4") + 
+                        scale_x_continuous(breaks = seq(1, 12, by = 1),
+                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
+                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
+                        scale_y_continuous(expand = c(0, 0), limits = c(0, 110)) +
+                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
+                        labs(x = "Declaration Month",
+                             y = "Number of Declarations",
+                             title = "Disaster Declarations per Month (2012 - 2021)")
+month_hist
+
+#save file
+month_fig_file = here("results","incidentmonth.png")
+ggsave(filename = month_fig_file, plot = month_hist)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident month
+req_fig_month <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentMonth)) +
+                      geom_bar(aes(y = ReqAmt),
+                               stat = "identity",
+                               fill = "tomato") +
+                      scale_y_continuous(expand = c(0, 0), 
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(breaks = seq(1, 12, by = 1),
+                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
+                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
+                      labs(x = "Declaration Month",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_month
+
+#repeat for obligated funds
+obl_fig_month <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentMonth)) +
+                      geom_bar(aes(y = OblAmt),
+                               stat = "identity",
+                               fill = "turquoise4") +
+                      scale_y_continuous(expand = c(0, 0),
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(breaks = seq(1, 12, by = 1),
+                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
+                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
+                      labs(x = "Declaration Month",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_month
+
+#combine into one plot
+plot_col1 <- cowplot::plot_grid(req_fig_month, 
+                                     obl_fig_month,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col1
+
+title2 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by declaration month",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_month <- cowplot::plot_grid(
+                              title2, plot_col1,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_month
+
+#save file
+funds_month_fig_file = here("results","funds-month.png")
+ggsave(filename = funds_month_fig_file, plot = funds_fig_month)
+
+#create a table where columns represent requested and obligated funds and rows are 3 most distaster prone-months
+decs_month_3 <- EDAdata %>%
+                    dplyr::add_count(IncidentMonth) %>%
+                    dplyr::filter(dense_rank(-n) < 4) %>%
+                    gtsummary::select(IncidentMonth, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = IncidentMonth,
+                      label = list(IncidentMonth ~ "Month",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>%
+                    gtsummary::modify_header(update = list(
+                                               stat_1 ~ "**Jan**",
+                                               stat_2 ~ "**Sept**",
+                                               stat_3 ~ "**Oct**"
+                                             )) %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Month**") %>%
+                    gtsummary::modify_caption("**Table 3. FEMA Funding for the 3 Costliest Months (2012 - 2021)**") %>%
+                    gtsummary::bold_labels()
+decs_month_3
+
+#save file
+decs_month_file = here("results", "decs-month-3.Rds")
+saveRDS(decs_month_3, file = decs_month_file)
+```
+
+---
+
+## Predictor: Declaration Type
+```{r}
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$declarationType, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#start with obligated amount
+obl_fig_DT <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = declarationType, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
+                      labs(x = "Declaration Type",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated for declaration types")
+obl_fig_DT
+
+#save file
+obl_DT_fig_file = here("results","obl-DT.png")
+ggsave(filename = obl_DT_fig_file, plot = obl_fig_DT)
+
+#repeat for requested funds
+req_fig_DT <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = declarationType, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
+                      labs(x = "Declaration Type",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested for declaration types")
+req_fig_DT
+
+#save file
+req_DT_fig_file = here("results","req-DT.png")
+ggsave(filename = req_DT_fig_file, plot = req_fig_DT)
+
+#create a table where rows represent requested and obligated funds and columns are declaration type
+decs_dtype <- EDAdata %>%
+                    gtsummary::select(declarationType, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = declarationType,
+                      label = list(declarationType ~ "Type",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_header(update = list(
+                                               stat_1 ~ "**Major Disaster**",
+                                               stat_2 ~ "**Emergency**")) %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Type**") %>%
+                    gtsummary::modify_caption("**Table 4. FEMA Funding for Declaration Types 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_dtype
+
+#save file
+decs_dtype_file = here("results", "decs-DT.Rds")
+saveRDS(decs_dtype, file = decs_dtype_file)
+```
+
+---
+
+## Predictor: Incident Type
+```{r}
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$incidentType, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#start with obligated amount
+obl_fig_IT <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = incidentType, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      labs(x = "Incident Type",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated for incident types") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+obl_fig_IT
+
+#save file
+obl_IT_fig_file = here("results","obl-IT.png")
+ggsave(filename = obl_IT_fig_file, plot = obl_fig_IT)
+
+#repeat for requested funds
+req_fig_IT <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = incidentType, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) + 
+                      labs(x = "Incident Type",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested for declaration types") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+req_fig_IT
+
+#save file
+req_IT_fig_file = here("results","req-IT.png")
+ggsave(filename = req_IT_fig_file, plot = req_fig_IT)
+
+#create a table where rows represent requested and obligated funds and columns are declaration type
+decs_itype3 <- EDAdata %>%
+                    dplyr::add_count(incidentType) %>%
+                    dplyr::filter(dense_rank(-n) < 4) %>%
+                    gtsummary::select(incidentType, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = incidentType,
+                      label = list(incidentType ~ "Type",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Type**") %>%
+                    gtsummary::modify_caption("**Table 4. FEMA Funding for Declaration Types 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_itype3
+
+#save file
+decs_itype3_file = here("results", "decs-IT-3.Rds")
+saveRDS(decs_itype3, file = decs_itype3_file)
+```
+
+---
+
+## Predictor: Incident Duration
+```{r}
+#convert class to numeric
+EDAdata$IncidentDuration <- as.numeric(EDAdata$IncidentDuration, units="days")
+
+#summary statistics since it's continuous
+base::summary(EDAdata$IncidentDuration)
+#NAs are ongoing events
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = IncidentDuration)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = IncidentDuration)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$IncidentDuration)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident month
+req_fig_dur <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentDuration)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Incident Duration (Days)",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_dur
+
+#repeat for obligated funds
+obl_fig_dur <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IncidentDuration)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Incident Duration (Days)",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_dur
+
+#combine into one plot
+plot_col2 <- cowplot::plot_grid(req_fig_dur, 
+                                     obl_fig_dur,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col2
+
+title3 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by Incident Duration (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_dur <- cowplot::plot_grid(
+                              title3, plot_col2,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_dur
+
+#save file
+funds_dur_fig_file = here("results","funds-dur.png")
+ggsave(filename = funds_dur_fig_file, plot = funds_fig_dur)
+```
+
+---
+
+## Predictor: Population
+```{r}
+#convert class to numeric
+EDAdata$Population <- as.numeric(gsub(",", "", EDAdata$Population))
+
+#summary statistics since it's continuous
+base::summary(EDAdata$Population)
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = Population)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = Population)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$Population)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per pop
+req_fig_pop <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = Population)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(label = comma,
+                                         limits = c(0, 40000000)) +
+                      labs(x = "State Population",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_pop
+
+#repeat for obligated funds
+obl_fig_pop <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = Population)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      scale_x_continuous(label = comma,
+                                         limits = c(0, 40000000)) +
+                      labs(x = "State Population",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_pop
+
+#combine into one plot
+plot_col3 <- cowplot::plot_grid(req_fig_pop, 
+                                     obl_fig_pop,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col3
+
+title4 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by State Population (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_pop <- cowplot::plot_grid(
+                              title4, plot_col3,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_pop
+
+#save file
+funds_pop_fig_file = here("results","funds-pop.png")
+ggsave(filename = funds_pop_fig_file, plot = funds_fig_pop)
+```
+
+---
+
+## Predictor: Total Number of Agencies Involved
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$TotalAgencies)
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = TotalAgencies)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = TotalAgencies)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$TotalAgencies)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident month
+req_fig_ag <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = TotalAgencies)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Number of Federal Agencies",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_ag
+
+#repeat for obligated funds
+obl_fig_ag <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = TotalAgencies)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Number of Federal Agencies",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_ag
+
+#combine into one plot
+plot_col4 <- cowplot::plot_grid(req_fig_ag, 
+                                     obl_fig_ag,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col4
+
+title5 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by number of federal agencies (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_ag <- cowplot::plot_grid(
+                              title5, plot_col4,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_ag
+
+#save file
+funds_ag_fig_file = here("results","funds-ag.png")
+ggsave(filename = funds_ag_fig_file, plot = funds_fig_ag)
+```
+
+---
+
+## Predictor: Number of Counties
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$Counties)
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = Counties)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = Counties)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$Counties)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident month
+req_fig_count <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = Counties)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Number of Counties",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_count
+
+#repeat for obligated funds
+obl_fig_count <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = Counties)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Number of Counties",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_count
+
+#combine into one plot
+plot_col5 <- cowplot::plot_grid(req_fig_count, 
+                                     obl_fig_count,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col5
+
+title6 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by number of counties (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_count <- cowplot::plot_grid(
+                              title6, plot_col5,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_count
+
+#save file
+funds_count_fig_file = here("results","funds-count.png")
+ggsave(filename = funds_count_fig_file, plot = funds_fig_count)
+```
+
+---
+
+## Predictor: FEMA Cost Share Average
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$FEMACSAvg)
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = FEMACSAvg)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = FEMACSAvg)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$FEMACSAvg)
+
+#examine graphical relationship with outcomes of interest
+#look at total cost per incident month
+req_fig_cs <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = FEMACSAvg)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Average FEMA Cost Share",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_cs
+
+#repeat for obligated funds
+obl_fig_cs <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = FEMACSAvg)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Average FEMA Cost Share",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_cs
+
+#combine into one plot
+plot_col6 <- cowplot::plot_grid(req_fig_cs, 
+                                     obl_fig_cs,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col6
+
+title7 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by avg FEMA cost share (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_cs <- cowplot::plot_grid(
+                              title7, plot_col6,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_cs
+
+#save file
+funds_cs_fig_file = here("results","funds-cs.png")
+ggsave(filename = funds_cs_fig_file, plot = funds_fig_cs)
+```
+
+---
+
+## Predictor: IH Program
+```{r}
+#convert to factor
+EDAdata$IH <- as.factor(EDAdata$IH)
+
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$IH, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#start with obligated amount
+obl_fig_IH <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IH, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "IH Program",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated when IH program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+obl_fig_IH
+
+#save file
+obl_IH_fig_file = here("results","obl-IH.png")
+ggsave(filename = obl_IH_fig_file, plot = obl_fig_IH)
+
+#repeat for requested funds
+req_fig_IH <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = IH, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) + 
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "IH Program",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested when IH program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+req_fig_IH
+
+#save file
+req_IH_fig_file = here("results","req-IH.png")
+ggsave(filename = req_IH_fig_file, plot = req_fig_IH)
+
+#create a table where rows represent requested and obligated funds and columns are declaration type
+decs_ih <- EDAdata %>%
+                    gtsummary::select(IH, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = IH,
+                      label = list(IH ~ "Individuals / Households Program",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_header(update = list(
+                                               stat_1 ~ "**Not Awarded**",
+                                               stat_2 ~ "**Awarded**")) %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Individuals/Households Program**") %>%
+                    gtsummary::modify_caption("**Table 5. FEMA Funding by IH Program Awardance 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_ih
+
+#save file
+decs_ih_file = here("results", "decs-IH.Rds")
+saveRDS(decs_ih, file = decs_ih_file)
+```
+
+---
+
+## Predictor: PA Program
+```{r}
+#convert to factor
+EDAdata$PA <- as.factor(EDAdata$PA)
+
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$PA, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#start with obligated amount
+obl_fig_PA <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = PA, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "PA Program",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated when PA program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+obl_fig_PA
+
+#save file
+obl_PA_fig_file = here("results","obl-PA.png")
+ggsave(filename = obl_PA_fig_file, plot = obl_fig_PA)
+
+#repeat for requested funds
+req_fig_PA <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = PA, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) + 
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "PA Program",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested when PA program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+req_fig_PA
+
+#save file
+req_PA_fig_file = here("results","req-PA.png")
+ggsave(filename = req_PA_fig_file, plot = req_fig_PA)
+
+#create a table where rows represent requested and obligated funds and columns are declaration type
+decs_pa <- EDAdata %>%
+                    gtsummary::select(PA, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = PA,
+                      label = list(PA ~ "Public Assistance Program",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_header(update = list(
+                                               stat_1 ~ "**Not Awarded**",
+                                               stat_2 ~ "**Awarded**")) %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Public Assistance Program**") %>%
+                    gtsummary::modify_caption("**Table 6. FEMA Funding by PA Program Awardance 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_pa
+
+#save file
+decs_pa_file = here("results", "decs-PA.Rds")
+saveRDS(decs_pa, file = decs_pa_file)
+```
+
+---
+
+## Predictor: HM Program
+```{r}
+#convert to factor
+EDAdata$HM <- as.factor(EDAdata$HM)
+
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata$HM, report.nas = FALSE)
+
+#examine graphical relationship with outcomes
+#include a jitter function to have a better idea of number of measurements and distribution
+#start with obligated amount
+obl_fig_HM <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = HM, y = OblAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "HM Program",
+                           y = "Obligated FEMA Funds",
+                           title = "FEMA funds obligated when HM program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+obl_fig_HM
+
+#save file
+obl_HM_fig_file = here("results","obl-HM.png")
+ggsave(filename = obl_HM_fig_file, plot = obl_fig_HM)
+
+#repeat for requested funds
+req_fig_HM <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = HM, y = ReqAmt)) +
+                      geom_boxplot() +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      scale_y_continuous(labels = scales::dollar_format()) + 
+                      scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
+                      labs(x = "HM Program",
+                           y = "Requested FEMA Funds",
+                           title = "FEMA funds requested when HM program is awarded") +
+                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+req_fig_HM
+
+#save file
+req_HM_fig_file = here("results","req-HM.png")
+ggsave(filename = req_HM_fig_file, plot = req_fig_HM)
+
+#create a table where rows represent requested and obligated funds and columns are declaration type
+decs_hm <- EDAdata %>%
+                    gtsummary::select(HM, ReqAmt, OblAmt) %>%
+                    gtsummary::tbl_summary(
+                      by = HM,
+                      label = list(HM ~ "Hazard Mitigation Program",
+                                   ReqAmt ~ "Requested Funds",
+                                   OblAmt ~ "Obligated Funds"),
+                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+                    gtsummary::modify_header(label ~ "**Variable**") %>% 
+                    gtsummary::modify_header(update = list(
+                                               stat_1 ~ "**Not Awarded**",
+                                               stat_2 ~ "**Awarded**")) %>% 
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Hazard Mitigation Program**") %>%
+                    gtsummary::modify_caption("**Table 7. FEMA Funding by HM Program Awardance 2012 - 2021**") %>%
+                    gtsummary::bold_labels()
+decs_hm
+
+#save file
+decs_hm_file = here("results", "decs-HM.Rds")
+saveRDS(decs_hm, file = decs_hm_file)
+```
+
+---
+
+## Create "Table 1"
+Make the summary table for the manuscript. Variables to include:
+
+* state
+* IncidentYear
+* IncidentMonth
+* declarationType
+* incidentType
+* IncidentDuration
+* IH
+* PA
+* HM
+* Population
+* Counties
+* TotalAgencies
+* FEMACSAvg
+
+```{r}
+#label factors
+EDAdata$IH <- factor(EDAdata$IH, 
+                     levels = c(0, 1),
+                     labels = c("Not Awarded", "Awarded"))
+EDAdata$PA <- factor(EDAdata$PA, 
+                     levels = c(0, 1),
+                     labels = c("Not Awarded", "Awarded"))
+EDAdata$HM <- factor(EDAdata$HM, 
+                     levels = c(0, 1),
+                     labels = c("Not Awarded", "Awarded"))
+
+#specify table labels
+table1::label(EDAdata$state) <- "State"
+table1::label(EDAdata$IncidentYear) <- "Incident Year"
+table1::label(EDAdata$IncidentMonth) <- "Declaration Month"
+table1::label(EDAdata$declarationType) <- "Declaration Type"
+table1::label(EDAdata$incidentType) <- "Incident Type"
+table1::label(EDAdata$IncidentDuration) <- "Incident Duration"
+table1::label(EDAdata$Population) <- "State Population"
+table1::label(EDAdata$Counties) <- "Counties per State"
+table1::label(EDAdata$TotalAgencies) <- "Federal Agencies Involved"
+table1::label(EDAdata$FEMACSAvg) <- "Average FEMA Cost Share"
+table1::label(EDAdata$IH) <- "Individuals/Households Program"
+table1::label(EDAdata$PA) <- "Public Assistance Program"
+table1::label(EDAdata$HM) <- "Hazard Mitigation Program"
+table1::label(EDAdata$ReqAmt) <- "Requested FEMA Funds"
+table1::label(EDAdata$OblAmt) <- "Obligated FEMA Funds"
+
+#specify units
+table1::units(EDAdata$IncidentDuration) <- "days"
+
+#create table
+table1 <- table1::table1(~ ReqAmt + OblAmt + state + IncidentYear + IncidentMonth + declarationType + incidentType +
+                           IncidentDuration + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
+                          data = EDAdata)
+
+#save file
+table1_file = here("results", "table1.Rds")
+saveRDS(table1, file = table1_file)
+```
+
+---
+
+## Save Data
+Since we've made some changes to the format of the processed data, let's save it for the analysis.
+```{r}
+#location to save processed file
+analysis_data_location <- here::here("data","processed_data","analysisdata.rds")
+
+#save as an RDS
+saveRDS(EDAdata, file = analysis_data_location)
+```

### 2021-11-11 code/analysis_code/analysis-linearmodels.Rmd
EDA complete, linear analysis started

---
title: "Analysis: Linear Models"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the processed analysis data and uses the tidymodels framework to fit linear models. Follow the previous markdowns to generate the data.

<br>

This analysis examines the predictors of allocation of FEMA funds during disaster declarations. The following definitions exist for the analysis:

* The two outcomes of interest are Requested Amount (`ReqAmt`) and Obligated Amount (`OblAmt`).
* The primary predictor of interest is Incident Duration (`IncidentDuration`)
* Whichever outcome is not currently fitted to the model will *not* be considered a predictor of interest

<br>

For each outcome of interest, the following steps will be completed:

1. Fit a simple linear regression with the primary predictor of interest
2. Fit a multivariable linear regression
3. Fit a LASSO linear regression

---

## Required Packages
The following R packages are required for this script:

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
* tidymodels: for data modeling

```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
library(tidymodels) #for data modeling

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)
```

---

## Load Processed Data
Load the data generated from the markdowns in the `processing_code` folder.
```{r load data}
#define file path
data_location <- here::here("data", "processed_data", "analysisdata.rds")

#load data
analysis_data <- readRDS(data_location)
```

---

## Data Overview
To better understand the data, let's use `summarytools` to better visualize the data.
```{r data overview}
skimr::skim(analysis_data)
```

---

## Data Filtering
We will make two datasets with the only the relevant variables for each:

* state
* IncidentYear
* IncidentMonth
* declarationType
* incidentType
* IncidentDuration
* IH
* PA
* HM
* Population
* Counties
* TotalAgencies
* FEMACSAvg

```{r data subsets}
#obligated amount
obl_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#requested amount
req_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
```

---

## Outcome of Interest: Requested FEMA Funds
We will start with requested FEMA funds as the outcome of interest.

<br>

### Data Set Up
Split the data randomly into train and test datasets, and define cross validation
```{r req data splitting}
#fix random numbers by setting the seed (helps reproducibility)
set.seed(123)

#put 3/4 of data into the training set
req_data_split <- rsample::initial_split(req_data, prop = 3/4)

#create dataframes for the two sets
req_train_data <- rsample::training(req_data_split)
req_test_data <- rsample::testing(req_data_split)
```

---

### Simple Linear Regression
Create the SLR using the `tidymodels` setup.
```{r req SLR}
#define lr mod
lr_mod <- parsnip::linear_reg() %>%
          parsnip::set_engine("lm")

#define recipe
req_SLR_rec <- recipes::recipe(ReqAmt ~ IncidentDuration, data = req_train_data)

#define SLR workflow
req_SLR_wflow <- workflows::workflow() %>%
                   workflows::add_model(lr_mod) %>%
                   workflows::add_recipe(req_SLR_rec)

#fit the training data to the workflow
req_SLR_fit <- req_SLR_wflow %>%
                  parsnip::fit(data = req_train_data)

#create tibble for model fit using broom and extract
req_SLR_fit %>%
  workflowsets::extract_fit_parsnip() %>%
  broom::tidy()
```

<br>

Create box and whisker plot for the fit output.
```{r}
req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_SLR_bp
```

<br>

Predict requested funds from FEMA  based on the model above for the training data set.
```{r req SLR predict}
#use predict to predict funds
stats::predict(req_SLR_fit, req_train_data)

#create df with model predictions and actual measures
req_SLR_aug <- augment(req_SLR_fit, req_train_data)

#add labeling variables for later summarization
req_SLR_train <- req_SLR_aug %>%
                    dplyr::mutate(data = "train",
                                  model = "SLR")
```

<br>

Assess goodness of fit measures.
```{r req SLR eval}
#using glace
modelsummary::glance(req_SLR_fit)

#using the prediction capacity
req_SLR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
```
This is a terrible model. Time to try adding more predictors.

---

### Multivariable Linear Regression
Create the MVR using the `tidymodels` setup.
```{r req SLR}
#define linear mod
lr_mod <- linear_reg() %>%
              set_engine("lm")

#define recipe
req_MVR_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data)

#define SLR workflow
req_MVR_wflow <- workflows::workflow() %>%
                   workflows::add_model(lr_mod) %>%
                   workflows::add_recipe(req_MVR_rec)

#fit the training data to the workflow
req_MVR_fit <- req_MVR_wflow %>%
                  parsnip::fit(data = req_train_data)

#create tibble for model fit using broom and extract
req_MVR_fit %>%
  workflowsets::extract_fit_parsnip() %>%
  broom::tidy()
```

<br>

Create box and whisker plot for the fit output.
```{r}
#basic box and whisker plot
req_MVR_bp <- broom.mixed::tidy(req_MVR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_bp

#there's a lot of information here, but hard to identify the ones that are significant due to volume

#box and whisker plot for significant predictors
#first filter significant results
req_MVR_fit_sig <- broom.mixed::tidy(req_MVR_fit) %>%
                      dplyr::filter(p.value < 0.05)

#box and whisker plot for lm_fit2 significant predictors
req_MVR_fit_bp2 <- req_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_fit_bp2
```

<br>

Predict requested funds from FEMA  based on the model above for the training data set.
```{r req SLR predict}
#use predict to predict funds
stats::predict(req_MVR_fit, req_train_data)

#create df with model predictions and actual measures
req_MVR_aug <- augment(req_MVR_fit, req_train_data)

#add labeling variables for later summarization
req_MVR_train <- req_MVR_aug %>%
                    dplyr::mutate(data = "train",
                                  model = "MVR")
```

<br>

Assess goodness of fit measures.
```{r req SLR eval}
#using glace
modelsummary::glance(req_SLR_fit)

#using the prediction capacity
req_SLR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
```
This is a terrible model. Time to try adding more predictors.

### 2021-11-11 code/processing_code/processing-combine.Rmd
EDA complete, linear analysis started

@@
 
-## Joining the Datasets
+## Preparing the datasets
 The two datasets could be joined by state or declaration number. 
 ```{r}
-# #first change name of states column in missions df
-# names(missions)[names(missions) == 'stateorTribe'] <- 'state'
-# 
-# #create df with states and declaration numbers from each df
-# missions_nums <- as.data.frame(table(missions$state, missions$disasterNumber))
-# declarations_nums <- as.data.frame(table(declarations$state, declarations$disasterNumber))
+# first change name of states column in missions df
+names(missions)[names(missions) == 'stateorTribe'] <- 'state'
 
-#we want a full join to keep all of the data
-combined <- dplyr::full_join(declarations, missions, by = 'disasterNumber')
+#find differences between disasterNumber and states included in each dataset
+declarations_sub <- declarations %>% dplyr::select(disasterNumber,
+                                                   state)
+
+missions_sub <- missions %>% dplyr::select(disasterNumber,
+                                           state)
+
+missing_declarations <- dplyr::setdiff(missions_sub, declarations_sub)
+
+#search in raw declarations
+decs_check <- raw_decs %>%
+                dplyr::filter(disasterNumber %in% missing_declarations$disasterNumber)
 ```
+By examining the `decs_check` df, we can see that the missing declarations from the missions assignments do not exist in the declarations summaries.
+We can move forward and accept the reduction in data.
 
@@
 ```{r}
+#use a left join to merge into the declarations df
+combined <- dplyr::left_join(declarations, missions, by = c("state" = "state", "disasterNumber" = "disasterNumber"))
+
+#the NAs for funding cost are declarations for which FEMA did not provide any funding
+#therefore they can be replaced with 0
+combined$TotalAgencies[is.na(combined$TotalAgencies)] <- 0
+combined$ReqAmt[is.na(combined$ReqAmt)] <- 0
+combined$OblAmt[is.na(combined$OblAmt)] <- 0
+combined$FEMACSAvg[is.na(combined$FEMACSAvg)] <- 0
+
+#summary of processed data
 summary(combined)
 
-#filter the rows that aren't in declarations data but are in missions data
-missing_nums <- combined %>% 
-                  filter(is.na(state)) %>%
-                  select(disasterNumber)
-
-#unique list of missing declarations
-missing_declarations <- unique(missing_nums)
-
-#check missing declarations in raw declarations
-decs_check <- raw_decs %>%
-                dplyr::filter(disasterNumber %in% missing_declarations$disasterNumber)
-
-#the missing values represent declarations in territories, not states
-#for now, remove - we can always come back and adjust as needed
-combined_processed <- combined %>%
-                        dplyr::filter(!is.na(IncidentYear))
 ```
@@
 #save as an RDS
-saveRDS(combined_processed, file = combined_data_location)
+saveRDS(combined, file = combined_data_location)
 ```

### 2021-11-11 code/processing_code/processing-missions.Rmd
EDA complete, linear analysis started

@@
 
-Currently, the dataset is a mixture of wide and long formatting. To integrate effectively with the Disaster Declaration Summaries dataframe, we need to convert it completely to long formatting.
+Instead of trying to use agencies as a categorical variable, let's convert it to total number of agencies involved in the response.
+Then we can calculate the sum of the FEMA funding for each disaster number and each state
 ```{r long format}
-#the gather function works positionally, so first reorder the columns to be in the right order
-#essentially grouping columns by type of value (e.g. cost share percent, requested funding, obligated funding)
-missions_6 <- missions_5[, c(1:3, 12, 6, 7, 4, 8, 9, 5, 10, 11)]
+#find count of distinct agencies for each disasterNumber and state
+missions_6 <- missions_5 %>%
+                dplyr::group_by(disasterNumber, stateorTribe) %>%
+                dplyr::mutate(TotalAgencies = n_distinct(agency)) %>%
+                subset(select = -c(agency))
 
-#trying piping several gather functions
+#convert NaN to na for variables we're keeping
+missions_6$FEMACSPct[is.nan(missions_6$FEMACSPct)] <- NA
+missions_6$FEMAOblAmt[is.nan(missions_6$FEMAOblAmt)] <- NA
+missions_6$FEMAReqAmt[is.nan(missions_6$FEMAReqAmt)] <- NA
+
+#find sum of FEMA funds
 missions_7 <- missions_6 %>%
-                tidyr::gather(PayOrg, CostSharePct, TotalCSPct:FEMACSPct) %>%
-                tidyr::gather(PayOrg2, ReqAmt, RequestedAmt:FEMAReqAmt) %>%
-                tidyr::gather(PayOrg3, OblAmt, ObligatedAmt:FEMAOblAmt) %>%
-                subset(select = -c(PayOrg2, PayOrg3))
+                dplyr::group_by(disasterNumber, stateorTribe, TotalAgencies) %>%
+                dplyr::summarize(ReqAmt = sum(FEMAReqAmt, na.rm = TRUE),
+                                 OblAmt = sum(FEMAOblAmt, na.rm = TRUE),
+                                 FEMACSAvg = mean(FEMACSPct, na.rm = TRUE))
 
-#relabel values of `PayOrg`
-missions_7$PayOrg[missions_7$PayOrg == "TotalCSPct"] <- "Total"
-missions_7$PayOrg[missions_7$PayOrg == "StateCSPct"] <- "State"
-missions_7$PayOrg[missions_7$PayOrg == "FEMACSPct"] <- "FEMA"
+#the NaN values in FEMACSAvg are when FEMA has no cost share (e.g. 0/0), so it can be replaced with 0
+missions_7$FEMACSAvg[is.nan(missions_7$FEMACSAvg)] <- 0
 ```
@@
 
-After integrating the data, I have realized I really only care about the FEMA funds. So, just subset `PayOrg` to represent FEMA and then drop the `PayOrg` variable.
-```{r FEMA filter}
-missions_8 <- missions_7 %>%
-                dplyr::filter(PayOrg == "FEMA") %>%
-                subset(select = -c(PayOrg))
-
-```
-
----
-
 ## Processed Data Overview
@@
 #create a mirror set for documentation purposes
-missions_processed <- missions_8
+missions_processed <- missions_7
 
@@
 summary(missions_processed)
-
-#not using skimr because it does not do well with long format data
 ```

### 2021-11-12 code/analysis_code/analysis-ML-RA.Rmd
Finished linear models, started ML models

---
title: "Analysis: Machine Learning - ReqAmt"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---

---

## Introduction

This script uses requested FEMA funds as the outcome of interest and fits the following models to the analysis data:

* Null
* Decision Tree
* LASSO
* Random Forest.

It compares the  models, and then finally fits the “best” model to the test data.

<br>

Follow the previous processing and analysis markdowns to generate the analysis data used here.

<br>

Each model will follow this process:

1. Model Specification
2. Workflow Definition
3. Tuning Grid Specification
4. Tuning Using Cross-Validation and the tune_grid() function
5. Identify Best Model
6. Model Evaluation

---

## Required Packages
The following R packages are required for this script:

* here: for data loading/saving
* tidyverse: for data management
* tidymodels: for data modeling
* skimr: for variable summaries
* broom.mixed: for converting bayesian models to tidy tibbles
* rpart.plot: for visualizing a decision tree
* vip: for variable importance plots
* glmnet: for lasso models
* doParallel: for parallel backend for tuning processes
* ranger: for random forest models
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #for data loading/saving
library(tidyverse) #for data management
library(tidymodels) #for data modeling
library(skimr) #for variable summaries
library(broom.mixed) #for converting bayesian models to tidy tibbles
library(rpart.plot) #for visualizing a decision tree
library(vip) #for variable importance plots
library(glmnet) #for lasso models
library(doParallel) #for parallel backend for tuning processes
library(ranger) #for random forest models

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)
```

---

## Load Data
Load the analysis data from the `processed_data` folder in the project file.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","analysisdata.rds")

#load data. 
analysisdata <- readRDS(data_location)

#summary of data using skimr package
skimr::skim(analysisdata)
```

---

## Data Setup
We will specify the following parameters:

* Drop the unused variables
* Transform Incident Month into ordinal variable
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on ReqAmt
* 5-fold cross validation, 5 times repeated, stratified on ReqAmt for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
req_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)

#transform incident month into ordinal variable
req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#set random seed
set.seed(123)

#split dataset into 70% training, 30% testing
#use ReqAmt as stratification
req_data_split <- rsample::initial_split(req_data, prop = 7/10,
                                     strata = c(ReqAmt, state, incidentType))

#optional tag row for later identification: 
req_data$rowid<-1:nrow(req_data)
dplyr_stratified <- function(df, percent, ...){
  columns<-enquos(...)
   #group then sample each group
  out<-df %>% group_by(!!!columns)  %>% slice( sample(1:n(), percent*n())) 
}

train_group <- dplyr_stratified(req_data, 0.9, state, incidentType)
train_group

test_group <- dplyr_stratified(req_data, 0.5, state, incidentType)
test_group

#create dataframes for the two sets:
req_train_data <- rsample::training(req_data_split)
req_test_data <- rsample::testing(req_data_split)

#training set proportions by ReqAmt
req_train_data %>%
      dplyr::count(ReqAmt) %>%
      dplyr::mutate(prop = n / sum(n))

#testing set proportions by ReqAmt
req_test_data %>%
      dplyr::count(ReqAmt) %>%
      dplyr::mutate(prop = n / sum(n))

#5-fold cross validation, 5 times repeated, stratified on `ReqAmt` for the CV folds
req_folds <- rsample::vfold_cv(req_train_data,
                               v = 5,
                               repeats = 5,
                               strata = ReqAmt)

#create recipe that codes categorical variables as dummy variables
RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

---

## Null Model
Determine the performance of a null model (i.e. one with no predictors) and compute the RMSE for both training and test data.
```{r null model}
#create null model
null_mod <- parsnip::null_model() %>%
            parsnip::set_engine("parsnip") %>%
            parsnip::set_mode("regression")

#add recipe and model into workflow
null_wflow <- workflows::workflow() %>%
              workflows::add_recipe(RA_rec) %>%
              workflows::add_model(null_mod)

#"fit" model to training data
req_null_train <- null_wflow %>%
                  parsnip::fit(data = req_train_data)

#summary of null model with training data to get mean (which in this case is the RMSE)
req_null_train_sum <- broom.mixed::tidy(req_null_train)
req_null_train_sum

#RMSE for training data for formatting later
req_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = req_train_data$ReqAmt,
                                              estimate = rep(mean(req_train_data$ReqAmt), nrow(req_train_data))),
                          SE = 0,
                          model = "Null - Train")

#"fit" model to test data
req_null_test <- null_wflow %>%
                  parsnip::fit(data = req_test_data)

#summary of null model with test data to get mean (which in this case is the RMSE)
req_null_test_sum <- broom.mixed::tidy(req_null_test)
req_null_test_sum

#RMSE for testing data for formatting later
req_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = req_test_data$ReqAmt,
                                              estimate = rep(mean(req_test_data$ReqAmt), nrow(req_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

---

## LASSO Model

<br>

### 1. Model Specification
```{r lasso model specification}
#define the lasso model
#mixture = 1 identifies the model to be a LASSO model
lasso_mod <-
  parsnip::linear_reg(mode = "regression",
                      penalty = tune(), 
                      mixture = 1) %>%
  parsnip::set_engine("glmnet")

#use the recipe specified earlier
```

<br>

### 2. Workflow Definition
```{r lasso wflow def}
#define workflow for lasso regression
lasso_wflow <- workflows::workflow() %>%
               workflows::add_model(lasso_mod) %>%
               workflows::add_recipe(RA_rec)
```

<br>

### 3. Tuning Grid Specification
```{r lasso tune grid spec}
#tuning grid specification
lasso_grid <- grid_regular(penalty(), levels = 10)

#5 lowest penalty values
lasso_grid %>%
  dplyr::top_n(-5)

#5 highest penalty values
lasso_grid %>%
  dplyr::top_n(5)
```

<br>

### 4. Tuning Using Cross-Validation and the `tune_grid()` function 
```{r lasso tuning}
#tune the model with previously specified cross-validation and RMSE as target metric
lasso_res <- lasso_wflow %>%
                tune::tune_grid(resamples = req_folds,
                                grid = lasso_grid,
                                control = control_grid(verbose = TRUE, 
                                                       save_pred = TRUE),
                                metrics = metric_set(rmse))

#look at 15 models with lowest RMSEs
top_lasso_models <- lasso_res %>%
                      tune::show_best("rmse", n = 15) %>%
                      dplyr::arrange(penalty)
top_lasso_models

#default visualization
lasso_res %>% autoplot()
```

### 2021-11-12 code/analysis_code/analysis-eda.Rmd
Finished linear models, started ML models

@@
 
-* There are 445 distinct disaster numbers across 2012 - 2021. 
-* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297.
-* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486.
-* `ResponseDays` are missing 72% of the time, which suggests it should not be included in the analysis.
+* There are 445 distinct disaster numbers across 2012 - 2021
+* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297
+* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486
 * There are 16 possible incident types
-* There is a clear effect of both incident month and incident year.
+* There is a clear effect of both incident month and incident year
 
@@
 
+## Predictor: Response Duration
+```{r}
+#summary statistics since it's continuous
+base::summary(EDAdata$ResponseDays)
+#NAs are ongoing events
+
+#density plot
+ggplot2::ggplot(data = EDAdata, aes(x = ResponseDays)) +
+  geom_density()
+
+#normal boxplot
+ggplot2::ggplot(data = EDAdata, aes(y = ResponseDays)) +
+  geom_boxplot()
+
+##QQ plot
+car::qqPlot(EDAdata$ResponseDays)
+
+#examine graphical relationship with outcomes of interest
+req_fig_resp <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = ResponseDays)) +
+                      geom_point(aes(y = ReqAmt),
+                                 color = "tomato",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Response Duration (Days)",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
+req_fig_resp
+
+#repeat for obligated funds
+obl_fig_resp <- EDAdata %>%
+                    ggplot2::ggplot(aes(x = ResponseDays)) +
+                      geom_point(aes(y = OblAmt),
+                                 color = "turquoise4",
+                                 show.legend = FALSE) +
+                      scale_y_continuous(expand = c(0, 0),
+                                         limits = c(0, 1000000000),
+                                         labels = scales::dollar_format()) +
+                      labs(x = "Response Duration (Days)",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_resp
+
+#combine into one plot
+plot_col7 <- cowplot::plot_grid(req_fig_resp, 
+                                     obl_fig_resp,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_col7
+
+title8 <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs obligated FEMA funds by response duration (Under $1B)",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme( plot.margin = margin(0, 0, 0, 7))
+
+funds_fig_resp <- cowplot::plot_grid(
+                              title8, plot_col7,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+funds_fig_resp
+
+#save file
+funds_resp_fig_file = here("results","funds-resp.png")
+ggsave(filename = funds_resp_fig_file, plot = funds_fig_resp)
+```
+
+---
+
 ## Create "Table 1"
@@
 * IncidentDuration
+* ResponseDays
 * IH
@@
 table1::label(EDAdata$IncidentDuration) <- "Incident Duration"
+table1::label(EDAdata$ResponseDays) <- "Response Duration"
 table1::label(EDAdata$Population) <- "State Population"
@@
 table1::units(EDAdata$IncidentDuration) <- "days"
+table1::units(EDAdata$ResponseDays) <- "days"
 
@@
 table1 <- table1::table1(~ ReqAmt + OblAmt + state + IncidentYear + IncidentMonth + declarationType + incidentType +
-                           IncidentDuration + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
+                           IncidentDuration + ResponseDays + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
                           data = EDAdata)

### 2021-11-12 code/analysis_code/analysis-linearmodels.Rmd
Finished linear models, started ML models

@@
 ---
-<br>
 
@@
 2. Fit a multivariable linear regression
-3. Fit a LASSO linear regression
+3. Comparison of the two models
 
@@
 ## Data Filtering
-We will make two datasets with the only the relevant variables for each:
+We first need to address the NA values, and then we will make two datasets with the only the relevant variables for each:
 
@@
 obl_data <- analysis_data %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
+              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                             Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)
@@
 req_data <- analysis_data %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
+              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                             Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
@@
 Create box and whisker plot for the fit output.
-```{r}
+```{r req SLR bp}
 req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
@@
 Create the MVR using the `tidymodels` setup.
-```{r req SLR}
+```{r req MVR}
 #define linear mod
@@
 Create box and whisker plot for the fit output.
-```{r}
+```{r req MVR bp}
 #basic box and whisker plot
@@
 ```
+This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of requested FEMA funding.
+It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.
 
@@
 Predict requested funds from FEMA  based on the model above for the training data set.
-```{r req SLR predict}
+```{r req MVR predict}
 #use predict to predict funds
@@
 Assess goodness of fit measures.
-```{r req SLR eval}
+```{r req MVR eval}
 #using glace
-modelsummary::glance(req_SLR_fit)
+modelsummary::glance(req_MVR_fit)
 
 #using the prediction capacity
-req_SLR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
+req_MVR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
+```
+While this is an improvement, it's still not a great model. Let's compare the two side-by-side using RMSE.
+
+---
+
+### Requested Funds: Model Evaluation
+Formally compare the two linear models
+```{r req lm eval}
+#First create comprehensive df with all of the predictions
+req_lm_preds <- bind_rows(req_MVR_train,
+                          req_SLR_train)
+
+#Use the yardstick package to calculate the metrics on resamples
+req_lm_metrics <- req_lm_preds %>%
+                      dplyr::group_by(data, model) %>%
+                      yardstick::metrics(truth = ReqAmt, estimate = .pred)
+
+#for interpretation: rmse = Root Mean Squared Error (RMSE), rsq = R^2, mae = Mean Absolute Error (MAE)
+#make a df that displays RMSE and R^2 for each model and data (all dplyr functions)
+req_lm_RMSE <- req_lm_metrics %>%
+                    filter(.metric == "rmse" | .metric == "rsq") %>%
+                    select(-c(.estimator, data))
+
+#save comparison table
+req_lm_eval_file = here("results", "lm-req-lm-eval.Rds")
+saveRDS(req_lm_RMSE, file = req_lm_eval_file)
+```
+In comparing the two models, the full model certainly improves the fit, but neither is ideal. Alternative models should be considered.
+
+---
+
+## Outcome of Interest: Obligated FEMA Funds
+We will now examine obligated FEMA funds as the outcome of interest.
+
+<br>
+
+### Data Set Up
+Split the data randomly into train and test datasets, and define cross validation
+```{r obl data splitting}
+#fix random numbers by setting the seed (helps reproducibility)
+set.seed(123)
+
+#put 3/4 of data into the training set
+obl_data_split <- rsample::initial_split(obl_data, prop = 3/4)
+
+#create dataframes for the two sets
+obl_train_data <- rsample::training(obl_data_split)
+obl_test_data <- rsample::testing(obl_data_split)
+```
+
+---
+
+### Simple Linear Regression
+Create the SLR using the `tidymodels` setup.
+```{r obl SLR}
+#define lr mod
+lr_mod <- parsnip::linear_reg() %>%
+          parsnip::set_engine("lm")
+
+#define recipe
+obl_SLR_rec <- recipes::recipe(OblAmt ~ IncidentDuration, data = obl_train_data)
+
+#define SLR workflow
+obl_SLR_wflow <- workflows::workflow() %>%
+                   workflows::add_model(lr_mod) %>%
+                   workflows::add_recipe(obl_SLR_rec)
+
+#fit the training data to the workflow
+obl_SLR_fit <- obl_SLR_wflow %>%
+                  parsnip::fit(data = obl_train_data)
+
+#create tibble for model fit using broom and extract
+obl_SLR_fit %>%
+  workflowsets::extract_fit_parsnip() %>%
+  broom::tidy()
+```
+
+<br>
+
+Create box and whisker plot for the fit output.
+```{r obl SLR bp}
+obl_SLR_bp <- broom.mixed::tidy(obl_SLR_fit) %>%
+                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
+                                   whisker_args = list(color = "blue"),
+                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
+obl_SLR_bp
+```
+
+<br>
+
+Predict requested funds from FEMA  based on the model above for the training data set.
+```{r obl SLR predict}
+#use predict to predict funds
+stats::predict(obl_SLR_fit, obl_train_data)
+
+#create df with model predictions and actual measures
+obl_SLR_aug <- augment(obl_SLR_fit, obl_train_data)
+
+#add labeling variables for later summarization
+obl_SLR_train <- obl_SLR_aug %>%
+                    dplyr::mutate(data = "train",
+                                  model = "SLR")
+```
+
+<br>
+
+Assess goodness of fit measures.
+```{r obl SLR eval}
+#using glace
+modelsummary::glance(obl_SLR_fit)
+
+#using the prediction capacity
+obl_SLR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
 ```
 This is a terrible model. Time to try adding more predictors.
+
+---
+
+### Multivariable Linear Regression
+Create the MVR using the `tidymodels` setup.
+```{r obl MVR}
+#define linear mod
+lr_mod <- linear_reg() %>%
+              set_engine("lm")
+
+#define recipe
+obl_MVR_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data)
+
+#define SLR workflow
+obl_MVR_wflow <- workflows::workflow() %>%
+                   workflows::add_model(lr_mod) %>%
+                   workflows::add_recipe(obl_MVR_rec)
+
+#fit the training data to the workflow
+obl_MVR_fit <- obl_MVR_wflow %>%
+                  parsnip::fit(data = obl_train_data)
+
+#create tibble for model fit using broom and extract
+obl_MVR_fit %>%
+  workflowsets::extract_fit_parsnip() %>%
+  broom::tidy()
+```
+
+<br>
+
+Create box and whisker plot for the fit output.
+```{r obl MVR bp}
+#basic box and whisker plot
+obl_MVR_bp <- broom.mixed::tidy(obl_MVR_fit) %>%
+                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
+                                   whisker_args = list(color = "blue"),
+                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
+obl_MVR_bp
+
+#there's a lot of information here, but hard to identify the ones that are significant due to volume
+
+#box and whisker plot for significant predictors
+#first filter significant results
+obl_MVR_fit_sig <- broom.mixed::tidy(obl_MVR_fit) %>%
+                      dplyr::filter(p.value < 0.05)
+
+#box and whisker plot for lm_fit2 significant predictors
+obl_MVR_fit_bp2 <- obl_MVR_fit_sig %>%
+                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
+                                           whisker_args = list(color = "blue"),
+                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
+obl_MVR_fit_bp2
+```
+This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of obligated FEMA funding.
+It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.
+
+<br>
+
+Predict obligated funds from FEMA  based on the model above for the training data set.
+```{r obl MVR predict}
+#use predict to predict funds
+stats::predict(obl_MVR_fit, obl_train_data)
+
+#create df with model predictions and actual measures
+obl_MVR_aug <- augment(obl_MVR_fit, obl_train_data)
+
+#add labeling variables for later summarization
+obl_MVR_train <- obl_MVR_aug %>%
+                    dplyr::mutate(data = "train",
+                                  model = "MVR")
+```
+
+<br>
+
+Assess goodness of fit measures.
+```{r obl MVR eval}
+#using glace
+modelsummary::glance(obl_MVR_fit)
+
+#using the prediction capacity
+obl_MVR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
+```
+While this is an improvement, it's still not a great model. Let's compare the two side-by-side using RMSE.
+
+---
+
+### Obligated Funds: Model Evaluation
+Formally compare the two linear models
+```{r obl lm eval}
+#First create comprehensive df with all of the predictions
+obl_lm_preds <- bind_rows(obl_MVR_train,
+                          obl_SLR_train)
+
+#Use the yardstick package to calculate the metrics on resamples
+obl_lm_metrics <- obl_lm_preds %>%
+                      dplyr::group_by(data, model) %>%
+                      yardstick::metrics(truth = OblAmt, estimate = .pred)
+
+#for interpretation: rmse = Root Mean Squared Error (RMSE), rsq = R^2, mae = Mean Absolute Error (MAE)
+#make a df that displays RMSE and R^2 for each model and data (all dplyr functions)
+obl_lm_RMSE <- obl_lm_metrics %>%
+                    filter(.metric == "rmse" | .metric == "rsq") %>%
+                    select(-c(.estimator, data))
+
+#save comparison table
+obl_lm_eval_file = here("results", "lm-obl-lm-eval.Rds")
+saveRDS(obl_lm_RMSE, file = obl_lm_eval_file)
+```
+In comparing the two models, the full model certainly improves the fit, but neither is ideal. Alternative models should be considered.
+
+---
+
+## Compare Requested and Obligated Models
+As one last step for the linear models, let's compare the metrics from the obligated and requested funds models.
+```{r lm comp}
+#add outcome labels
+req_lm_RMSE$outcome <- "Requested"
+obl_lm_RMSE$outcome <- "Obligated"
+
+#combine into one table
+lm_comp <- dplyr::bind_rows(req_lm_RMSE,
+                            obl_lm_RMSE)
+
+#clean up the table
+lm_comp$label <- paste(lm_comp$model, lm_comp$.metric, sep=": ")
+
+lm_comp_tbl <- lm_comp %>%
+                  subset(select = -c(1:2))
+
+lm_comp_table <- xtabs(.estimate ~ label + outcome, lm_comp_tbl)
+lm_comp_table
+
+#save file
+lm_eval_file = here("results", "lm-eval.Rds")
+saveRDS(lm_comp_table, file = lm_eval_file)
+```
+The two outcomes yield virtually identical models.

### 2021-11-12 code/processing_code/processing-declarations.Rmd
Finished linear models, started ML models

@@
 declarations_3 <- declarations_2 %>%
-                    dplyr::mutate(IncidentDuration = difftime(incidentEndDate,
+                    dplyr::mutate(IncidentDuration = ifelse(is.na(incidentEndDate),
+                                                            difftime("2021-11-10T00:00:00.000Z",
                                                                      incidentBeginDate,
                                                                      units = c("days")),
-                                  ResponseDays = difftime(disasterCloseoutDate,
+                                                            difftime(incidentEndDate,
                                                               incidentBeginDate,
-                                                          units = c("days")))
+                                                              units = c("days"))),
+                                  ResponseDays = ifelse(is.na(disasterCloseoutDate),
+                                                        difftime("2021-11-10T00:00:00.000Z",
+                                                                 incidentBeginDate,
+                                                                 units = c("days")),
+                                                        difftime(disasterCloseoutDate,
+                                                                incidentBeginDate,
+                                                                units = c("days"))))
 ```

### 2021-11-12 archive/Part1-Proposal.Rmd
Finished modeling analysis

---
title: "MADA Project - Part 1"
author: Morgan Taylor
output: html_document
---

```{r setup, include=FALSE}
library(here) #to set paths
knitr::opts_chunk$set(echo = TRUE)
```

---

# Part 1: Proposal and Data Identification

<br>

# Primary Proposal: Predicting Outcomes for Burn Patients

<br>

## Relevant Background Information
A nuclear catastrophe would generate an unprecedented volume of patients, with injuries resulting from the blast itself, the radiation emitted, and thermal burns. Of all the injury etiologies, burns may be the least understood and have the biggest resource constraints. As a result, prediction models characterizing burn outcomes for patients will better equip healthcare providers and other responders to manage these specific injuries. Logistic regression models have traditionally been employed to develop prediction scores for mortality of all burn patients (Halgas et al., 2018). However, other healthcare disciplines have increasingly transitioned to ML models, which are automatically generated and continually improved, potentially increasing predictive accuracy (Rojas et al., 2018; Taylor et al., 2016). Preliminary research suggests ML models using artificial neural network (ANN) models can predict burn patient mortality more accurately than commonly used prediction scores (Liu & Salinas, 2015), but there is evidence to suggest random forest models may be more advantageous in emergent applications (Taylor et al., 2016).

<br>

## Data Source
American Burn Association (ABA) provides a dataset called the National Burn Repository (NBR), which includes patients treated or admitted to an ABA-verified burn hospital for the years 2009 - 2018. Each year, the ABA-verfied burn hospitals submit the appropriate patient records, and the NBR Committee reviews and validates the data. The dataset includes information regarding patient demographics, injury, injury severity, diagnoses, burn center admission, hospital procedure, outcome, and finance.

<br>

## Data Overview
For the study period, there are 82,404 burn patients with a thermal etiology identified in the analysis. Full access to the data is waiting on approval from the legal consultant at the Office of Research for UGA and access to a HIPAA-compliant server. This access is expected soon, but there is a second proposal that follows in the event accessing the data takes longer than the timeline for this course allows.

<br>

## Questions of Interest
There are two questions this study will endeavor to answer:

1. What are the variables that significantly predict thermal burn patient mortality for individuals treated at an ABA-verified burn center?

2. What are the variables that significantly predict thermal burn patient length of stay at an ABA-verified burn center?

All of the predictors will be considered initially, if the infrastructure allows for it. A second iteration of the modeling may focus specifically on predictors that can be determined at initial patient presentation to increase applicability in emergent settings.

<br>

## Proposed Methods
As logistic regression prediction models are already well established in the field, this project will utilize machine learning methods, or more specifically random forest and artificial neural network. A bootstrap sampling methodology will also be used with a 67%:33% sampling ratio to allow for model training and internal validation. The accuracy of the models will also be compared to the currently accepted Modified-Baux score.

<br>
<br>

---

<br>
<br>

# Secondary Proposal: Tracking Radiation from Chernobyl
In the event that the data for the primary proposal does not make it through regulation protocols in time, the following section details a backup project using publicly available data.

<br>

## Relevant Background Information
On April 26, 1989, a power failure at the number 4 reactor at the Chernobyl Nuclear Power Plant in the Soviet Union caused the worst nuclear disaster in history, resulting in an estimated $68 billion (adjusted for inflation) of damages. Due to the unprecedented nature of the event and the delay in notification of the general public, many questions still remain regarding the concentration and distribution of fallout radiation across several countries. The controversy surrounding the true impact of the event remains even more than thirty years later, so this study aims to better understand the distribution of the different radioisotopes from a nuclear reactor meltdown.

<br>

## Data Source
A publicly available dataset identifying sampling locations from across several European countries can be accessed at the following link:

https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data

There are nine variables included: country abbreviation, country numerical code, sampling location (city or state), latitude of sampling location, longitude of sampling location, date of sampling, and concentration of three radioisotopes (Iodine 131, Caesium 134, and Ceasium 137). There are 2051 observations included, but not all observations have complete data.

<br>

## Data Overview
As the data is publicly available for download, summary descriptions of the data can be done. Radioisotopes are reported in the concentration of rad/m^3.

```{r}
#Import data
#define path to data
data_location <- here::here("data","raw_data","Chernobyl_ Chemical_Radiation.csv")

#load data. 
rawdata <- utils::read.csv(data_location)

#rename variable columns
processeddata <- rawdata
colnames(processeddata) <- c("Country-Abbrev", "Country-Code", "Sample-Location", "Longitude", "Latitude", "Date", "Iodine131", "Caesium134", "Caesium137")

#convert radioisotope variables to numeric variables
processeddata$Caesium134 <- as.numeric(processeddata$Caesium134)
processeddata$Caesium137 <- as.numeric(processeddata$Caesium137)
processeddata$Iodine131 <- as.numeric(processeddata$Iodine131)
#NAs introduced by coercion with this coding - will require more thorough cleaning for actual analysis

#overview
utils::str(processeddata)

#summary
summary(processeddata)
```

<br>

## Questions of Interest
There are three primary questions this study will endeavor to answer:

1. Is there a significant reduction in radioisotope concentration captured within the study period? If so, do the results represent known half-life decomposition for each radioisotope type?

2. What is the geospatial dispersion of fallout radiation captured through this publicly available sampling dataset?

3. Does the geospatial dispersion reported in the dataset match commonly accepted models for nuclear reactor meltdown? (Pending approval)

<br>

## Proposed Methods
The first portion of this analysis will involve fitting the data to a nuclear half-life decay model. Further research is required to understand how to account for potential interaction between the three isotopes during the process of radioactive decay.

The second research question will involve an in-depth exploration of geospatial modeling capabilities of R. The latitude and longitude values will be used to identify the sampling locations on a map, and the concentration of each radioisotope will be the values of interest.

The final research question hinges upon approval from the appropriate authorities regarding the modeling software and whether the methods and results can be publicly available. However, if this is allowed, the two geospatial files can be overlayed and k-nearest methods could potentially be used to determine discrepancies between the two models.

### 2021-11-12 code/analysis_code/analysis-ML-OA.Rmd
Finished modeling analysis

---
title: "Analysis: Machine Learning - OblAmt"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---

---

## Introduction

This script uses obligated FEMA funds as the outcome of interest and fits the following models to the analysis data:

* Null
* Decision Tree
* Random Forest
* Bagged Tree

It compares the  models, and then finally fits the “best” model to the test data.

<br>

Follow the previous processing and analysis markdowns to generate the analysis data used here.

<br>

Each model will follow this process:

1. Model Specification
2. Workflow Definition
3. Tuning Grid Specification
4. Tuning Using Cross-Validation and the tune_grid() function
5. Identify Best Model
6. Model Evaluation

---

## Required Packages
The following R packages are required for this script:

* here: for data loading/saving
* tidyverse: for data management
* tidymodels: for data modeling
* skimr: for variable summaries
* broom.mixed: for converting bayesian models to tidy tibbles
* rpart.plot: for visualizing a decision tree
* vip: for variable importance plots
* glmnet: for lasso models
* doParallel: for parallel backend for tuning processes
* ranger: for random forest models
* baguette: for bagged decision tree
```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #for data loading/saving
library(tidyverse) #for data management
library(tidymodels) #for data modeling
library(skimr) #for variable summaries
library(broom.mixed) #for converting bayesian models to tidy tibbles
library(rpart.plot) #for visualizing a decision tree
library(vip) #for variable importance plots
library(glmnet) #for lasso models
library(doParallel) #for parallel backend for tuning processes
library(ranger) #for random forest models
library(baguette) #for bagged decision tree

#global environment options
# formatting for script to avoid scientific notation output
options(scipen=999)
```

---

## Load Data
Load the analysis data from the `processed_data` folder in the project file.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","analysisdata.rds")

#load data. 
analysisdata <- readRDS(data_location)

#summary of data using skimr package
skimr::skim(analysisdata)
```

---

## Data Setup
We will specify the following parameters:

* Drop the unused variables
* Perform feature engineering
* Transform Incident Month into ordinal variable
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on incidentType
* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
obl_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#transform incident month into ordinal variable
obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#not enough values per state, so create a variable that is FEMA Regions
obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
                                 obl_data$state == "ME" |
                                 obl_data$state == "MA" |
                                 obl_data$state == "NH" |
                                 obl_data$state == "RI" |
                                 obl_data$state == "VT"),
                              "Region I",
                              ifelse((obl_data$state == "NJ" |
                                         obl_data$state == "NY" |
                                         obl_data$state == "PR" |
                                         obl_data$state == "VI"),
                                      "Region II",
                                       ifelse((obl_data$state == "DE" |
                                                 obl_data$state == "MD" |
                                                 obl_data$state == "PA" |
                                                 obl_data$state == "VA" |
                                                 obl_data$state == "DC" |
                                                 obl_data$state == "WV"),
                                              "Region III",
                                               ifelse((obl_data$state == "AL" |
                                                         obl_data$state == "FL" |
                                                         obl_data$state == "GA" |
                                                         obl_data$state == "KY" |
                                                         obl_data$state == "MS" |
                                                         obl_data$state == "NC" |
                                                         obl_data$state == "SC" |
                                                         obl_data$state == "TN"),
                                                      "Region IV",
                                                      ifelse((obl_data$state == "IL" |
                                                               obl_data$state == "IN" |
                                                               obl_data$state == "MI" |
                                                               obl_data$state == "MN" |
                                                               obl_data$state == "OH" |
                                                               obl_data$state == "WI"),
                                                            "Region V",
                                                             ifelse((obl_data$state == "AR" |
                                                                         obl_data$state == "LA" |
                                                                         obl_data$state == "NM" |
                                                                         obl_data$state == "OK" |
                                                                         obl_data$state == "TX"),
                                                                      "Region VI",
                                                                       ifelse((obl_data$state == "IA" |
                                                                                 obl_data$state == "KS" |
                                                                                 obl_data$state == "MO" |
                                                                                 obl_data$state == "NE"),
                                                                              "Region VII",
                                                                               ifelse((obl_data$state == "CO" |
                                                                                         obl_data$state == "MT" |
                                                                                         obl_data$state == "ND" |
                                                                                         obl_data$state == "SD" |
                                                                                         obl_data$state == "UT" |
                                                                                         obl_data$state == "WY"),
                                                                                      "Region VIII",
                                                                                       ifelse((obl_data$state == "AZ" |
                                                                                                 obl_data$state == "CA" |
                                                                                                 obl_data$state == "HI" |
                                                                                                 obl_data$state == "NV"),
                                                                                              "Region IX",
                                                                                               ifelse((obl_data$state == "AK" |
                                                                                                         obl_data$state == "ID" |
                                                                                                         obl_data$state == "OR" |
                                                                                                         obl_data$state == "WA"),
                                                                                                      "Region X",
                                                                                                      NA
                                                                                                      )
                                                                                              )
                                                                                      )
                                                                              )
                                                                      )
                                                            )
                                                      )
                                              )
                                     )
                              )

#drop state variable
obl_data <- obl_data %>%
              dplyr::select(-c(state))

#add to "other" category for incident Types that are infrequent
obl_data$incidentType[(obl_data$incidentType == "Chemical" |
                        obl_data$incidentType == "Coastal Storm" |
                        obl_data$incidentType == "Dam/Levee Break" |
                        obl_data$incidentType == "Mud/Landslide" |
                        obl_data$incidentType == "Volcano" |
                        obl_data$incidentType == "Terrorist" |
                        obl_data$incidentType == "Snow")] <- "Other"

#set random seed
set.seed(123)

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
obl_data_split <- rsample::initial_split(obl_data, prop = 2/3,
                                     strata = incidentType)

#create dataframes for the two sets:
obl_train_data <- rsample::training(obl_data_split)
obl_test_data <- rsample::testing(obl_data_split)

#5-fold cross validation, repeated 5 times, stratified by incidentType
obl_folds <- rsample::vfold_cv(obl_train_data,
                               strata = incidentType,
                               v = 5,
                               repeats = 5)

#create recipe that codes categorical variables as dummy variables
OA_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

---

## Null Model
Determine the performance of a null model (i.e. one with no predictors) and compute the RMSE for both training and test data.
```{r null model}
#create null model
null_mod <- parsnip::null_model() %>%
            parsnip::set_engine("parsnip") %>%
            parsnip::set_mode("regression")

#add recipe and model into workflow
null_wflow <- workflows::workflow() %>%
              workflows::add_recipe(OA_rec) %>%
              workflows::add_model(null_mod)

#"fit" model to training data
obl_null_train <- null_wflow %>%
                  parsnip::fit(data = obl_train_data)

#summary of null model with training data to get mean (which in this case is the RMSE)
obl_null_train_sum <- broom.mixed::tidy(obl_null_train)
obl_null_train_sum

#RMSE for training data for formatting later
obl_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_train_data$OblAmt,
                                              estimate = rep(mean(obl_train_data$OblAmt), nrow(obl_train_data))),
                          SE = 0,
                          model = "Null - Train")

#"fit" model to test data
obl_null_test <- null_wflow %>%
                  parsnip::fit(data = obl_test_data)

#summary of null model with test data to get mean (which in this case is the RMSE)
obl_null_test_sum <- broom.mixed::tidy(obl_null_test)
obl_null_test_sum

#RMSE for testing data for formatting later
obl_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_test_data$OblAmt,
                                              estimate = rep(mean(obl_test_data$OblAmt), nrow(obl_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

---

## Decision Tree

<br>

### 1. Model Specification
```{r}
#run parallels to determine number of cores
cores <- parallel::detectCores() - 1
cores

cl <- makeCluster(cores)

registerDoParallel(cl)

#define the tree model
tree_mod <-
  parsnip::decision_tree(cost_complexity = tune(),
                         tree_depth = tune(),
                         min_n = tune()) %>%
  parsnip::set_engine("rpart") %>%
  parsnip::set_mode("regression")

#use the recipe specified earlier (line 133)
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
tree_wflow <- workflows::workflow() %>%
               workflows::add_model(tree_mod) %>%
               workflows::add_recipe(OA_rec)
```

---

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
tree_grid <- dials::grid_regular(cost_complexity(),
                                 tree_depth(),
                                 min_n(),
                                 levels = 5)

#tree depth
tree_grid %>%
  dplyr::count(tree_depth)
```

---

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = obl_folds,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

#collect metrics
tree_res %>% workflowsets::collect_metrics()

#default visualization
DT_auto_plot <- tree_res %>% autoplot()
DT_auto_plot

#save file
DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

#more detailed plot
DT_detail_plot <- tree_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
DT_detail_plot

#save file
DT_detail_fig_file = here("results","DT-det-plot-obl.png")
ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)

```

---

### 5. Identify Best Model
```{r}
#select the tree model with the lowest rmse
tree_lowest_rmse <- tree_res %>%
                        tune::select_best("rmse")

#finalize the workflow by using the selected DT model
best_tree_wflow <- tree_wflow %>%
                      tune::finalize_workflow(tree_lowest_rmse)
best_tree_wflow

#one last fit on the training data
best_tree_fit <- best_tree_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

---

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                         roundint = F,
                                         type = 5,
                                         digits = 5,
                                         main = "Selected Decision Tree Model")
#save file manually -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
DT_pred_act

#save file
DT_pred_act_file = here("results","DT-pred-act-obl.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
                     geom_hline(yintercept = 0, color = "red", lty = 2) +
                     geom_point() +
                     labs(title = "Decision Tree Fit: Residuals",
                          x = "Observation Number",
                          y = "Residual")
DT_resid

#save file
DT_resid_file = here("results","DT-resid-obl.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
DT_pred_resid

#save file
DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

#print model performance
#print 10 best performing hyperparameter sets
tree_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
  dplyr::mutate(rmse = round(rmse, 3),
                std_err = round(std_err, 4),
                cost_complexity = scales::scientific(cost_complexity))

#print the best model performance
tree_performance <- tree_res %>% tune::show_best(n = 1)
print(tree_performance)

#compare model performance to null model
tree_RMSE <- tree_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Tree") %>%
               dplyr::bind_rows(obl_null_RMSE_train)
tree_RMSE
```

<br>

---

## Random Forest Model

<br>

### 1. Model Specification
```{r}
#run parallels to determine number of cores
cores <- parallel::detectCores() - 1
cores

cl <- makeCluster(cores)

registerDoParallel(cl)

#define the RF model
RF_mod <-
  parsnip::rand_forest(mtry = tune(),
                       min_n = tune(),
                       trees = tune()) %>%
  parsnip::set_engine("ranger",
                      importance = "permutation") %>%
  parsnip::set_mode("regression")

#use the recipe specified earlier (line 133)

#check to make sure identified parameters will be tuned
RF_mod %>% tune::parameters()
```

---

### 2. Workflow Definition
```{r}
#define workflow for RF regression
RF_wflow <- workflows::workflow() %>%
               workflows::add_model(RF_mod) %>%
               workflows::add_recipe(OA_rec)
```

---

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
RF_grid <- expand.grid(mtry = c(3, 4, 5, 6),
                       min_n = c(40, 50, 60),
                       trees = c(500,1000))
```

---

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = obl_folds,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

#look at top 5 RF models
top_RF_models <- RF_res %>%
                    tune::show_best("rmse", n = 5)
top_RF_models

#default visualization
RF_auto_plot <- RF_res %>% autoplot()
RF_auto_plot

#save file
RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

---

### 5. Identify Best Model
```{r}
#select the RF model with the lowest rmse
RF_lowest_rmse <- RF_res %>%
                      tune::select_best("rmse")

#finalize the workflow by using the selected RF model
best_RF_wflow <- RF_wflow %>%
                      tune::finalize_workflow(RF_lowest_rmse)
best_RF_wflow

#one last fit on the training data
best_RF_fit <- best_RF_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

---

### 6. Model evaluation
```{r}
#extract model from final fit
x_RF <- best_RF_fit$fit$fit$fit

#plot most important predictors in the model
RF_plot <- vip::vip(x_RF, num_features = 20)
RF_plot

#save  manually - ggsave throws an error

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = obl_train_data) %>%
                dplyr::select(.pred, OblAmt) %>%
                dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point() +
                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")

#save file
RF_pred_act_file = here("results","RF-pred-act-obl.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point() +
                            labs(title = "RF Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
RF_resids

#save file
RF_resid_file = here("results","RF-resid-obl.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Obligated Funds",
                                x = "Fitted Requested Funds ($)",
                                y = "Residual")
RF_pred_resid

#save file
RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

#print the 10 best performing hyperparameter sets
RF_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err)

#print the best model
RF_performance <- RF_res %>% tune::show_best(n = 1)
RF_performance

#compare model performance to null model (and other models)
RF_RMSE <- RF_res %>%
              tune::show_best(n = 1) %>%
              dplyr::transmute(
                rmse = round(mean, 3),
                SE = round(std_err, 4),
                model = "RF") %>%
             dplyr::bind_rows(tree_RMSE)
RF_RMSE
```

---

## Bagged Decision Tree Model

<br>

### 1. Model Specification
```{r}
#run parallels to determine number of cores
cores <- parallel::detectCores() - 1
cores

cl <- makeCluster(cores)

registerDoParallel(cl)

#start over with data split to allow for bootstrapping
#set random seed
set.seed(20140102)

#use bootstrap instead of CV
obl_data_bootsplit <- obl_train_data %>%
                        rsample::bootstraps(times = 10,
                                            strata = "incidentType")

#define model
bag_tree_mod <- baguette::bag_tree(cost_complexity = tune(),
                                   tree_depth = tune(),
                                   min_n = tune()) %>%
                parsnip::set_engine("rpart",
                                    times = 10) %>%
                parsnip::set_mode("regression")

#create recipe that codes categorical variables as dummy variables
BT_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          step_string2factor(all_nominal())
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
BT_wflow <- workflows::workflow() %>%
               workflows::add_model(bag_tree_mod) %>%
               workflows::add_recipe(BT_rec)
```

---

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
tree_grid <- dials::grid_regular(cost_complexity(),
                                 tree_depth(),
                                 min_n(),
                                 levels = 5)
```

---

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
BT_res <- baguette::bag_tree(cost_complexity = tune(),
                                   tree_depth = tune(),
                                   min_n = tune()) %>%
          parsnip::set_engine("rpart",
                              times = 10) %>%
          parsnip::set_mode("regression") %>%
          tune::tune_grid(preprocessor = BT_rec,
                          resamples = obl_data_bootsplit,
                          grid = tree_grid,
                          metrics = metric_set(rmse))

#collect metrics
BT_res %>% workflowsets::collect_metrics()

#default visualization
BT_auto_plot <- BT_res %>% autoplot()
BT_auto_plot

#save file
BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)

#more detailed plot
BT_det_plot <- BT_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
BT_det_plot

#save file
BT_det_fig_file = here("results","BT-det-plot-obl.png")
ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
```

---

### 5. Identify Best Model
```{r}
#select the BT model with the lowest rmse
BT_lowest_rmse <- BT_res %>%
                        tune::select_best("rmse")

#finalize the workflow by using the selected BT model
best_BT_wflow <- BT_wflow %>%
                      tune::finalize_workflow(BT_lowest_rmse)
best_BT_wflow

#one last fit on the training data
best_BT_fit <- best_BT_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

---

### 6. Model evaluation
```{r}
#extract model from final fit
x_BT <- best_BT_fit$fit$fit$fit

#vip package doesn't support BT

#find predictions and intervals
BT_resid <- best_BT_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
BT_pred_act

#save file
BT_pred_act_file = here("results","BT-pred-act-obl.png")
ggsave(filename = BT_pred_act_file, plot = BT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point() +
                      labs(title = "Bagged Tree Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
BT_resids

#save file
BT_resid_file = here("results","BT-resids-obl.png")
ggsave(filename = BT_resid_file, plot = BT_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
BT_pred_resid

#save file
BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)

#print model performance
#print 10 best performing hyperparameter sets
BT_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
  dplyr::mutate(rmse = round(rmse, 3),
                std_err = round(std_err, 4),
                cost_complexity = scales::scientific(cost_complexity))

#print the best model performance
BT_performance <- BT_res %>% tune::show_best(n = 1)
print(BT_performance)

#compare model performance to null model
BT_RMSE <- BT_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Bagged Tree") %>%
               dplyr::bind_rows(RF_RMSE)
BT_RMSE

#save combined table
model_comp_file = here("results", "ML-model-comp-obl.Rds")
saveRDS(BT_RMSE, file = model_comp_file)
```

In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.

--

## Model Selection and Evaluation
Using the bagged tree model
```{r}
#fit model to training set but evaluate on the test data
BT_fit_test <- best_BT_wflow %>%
                    tune::last_fit(split = obl_data_split)

#compare test performance against training performance
BT_rmse_test <- collect_metrics(BT_fit_test) %>%
  dplyr::select(rmse = .estimate) %>%
  dplyr::mutate(data = "test")

BT_comp <- BT_RMSE %>%
                  dplyr::filter(model == "Bagged Tree") %>%
                  dplyr::transmute(rmse, data = "train") %>%
                  dplyr::bind_rows(BT_rmse_test) %>%
                  dplyr::slice(-3) #don't know why the third row shows up
BT_comp
#RMSEs are incredibly different...

#find predictions and intervals
BT_resid_fit <- BT_fit_test %>%
                  broom.mixed::augment() %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
BT_pred_act_test

#save file
BT_pred_act_test_file = here("results","BT-pred-act-test-obl.png")
ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
BT_resids_test <- ggplot2::ggplot(BT_resid_fit, aes(x = as.numeric(row.names(BT_resid_fit)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point() +
                      labs(title = "Bagged Tree Test Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
BT_resids_test

#save file
BT_resid_test_file = here("results","BT-resid-test-obl.png")
ggsave(filename = BT_resid_test_file, plot = BT_resids_test)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
BT_pred_resid_test

#save file
BT_pred_resid_test_file = here("results","BT-pred-resid-test-obl.png")
ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)
```

### 2021-11-12 products/manuscript/Manuscript_Disasters.Rmd
Finished modeling analysis

--- 2021-10-29 products/manuscript/Manuscript_Fallout.Rmd
@@
 ---
-title: "Examining the Fallout Radioactivity Across Europe After the Chernobyl Meltdown"
+title: "Predicting Federal Emergency Management Agency Funding for U.S. Disasters"
 subtitle: ""
@@
 _Provide enough background on your topic that others can understand the why and how of your analysis_
-On April 26, 1989, a power failure at the number 4 reactor at the Chernobyl Nuclear Power Plant in the Soviet Union caused the worst nuclear disaster in history, resulting in an estimated $68 billion (adjusted for inflation) of damages.
-Due to the unprecedented nature of the event and the delay in notification of the general public, many questions still remain regarding the concentration and distribution of fallout radiation across several countries.
-As the controversy surrounding the true impact of the event remains even more than thirty years later, this study aims to better understand the distribution of the different radioisotopes from a nuclear reactor meltdown.
 
@@
 _Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-A publicly available dataset identifying sampling locations from across several European countries can be accessed [here](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data).
-
-There are nine variables included: country abbreviation, country numerical code, sampling location (city or state), latitude of sampling location, longitude of sampling location, date of sampling, and concentration of three radioisotopes (Iodine 131, Caesium 134, and Ceasium 137).
-There are 2051 observations included, but not all observations have complete data.
 
@@
 _State the research questions you plan to answer with this analysis._
-There are three primary questions this study will endeavor to answer:
 
-1. Is there a significant reduction in radioisotope concentration captured within the study period? If so, do the results represent known half-life decomposition for each radioisotope type?
+The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019.
+These laws require FEMA to publicly provide the following information for each mission assignment:
 
-2. What is the geospatial dispersion of fallout radiation captured through this publicly available sampling dataset?
+When disasters occur, one of the most complicated and difficult challenges becomes financing the response and recovery efforts.
+In 1988, United States (US) President Ronald Reagan signed the Robert T.
+Stafford Disaster Relief and Emergency Act into law to formalize the financial responsibilities of each tier of US government.
+Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government.
+That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither.
+FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type.
+However, the federal government does not necessarily incur all of the cost of these resources.
+Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred.
+However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request.
+The accumulation of this process, including characteristics of the incident itself, are extraordinarily complex, and therefore make it difficult to predict the cost and true impact.
 
-3. Does the geospatial dispersion reported in the dataset match commonly accepted models for nuclear reactor meltdown? (Pending approval)
+Moreover, the combination of rapidly increasing inflation and worsening effects of climate change has caused disasters to become more frequent and devastating on an annual basis, and the financial impacts are becoming increasingly difficult to absorb within the United States (US) government infrastructure.
+It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices.
+Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster.
+As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster.
+In particular, a decision tree, bagged tree, and random forest model will be used to predict obligated and requested FEMA funds.
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
-The raw data was obtained from [this online repository](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data), but it is unclear from where the original poster obtained this data.
-The data were then imported into [RStudio 4.1](https://www.rstudio.com/), and the [tidyverse](https://www.tidyverse.org/) package was used to adjust variable formats and create subsets for future analyses.
-In lieu of using the calendar date for the sample collection, a new variable representing the number of days since the Chernobyl meltdown was created.
-This decision was motivated by the fact that radioactive decay is a relative measure of elapsed time rather than a calendar date.
-The [geosphere](https://cran.r-project.org/web/packages/geosphere/geosphere.pdf) package was used in a different feature engineering application to determine the distance between the Chernobyl nuclear power plant and each of the sampling locations.
-The full R code for data import and cleaning is available in the supplementary information.
+The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website.
+The "Mission Assignment" data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory.
+Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.
+
+The data obtained from the US Department of Agriculture Economic Research was utilized to generate state population estimates and the number of counties per state.
+
+The data summarizing disaster declarations was cleaned to represent the unique declarations per state
 

### 2021-11-13 code/analysis_code/analysis-eda.Rmd
Manuscript updates

@@
                     gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
-                    gtsummary::modify_caption("**Table 1. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
+                    gtsummary::modify_caption("**Table 4. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
                     gtsummary::bold_labels()

### 2021-11-13 code/analysis_code/analysis-linearmodels.Rmd
Manuscript updates

@@
 req_SLR_bp
+
+#save image
+req_SLR_bp_file = here("results","req_SLR_bp.png")
+ggsave(filename = req_SLR_bp_file, plot = req_SLR_bp)
 ```
@@
 req_MVR_fit_bp2
+
+#save file
+req_MVR_bp_file = here("results","req_MVR_bp.png")
+ggsave(filename = req_MVR_bp_file, plot = req_MVR_fit_bp2)
 ```
@@
 obl_SLR_bp
+
+#save image
+obl_SLR_bp_file = here("results","obl_SLR_bp.png")
+ggsave(filename = obl_SLR_bp_file, plot = obl_SLR_bp)
 ```
@@
 obl_MVR_fit_bp2
+
+#save file
+obl_MVR_bp_file = here("results","obl_MVR_bp.png")
+ggsave(filename = obl_MVR_bp_file, plot = obl_MVR_fit_bp2)
 ```

### 2021-11-13 code/processing_code/processing-populations.Rmd
Manuscript updates

@@
 This markdown imports the raw data and performs some cleaning as well as feature engineering.
-The raw data comes from  [the U.S> Department of Agriculture Economic Research Service](https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/).
+The raw data comes from  [the U.S.
+Department of Agriculture Economic Research Service](https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/).
 

### 2021-11-13 products/manuscript/Manuscript_Disasters.Rmd
Manuscript updates

@@
 # Summary/Abstract
-_Write a summary of your project._
+Rapidly increasing inflation rates and worsening effects of climate change are dramatically increasing the financial impacts of disasters.
+Within the United States, the Federal Emergency Management Agency (FEMA) is responsible for providing federal resources to support states and territories during the response phase as well as tracking the financial costs.
+All tiers of government would benefit from knowing which disaster declaration characteristics would change the amount of financial support from FEMA.
+This analysis utilizes machine learning technology to predict the funds requested by states or territories from FEMA as well as the funds eventually obligated to states or territories from FEMA for a disaster declaration.
+In particular, three models are considered: decision tree, bagged tree, and random forest.
+Cross-validation and bootstrap resampling structures are utilized, depending on the model, and the fits of the models are evaluated using the root mean square error (RMSE) statistic.
+The bagged tree model employing bootstrap resampling best fit both the requested and obligated fund models; however, future studies should consider other machine learning models as well as further refinement of the tuning paramters for the ones considered in this analysis.
+The prediction capabilities created by such models have the potential to dramatically improve comprehension of the key factors affecting financial disaster relief support.
 
@@
 
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
-
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-
-## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
-
-The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019.
-These laws require FEMA to publicly provide the following information for each mission assignment:
-
-When disasters occur, one of the most complicated and difficult challenges becomes financing the response and recovery efforts.
-In 1988, United States (US) President Ronald Reagan signed the Robert T.
-Stafford Disaster Relief and Emergency Act into law to formalize the financial responsibilities of each tier of US government.
+On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T.
+Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government.
 Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government.
@@
 However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request.
-The accumulation of this process, including characteristics of the incident itself, are extraordinarily complex, and therefore make it difficult to predict the cost and true impact.
+The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019.
+These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration.
+As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments.
+While this has improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.
 
-Moreover, the combination of rapidly increasing inflation and worsening effects of climate change has caused disasters to become more frequent and devastating on an annual basis, and the financial impacts are becoming increasingly difficult to absorb within the United States (US) government infrastructure.
 It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices.
@@
 As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster.
-In particular, a decision tree, bagged tree, and random forest model will be used to predict obligated and requested FEMA funds.
+In particular, a decision tree, bagged tree, and random forest model will be used to predict requested and obligated FEMA funds.
 
@@
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
 The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website.
-The "Mission Assignment" data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory.
+The mission assignments data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory.
 Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.
-
-The data obtained from the US Department of Agriculture Economic Research was utilized to generate state population estimates and the number of counties per state.
-
-The data summarizing disaster declarations was cleaned to represent the unique declarations per state
+Similarly, the disaster declaration summaries were also summarized to represent the unique disaster declarations for each state, bounded to the same time frame as the mission assignments data.
+Feature engineering was performed to create variables for incident duration, response duration, month of declaration, and year of the incident.
+The population data from the US Department of Agriculture Economic Research Service was incorporated into the disaster summary data, after creating a variable for state.
+The mission assignment data and declaration data were then joined to create a comprehensive dataset that was used for analysis.
+The supplementary files contain the RStudio 4.1 code used to process the data.
+The following R packages were utilized for importing and cleaning the data: here, tidyverse, skimr, and lubridate.
 
 ## Exploratory analysis
+The processed data were then explored to characterize and summarize the relevant features in RStudio 4.1 with the following packages: here, tidyverse, skimr, summarytools, ggplot2, car, gtsummary, scales, cowplot, and table1.
+The supplementary files contain the code that generates the following results.
 
-_Use a combination of text/tables/figures to explore and describe your data.
-You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-FIgures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data._
+Table \@ref(tab:table1) reflects the results of the exploratory data analysis.
+Between February 2012 and November 2021, there were 446 disaster declarations for states and territories.
+Funds requested by states or territories for FEMA ranged from -\$855,200.3 to \$4,388,540,486, with an average of \$29,673,687, and FEMA funds obligated to the states and territories after reconciliation ranged from -\$855,200.3 to \$440,842,4297 with an average of \$29,744,109.
 
-_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-Plots produced here can be scatterplots, boxplots, violinplots, etc.
-Tables can be simple 2x2 tables or larger ones._
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-A total 2,051 observations were included in the original dataset, but 20 were removed for having no concentration recorded for any of the radioisotopes in question.
-
-Table \@ref(tab:isotope_table) describes the three radioisotopes captured in the data.
-
-```{r isotopetable,  echo=FALSE}
-#load data for entire document
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","processeddata.rds")
-data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")
-
-#load data
-mydata <- readRDS(data_location)
-mydata_long <- readRDS(data_location2)
-
-#define row labels
-table1::label(mydata$Iodine131) <- "Iodine 131 (Bq/m^3)"
-table1::label(mydata$Caesium134) <- "Caesium 134 (Bq/m^3)"
-table1::label(mydata$Caesium137) <- "Caesium 137 (Bq/m^3)"
-
-#now combine into table
-table1::table1(~ Iodine131 + Caesium134 + Caesium137, data = mydata)
+```{r summarytable,  echo=FALSE, message = FALSE}
+readRDS("../../results/table1.rds")
 ```
 
-Figure \@ref(fig:radioisotopesdays_50) depicts the relationship between the first 50 days after the incident and the measured concentration of each of the three radioisotopes.
-Radioactive decay is often used as an example of exponential decay; however, this pattern is not represented with this data.
-This phenomenon can be explained by the fact that fallout radiation instead follows a Poisson distribution pattern, due to the geographical dispersion of the particles.
+Figure \@ref(fig:incidentyears) shows the number of disaster declarations per year, and Figure \@ref(fig:funds-years) shows the total FEMA funds (both requested and obligated) by states per year.
+The substantial peak in 2020 corresponds to the COVID-19 pandemic, and the one in 2017 represents the hurricane season with Hurricanes Irma, Harvey, and Maria.
+These characteristics can also be seen in Table \@ref(tab:decs-year-3), which represents FEMA Funding for the 3 Costliest Years (2012 - 2021).
 
-```{r radioisotopesdays_50,  fig.cap='Radioisotope concentrations in European countries by days after Chernobyl..', echo=FALSE}
-knitr::include_graphics("../../results/radioisotopesdays_50.png")
-```
-Fallout radioisotope concentrations were collected from 96 different locations across 96 countries.
-
-Figure \@ref(fig:i131samplelocations) depicts the relationship between time elapsed and the measured concentration of I-131 across the different countries.
-Figures \@ref(fig:cs134samplelocations) and \@ref(fig:cs137samplelocations) are identical representations but for Cs-134 and Cs-137 respectively.
-
-```{r i131locations,  fig.cap='Concentration of I-131 in European countries by days after Chernobyl.', echo=FALSE}
-knitr::include_graphics("../../results/i131samplelocations.png")
+```{r incidentyears,  fig.cap='Number of disaster declarations per year 2012 - 2021', echo=FALSE, message = FALSE}
+knitr::include_graphics("../../results/incidentyear.png")
 ```
 
-```{r cs134locations,  fig.cap='Concentration of Cs-134 in European countries by days after Chernobyl.', echo=FALSE}
-knitr::include_graphics("../../results/cs134samplelocations.png")
+```{r funds-years,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by year 2012 - 2021', echo=FALSE, message = FALSE}
+knitr::include_graphics("../../results/funds-years.png")
 ```
 
-```{r cs137locations,  fig.cap='Concentration of Cs-137 in European countries by days after Chernobyl.', echo=FALSE}
-knitr::include_graphics("../../results/cs137samplelocations.png")
+```{r decs-year-3,  echo=FALSE, message = FALSE}
+readRDS("../../results/decs-year-3.Rds")
+```
+
+Figure \@ref(fig:incidentmonth) shows the number of disaster declarations per month, and Figure \@ref(fig:funds-month) shows the total FEMA funds (both requested and obligated) by states per month.
+The substantial peak in January likely corresponds to the COVID-19 pandemic, and the ones in August, September, and October represent the peak hurricane season.
+These characteristics can also be seen in Table \@ref(tab:decs-month-3), which represents FEMA Funding for the 3 Costliest Months.
+
+```{r incidentmonth,  fig.cap='Number of disaster declarations per month for years 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/incidentmonth.png")
+```
+
+```{r funds-month,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by month for 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/funds-month.png")
+```
+
+```{r decs-month-3,  echo=FALSE, message = FALSE}
+readRDS("../../results/decs-month-3.Rds")
+```
+
+All 50 states as well as Washington, DC were included in the final, combined dataset.
+Table \@ref(tab:decs-state-5) provides the five most disaster-prone states and summarizes their requested FEMA funding and received FEMA funding over the study period.
+```{r decs-state-5,  echo=FALSE, message = FALSE}
+readRDS("../../results/decs-state-5.Rds")
+```
+
+Incident duration ranged from zero days to 660 days (or almost two years), as seen by Figure \@ref(fig:funds-dur).
+The incident duration of 660 days likely corresponds to the COVID-19 pandemic, which is an ongoing incident of almost two years.
+Further analysis of all variables can be found in the supplementary information.
+
+```{r funds-dur,  fig.cap='Incident duration for requested FEMA funds and obligated FEMA funds under $1 billion for 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/funds-dur.png")
+```
+
+Prior to fitting the data to machine learning models, standard simple and multivariable regression models were fitted using the tidymodels framework in RStudio 4.1.
+Both outcomes of interest were fitted, using incident duration as the primary predictor of interest.
+Figure \@ref(fig:req_MVR_bp) and \@ref(fig:obl_MVR_bp) show the significant variables fitted in the multivariable regression models for requested and obligated FEMA funds, respectively.
+
+```{r req_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/req_MVR_bp.png")
+```
+
+```{r obl_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/obl_MVR_bp.png")
+```
+
+Table \@ref(tab:lm-eval) shows the comparisons of the models, using the root mean square error (RMSE) and r-squared statistics.
+The requested and obligated funding models performed almost identically.
+None of the models fit the data well, which suggests more sophisticated modeling practices are necessary to adequately characterize the data.
+
+```{r lm-eval,  echo=FALSE, message = FALSE}
+resulttable2=readRDS("../../results/lm-eval.rds")
+knitr::kable(resulttable2, caption = 'Linear model comparisons')
 ```
@@
 
+Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g.
+requested and obligated funds).
+Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed.
+Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category.
+The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type.
+Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type.
+The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type.
+The models were then compared based on the root mean square error (RMSE), the standard metric for regression models.
+The analysis was completed in RStudio 4.1 with the following packages: here, tidyverse, tidymodels, skmir, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.
+
+The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot) and  \@ref(fig:DT-plot-obl).
+Both outcomes identified the same decision point features: total number of federal agencies involved in the response, duration of the incident, state population, state counties, and whether the individual / housing assistance program was awarded.
+
+```{r DT-plot,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/DT-plot.png")
+```
+
+```{r DT-plot-obl,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/DT-plot-obl")
+```
+
+The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot) and  \@ref(fig:RF-plot-obl).
+Both outcomes identified the same important features: incident duration, total number of agencies involved, individual/housing assistance program awardance, and incident month.
+Interestingly, however, the obligated funds model includes days of the entire response as the fifth most important feature, whereas the requested funds model includes it as the eleventh most important feature.
+Unfortunately, the tidymodels framework does not support visualization of the bagged tree model.
+
+```{r RF-plot,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot.png")
+```
+
+```{r RF-plot-obl,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot-obl")
+```
+
+Table \@ref(tab:ML-model-comp) shows the model comparisons for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding.
+In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.
+
+```{r ML-model-comp,  echo=FALSE, message = FALSE}
+resulttable=readRDS("../../results/ML-model-comp.rds")
+knitr::kable(resulttable, caption = 'ML model comparison for requested FEMA funding')
+```
+
+```{r ML-model-comp-obl,  echo=FALSE}
+resulttable1=readRDS("../../results/ML-model-comp-obl.rds")
+knitr::kable(resulttable1, caption = 'ML model comparison for obligated FEMA funding')
+```
+
+The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample.
+The RMSEs for the testing and training datasets were substantially different.
+Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds.
+The supplementary information contains the code for the models as well as all diagnostic plots.
+```{r BT_pred_act_test,  fig.cap='Predicted vs actual requested FEMA funds for the bagged tree fit', echo=FALSE}
+knitr::include_graphics("../../results/BT-pred-act-test.png")
+```
+
+```{r BT_pred_act_test-obl,  fig.cap='Predicted vs actual obligated FEMA funds for the bagged tree fit', echo=FALSE}
+knitr::include_graphics("../../results/BT-pred-act-test-obl.png")
+```
+
 # Discussion
+This analysis applies machine learning modeling techniques to predict the requested and obligated funds from FEMA in the aftermath of a disaster.
 
@@
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+This analysis shows promise but requires further tuning and validation.
+THe usefulness of the model is limited by the broad variance present in the relatively small dataset, which potentially leads to biases.
+Moreover, the COVID-19 pandemic created extensive outliers in the dataset, which further complicate the prediction algorithms.
+While it is tempting to remove the outlier from a statistics perspective, it makes less sense scientifically, as the outlier disasters are often the ones where funding prediction would be the most useful.
 
@@
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@Leek2015a] discusses types of analyses.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
 # References

### 2021-11-28 code/analysis_code/analysis-ML-OA-outlier.Rmd
Outlier Analysis

--- 2021-11-12 code/analysis_code/analysis-ML-OA.Rmd
@@
 ---
-title: "Analysis: Machine Learning - OblAmt"
+title: "Analysis: Machine Learning - OblAmt - Outliers"
 output: 
@@
 
-It compares the  models, and then finally fits the “best” model to the test data.
+It compares the  models, and then finally fits the “best” model to the test data. It removes the $4B outlier that corresponds to Hurricane Maria in Puerto Rico in 2017.
 
@@
 * Transform Incident Month into ordinal variable
+* Remove outlier
 * Set the random seed to 123
@@
 
+#remove Hurricane Maria outlier
+obl_data <-subset(obl_data, OblAmt< 4000000000)
+
 #set random seed
@@
 
-#use the recipe specified earlier (line 133)
+#use the recipe specified earlier
 ```
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-          parsnip::set_engine("rpart",
-                              times = 10) %>%
-          parsnip::set_mode("regression") %>%
-          tune::tune_grid(preprocessor = BT_rec,
-                          resamples = obl_data_bootsplit,
+BT_res <- BT_wflow %>%
+          tune::tune_grid(resamples = obl_data_bootsplit,
                           grid = tree_grid,
-                          metrics = metric_set(rmse))
+                          metrics = metric_set(rmse),
+                          control = control_resamples(save_pred = TRUE,
+                                                      extract = function(x) extract_model(x)))
 
@@
 
-#vip package doesn't support BT
+#bag roots function to extract fit
+bag_roots <-  function(x){
+  x %>% 
+    dplyr::select(.extracts) %>% 
+    unnest(cols = c(.extracts)) %>% 
+    dplyr::mutate(models = map(.extracts,
+                               ~.x$model_df)) %>% 
+    dplyr::select(-.extracts) %>% 
+    unnest(cols = c(models)) %>% 
+    mutate(root = map_chr(model,
+                          ~as.character(.x$fit$frame[1, 1]))) %>%
+    dplyr::select(root)  
+}
+
+#plot most important variables
+bag_roots(BT_res) %>% 
+  ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
+  geom_bar() + 
+  coord_flip() + 
+  labs(x = "root", y = "count")
 

### 2021-11-28 code/analysis_code/analysis-ML-RA-outlier.Rmd
Outlier Analysis

--- 2021-11-12 code/analysis_code/analysis-ML-OA.Rmd
@@
 ---
-title: "Analysis: Machine Learning - OblAmt"
+title: "Analysis: Machine Learning - ReqAmt"
 output: 
@@
 
-This script uses obligated FEMA funds as the outcome of interest and fits the following models to the analysis data:
+This script uses requested FEMA funds as the outcome of interest and fits the following models to the analysis data:
 
@@
 #requested amount filter
-obl_data <- analysisdata %>%
+req_data <- analysisdata %>%
               dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)
+                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
 
 #transform incident month into ordinal variable
-obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
+req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
                                   labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
@@
 #not enough values per state, so create a variable that is FEMA Regions
-obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
-                                 obl_data$state == "ME" |
-                                 obl_data$state == "MA" |
-                                 obl_data$state == "NH" |
-                                 obl_data$state == "RI" |
-                                 obl_data$state == "VT"),
+req_data$FEMARegion <- ifelse((req_data$state == "CT" |
+                                 req_data$state == "ME" |
+                                 req_data$state == "MA" |
+                                 req_data$state == "NH" |
+                                 req_data$state == "RI" |
+                                 req_data$state == "VT"),
                               "Region I",
-                              ifelse((obl_data$state == "NJ" |
-                                         obl_data$state == "NY" |
-                                         obl_data$state == "PR" |
-                                         obl_data$state == "VI"),
+                              ifelse((req_data$state == "NJ" |
+                                         req_data$state == "NY" |
+                                         req_data$state == "PR" |
+                                         req_data$state == "VI"),
                                       "Region II",
-                                       ifelse((obl_data$state == "DE" |
-                                                 obl_data$state == "MD" |
-                                                 obl_data$state == "PA" |
-                                                 obl_data$state == "VA" |
-                                                 obl_data$state == "DC" |
-                                                 obl_data$state == "WV"),
+                                       ifelse((req_data$state == "DE" |
+                                                 req_data$state == "MD" |
+                                                 req_data$state == "PA" |
+                                                 req_data$state == "VA" |
+                                                 req_data$state == "DC" |
+                                                 req_data$state == "WV"),
                                               "Region III",
-                                               ifelse((obl_data$state == "AL" |
-                                                         obl_data$state == "FL" |
-                                                         obl_data$state == "GA" |
-                                                         obl_data$state == "KY" |
-                                                         obl_data$state == "MS" |
-                                                         obl_data$state == "NC" |
-                                                         obl_data$state == "SC" |
-                                                         obl_data$state == "TN"),
+                                               ifelse((req_data$state == "AL" |
+                                                         req_data$state == "FL" |
+                                                         req_data$state == "GA" |
+                                                         req_data$state == "KY" |
+                                                         req_data$state == "MS" |
+                                                         req_data$state == "NC" |
+                                                         req_data$state == "SC" |
+                                                         req_data$state == "TN"),
                                                       "Region IV",
-                                                      ifelse((obl_data$state == "IL" |
-                                                               obl_data$state == "IN" |
-                                                               obl_data$state == "MI" |
-                                                               obl_data$state == "MN" |
-                                                               obl_data$state == "OH" |
-                                                               obl_data$state == "WI"),
+                                                      ifelse((req_data$state == "IL" |
+                                                               req_data$state == "IN" |
+                                                               req_data$state == "MI" |
+                                                               req_data$state == "MN" |
+                                                               req_data$state == "OH" |
+                                                               req_data$state == "WI"),
                                                             "Region V",
-                                                             ifelse((obl_data$state == "AR" |
-                                                                         obl_data$state == "LA" |
-                                                                         obl_data$state == "NM" |
-                                                                         obl_data$state == "OK" |
-                                                                         obl_data$state == "TX"),
+                                                             ifelse((req_data$state == "AR" |
+                                                                         req_data$state == "LA" |
+                                                                         req_data$state == "NM" |
+                                                                         req_data$state == "OK" |
+                                                                         req_data$state == "TX"),
                                                                       "Region VI",
-                                                                       ifelse((obl_data$state == "IA" |
-                                                                                 obl_data$state == "KS" |
-                                                                                 obl_data$state == "MO" |
-                                                                                 obl_data$state == "NE"),
+                                                                       ifelse((req_data$state == "IA" |
+                                                                                 req_data$state == "KS" |
+                                                                                 req_data$state == "MO" |
+                                                                                 req_data$state == "NE"),
                                                                               "Region VII",
-                                                                               ifelse((obl_data$state == "CO" |
-                                                                                         obl_data$state == "MT" |
-                                                                                         obl_data$state == "ND" |
-                                                                                         obl_data$state == "SD" |
-                                                                                         obl_data$state == "UT" |
-                                                                                         obl_data$state == "WY"),
+                                                                               ifelse((req_data$state == "CO" |
+                                                                                         req_data$state == "MT" |
+                                                                                         req_data$state == "ND" |
+                                                                                         req_data$state == "SD" |
+                                                                                         req_data$state == "UT" |
+                                                                                         req_data$state == "WY"),
                                                                                       "Region VIII",
-                                                                                       ifelse((obl_data$state == "AZ" |
-                                                                                                 obl_data$state == "CA" |
-                                                                                                 obl_data$state == "HI" |
-                                                                                                 obl_data$state == "NV"),
+                                                                                       ifelse((req_data$state == "AZ" |
+                                                                                                 req_data$state == "CA" |
+                                                                                                 req_data$state == "HI" |
+                                                                                                 req_data$state == "NV"),
                                                                                               "Region IX",
-                                                                                               ifelse((obl_data$state == "AK" |
-                                                                                                         obl_data$state == "ID" |
-                                                                                                         obl_data$state == "OR" |
-                                                                                                         obl_data$state == "WA"),
+                                                                                               ifelse((req_data$state == "AK" |
+                                                                                                         req_data$state == "ID" |
+                                                                                                         req_data$state == "OR" |
+                                                                                                         req_data$state == "WA"),
                                                                                                       "Region X",
@@
 #drop state variable
-obl_data <- obl_data %>%
+req_data <- req_data %>%
               dplyr::select(-c(state))
@@
 #add to "other" category for incident Types that are infrequent
-obl_data$incidentType[(obl_data$incidentType == "Chemical" |
-                        obl_data$incidentType == "Coastal Storm" |
-                        obl_data$incidentType == "Dam/Levee Break" |
-                        obl_data$incidentType == "Mud/Landslide" |
-                        obl_data$incidentType == "Volcano" |
-                        obl_data$incidentType == "Terrorist" |
-                        obl_data$incidentType == "Snow")] <- "Other"
+req_data$incidentType[(req_data$incidentType == "Chemical" |
+                        req_data$incidentType == "Coastal Storm" |
+                        req_data$incidentType == "Dam/Levee Break" |
+                        req_data$incidentType == "Mud/Landslide" |
+                        req_data$incidentType == "Volcano" |
+                        req_data$incidentType == "Terrorist" |
+                        req_data$incidentType == "Snow")] <- "Other"
+
+#remove Hurricane Maria outlier
+req_data <-subset(req_data, ReqAmt< 4000000000)
 
@@
 #use incidentType as stratification
-obl_data_split <- rsample::initial_split(obl_data, prop = 2/3,
+req_data_split <- rsample::initial_split(req_data, prop = 2/3,
                                      strata = incidentType)
@@
 #create dataframes for the two sets:
-obl_train_data <- rsample::training(obl_data_split)
-obl_test_data <- rsample::testing(obl_data_split)
+req_train_data <- rsample::training(req_data_split)
+req_test_data <- rsample::testing(req_data_split)
 
 #5-fold cross validation, repeated 5 times, stratified by incidentType
-obl_folds <- rsample::vfold_cv(obl_train_data,
+req_folds <- rsample::vfold_cv(req_train_data,
                                strata = incidentType,
@@
 #create recipe that codes categorical variables as dummy variables
-OA_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
+RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
           recipes::step_dummy(all_nominal_predictors())
@@
 null_wflow <- workflows::workflow() %>%
-              workflows::add_recipe(OA_rec) %>%
+              workflows::add_recipe(RA_rec) %>%
               workflows::add_model(null_mod)
@@
 #"fit" model to training data
-obl_null_train <- null_wflow %>%
-                  parsnip::fit(data = obl_train_data)
+req_null_train <- null_wflow %>%
+                  parsnip::fit(data = req_train_data)
 
 #summary of null model with training data to get mean (which in this case is the RMSE)
-obl_null_train_sum <- broom.mixed::tidy(obl_null_train)
-obl_null_train_sum
+req_null_train_sum <- broom.mixed::tidy(req_null_train)
+req_null_train_sum
 
 #RMSE for training data for formatting later
-obl_null_RMSE_train <- tibble::tibble(
-                          rmse = rmse_vec(truth = obl_train_data$OblAmt,
-                                              estimate = rep(mean(obl_train_data$OblAmt), nrow(obl_train_data))),
+req_null_RMSE_train <- tibble::tibble(
+                          rmse = rmse_vec(truth = req_train_data$ReqAmt,
+                                              estimate = rep(mean(req_train_data$ReqAmt), nrow(req_train_data))),
                           SE = 0,
@@
 #"fit" model to test data
-obl_null_test <- null_wflow %>%
-                  parsnip::fit(data = obl_test_data)
+req_null_test <- null_wflow %>%
+                  parsnip::fit(data = req_test_data)
 
 #summary of null model with test data to get mean (which in this case is the RMSE)
-obl_null_test_sum <- broom.mixed::tidy(obl_null_test)
-obl_null_test_sum
+req_null_test_sum <- broom.mixed::tidy(req_null_test)
+req_null_test_sum
 
 #RMSE for testing data for formatting later
-obl_null_RMSE_test <- tibble::tibble(
-                          rmse = rmse_vec(truth = obl_test_data$OblAmt,
-                                              estimate = rep(mean(obl_test_data$OblAmt), nrow(obl_test_data))),
+req_null_RMSE_test <- tibble::tibble(
+                          rmse = rmse_vec(truth = req_test_data$ReqAmt,
+                                              estimate = rep(mean(req_test_data$ReqAmt), nrow(req_test_data))),
                           SE = 0,
@@
                workflows::add_model(tree_mod) %>%
-               workflows::add_recipe(OA_rec)
+               workflows::add_recipe(RA_rec)
 ```
@@
 tree_res <- tree_wflow %>%
-                tune::tune_grid(resamples = obl_folds,
+                tune::tune_grid(resamples = req_folds,
                                 grid = tree_grid,
@@
 #save file
-DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
+DT_auto_fig_file = here("results","DT-auto-plot.png")
 ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)
@@
 #save file
-DT_detail_fig_file = here("results","DT-det-plot-obl.png")
-ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)
+DT_det_fig_file = here("results","DT-det-plot.png")
+ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)
 
@@
 best_tree_fit <- best_tree_wflow %>%
-                    parsnip::fit(data = obl_train_data)
+                    parsnip::fit(data = req_train_data)
 ```
@@
                                                      main = "Selected Decision Tree Model")
-#save file manually -- ggsave throws an error
+#save manually -- ggsave throws an error
 
@@
 tree_resid <- best_tree_fit %>%
-                  broom.mixed::augment(new_data = obl_train_data) %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                  broom.mixed::augment(new_data = req_train_data) %>%
+                  dplyr::select(.pred, ReqAmt) %>%
+                  dplyr::mutate(.resid = .pred - ReqAmt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = OblAmt)) +
+DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = ReqAmt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                         geom_point() +
-                        labs(title = "Decision Tree Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        labs(title = "Decision Tree Fit: Predicted vs. Actual Requested Funds",
+                              x = "Predicted Requested Funds ($)",
+                              y = "Actual Requested Funds ($)")
 DT_pred_act
@@
 #save file
-DT_pred_act_file = here("results","DT-pred-act-obl.png")
+DT_pred_act_file = here("results","DT-pred-act.png")
 ggsave(filename = DT_pred_act_file, plot = DT_pred_act)
@@
 #save file
-DT_resid_file = here("results","DT-resid-obl.png")
+DT_resid_file = here("results","DT-resid.png")
 ggsave(filename = DT_resid_file, plot = DT_resid)
@@
                           geom_point() +
-                          labs(title = "Decision Tree Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          labs(title = "Decision Tree Fit: Residuals vs Fitted Requested Funds",
+                                x = "Predicted Requested Funds ($)",
                                 y = "Residual")
@@
 #save file
-DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
+DT_pred_resid_file = here("results","DT-pred-resid.png")
 ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)
@@
                   model = "Tree") %>%
-               dplyr::bind_rows(obl_null_RMSE_train)
+               dplyr::bind_rows(req_null_RMSE_train)
 tree_RMSE
 ```
+
+The identified best performing tree model has some interesting results.
+It identifies incident duration, state population, total number of federal agencies, total number of state counties, and individual/households program awardance as significant decision points for predicting requesting FEMA funding.
+All of these variables are to be expected, but it is interesting that incident type or declaration type are not included as significant.
+The RMSE is improved from the Null model, but there's still opportunity for improvement.
+Let's try another model.
 
@@
                workflows::add_model(RF_mod) %>%
-               workflows::add_recipe(OA_rec)
+               workflows::add_recipe(RA_rec)
 ```
@@
 RF_res <- RF_wflow %>%
-              tune::tune_grid(resamples = obl_folds,
+              tune::tune_grid(resamples = req_folds,
                               grid = RF_grid,
@@
 #save file
-RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
+RF_auto_fig_file = here("results","RF-auto-plot.png")
 ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
@@
 best_RF_fit <- best_RF_wflow %>%
-                    parsnip::fit(data = obl_train_data)
+                    parsnip::fit(data = req_train_data)
 ```
@@
 
-#save  manually - ggsave throws an error
+#save manually-- ggsave throws errors
 
@@
 RF_resid <- best_RF_fit %>%
-                broom.mixed::augment(new_data = obl_train_data) %>%
-                dplyr::select(.pred, OblAmt) %>%
-                dplyr::mutate(.resid = .pred - OblAmt)
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, ReqAmt) %>%
+                dplyr::mutate(.resid = .pred - ReqAmt)
 
@@
 #geom_abline is a 45 degree line, along which the results should fall
-RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
+RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = ReqAmt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                         geom_point() +
-                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        labs(title = "RF Fit: Actual vs. Predicted Requested Funds",
+                              x = "Predicted Requested Funds ($)",
+                              y = "Actual Requested Funds ($)")
+RF_pred_act
 
 #save file
-RF_pred_act_file = here("results","RF-pred-act-obl.png")
+RF_pred_act_file = here("results","RF-pred-act.png")
 ggsave(filename = RF_pred_act_file, plot = RF_pred_act)
@@
 #save file
-RF_resid_file = here("results","RF-resid-obl.png")
+RF_resid_file = here("results","RF-resid.png")
 ggsave(filename = RF_resid_file, plot = RF_resids)
@@
                           geom_point() +
-                          labs(title = "Random Forest Fit: Residuals vs Fitted Obligated Funds",
+                          labs(title = "Random Forest Fit: Residuals vs Fitted Requested Funds",
                                 x = "Fitted Requested Funds ($)",
@@
 #save file
-RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
+RF_pred_resid_file = here("results","RF-pred-resid.png")
 ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)
@@
 
+In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
+Interestingly, it includes declaration month of Jan - March as more important than state population.
+Let's try one more model.
+
 ---
@@
 #use bootstrap instead of CV
-obl_data_bootsplit <- obl_train_data %>%
+req_data_bootsplit <- req_train_data %>%
                         rsample::bootstraps(times = 10,
@@
 #create recipe that codes categorical variables as dummy variables
-BT_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
+BT_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
           step_string2factor(all_nominal())
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-          parsnip::set_engine("rpart",
-                              times = 10) %>%
-          parsnip::set_mode("regression") %>%
-          tune::tune_grid(preprocessor = BT_rec,
-                          resamples = obl_data_bootsplit,
+BT_res <- BT_wflow %>%
+          tune::tune_grid(resamples = req_data_bootsplit,
                           grid = tree_grid,
-                          metrics = metric_set(rmse))
+                          metrics = metric_set(rmse),
+                          control = control_resamples(save_pred = TRUE,
+                                                      extract = function(x) extract_model(x)))
 
@@
 #save file
-BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
+BT_auto_fig_file = here("results","BT-auto-plot.png")
 ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)
@@
 #save file
-BT_det_fig_file = here("results","BT-det-plot-obl.png")
+BT_det_fig_file = here("results","BT-det-plot.png")
 ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
@@
 best_BT_fit <- best_BT_wflow %>%
-                    parsnip::fit(data = obl_train_data)
+                    parsnip::fit(data = req_train_data)
 ```
@@
 ```{r}
-#extract model from final fit
-x_BT <- best_BT_fit$fit$fit$fit
+#bag roots function to extract most important variables
+bag_roots <-  function(x){
+  x %>% 
+    dplyr::select(.extracts) %>% 
+    unnest(cols = c(.extracts)) %>% 
+    dplyr::mutate(models = map(.extracts,
+                               ~.x$model_df)) %>% 
+    dplyr::select(-.extracts) %>% 
+    unnest(cols = c(models)) %>% 
+    mutate(root = map_chr(model,
+                          ~as.character(.x$fit$frame[1, 1]))) %>%
+    dplyr::select(root)  
+}
 
-#vip package doesn't support BT
+#plot most important variables
+bag_roots(BT_res) %>% 
+  ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
+  geom_bar() + 
+  coord_flip() + 
+  labs(x = "root", y = "count")
 
@@
 BT_resid <- best_BT_fit %>%
-                  broom.mixed::augment(new_data = obl_train_data) %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                  broom.mixed::augment(new_data = req_train_data) %>%
+                  dplyr::select(.pred, ReqAmt) %>%
+                  dplyr::mutate(.resid = .pred - ReqAmt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
+BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = ReqAmt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                         geom_point() +
-                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Requested Funds",
+                              x = "Predicted Requested Funds ($)",
+                              y = "Actual Requested Funds ($)")
 BT_pred_act
@@
 #save file
-BT_pred_act_file = here("results","BT-pred-act-obl.png")
+BT_pred_act_file = here("results","BT-pred-act.png")
 ggsave(filename = BT_pred_act_file, plot = BT_pred_act)
@@
 #save file
-BT_resid_file = here("results","BT-resids-obl.png")
+BT_resid_file = here("results","BT-resid.png")
 ggsave(filename = BT_resid_file, plot = BT_resids)
@@
                           geom_point() +
-                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Requested Funds",
+                                x = "Predicted Requested Funds ($)",
                                 y = "Residual")
@@
 #save file
-BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
+BT_pred_resid_file = here("results","BT-pred-resid.png")
 ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)
@@
 #save combined table
-model_comp_file = here("results", "ML-model-comp-obl.Rds")
+model_comp_file = here("results", "ML-model-comp.Rds")
 saveRDS(BT_RMSE, file = model_comp_file)
@@
 BT_fit_test <- best_BT_wflow %>%
-                    tune::last_fit(split = obl_data_split)
+                    tune::last_fit(split = req_data_split)
 
@@
                   broom.mixed::augment() %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                  dplyr::select(.pred, ReqAmt) %>%
+                  dplyr::mutate(.resid = .pred - ReqAmt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = OblAmt)) +
+BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = ReqAmt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                         geom_point() +
-                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Requested Funds",
+                              x = "Predicted Requested Funds ($)",
+                              y = "Actual Requested Funds ($)")
 BT_pred_act_test
@@
 #save file
-BT_pred_act_test_file = here("results","BT-pred-act-test-obl.png")
+BT_pred_act_test_file = here("results","BT-pred-act-test.png")
 ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)
@@
 #save file
-BT_resid_test_file = here("results","BT-resid-test-obl.png")
+BT_resid_test_file = here("results","BT-resid-test.png")
 ggsave(filename = BT_resid_test_file, plot = BT_resids_test)
@@
                           geom_point() +
-                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Requested Funds",
+                                x = "Predicted Requested Funds ($)",
                                 y = "Residual")
@@
 #save file
-BT_pred_resid_test_file = here("results","BT-pred-resid-test-obl.png")
+BT_pred_resid_test_file = here("results","BT-pred-resid-test.png")
 ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)

### 2021-11-28 code/processing_code/processing-populations.Rmd
Outlier Analysis

@@
 pops_1 <- populations_raw %>%
-                dplyr::select(FIPStxt,
+                dplyr::select(ï..FIPStxt,
                               State,
                               Area.name,
-                              Population.2020)
+                              Population.2020) %>%
+               dplyr::rename(FIPStxt = ï..FIPStxt)
 ```

### 2021-11-29 code/analysis_code/analysis-ML-OA-outlier.Rmd
Part 5 Update

@@
 #plot most important variables
-bag_roots(BT_res) %>% 
+BT_IV <- bag_roots(BT_res) %>% 
             ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
@@
             labs(x = "root", y = "count")
+BT_IV
+
+#save file
+BT_IV_file = here("results","BT-IV-obl-outlier.png")
+ggsave(filename = BT_IV_file, plot = BT_IV)
 

### 2021-11-29 code/analysis_code/analysis-ML-RA-outlier.Rmd
Part 5 Update

@@
 ---
-title: "Analysis: Machine Learning - ReqAmt"
+title: "Analysis: Machine Learning - ReqAmt - Outliers"
 output: 
@@
 
-#plot most important variables
-bag_roots(BT_res) %>% 
+BT_IV <- bag_roots(BT_res) %>% 
             ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
@@
             labs(x = "root", y = "count")
+BT_IV
+
+#save file
+BT_IV_file = here("results","BT-IV-outlier.png")
+ggsave(filename = BT_IV_file, plot = BT_IV)
 

### 2021-11-29 products/manuscript/Manuscript_Disasters.Rmd
started fixing references
Part 5 Update

@@
 ---
-
 # Summary/Abstract
@@
 In particular, three models are considered: decision tree, bagged tree, and random forest.
-Cross-validation and bootstrap resampling structures are utilized, depending on the model, and the fits of the models are evaluated using the root mean square error (RMSE) statistic.
-The bagged tree model employing bootstrap resampling best fit both the requested and obligated fund models; however, future studies should consider other machine learning models as well as further refinement of the tuning paramters for the ones considered in this analysis.
+Cross-validation and bootstrap re-sampling structures are utilized, depending on the model, and the fits of the models are evaluated using the root mean square error (RMSE) statistic.
+The bagged tree model employing bootstrap re-sampling best fit both the requested and obligated fund models; however, future studies should consider other machine learning models as well as further refinement of the tuning parameters for the ones considered in this analysis.
 The prediction capabilities created by such models have the potential to dramatically improve comprehension of the key factors affecting financial disaster relief support.
@@
 # Introduction
-
 On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T.
-Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government.
+Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government [@moss_stafford_2009].
 Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government.
-That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither.
+That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither [@mccarthy_femas_2009].
 FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type.
@@
 However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request.
-The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019.
+The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019 [@fema_guide_2021].
 These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration.
@@
 Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster.
+Modeling approaches that include machine learning algorithms have been demonstrated to have extensive prediction capability for large datasets.
 As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster.
@@
 
-# Methods and Results
+## Disaster Declaration Process
+When a disaster occurs, the onus for requesting assistance begins at the local government.
+The local officials must have enacted their own emergency operations plans and demonstrate the need for more resources than the local entities can provide.
+Once this threshold is crossed, the municipal or county leader submits a formal request to the state governor for assistance by requesting a state disaster declaration [@mccarthy_femas_2009].
+If the governor approves such a request, state resources are mobilized to assist the affected area, but the local government is still the primary leading entity for the emergency management response.
+If the state has fully enacted  their emergency response plans and further assistance is still required, the governor requests the state emergency management agency to conduct a state-level Preliminary Damage Assessment (PDA).
+These assessments aim to estimate the impact of the disaster on individuals, infrastructure, and government agencies [@moss_stafford_2009].
+As federal resources are extraordinarily expensive, state governors utilize PDAs to ensure federal assistance is only requested when absolutely necessary.
+If the governor does decide that federal resources are required, a formal request is submitted by the governor's office through the regional FEMA office.
+This request includes the PDA, barring a few exceptional circumstances for catastrophic events, and a list of requested federal resources, including the anticipated duration and cost of each resource [@moss_stafford_2009].
+The regional FEMA office ensures the request is complete before sending it to a special declaration processing unit in FEMA headquarters, which further add to the request by incorporating detailed statistics about the region and their own assessment of needs and costs [@mccarthy_femas_2009].
+The declaration processing unit then submits the request to the FEMA Administrator, who must make a recommendation for the President.
+The request and the recommendation are then submitted to the President, who ultimately decides whether or not to grant the declaration request.
+The President is not bound by the FEMA Administrator's recommendation, and, in the event the President denies the request, the governor is permitted to submit an appeal based on further damage assessments [@moss_stafford_2009].
+If the request is approved, FEMA becomes the primary agency to coordinate resource allocation and support the local response.
+The approval from the President also specifies which programs are authorized as well as the federal/state cost-sharing ratio [@fema_guide_2021].
 
-_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._
+## Machine Learning Models
+There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on three different types of tree models.
+This type of model was chosen as the primary users of the prediction algorithms are non-statisticians, and tree models are among the easiest machine learning models to interpret and understand [@james_introduction_2021].
 
-## Data import and cleaning
-The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website.
+* Decision Tree (DT): Also known as a classification and regression tree (CART), the DT examines each predictor and splits the outcome at a value of the predictor that leads to the best performance increase in the model.
+The tree development continues until a certain threshold criteria is met, such as number of observations in each leaf of the tree.
+A DT model is often the most intuitive machine learning algorithm, but it also typically has a reduced performance compared to other models [@boehmke_hands-machine_2019].
+
+* Bagged Tree (BT): A slightly more sophisticated model, a BT aims to reduce the variance by re-sampling the existing data (otherwise known as bootstrapping) and builds a tree for each sample.
+The final model then averages all of the individually sampled trees.
+However, the reducing the variance in the model also reduces the ease of interpretation of the model [@dietterich_experimental_nodate].
+
+* Random Forest (RF): A RF model follows similar principles to the BT, but instead of considering all possible predictors at each decision tree split, the split occurs based on the best predictors of a random sample of all predictors.
+This method allows for de-correlation in the trees by avoiding the default nature of decision trees that aim to include as many predictors as possible.
+Each individual tree is then averaged to find the final model.
+RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].
+
+# Materials and Methods
+
+## Data and Processing
+The data were obtained from the OpenFEMA Data Sets website [@noauthor_openfema_nodate] as well as the US Department of Agriculture Economic Research Service website [@noauthor_usda_nodate].
 The mission assignments data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory.
@@
 The mission assignment data and declaration data were then joined to create a comprehensive dataset that was used for analysis.
-The supplementary files contain the RStudio 4.1 code used to process the data.
-The following R packages were utilized for importing and cleaning the data: here, tidyverse, skimr, and lubridate.
 
-## Exploratory analysis
-The processed data were then explored to characterize and summarize the relevant features in RStudio 4.1 with the following packages: here, tidyverse, skimr, summarytools, ggplot2, car, gtsummary, scales, cowplot, and table1.
-The supplementary files contain the code that generates the following results.
+Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed.
+Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category.
+Additionally, the relief provided to Puerto Rico in the aftermath of Hurricane Maria in 2017 was removed as an outlier, as it was more than $2 billion more than the next largest allocation of funding.
 
-Table \@ref(tab:table1) reflects the results of the exploratory data analysis.
-Between February 2012 and November 2021, there were 446 disaster declarations for states and territories.
-Funds requested by states or territories for FEMA ranged from -\$855,200.3 to \$4,388,540,486, with an average of \$29,673,687, and FEMA funds obligated to the states and territories after reconciliation ranged from -\$855,200.3 to \$440,842,4297 with an average of \$29,744,109.
-
-```{r summarytable,  echo=FALSE, message = FALSE}
-readRDS("../../results/table1.rds")
-```
-
-Figure \@ref(fig:incidentyears) shows the number of disaster declarations per year, and Figure \@ref(fig:funds-years) shows the total FEMA funds (both requested and obligated) by states per year.
-The substantial peak in 2020 corresponds to the COVID-19 pandemic, and the one in 2017 represents the hurricane season with Hurricanes Irma, Harvey, and Maria.
-These characteristics can also be seen in Table \@ref(tab:decs-year-3), which represents FEMA Funding for the 3 Costliest Years (2012 - 2021).
-
-```{r incidentyears,  fig.cap='Number of disaster declarations per year 2012 - 2021', echo=FALSE, message = FALSE}
-knitr::include_graphics("../../results/incidentyear.png")
-```
-
-```{r funds-years,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by year 2012 - 2021', echo=FALSE, message = FALSE}
-knitr::include_graphics("../../results/funds-years.png")
-```
-
-```{r decs-year-3,  echo=FALSE, message = FALSE}
-readRDS("../../results/decs-year-3.Rds")
-```
-
-Figure \@ref(fig:incidentmonth) shows the number of disaster declarations per month, and Figure \@ref(fig:funds-month) shows the total FEMA funds (both requested and obligated) by states per month.
-The substantial peak in January likely corresponds to the COVID-19 pandemic, and the ones in August, September, and October represent the peak hurricane season.
-These characteristics can also be seen in Table \@ref(tab:decs-month-3), which represents FEMA Funding for the 3 Costliest Months.
-
-```{r incidentmonth,  fig.cap='Number of disaster declarations per month for years 2012 - 2021', echo=FALSE}
-knitr::include_graphics("../../results/incidentmonth.png")
-```
-
-```{r funds-month,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by month for 2012 - 2021', echo=FALSE}
-knitr::include_graphics("../../results/funds-month.png")
-```
-
-```{r decs-month-3,  echo=FALSE, message = FALSE}
-readRDS("../../results/decs-month-3.Rds")
-```
-
-All 50 states as well as Washington, DC were included in the final, combined dataset.
-Table \@ref(tab:decs-state-5) provides the five most disaster-prone states and summarizes their requested FEMA funding and received FEMA funding over the study period.
-```{r decs-state-5,  echo=FALSE, message = FALSE}
-readRDS("../../results/decs-state-5.Rds")
-```
-
-Incident duration ranged from zero days to 660 days (or almost two years), as seen by Figure \@ref(fig:funds-dur).
-The incident duration of 660 days likely corresponds to the COVID-19 pandemic, which is an ongoing incident of almost two years.
-Further analysis of all variables can be found in the supplementary information.
-
-```{r funds-dur,  fig.cap='Incident duration for requested FEMA funds and obligated FEMA funds under $1 billion for 2012 - 2021', echo=FALSE}
-knitr::include_graphics("../../results/funds-dur.png")
-```
-
-Prior to fitting the data to machine learning models, standard simple and multivariable regression models were fitted using the tidymodels framework in RStudio 4.1.
-Both outcomes of interest were fitted, using incident duration as the primary predictor of interest.
-Figure \@ref(fig:req_MVR_bp) and \@ref(fig:obl_MVR_bp) show the significant variables fitted in the multivariable regression models for requested and obligated FEMA funds, respectively.
-
-```{r req_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for requested FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/req_MVR_bp.png")
-```
-
-```{r obl_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for obligated FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/obl_MVR_bp.png")
-```
-
-Table \@ref(tab:lm-eval) shows the comparisons of the models, using the root mean square error (RMSE) and r-squared statistics.
-The requested and obligated funding models performed almost identically.
-None of the models fit the data well, which suggests more sophisticated modeling practices are necessary to adequately characterize the data.
-
-```{r lm-eval,  echo=FALSE, message = FALSE}
-resulttable2=readRDS("../../results/lm-eval.rds")
-knitr::kable(resulttable2, caption = 'Linear model comparisons')
-```
-
-## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
+## Model Development
 Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g.
 requested and obligated funds).
-Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed.
-Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category.
 The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type.
@@
 The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type.
-The models were then compared based on the root mean square error (RMSE), the standard metric for regression models.
-The analysis was completed in RStudio 4.1 with the following packages: here, tidyverse, tidymodels, skmir, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.
+Incident type was chosen as the stratification variable to address the fact that the frequency of incident types varies across FEMA regions.
 
-The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot) and  \@ref(fig:DT-plot-obl).
-Both outcomes identified the same decision point features: total number of federal agencies involved in the response, duration of the incident, state population, state counties, and whether the individual / housing assistance program was awarded.
+## Model Performance Evaluation
+The standard metric for evaluating regression models is the root mean square error (RMSE), which is the square root of the mean of the square of all error. It is formally defined as:
+$$
+RMSE = \sqrt{\frac {1}{n} \sum_{i = 1}^{n}{(S_i - O_i)^2}}
+$$
+where O_i are the observations, S_i are the predicted values of the outcome, and n is the number of observations in the analysis.
+RMSE is scale dependent, but the lower the RMSE, the better the model performance.
 
-```{r DT-plot,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/DT-plot.png")
-```
+## Software
+This analysis was conducted in Rstudio 4.1 on a Windows Server operating system [@r_core_team_r_2021].
+The following R packages were utilized: here, tidyverse, skimr, lubridate, tidymodels, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.
+All processing and analysis code can be found in the supplementary materials.
 
-```{r DT-plot-obl,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/DT-plot-obl")
-```
+# Results
+A total of 446 disaster declarations were included in this analysis, ranging from February 2010 to November 2021.
+On average, \$29,700,000 (SD = \$219,000,000) were requested and obligated in FEMA funding, with requests ranging from \-$855,000 to $4,390,000,000.
+All states and territories were represented in the analysis, and on average, the FEMA cost-share was 80.6% (SD = 35.4%).
+The year 2020 had more than double the average number of disaster declarations, which likely can be attributed to the COVID-19 pandemic.
+Further information about the distribution of the data can be found in the supplementary documentation.
 
-The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot) and  \@ref(fig:RF-plot-obl).
-Both outcomes identified the same important features: incident duration, total number of agencies involved, individual/housing assistance program awardance, and incident month.
-Interestingly, however, the obligated funds model includes days of the entire response as the fifth most important feature, whereas the requested funds model includes it as the eleventh most important feature.
-Unfortunately, the tidymodels framework does not support visualization of the bagged tree model.
-
-```{r RF-plot,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/RF-plot.png")
-```
-
-```{r RF-plot-obl,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/RF-plot-obl")
-```
-
-Table \@ref(tab:ML-model-comp) shows the model comparisons for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding.
+Three models were fit to predict requested and obligated FEMA funding in the event of a disaster declaration: decision tree, bagged tree, and random forest.
+Table \@ref(tab:ML-model-comp) shows the model comparison for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding.
 In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.
@@
 The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample.
-The RMSEs for the testing and training datasets were substantially different.
+Figure \@ref(fig:BT-plot-outlier) shows the identified most important variables in the bagged tree model for requested FEMA funds, and Figure \@ref(fig:BT-plot-obl-outlier) illustrates the identical principle but for obligated FEMA funds.
+In both models, total number of federal agencies involved, incident type, incident duration, state population, awardance of the individual / household assistance program, and FEMA region were identified as the most important predictors of FEMA funding.
+
+```{r BT-plot-outlier,  fig.cap='Most important variables to predict requested FEMA funds for the bagged tree fit', echo=FALSE}
+knitr::include_graphics("../../results/BT-IV-outlier.png")
+```
+
+```{r BT-plot-obl-outlier,  fig.cap='Most important variables to predict obligated FEMA funds for the bagged tree fit', echo=FALSE}
+knitr::include_graphics("../../results/BT-IV-obl-outlier.png")
+```
+
+The RMSEs for the testing and training datasets were substantially different, with an RMSE of 45003235 for the training data and 69378994 for the testing data.
 Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds.
@@
 # Discussion
-This analysis applies machine learning modeling techniques to predict the requested and obligated funds from FEMA in the aftermath of a disaster.
+In order to identify the most important predictors of requested and obligated FEMA funding in the aftermath of a disaster, three machine learning algorithms were fit to FEMA's disaster declaration data: decision tree, bagged tree, and random forest.
+Of the three models, the bagged tree model demonstrated the best performance with the lowest RMSE, the identified metric of evaluation.
+The bagged tree model identified total number of federal agencies involved, incident type, incident duration, state population, awardance of individual / household assistance program, and FEMA region.
+These variables are relatively expected, as the each of them plays a critical role in FEMA's decision of how much funding to award a state.
+However, it is interesting the state population is identified as such a critical predictor; it suggests that the number of individuals affected by a disaster plays a crucial role in the President's decision to provide assistance.
 
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+While this model does take a measured step towards identifying predictors for FEMA funding, there are a number of limitations to the study.
+First and foremost, the limited data available about disaster declarations leaves a number of important variables out of the analysis.
+Aspects such as the number of casualties or damage to critical infrastructure inevitably changes the amount of federal government assistance.
+However, no such data widely exists for casualties by federal disaster declaration.
+Moreover, the materials included in the recommendation to the President, such as the PDA, are protected by executive privilege and are therefore unavailable to the public.
+In order to improve the meaningful nature of these prediction models, such information should be incorporated.
 
-## Strengths and Limitations
-This analysis shows promise but requires further tuning and validation.
-THe usefulness of the model is limited by the broad variance present in the relatively small dataset, which potentially leads to biases.
-Moreover, the COVID-19 pandemic created extensive outliers in the dataset, which further complicate the prediction algorithms.
-While it is tempting to remove the outlier from a statistics perspective, it makes less sense scientifically, as the outlier disasters are often the ones where funding prediction would be the most useful.
+The distribution of federal government funding also provides a limitation to this analysis.
+The range is extensive, and exploratory results suggest the data follow a logarithmic curve rather than a linear increase.
+However, as there are disasters where the federal government does not provide any funding, so the state is responsible for covering the cost of federal resources.
+In these instances, the requested and obligated FEMA funding is reported as a negative cost, which cannot be log-transformed.
+Future analyses could divide the data into two analyses: one that addresses disasters where the federal government does provide cost-sharing support, and one that addresses disasters where the state provides all funding.
 
-## Conclusions
-_What are the main take-home messages?_
+Ultimately, this analysis is the first of its kind to apply prediction algorithms to federal funding for disaster declarations in the United States.
+It demonstrates the clear potential of such modeling capabilities not only to allow local and state emergency managers to predict which aspects of a PDA are critical in obtaining financial federal relief, but also to allow the regional FEMA offices to identify such predictors while developing the formal requests.
+Continued research in this area will support emergency managers of all levels in predicting financial disaster relief support.
 

### 2021-11-29 products/manuscript/Manuscript_Supplemental.Rmd
Part 5 Update

---
title: "Predicting Federal Emergency Management Agency Funding for U.S. Disasters"
subtitle: ""
author: Morgan Taylor
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---
# Supplemental Information
This document represents additional information about the data and results.

## Exploratory analysis
The processed data were then explored to characterize and summarize the relevant features in RStudio 4.1 with the following packages: here, tidyverse, skimr, summarytools, ggplot2, car, gtsummary, scales, cowplot, and table1. The supplementary files contain the code that generates the following results.

Table \@ref(tab:table1) reflects the results of the exploratory data analysis. Between February 2012 and November 2021, there were 446 disaster declarations for states and territories. Funds requested by states or territories for FEMA ranged from -\$855,200.3 to \$4,388,540,486, with an average of \$29,673,687, and FEMA funds obligated to the states and territories after reconciliation ranged from -\$855,200.3 to \$440,842,4297 with an average of \$29,744,109.

```{r summarytable,  echo=FALSE, message = FALSE}
readRDS("../../results/table1.rds")
```

Figure \@ref(fig:incidentyears) shows the number of disaster declarations per year, and Figure \@ref(fig:funds-years) shows the total FEMA funds (both requested and obligated) by states per year. The substantial peak in 2020 corresponds to the COVID-19 pandemic, and the one in 2017 represents the hurricane season with Hurricanes Irma, Harvey, and Maria. These characteristics can also be seen in Table \@ref(tab:decs-year-3), which represents FEMA Funding for the 3 Costliest Years (2012 - 2021).

```{r incidentyears,  fig.cap='Number of disaster declarations per year 2012 - 2021', echo=FALSE, message = FALSE}
knitr::include_graphics("../../results/incidentyear.png")
```

```{r funds-years,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by year 2012 - 2021', echo=FALSE, message = FALSE}
knitr::include_graphics("../../results/funds-years.png")
```

```{r decs-year-3,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-year-3.Rds")
```

Figure \@ref(fig:incidentmonth) shows the number of disaster declarations per month, and Figure \@ref(fig:funds-month) shows the total FEMA funds (both requested and obligated) by states per month. The substantial peak in January likely corresponds to the COVID-19 pandemic, and the ones in August, September, and October represent the peak hurricane season. These characteristics can also be seen in Table \@ref(tab:decs-month-3), which represents FEMA Funding for the 3 Costliest Months.

```{r incidentmonth,  fig.cap='Number of disaster declarations per month for years 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/incidentmonth.png")
```

```{r funds-month,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by month for 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/funds-month.png")
```

```{r decs-month-3,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-month-3.Rds")
```

All 50 states as well as Washington, DC were included in the final, combined dataset. Table \@ref(tab:decs-state-5) provides the five most disaster-prone states and summarizes their requested FEMA funding and received FEMA funding over the study period.
```{r decs-state-5,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-state-5.Rds")
```

Incident duration ranged from zero days to 660 days (or almost two years), as seen by Figure \@ref(fig:funds-dur). The incident duration of 660 days likely corresponds to the COVID-19 pandemic, which is an ongoing incident of almost two years. Further analysis of all variables can be found in the supplementary information.

```{r funds-dur,  fig.cap='Incident duration for requested FEMA funds and obligated FEMA funds under $1 billion for 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/funds-dur.png")
```

Prior to fitting the data to machine learning models, standard simple and multivariable regression models were fitted using the tidymodels framework in RStudio 4.1. Both outcomes of interest were fitted, using incident duration as the primary predictor of interest. Figure \@ref(fig:req_MVR_bp) and \@ref(fig:obl_MVR_bp) show the significant variables fitted in the multivariable regression models for requested and obligated FEMA funds, respectively.

```{r req_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/req_MVR_bp.png")
```

```{r obl_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/obl_MVR_bp.png")
```

Table \@ref(tab:lm-eval) shows the comparisons of the models, using the root mean square error (RMSE) and r-squared statistics. The requested and obligated funding models performed almost identically. None of the models fit the data well, which suggests more sophisticated modeling practices are necessary to adequately characterize the data.

```{r lm-eval,  echo=FALSE, message = FALSE}
resulttable2=readRDS("../../results/lm-eval.rds")
knitr::kable(resulttable2, caption = 'Linear model comparisons')
```

## Full analysis

Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g. requested and obligated funds). Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed. Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category. The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type. Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type. The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type. The models were then compared based on the root mean square error (RMSE), the standard metric for regression models. The analysis was completed in RStudio 4.1 with the following packages: here, tidyverse, tidymodels, skmir, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.

The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot) and  \@ref(fig:DT-plot-obl). Both outcomes identified the same decision point features: total number of federal agencies involved in the response, duration of the incident, state population, state counties, and whether the individual / housing assistance program was awarded.

```{r DT-plot,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/DT-plot.png")
```

```{r DT-plot-obl,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/DT-plot-obl")
```

The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot) and  \@ref(fig:RF-plot-obl). Both outcomes identified the same important features: incident duration, total number of agencies involved, individual/housing assistance program awardance, and incident month. Interestingly, however, the obligated funds model includes days of the entire response as the fifth most important feature, whereas the requested funds model includes it as the eleventh most important feature. Unfortunately, the tidymodels framework does not support visualization of the bagged tree model.

```{r RF-plot,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/RF-plot.png")
```

```{r RF-plot-obl,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/RF-plot-obl")
```

Table \@ref(tab:ML-model-comp) shows the model comparisons for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding. In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.

```{r ML-model-comp,  echo=FALSE, message = FALSE}
resulttable=readRDS("../../results/ML-model-comp.rds")
knitr::kable(resulttable, caption = 'ML model comparison for requested FEMA funding')
```

```{r ML-model-comp-obl,  echo=FALSE}
resulttable1=readRDS("../../results/ML-model-comp-obl.rds")
knitr::kable(resulttable1, caption = 'ML model comparison for obligated FEMA funding')
```

The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample. The RMSEs for the testing and training datasets were substantially different. Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds. The supplementary information contains the code for the models as well as all diagnostic plots.
```{r BT_pred_act_test,  fig.cap='Predicted vs actual requested FEMA funds for the bagged tree fit', echo=FALSE}
knitr::include_graphics("../../results/BT-pred-act-test.png")
```

```{r BT_pred_act_test-obl,  fig.cap='Predicted vs actual obligated FEMA funds for the bagged tree fit', echo=FALSE}
knitr::include_graphics("../../results/BT-pred-act-test-obl.png")
```

### 2021-11-29 products/manuscript/Manuscript_Disasters.Rmd
Manuscript Fix

@@
 $$
-where O_i are the observations, S_i are the predicted values of the outcome, and n is the number of observations in the analysis.
+where O<sub>i</sub> are the observations, S<sub>i</sub> are the predicted values of the outcome, and n is the number of observations in the analysis.
 RMSE is scale dependent, but the lower the RMSE, the better the model performance.
@@
 The RMSEs for the testing and training datasets were substantially different, with an RMSE of 45003235 for the training data and 69378994 for the testing data.
-Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds.
+Figure \@ref(fig:BT_pred_act_test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT_pred_act_test-obl) does the same for obligated FEMA funds.
 The supplementary information contains the code for the models as well as all diagnostic plots.

### 2021-12-02 Zane_Project_Review_Template.Rmd
update review

---
title: Project Review Template 
date: "`r file.mtime(knitr::current_input())`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
    number_sections: true
---

# Overview

Title of project: Predicting Federal Emergency Management Agency Funding for U.S. Disasters

Name of project author(s): Morgan Taylor

Name of project reviewer: Zane Billings

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

* The paper includes a fairly comprehensive background of the disaster response process and how FEMA money is requested and allocated. The motivation for the project is clear.
* The only previous context for the research question is that "little has been done to draw conclusions from these data for predicting the financial impacts of disasters". I think it would benefit the paper to include a section on previous work for modeling disease costs, or mentioning any studies that have previously used this data in a similar way.
* The new information the project hopes to provide is explained well.
* Other than the lack of previous research context, I think the introduction is comprehensive. If there are truly no studies in this area, that should be noted as well.
* The section on machine learning models should be moved to the Methods section.

### Summary assessment

Some contextualization and motivation.

## Question description

### Feedback and Comments

* The question is explained well. What I understood as the research question is:

> Can the amount of allocated or requested FEMA funding be predicted for a given disaster?

> If this is not the intended research question, I advise revision of the introduction section.

* For me, the introduction would flow better if the paragraph beginning "It would greatly benefit the federal government to be able to predict..." was at the end of the introduction as this paragraph contains the statement of the research question.
* I think the research question is clear, but I think it could be made by more detailed by saying something to the effect of "We aim to predict allocated/predicted FEMA funds using...[list of predictors]." Not every predictor needs to be listed but I think this would be nice (overlaps somewhat with a point I make in the next section).

### Summary assessment
Question/hypotheses fully clear.

## Data description

### Feedback and Comments

* The data source is explicitly listed.
* The data are described well but I would prefer an explicit list of candidate predictors used for the models.
* I wasn't aware that codebooks were a requirement for the project, but codebooks are provided for two of the three raw data sheets, and links to the data sources are provided in the README.

### Summary assessment
Source and overall structure of data well explained.

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* major weaknesses in wrangling and exploratory component
* some weaknesses in wrangling and exploratory component
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

If time permits, I would recommend consideration of a few other modeling methods. I think regression with the elastic net should be included--this tends to be a logical "next step" from multiple linear regression, and is typically quite fast to fit. The model is also easy to interpret, and usually I think if an elastic net model fits "about as well" as a tree-based model, the elastic net model should be preferred.

KNN or a gradient-boosted tree might also be a nice addition. In some cases, gradient-boosted trees can generate more accurate predictions than random forests, but can also be easier to overfit and typically take a long time to train. On the other hand, KNN is very easy to train (one parameter, unless you count the distance function) and the performance depends entirely on the amount of local structure in the data, so it might or might not be useful.

That being said,

### Summary assessment
* wrong/inadequate analysis
* defensible but not optimal analysis 
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* results are poorly presented, hard to understand, poor quality
* results are presented ok, with room for improvement
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* major parts of discussion missing or wrong 
* minor parts wrong, missing or unclear
* strong, complete and clear discussion

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* poor and confusing structure
* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* poorly documented
* decently documented with some gaps
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* major parts not reproducible 
* small parts not reproducible or required manual intervention 
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment
* weak level of thoroughness
* decent level of thoroughness
* strong level of thorougness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

* Best not to index temporary Office files with Git, e.g. `~$nuscript_Disasters.docx`. These are indexed for Git tracking when you commit all staged changes to Git while you have an unsaved Office document open. Assuming that you don't use `$` in file names, you can add `*$*` to the `.gitignore` (or to be safer, `*$*.docx`).
* You can format an abstract in your document using YAML by adding the lines
    `abstract: |`
    `text of abstract goes down here`
*

### 2021-12-07 Veytsel_Review (Taylor).Rmd

---
title: Project Review Template 
date: "`r file.mtime(knitr::current_input())`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
    number_sections: true
---

#Overview
Title of project: Predicting Federal Emergency Management Agency Funding for U.S. Disasters

Name of project author(s): Morgan Taylor

Name of project reviewer: Gabriella Veytsel

# Specific project content evaluation
**Background, Context and Motivation**
A few small comments. I think you should cite your first sentence of the summary/abstract; in fact, it would be pertinent to include some statistics from the literature. You spend a lot of text discussing the process of disaster assistance and one sentence on the gap in knowledge, but I think there should be a little bit more justification/signficance of your study. For example, when you say that this information is important to states and the federal govt., I think you could be more explicit about how exactly this knowledge would benefit states. It might also be benficial to have a brief background on disasters themselves: how often the occur, categories, the costs involved, etc. Ultimately, strong contexualization, but I would recommend hitting more on what has been done already, the need, and how this research addresses it.

**Summary assessment**
* strong contextualization and motivation

**Question description**
Clear and straightforward question.

**Summary assessment**
* question/hypotheses fully clear

**Data description**
I think the data was well described. A data dictionary in the supplemental section would probably be useful, but I'm not sure how common that really is. You described the leaning, feature engineering done for each dataset, and merging into final datset for analysis.

**Summary assessment**
* source and overall structure of data well explained

**Data wrangling and exploratory analysis**
Well explained and extremely through exploratory analysis. Generated a ton of figures, but results maybe could have been more described in the code, as well as reasoning behind each one.

**Summary assessment**
* essentially no weaknesses in wrangling and exploratory component

**Appropriateness of Analysis**
I was wondering, for the bagged tree model, is it possible for the RMSE to be in the millions? That seems kind of strange, maybe. Does it make sense to remove Hurricane Maria from the analysis? Overall, really strong analysis and data exploration, just so thorough and well done. She used many models and each one had a detailed investigation into performance metrics and diagnostic plots.

**Summary assessment**
* strong and reasonable analysis

**Presentation**
Might want to change your y-axis label for figures 4.1 and 4.2. Your model comparison tables look good to me. You should change the @ref you're getting below your figures. Your supplemental section also looks great and thorough.

**Summary assessment**
* results are very well presented

**Discussion/Conclusions**
Regarding your statement about state population: I feel like the state population is a different variable than the number of people directly/indirectly affected. Strengths and limitations were really well done.

**Summary assessment**
* strong, complete and clear discussion

#Overall project content evaluation

**Structure**
Your readme files are really nice and thorough, explaining the organization of everything in your folder and a summary of the steps and data.

**Summary assessment**
* well structured

**Documentation**
The scripts are clean and well organized, which is super important as I'm trying to follow along with your steps, reasoning, and interpretation. I think there still is some room to comment on your decision-making throughout the code.

**Summary assessment**
* fully and well documented

**Reproducibility**
Yes, everything ran without issues! I think there should be explicit instructions though for executing the code.

**Summary assessment**
* fully reproducible without issues

**Thoroughness**
This was excellent! REALLY well done. It is really clear that you put in an incredible amount of work into everything and had really great understanding of your data and the methods. Your organization was so well documented. SO thorough and thoughtful, used many models that we discussed in class. Honestly, this seemed publication quality to me, like you didn't take any shortcuts--everything is so neat. I can only imagine how long it took you to do this!

**Summary assessment**
* strong level of thorougness

### 2021-12-08 reviews/Zane_Project_Review_Template.Rmd
Documentation Updates

---
title: Project Review Template 
date: "`r file.mtime(knitr::current_input())`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
    number_sections: true
---

```{r}
require(emoji)
```

# Overview

Title of project: Predicting Federal Emergency Management Agency Funding for U.S. Disasters

Name of project author(s): Morgan Taylor

Name of project reviewer: Zane Billings

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

* The paper includes a fairly comprehensive background of the disaster response process and how FEMA money is requested and allocated. The motivation for the project is clear.
* The only previous context for the research question is that "little has been done to draw conclusions from these data for predicting the financial impacts of disasters". I think it would benefit the paper to include a section on previous work for modeling disease costs, or mentioning any studies that have previously used this data in a similar way.
* The new information the project hopes to provide is explained well.
* Other than the lack of previous research context, I think the introduction is comprehensive. If there are truly no studies in this area, that should be noted as well.
* The section on machine learning models should be moved to the Methods section.

### Summary assessment

Some contextualization and motivation.

## Question description

### Feedback and Comments

* The question is explained well. What I understood as the research question is:

> Can the amount of allocated or requested FEMA funding be predicted for a given disaster?

If this is not the intended research question, I advise revision of the introduction section.

* For me, the introduction would flow better if the paragraph beginning "It would greatly benefit the federal government to be able to predict..." was at the end of the introduction as this paragraph contains the statement of the research question.
* I think the research question is clear, but I think it could be made by more detailed by saying something to the effect of "We aim to predict allocated/predicted FEMA funds using...[list of predictors]." Not every predictor needs to be listed but I think this would be nice (overlaps somewhat with a point I make in the next section).

### Summary assessment
Question/hypotheses fully clear.

## Data description

### Feedback and Comments

* The data source is explicitly listed.
* The data are described well but I would prefer an explicit list of candidate predictors used for the models.
* I wasn't aware that codebooks were a requirement for the project, but codebooks are provided for two of the three raw data sheets, and links to the data sources are provided in the README.

### Summary assessment
Source and overall structure of data well explained.

## Data wrangling and exploratory analysis

### Feedback and Comments

* Coverage of data wrangling is sparse, but what I would expect from a typical manuscript.
* Meaningful exploratory results are shown in the supplement.
  + I think a table stratifying by state is unnecessary. The information would be easier to see as a map.
  + Table numbers in the supplement aren't working correctly. (I.e. some crossrefs show up as "Table ??".)
  + Packages should be cited as references. I think citing RStudio is important but it's much more important to cite the R language--the analysis can be repeated without RStudio, but not without R. The supplement text says "RStudio 4.1" but RStudio is only on 1.4 (as opposed to R 4.1.x) so I think this is what is meant.
  + I don't understand how a state can request a negative amount of money. This should be explained in the text.
  + I think funds should probably always be plotted on a log scale. Plotting on a normal scale seems likely to distort the results. Maybe I'll discuss more in the modeling section.
  + If number of disasters per month and per year are in the data, I'd also like to see a time series plot of the number of disasters.
  + Table 1.3 runs off the edge of the page.
  + Supplement says "Further analysis of all variables can be found in the supplementary information." Well, that's where I am, so where is it? Haha just kind of funny `r emoji::emoji("grin")`
  + Glad you discussed the number of observations at 660 days incident duration. Maybe comment on if these were left in during analysis or not, as these observations will likely have overdue leverage in a regression model.
  + After the incident duration plots, "RStudio 4.1" is referenced again. This is a personal pet peeve, but I really recommend fixing this. The models weren't fit using RStudio, they were fit using R. RStudio is just a graphical interface, it does none of the heavy lifting.
  + Figure references are messy in the same paragraph, it looks like you forgot backslashes.
  + Anyways, nice consistent use of color, I really like the red and blue to distinguish between the two outcomes. However, I'd maybe consider switching to a perceptually uniform palette--one of the colors should be lighter than the other so they are still distinguishable if the document is printed in grayscale. The `scico` and `viridis` packages offer good examples.
  + Please show coefficients for all regression coefficients instead of just significant coefficients. Yes they will be clustered around zero, but I'd still like to see confidence intervals. I'd also recommend using the same color scheme as with the bar plots here.
  + I'd like to see histograms of the outcomes in the supplement (or density plot, but I like histograms more). If they are heavily skewed, this can affect your interpretation of the linear model. My suspicion is that modeling the log outcome would be more appropriate in this case, but it's hard to tell without seeing the data. Of course there are negative numbers in the range which can't be logged, but like I commented earlier I don't really understand why there are negative money amounts.
* You could discuss the overall cleaning/wrangling process in the manuscript a bit more if you wanted to, but I think what you have is pretty typical, usually we spend 90% of our time cleaning data only to not mention that at all in the paper :)
* For example, I think you could explain more what you mean by "feature engineering was performed to...", this phrase doesn't mean anything to me.\
* The only real "weaknesses" I see here are what I described about the beginning of the supplement, there are a few more things I think you need to justify a bit better like I mentioned before.

### Summary assessment
Some weaknesses in wrangling and exploratory component.

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

* I think the results of the multivariate regression should be reported in the main text along with the RMSE for the model.
* I don't think you need to include the definition of the RMSE in the main text.
* If your data are spatially stratified, I think you could consider using either spatial resampling (e.g. https://spatialsample.tidymodels.org/reference/spatial_clustering_cv.html) or justifying why you use normal cross-validation. For spatially-clustered data, points in the same cluster tend to be more similar than points in different clusters, and reporting metrics on cross-validation folds which don't take clusters into account can be overly optimistic.
* Including all the states as individual predictors in the linear model is a valid approach but maybe not the most efficient approach. E.g. you could get the centroid latitude/longitude for each state and include those as predictors instead. Nothing wrong with this and I don't know that much about spatial data, you just have a lot of predictors which can often be annoying.
* If time permits, I would recommend consideration of a few other modeling methods. I think regression with the elastic net should be included--this tends to be a logical "next step" from multiple linear regression, and is typically quite fast to fit. The model is also easy to interpret, and usually I think if an elastic net model fits "about as well" as a tree-based model, the elastic net model should be preferred. 
* KNN might also be a nice model to try. Sometimes it performs really well despite being so simple if the data have local structure. It's also easy and fast to tune most of the time.
* More complicated additions might be a gradient boosted tree or ensemble model. But I think the dataset is too small to reliably prevent overfitting by a boosted tree, and we didn't discuss ensembles much in class. Maybe a cubist model would be nice too but also not discovered in class and the tidymodels implementation is currently a bit sketchy.
* Are your reported metrics from fitting the best model on the entire training data set, or from the cross-validated predictions? Since they have SEs, I assume from cross-validation but you should clarify this in the table captions. I also think the raw RMSEs are quite hard to interpret and I really recommend showing the log RMSE instead of the raw RMSE. log is a monotonic function so this won't mess up your results or cause you to pick a different model, it just makes the numbers easier to compare.
* I think your analysis would really benefit from including an abbreviated Table 1 or at least a list of the predictors that were included in the models.
* How was variable importance calculated? It's weird to me that you have "count" on the x-axis for the VI plot, usually VI is in a pretty arbitrary metric. Also not sure what the "root" label on this plot means. Using the same color stratification for the two outcomes as you did in the supplement would be nice.
* Your discussion suggests to me that potentially requested/obligated funds per capita may be better to predict than the raw numbers.
* Can you provide the correlation between the request/obligated funds? From the prediction plots they appear to be highly correlated. It's still ok to model them independently, it just indicates that you could be losing information by modeling them separately. 
* Can you comment on the outliers in the diagnostic plots? It appears that your best model has a difficult time predicting extremely high values. I'd also like to see the residual vs predicted plots.
* Overall I think your analysis is nice, but could definitely be improved with a few additional explanations and maybe transforming the response variables. The only thing I think would be difficult to defend would be not using spatial resampling but I know that we didn't talk about this in class so `r emoji::emoji("man_shrugging")`

### Summary assessment
Strong and reasonable analysis.

## Presentation

### Feedback and Comments

* I don't like the ggplot2 default theme very much, the gray lines make the plots difficult to read IMO. If you like them that's fine but I think you should consider using a theme with less non-data ink.
* Plot text needs to be larger, it's quite difficult to read.
* I think you could really benefit from using color (preferably a perceptually uniform palette) to distinguish between the two outcomes.
* Also, I think the plots/table I mentioned earlier would be nice additions.
* I think if you have spatial data you can't go wrong with including at least one map if you have a place for it, it usually tends to impress readers (and editors). But if you don't have a place for it that's ok too, just thought I'd say this!
* Inclusion of the reference lines on the predicted vs observed plots is nice.
* Sorting by VI on the VIPs is nice, they are quite easy to interpret.
* Finally, a table of RMSEs on the test data should be included so you can discuss any potential over or underfitting.
* Overall I think your plots are good and effectively communicate what you want them to, but there's a bit of room for polishing.

### Summary assessment
Results are presented ok, with room for improvement.

## Discussion/Conclusions

### Feedback and Comments

* I agree with your assessment of variable importances
* More discussion of outliers in prediction and test data results would be good. Do you suspect overfitting? Underfitting?
* I agree with your second paragraph in the discussion completely. That data would be good but I understand why it's not feasible. Great explanation of a limitation that you are working around as best you can.
* I just read this section and now see why you report a negative cost, sorry I didn't catch this earlier. I think this should also be explained earlier or in the supplement. I also think that you would be justified in only modeling the positive outcomes, or using a model with two stages (similar to a hurdle/zero-inflated model) where you first predict whether a data point should be negative or positive based on the covariates and then predict the outcome. Your point about two separate models is also valid.
* I think your conclusion is fairly strong, but I think there is a bit of results that you are missing. Checking the VIP and metrics on the test data would be good, as would be a comparison of diagnostic plots on the test vs train data.

### Summary assessment
Minor parts wrong, missing or unclear.

# Overall project content evaluation
Evaluate overall features of the project by filling in the sections below.

## Structure

### Feedback and Comments

* Best not to index temporary Office files with Git, e.g. `~$nuscript_Disasters.docx`. These are indexed for Git tracking when you commit all staged changes to Git while you have an unsaved Office document open. Assuming that you don't use `$` in file names, you can add `*$*` to the `.gitignore` (or to be safer, `*$*.docx`).
* Project structure is fine, based of the template so nothing to really say about it.
* Archive is a bit messy, maybe stuff from previous project ideas should be kept in a different repo instead of in here.

### Summary assessment
Mostly clear, but some confusing parts (confusing files--useless is a bit too strong of a word for this).

## Documentation

### Feedback and Comments

* READMEs are good. Understood which scripts to run in what order.
* No (runtime) errors in processing scripts, code is well-documented and easy to understand.
* No (runtime) errors in analysis scripts, code is well-documented and easy to understand.
* No (runtime) errors in manuscript or supplement.
* Didn't check for semantic errors in code but results reported in manuscript seem to make sense.

### Summary assessment
Fully and well documented.

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Yes, see above section.

### Summary assessment
Fully reproducible without issues.
`analysis-ML-RA` generated 50+ warnings, but they all appear to be due to the parallel processor so they aren't actual issues.

## Thoroughness

### Feedback and Comments

Study appeared to be fairly thorough although alternatives like only modeling positive values (with/without log transform), per capita, or modeling both outcomes simultaneously could be considered. The latter is outside the scope of the class but e.g. the difference between the two outcomes could be modeled. Consideration of outliers could be more robust but I think it is fine as is because I rarely think outlier analysis is worthwhile.

### Summary assessment
Decent level of thoroughness.

## Further comments

* If you want, you can format an abstract in your document using YAML by adding the lines
    `abstract: |`
    `text of abstract goes down here`

### 2021-12-10 code/processing_code/processing-missions.Rmd
Documentation Updates
Missions Processing Update

@@
 * Summarization of mission costs for each state and disaster declaration
-* Feature engineering to create cost share proportions
+* Feature engineering to create cost share proportions and number of government agencies involved in the response
+* Cleaning the state variable names to match the other data sources
 * Saving the processed dataset as an .rds
@@
 ## Load Raw Data
-Load the raw data downloaded from provided hyerlink.
+Load the raw data downloaded from provided hyperlink.
 ```{r load data}

### 2021-12-10 code/analysis_code/analysis-eda.Rmd
Cleaning/Processing Update

@@
 
-#specify table labels
-table1::label(EDAdata$state) <- "State"
-table1::label(EDAdata$IncidentYear) <- "Incident Year"
-table1::label(EDAdata$IncidentMonth) <- "Declaration Month"
-table1::label(EDAdata$declarationType) <- "Declaration Type"
-table1::label(EDAdata$incidentType) <- "Incident Type"
-table1::label(EDAdata$IncidentDuration) <- "Incident Duration"
-table1::label(EDAdata$ResponseDays) <- "Response Duration"
-table1::label(EDAdata$Population) <- "State Population"
-table1::label(EDAdata$Counties) <- "Counties per State"
-table1::label(EDAdata$TotalAgencies) <- "Federal Agencies Involved"
-table1::label(EDAdata$FEMACSAvg) <- "Average FEMA Cost Share"
-table1::label(EDAdata$IH) <- "Individuals/Households Program"
-table1::label(EDAdata$PA) <- "Public Assistance Program"
-table1::label(EDAdata$HM) <- "Hazard Mitigation Program"
-table1::label(EDAdata$ReqAmt) <- "Requested FEMA Funds"
-table1::label(EDAdata$OblAmt) <- "Obligated FEMA Funds"
-
 #specify units

### 2021-12-10 code/processing_code/processing-ML.Rmd
Cleaning/Processing Update

---
title: "Data Processing: Creating the Analysis Dataframe"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

---

## Introduction
This markdown imports the processed, combined dataframe, and conducts the following data cleaning / wrangling steps:

* Adding coordinates and FEMA Region for each state and territory
* Subsetting the outcome
* Cleaning variables (e.g. "other" categorization and ordinal transformation of incident month)
* Labeling variables

It also creates the following dataframes for analysis:

* `analysisdata.rds` : used for exploratory data analysis
* `analysisdata-RA.rds`: used for modeling request amount as the outcome
* `analysisdata-OA.rds`: used for modeling obligated amount as the outcome

---

## Required Packages
The following R packages are required for this script:

* here: for data loading/saving
* tidyverse: for data management
* table1: for labeling variables

```{r libraries, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(here) #for data loading/saving
library(tidyverse) #for data management
library(table1) #for labeling the variables
```

---

## Load Data
Load the analysis data from the `processed_data` folder in the project file.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","analysisdata.rds")

#load data. 
analysisdata <- readRDS(data_location)

#summary of data using skimr package
skimr::skim(analysisdata)
```

---

## Further Data Cleaning / Processing
As I worked through the analysis, I realized there were additional needs for the data cleaning/processing step. As they all were based on the combined dataset, I decided to add them here instead of continuously repeating it in each analysis script.

---

### Geographic Data
For the exploratory analysis as well as for spatial cross-validation, we need some sort of geographic data for each state. Google Developers have a dataset that represents the latitude and longitude for each state as well as DC and PR that can be found [here](https://developers.google.com/public-data/docs/canonical/states_csv). It is also saved in the `raw_data` folder along with a data dictionary.

First import the data.
```{r load states}
#load states data
data_location_states <- here::here("data", "raw_data", "states.csv")

#load data
states_raw <- utils::read.csv(data_location_states, na.strings = c("", "NA"))

#summary of data using skimr package
skimr::skim(states_raw)
```

Now we can do an inner join based on the state abbreviation.
```{r state join}
analysisdata_1 <- dplyr::left_join(analysisdata, states_raw, by = "state")
```

We now have geographic data for each state. This step also added a full length version of state name, but we will wait until the end of this script to remove any unnecessary variables.

---

### Subsetting Outcome
In earlier versions of the analysis, requested and obligated amounts were both positive and negative. Conceptually, the negative amounts correspond to circumstances when the states owe FEMA money (e.g. states requested federal assistance but do not meet the damage threshold for federal relief), while positive amounts correspond to circumstances where FEMA is providing relief to the states.

The initial exploratory data analysis iteration suggested the data would be better modeled if log transformed, so I'm deciding to subset the outcomes by focusing only on circumstances where FEMA provides funding (e.g. only values greater than zero of `ReqAmt` and `OblAmt`). This will also ease the conceptual interpretation in the manuscript.
```{r positive values only}
#only keep rows with positive values of ReqAmt and OblAmt
analysisdata_2 <- analysisdata_1 %>%
                    dplyr::filter(ReqAmt > 0 & OblAmt > 0)
```

I'm choosing not to log transform the `ReqAmt` and `OblAmt` variables in the dataset itself, as I want to retain the flexibility to either use the raw or transformed values. It is not difficult to add log(variable) in models in R to include this transformation.

---

### FEMA Region
In the event the spatial cross-validation does not work at a state-level, I want to create a variable that represents each FEMA Region. There are ten FEMA Regions, which could serve as a proxy for the state variable, if necessary.
```{r FEMA Region}
analysisdata_2$FEMARegion <- ifelse((analysisdata_2$state == "CT" |
                                 analysisdata_2$state == "ME" |
                                 analysisdata_2$state == "MA" |
                                 analysisdata_2$state == "NH" |
                                 analysisdata_2$state == "RI" |
                                 analysisdata_2$state == "VT"),
                              "Region I",
                              ifelse((analysisdata_2$state == "NJ" |
                                         analysisdata_2$state == "NY" |
                                         analysisdata_2$state == "PR" |
                                         analysisdata_2$state == "VI"),
                                      "Region II",
                                       ifelse((analysisdata_2$state == "DE" |
                                                 analysisdata_2$state == "MD" |
                                                 analysisdata_2$state == "PA" |
                                                 analysisdata_2$state == "VA" |
                                                 analysisdata_2$state == "DC" |
                                                 analysisdata_2$state == "WV"),
                                              "Region III",
                                               ifelse((analysisdata_2$state == "AL" |
                                                         analysisdata_2$state == "FL" |
                                                         analysisdata_2$state == "GA" |
                                                         analysisdata_2$state == "KY" |
                                                         analysisdata_2$state == "MS" |
                                                         analysisdata_2$state == "NC" |
                                                         analysisdata_2$state == "SC" |
                                                         analysisdata_2$state == "TN"),
                                                      "Region IV",
                                                      ifelse((analysisdata_2$state == "IL" |
                                                               analysisdata_2$state == "IN" |
                                                               analysisdata_2$state == "MI" |
                                                               analysisdata_2$state == "MN" |
                                                               analysisdata_2$state == "OH" |
                                                               analysisdata_2$state == "WI"),
                                                            "Region V",
                                                             ifelse((analysisdata_2$state == "AR" |
                                                                         analysisdata_2$state == "LA" |
                                                                         analysisdata_2$state == "NM" |
                                                                         analysisdata_2$state == "OK" |
                                                                         analysisdata_2$state == "TX"),
                                                                      "Region VI",
                                                                       ifelse((analysisdata_2$state == "IA" |
                                                                                 analysisdata_2$state == "KS" |
                                                                                 analysisdata_2$state == "MO" |
                                                                                 analysisdata_2$state == "NE"),
                                                                              "Region VII",
                                                                               ifelse((analysisdata_2$state == "CO" |
                                                                                         analysisdata_2$state == "MT" |
                                                                                         analysisdata_2$state == "ND" |
                                                                                         analysisdata_2$state == "SD" |
                                                                                         analysisdata_2$state == "UT" |
                                                                                         analysisdata_2$state == "WY"),
                                                                                      "Region VIII",
                                                                                       ifelse((analysisdata_2$state == "AZ" |
                                                                                                 analysisdata_2$state == "CA" |
                                                                                                 analysisdata_2$state == "HI" |
                                                                                                 analysisdata_2$state == "NV"),
                                                                                              "Region IX",
                                                                                               ifelse((analysisdata_2$state == "AK" |
                                                                                                         analysisdata_2$state == "ID" |
                                                                                                         analysisdata_2$state == "OR" |
                                                                                                         analysisdata_2$state == "WA"),
                                                                                                      "Region X",
                                                                                                      NA
                                                                                                      )
                                                                                              )
                                                                                      )
                                                                              )
                                                                      )
                                                            )
                                                      )
                                              )
                                     )
                              )
```

---

### Incident Type Update
In early iterations of the analysis, it became clear that there were a handful of incident types that were uncommon, which prevented any cross-validation sampling. There is already an existing "other" category in the raw data, so I'm moving those uncommon incident types into the "other" category.
```{r other category}
#add to "other" category for incident Types that are infrequent
analysisdata_2$incidentType[(analysisdata_2$incidentType == "Chemical" |
                        analysisdata_2$incidentType == "Coastal Storm" |
                        analysisdata_2$incidentType == "Dam/Levee Break" |
                        analysisdata_2$incidentType == "Mud/Landslide" |
                        analysisdata_2$incidentType == "Volcano" |
                        analysisdata_2$incidentType == "Terrorist" |
                        analysisdata_2$incidentType == "Snow")] <- "Other"
```

---

### Incident Month Update
For the machine learning algorithms, I need to transform the incident month into an ordinal variable.
```{r incident month}
analysisdata_2$IncidentMonth <- ordered(analysisdata_2$IncidentMonth, 
                                        levels = 1:12,
                                        labels = c("January", 
                                                   "February", 
                                                   "March", 
                                                   "April", 
                                                   "May", 
                                                   "June", 
                                                   "July", 
                                                   "August", 
                                                   "September", 
                                                   "October", 
                                                   "November", 
                                                   "December"))
```

---

### Variable Labels
This isn't a strictly necessary step, really just a personal preference. I like to embed the labels for the variables within the dataframe prior to any analysis, to make any visualization or table summarization steps easier. There are many different ways to do this, but I use the table1 package, as I find it the easiest.
```{r labels}
#specify variable labels
table1::label(analysisdata_2$state) <- "State Abbreviation"
table1::label(analysisdata_2$disasterNumber) <- "Declaration Number"
table1::label(analysisdata_2$IncidentYear) <- "Incident Year"
table1::label(analysisdata_2$IncidentMonth) <- "Declaration Month"
table1::label(analysisdata_2$declarationType) <- "Declaration Type"
table1::label(analysisdata_2$incidentType) <- "Incident Type"
table1::label(analysisdata_2$declarationTitle) <- "Declaration Title"
table1::label(analysisdata_2$IncidentDuration) <- "Incident Duration"
table1::label(analysisdata_2$ResponseDays) <- "Response Duration"
table1::label(analysisdata_2$IH) <- "Individuals/Households Program Awarded"
table1::label(analysisdata_2$IH_pct) <- "Percent of Counties Awarded IH"
table1::label(analysisdata_2$PA) <- "Public Assistance Program Awarded"
table1::label(analysisdata_2$PA_pct) <- "Percent of Counties Awarded PA"
table1::label(analysisdata_2$HM) <- "Hazard Mitigation Program"
table1::label(analysisdata_2$HM_pct) <- "Percent of Counties Awarded HM"
table1::label(analysisdata_2$Population) <- "State Population"
table1::label(analysisdata_2$Counties) <- "Counties per State"
table1::label(analysisdata_2$TotalAgencies) <- "Federal Agencies Involved"
table1::label(analysisdata_2$ReqAmt) <- "Requested FEMA Funds"
table1::label(analysisdata_2$OblAmt) <- "Obligated FEMA Funds"
table1::label(analysisdata_2$FEMACSAvg) <- "Average FEMA Cost Share"
table1::label(analysisdata_2$latitude) <- "Centroid Latitude of State"
table1::label(analysisdata_2$longitude) <- "Centroid Longitude of State"
table1::label(analysisdata_2$name) <- "State Name"
table1::label(analysisdata_2$FEMARegion) <- "FEMA Region"
```

---

### Rerrange Columns
This is also just a personal preference step, but I'm rearranging the columns so that the related variables are near each other and the outcomes of interest are listed last.
```{r rearrange}
analysisdata_3 <- analysisdata_2 %>%
                    dplyr::select(name,
                                  state,
                                  latitude,
                                  longitude,
                                  Population,
                                  Counties,
                                  FEMARegion,
                                  disasterNumber,
                                  declarationType,
                                  declarationTitle,
                                  incidentType,
                                  IncidentYear,
                                  IncidentMonth,
                                  IncidentDuration,
                                  ResponseDays,
                                  IH,
                                  IH_pct,
                                  PA,
                                  PA_pct,
                                  HM,
                                  HM_pct,
                                  TotalAgencies,
                                  FEMACSAvg,
                                  ReqAmt,
                                  OblAmt
                                  )
```

---

## Save Datasets
Now that the relevant cleaning steps have been taken, we can save the processed dataset for analysis as an RDS, which maintains all of the variable formatting.
```{r save rds}
#location to save processed file
analysis_data_location <- here::here("data","processed_data","analysisdata.rds")

#save as an RDS
saveRDS(analysisdata_3, file = analysis_data_location)
```

<br>

For the modeling analysis (i.e. everything except the exploratory data analysis), I want two dataframes: one with only `ReqAmt` and one with only `OblAmt` as the outcomes to avoid collinearity. To that end, I'm also going to remove state name. There may be others depending on the spatial cross-validation, but for now, I'm leaving it there.
```{r subsets}
#for the requested amount df, dropping OblAmt and state name
req_data <- analysisdata_3 %>%
               dplyr::select(-c(name, OblAmt))

#location to save processed file
req_data_location <- here::here("data","processed_data","analysisdata-RA.rds")

#save as an RDS
saveRDS(req_data, file = req_data_location)

#for the obligated amount df, dropping ReqAmt and state name
obl_data <- analysisdata_3 %>%
              dplyr::select(-c(name, ReqAmt))

#location to save processed file
obl_data_location <- here::here("data","processed_data","analysisdata-OA.rds")

#save as an RDS
saveRDS(obl_data, file = obl_data_location)
```

### 2021-12-10 code/processing_code/processing-declarations.Rmd
Cleaning/Processing Update

@@
 * Removal of irrelevant variables
-* Feature engineering to create duration of incident and response
+* Feature engineering to create duration of incident and response as well as month and year of event
 * Summarization of key variables for each disaster declaration and state

### 2021-12-12 archive/analysis-ML-RA-old.Rmd
ML update - new models, started outliers

--- 2021-11-28 code/analysis_code/analysis-ML-RA-outlier.Rmd
@@
 ## Load Data
-Load the analysis data from the `processed_data` folder in the project file.
+Load the ReqAmt analysis data from the `processed_data` folder in the project file.
 ```{r load data}
@@
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","analysisdata.rds")
+data_location <- here::here("data","processed_data","analysisdata-RA.rds")
 
@@
 
-* Drop the unused variables
-* Perform feature engineering
-* Transform Incident Month into ordinal variable
-* Remove outlier
 * Set the random seed to 123
@@
 ```{r data setup}
-#requested amount filter
-req_data <- analysisdata %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
-
-#transform incident month into ordinal variable
-req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
-                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
-
-#not enough values per state, so create a variable that is FEMA Regions
-req_data$FEMARegion <- ifelse((req_data$state == "CT" |
-                                 req_data$state == "ME" |
-                                 req_data$state == "MA" |
-                                 req_data$state == "NH" |
-                                 req_data$state == "RI" |
-                                 req_data$state == "VT"),
-                              "Region I",
-                              ifelse((req_data$state == "NJ" |
-                                         req_data$state == "NY" |
-                                         req_data$state == "PR" |
-                                         req_data$state == "VI"),
-                                      "Region II",
-                                       ifelse((req_data$state == "DE" |
-                                                 req_data$state == "MD" |
-                                                 req_data$state == "PA" |
-                                                 req_data$state == "VA" |
-                                                 req_data$state == "DC" |
-                                                 req_data$state == "WV"),
-                                              "Region III",
-                                               ifelse((req_data$state == "AL" |
-                                                         req_data$state == "FL" |
-                                                         req_data$state == "GA" |
-                                                         req_data$state == "KY" |
-                                                         req_data$state == "MS" |
-                                                         req_data$state == "NC" |
-                                                         req_data$state == "SC" |
-                                                         req_data$state == "TN"),
-                                                      "Region IV",
-                                                      ifelse((req_data$state == "IL" |
-                                                               req_data$state == "IN" |
-                                                               req_data$state == "MI" |
-                                                               req_data$state == "MN" |
-                                                               req_data$state == "OH" |
-                                                               req_data$state == "WI"),
-                                                            "Region V",
-                                                             ifelse((req_data$state == "AR" |
-                                                                         req_data$state == "LA" |
-                                                                         req_data$state == "NM" |
-                                                                         req_data$state == "OK" |
-                                                                         req_data$state == "TX"),
-                                                                      "Region VI",
-                                                                       ifelse((req_data$state == "IA" |
-                                                                                 req_data$state == "KS" |
-                                                                                 req_data$state == "MO" |
-                                                                                 req_data$state == "NE"),
-                                                                              "Region VII",
-                                                                               ifelse((req_data$state == "CO" |
-                                                                                         req_data$state == "MT" |
-                                                                                         req_data$state == "ND" |
-                                                                                         req_data$state == "SD" |
-                                                                                         req_data$state == "UT" |
-                                                                                         req_data$state == "WY"),
-                                                                                      "Region VIII",
-                                                                                       ifelse((req_data$state == "AZ" |
-                                                                                                 req_data$state == "CA" |
-                                                                                                 req_data$state == "HI" |
-                                                                                                 req_data$state == "NV"),
-                                                                                              "Region IX",
-                                                                                               ifelse((req_data$state == "AK" |
-                                                                                                         req_data$state == "ID" |
-                                                                                                         req_data$state == "OR" |
-                                                                                                         req_data$state == "WA"),
-                                                                                                      "Region X",
-                                                                                                      NA
-                                                                                                      )
-                                                                                              )
-                                                                                      )
-                                                                              )
-                                                                      )
-                                                            )
-                                                      )
-                                              )
-                                     )
-                              )
-
-#drop state variable
-req_data <- req_data %>%
-              dplyr::select(-c(state))
-
-#add to "other" category for incident Types that are infrequent
-req_data$incidentType[(req_data$incidentType == "Chemical" |
-                        req_data$incidentType == "Coastal Storm" |
-                        req_data$incidentType == "Dam/Levee Break" |
-                        req_data$incidentType == "Mud/Landslide" |
-                        req_data$incidentType == "Volcano" |
-                        req_data$incidentType == "Terrorist" |
-                        req_data$incidentType == "Snow")] <- "Other"
-
-#remove Hurricane Maria outlier
-req_data <-subset(req_data, ReqAmt< 4000000000)
-
 #set random seed
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- BT_wflow %>%
-          tune::tune_grid(resamples = req_data_bootsplit,
+BT_res <- baguette::bag_tree(cost_complexity = tune(),
+                                   tree_depth = tune(),
+                                   min_n = tune()) %>%
+          parsnip::set_engine("rpart",
+                              times = 10) %>%
+          parsnip::set_mode("regression") %>%
+          tune::tune_grid(preprocessor = BT_rec,
+                          resamples = req_data_bootsplit,
                           grid = tree_grid,
-                          metrics = metric_set(rmse),
-                          control = control_resamples(save_pred = TRUE,
-                                                      extract = function(x) extract_model(x)))
+                          metrics = metric_set(rmse))
 
@@
 ```{r}
-#bag roots function to extract most important variables
-bag_roots <-  function(x){
-  x %>% 
-    dplyr::select(.extracts) %>% 
-    unnest(cols = c(.extracts)) %>% 
-    dplyr::mutate(models = map(.extracts,
-                               ~.x$model_df)) %>% 
-    dplyr::select(-.extracts) %>% 
-    unnest(cols = c(models)) %>% 
-    mutate(root = map_chr(model,
-                          ~as.character(.x$fit$frame[1, 1]))) %>%
-    dplyr::select(root)  
-}
+#extract model from final fit
+x_BT <- best_BT_fit$fit$fit$fit
 
-#plot most important variables
-bag_roots(BT_res) %>% 
-  ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
-  geom_bar() + 
-  coord_flip() + 
-  labs(x = "root", y = "count")
+#vip package doesn't support BT
 

### 2021-12-12 code/analysis_code/analysis-ML-RA-outlier.Rmd
ML update - new models, started outliers

@@
 * Random Forest
-* Bagged Tree
+* Elastic Net
+* LASSO
 
-It compares the  models, and then finally fits the “best” model to the test data. It removes the $4B outlier that corresponds to Hurricane Maria in Puerto Rico in 2017.
+It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis.
 
@@
 * skimr: for variable summaries
+* spatialsample: for spatial cross-validation
 * broom.mixed: for converting bayesian models to tidy tibbles
@@
 * baguette: for bagged decision tree
+* viridis: for a consistent color scheme in figures
 ```{r libraries, include = FALSE}
@@
 library(skimr) #for variable summaries
+library(spatialsample) #for spatial cross validation
 library(broom.mixed) #for converting bayesian models to tidy tibbles
@@
 library(baguette) #for bagged decision tree
+library(viridis) #for a consistent color scheme in figures
 
@@
 options(scipen=999)
+
+#set ggplot theme to classic
+ggplot2::theme_set(theme_classic())
+
+#center plot titles
+theme_update(plot.title = element_text(hjust = 0.5))
 ```
@@
 
-* Drop the unused variables
-* Perform feature engineering
-* Transform Incident Month into ordinal variable
-* Remove outlier
+* Log transform the outcome
+* Drop `ReqAmt`, `disasterNumber`, `declarationTitle`
+* Remove the obvious outliers (-log(ReqAmt))
 * Set the random seed to 123
-* Split the dataset into 70% training, 30% testing, stratified on incidentType
-* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
+* Split the dataset into 70% training, 30% testing
+* 5-fold cross validation for the spatial CV folds
 * Create a recipe for data and fitting that codes categorical variables as dummy variables
 ```{r data setup}
-#requested amount filter
-req_data <- analysisdata %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
+#log transform ReqAmt
+analysisdata$logReqAmt <- log(analysisdata$ReqAmt)
 
-#transform incident month into ordinal variable
-req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
-                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
+#drop unnecessary variables
+req_data_2 <- analysisdata %>%
+              dplyr::select(-c(ReqAmt, disasterNumber, declarationTitle))
 
-#not enough values per state, so create a variable that is FEMA Regions
-req_data$FEMARegion <- ifelse((req_data$state == "CT" |
-                                 req_data$state == "ME" |
-                                 req_data$state == "MA" |
-                                 req_data$state == "NH" |
-                                 req_data$state == "RI" |
-                                 req_data$state == "VT"),
-                              "Region I",
-                              ifelse((req_data$state == "NJ" |
-                                         req_data$state == "NY" |
-                                         req_data$state == "PR" |
-                                         req_data$state == "VI"),
-                                      "Region II",
-                                       ifelse((req_data$state == "DE" |
-                                                 req_data$state == "MD" |
-                                                 req_data$state == "PA" |
-                                                 req_data$state == "VA" |
-                                                 req_data$state == "DC" |
-                                                 req_data$state == "WV"),
-                                              "Region III",
-                                               ifelse((req_data$state == "AL" |
-                                                         req_data$state == "FL" |
-                                                         req_data$state == "GA" |
-                                                         req_data$state == "KY" |
-                                                         req_data$state == "MS" |
-                                                         req_data$state == "NC" |
-                                                         req_data$state == "SC" |
-                                                         req_data$state == "TN"),
-                                                      "Region IV",
-                                                      ifelse((req_data$state == "IL" |
-                                                               req_data$state == "IN" |
-                                                               req_data$state == "MI" |
-                                                               req_data$state == "MN" |
-                                                               req_data$state == "OH" |
-                                                               req_data$state == "WI"),
-                                                            "Region V",
-                                                             ifelse((req_data$state == "AR" |
-                                                                         req_data$state == "LA" |
-                                                                         req_data$state == "NM" |
-                                                                         req_data$state == "OK" |
-                                                                         req_data$state == "TX"),
-                                                                      "Region VI",
-                                                                       ifelse((req_data$state == "IA" |
-                                                                                 req_data$state == "KS" |
-                                                                                 req_data$state == "MO" |
-                                                                                 req_data$state == "NE"),
-                                                                              "Region VII",
-                                                                               ifelse((req_data$state == "CO" |
-                                                                                         req_data$state == "MT" |
-                                                                                         req_data$state == "ND" |
-                                                                                         req_data$state == "SD" |
-                                                                                         req_data$state == "UT" |
-                                                                                         req_data$state == "WY"),
-                                                                                      "Region VIII",
-                                                                                       ifelse((req_data$state == "AZ" |
-                                                                                                 req_data$state == "CA" |
-                                                                                                 req_data$state == "HI" |
-                                                                                                 req_data$state == "NV"),
-                                                                                              "Region IX",
-                                                                                               ifelse((req_data$state == "AK" |
-                                                                                                         req_data$state == "ID" |
-                                                                                                         req_data$state == "OR" |
-                                                                                                         req_data$state == "WA"),
-                                                                                                      "Region X",
-                                                                                                      NA
-                                                                                                      )
-                                                                                              )
-                                                                                      )
-                                                                              )
-                                                                      )
-                                                            )
-                                                      )
-                                              )
-                                     )
-                              )
+#specify all character variables as factors and clean variable names
+req_data_1 <- req_data_2 %>% 
+                janitor::clean_names() %>%
+                dplyr::mutate_if(is.character, factor)
 
-#drop state variable
-req_data <- req_data %>%
-              dplyr::select(-c(state))
-
-#add to "other" category for incident Types that are infrequent
-req_data$incidentType[(req_data$incidentType == "Chemical" |
-                        req_data$incidentType == "Coastal Storm" |
-                        req_data$incidentType == "Dam/Levee Break" |
-                        req_data$incidentType == "Mud/Landslide" |
-                        req_data$incidentType == "Volcano" |
-                        req_data$incidentType == "Terrorist" |
-                        req_data$incidentType == "Snow")] <- "Other"
-
-#remove Hurricane Maria outlier
-req_data <-subset(req_data, ReqAmt< 4000000000)
+#remove outliers (rows that have a -log(reqamt))
+req_data <- req_data_1 %>%
+              dplyr::filter(log_req_amt > 0)
 
@@
 #use incidentType as stratification
-req_data_split <- rsample::initial_split(req_data, prop = 2/3,
-                                     strata = incidentType)
+req_data_split <- rsample::initial_split(req_data, prop = 2/3)
 
@@
 
-#5-fold cross validation, repeated 5 times, stratified by incidentType
-req_folds <- rsample::vfold_cv(req_train_data,
-                               strata = incidentType,
-                               v = 5,
-                               repeats = 5)
+#5-fold spatial cross validation
+req_folds_spatial <- spatialsample::spatial_clustering_cv(req_train_data,
+                                                          coords = c(latitude, longitude),
+                                                          v = 5)
 
 #create recipe that codes categorical variables as dummy variables
-RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
+RA_rec <- recipes::recipe(log_req_amt ~ ., data = req_train_data) %>%
+          recipes::step_unknown(state, new_level = "unknown state") %>%
           recipes::step_dummy(all_nominal_predictors())
@@
 req_null_RMSE_train <- tibble::tibble(
-                          rmse = rmse_vec(truth = req_train_data$ReqAmt,
-                                              estimate = rep(mean(req_train_data$ReqAmt), nrow(req_train_data))),
+                          rmse = rmse_vec(truth = req_train_data$log_req_amt,
+                                              estimate = rep(mean(req_train_data$log_req_amt), nrow(req_train_data))),
                           SE = 0,
@@
 req_null_RMSE_test <- tibble::tibble(
-                          rmse = rmse_vec(truth = req_test_data$ReqAmt,
-                                              estimate = rep(mean(req_test_data$ReqAmt), nrow(req_test_data))),
+                          rmse = rmse_vec(truth = req_test_data$log_req_amt,
+                                              estimate = rep(mean(req_test_data$log_req_amt), nrow(req_test_data))),
                           SE = 0,
@@
   parsnip::set_mode("regression")
+```
 
-#use the recipe specified earlier
-```
 ---
@@
 tree_res <- tree_wflow %>%
-                tune::tune_grid(resamples = req_folds,
+                tune::tune_grid(resamples = req_folds_spatial,
                                 grid = tree_grid,
@@
                              geom_point(size = 2) +
-                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
+                             facet_wrap(~ .metric, scales = "free") +
                              scale_x_log10(labels = scales::label_number()) +
-                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
+                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0) +
+                             labs(x = "cost complexity",
+                                  y = "mean",
+                                  color = "Tree Depth")
 DT_detail_plot
@@
                                                      digits = 5,
-                                                     main = "Selected Decision Tree Model")
-#save manually -- ggsave throws an error
+                                                     main = "Selected Decision Tree Model",
+                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
+                                                     shadow.col = "grey65",
+                                                     col = "grey99")
+#save manually as "DT-plot.png" -- ggsave throws an error
 
@@
                   broom.mixed::augment(new_data = req_train_data) %>%
-                  dplyr::select(.pred, ReqAmt) %>%
-                  dplyr::mutate(.resid = .pred - ReqAmt)
+                  dplyr::select(.pred, log_req_amt) %>%
+                  dplyr::mutate(.resid = .pred - log_req_amt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = ReqAmt)) +
+DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = log_req_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Decision Tree Fit: Predicted vs. Actual Requested Funds",
-                              x = "Predicted Requested Funds ($)",
-                              y = "Actual Requested Funds ($)")
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "Decision Tree Fit: Predicted vs. Actual Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)") 
 DT_pred_act
@@
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
-                     geom_point() +
+                     geom_point(color = "#3b528b") +
                      labs(title = "Decision Tree Fit: Residuals",
@@
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Decision Tree Fit: Residuals vs Fitted Requested Funds",
-                                x = "Predicted Requested Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Decision Tree Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Predicted Log(Requested Funds) ($)",
                                 y = "Residual")
@@
 The identified best performing tree model has some interesting results.
-It identifies incident duration, state population, total number of federal agencies, total number of state counties, and individual/households program awardance as significant decision points for predicting requesting FEMA funding.
-All of these variables are to be expected, but it is interesting that incident type or declaration type are not included as significant.
-The RMSE is improved from the Null model, but there's still opportunity for improvement.
-Let's try another model.
+It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding.
+New York and November are particularly surprising key predictors.
+It is interesting that incident type or declaration type are not included as significant.
+The RMSE actually increases from the Null, which suggests this is not an appropriate model.
+Moving on.
 
@@
 RF_res <- RF_wflow %>%
-              tune::tune_grid(resamples = req_folds,
+              tune::tune_grid(resamples = req_folds_spatial,
                               grid = RF_grid,
@@
 #plot most important predictors in the model
-RF_plot <- vip::vip(x_RF, num_features = 20)
+RF_plot <- vip::vip(x_RF, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
 RF_plot
 
-#save manually-- ggsave throws errors
+#save manually as "RF-plot.png" -- ggsave throws errors
 
@@
                 broom.mixed::augment(new_data = req_train_data) %>%
-                dplyr::select(.pred, ReqAmt) %>%
-                dplyr::mutate(.resid = .pred - ReqAmt)
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
 
@@
 #geom_abline is a 45 degree line, along which the results should fall
-RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = ReqAmt)) +
+RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_req_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
-                        geom_point() +
-                        labs(title = "RF Fit: Actual vs. Predicted Requested Funds",
-                              x = "Predicted Requested Funds ($)",
-                              y = "Actual Requested Funds ($)")
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "RF Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
 RF_pred_act
@@
                             geom_hline(yintercept = 0, color = "red", lty = 2) +
-                            geom_point() +
+                            geom_point(color = "#3b528b") +
                             labs(title = "RF Fit: Residuals",
@@
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Random Forest Fit: Residuals vs Fitted Requested Funds",
-                                x = "Fitted Requested Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Random Forest Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Fitted Log(Requested Funds) ($)",
                                 y = "Residual")
@@
 In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
-Interestingly, it includes declaration month of Jan - March as more important than state population.
-Let's try one more model.
+It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and whether or not IH was awarded to any part of the state as the five most important predictors.
+Let's try another model to see if we can find an improvement in the RMSE.
 
@@
 
-## Bagged Decision Tree Model
+## Elastic Net Model
 
@@
 
-#start over with data split to allow for bootstrapping
-#set random seed
-set.seed(20140102)
+#define the elastic net model
+elastic_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = tune()) %>%
+                parsnip::set_engine("glmnet")
+```
 
-#use bootstrap instead of CV
-req_data_bootsplit <- req_train_data %>%
-                        rsample::bootstraps(times = 10,
-                                            strata = "incidentType")
-
-#define model
-bag_tree_mod <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-                parsnip::set_engine("rpart",
-                                    times = 10) %>%
-                parsnip::set_mode("regression")
-
-#create recipe that codes categorical variables as dummy variables
-BT_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
-          step_string2factor(all_nominal())
-```
 ---
@@
 ```{r}
-#define workflow for tree
-BT_wflow <- workflows::workflow() %>%
-               workflows::add_model(bag_tree_mod) %>%
-               workflows::add_recipe(BT_rec)
+#define elastic net workflow
+elastic_wflow <- workflows::workflow() %>%
+                 workflows::add_model(elastic_mod) %>%
+                 workflows::add_recipe(RA_rec)
 ```
@@
 #tuning grid specification
-tree_grid <- dials::grid_regular(cost_complexity(),
-                                 tree_depth(),
-                                 min_n(),
-                                 levels = 5)
+EN_grid <- dials::grid_regular(penalty(),
+                               mixture(),
+                               levels = 50)
 ```
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- BT_wflow %>%
-          tune::tune_grid(resamples = req_data_bootsplit,
-                          grid = tree_grid,
-                          metrics = metric_set(rmse),
-                          control = control_resamples(save_pred = TRUE,
-                                                      extract = function(x) extract_model(x)))
+EN_res <- elastic_wflow %>%
+              tune::tune_grid(resamples = req_folds_spatial,
+                              grid = EN_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
 
-#collect metrics
-BT_res %>% workflowsets::collect_metrics()
+#look at top 5 EN models
+top_EN_models <- EN_res %>%
+                    tune::show_best("rmse", n = 5)
+top_EN_models
 
 #default visualization
-BT_auto_plot <- BT_res %>% autoplot()
-BT_auto_plot
+EN_auto_plot <- EN_res %>% autoplot()
+EN_auto_plot
 
 #save file
-BT_auto_fig_file = here("results","BT-auto-plot.png")
-ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)
-
-#more detailed plot
-BT_det_plot <- BT_res %>%
-                    workflowsets::collect_metrics() %>%
-                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
-                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
-                             geom_line(size = 1.5, alpha = 0.6) +
-                             geom_point(size = 2) +
-                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
-                             scale_x_log10(labels = scales::label_number()) +
-                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
-BT_det_plot
-
-#save file
-BT_det_fig_file = here("results","BT-det-plot.png")
-ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
+EN_auto_fig_file = here("results","EN-auto-plot.png")
+ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
 ```
@@
 ```{r}
-#select the BT model with the lowest rmse
-BT_lowest_rmse <- BT_res %>%
+#select the EN model with the lowest rmse
+EN_lowest_rmse <- EN_res %>%
                       tune::select_best("rmse")
 
-#finalize the workflow by using the selected BT model
-best_BT_wflow <- BT_wflow %>%
-                      tune::finalize_workflow(BT_lowest_rmse)
-best_BT_wflow
+#finalize the workflow by using the selected RF model
+best_EN_wflow <- elastic_wflow %>%
+                      tune::finalize_workflow(EN_lowest_rmse)
+best_EN_wflow
 
 #one last fit on the training data
-best_BT_fit <- best_BT_wflow %>%
+best_EN_fit <- best_EN_wflow %>%
                     parsnip::fit(data = req_train_data)
@@
 ```{r}
-#bag roots function to extract most important variables
-bag_roots <-  function(x){
-  x %>% 
-    dplyr::select(.extracts) %>% 
-    unnest(cols = c(.extracts)) %>% 
-    dplyr::mutate(models = map(.extracts,
-                               ~.x$model_df)) %>% 
-    dplyr::select(-.extracts) %>% 
-    unnest(cols = c(models)) %>% 
-    mutate(root = map_chr(model,
-                          ~as.character(.x$fit$frame[1, 1]))) %>%
-    dplyr::select(root)  
-}
+#extract model from final fit
+x_EN <- best_EN_fit$fit$fit$fit
 
-BT_IV <- bag_roots(BT_res) %>% 
-            ggplot(mapping = aes(x = fct_rev(fct_infreq(root)))) + 
-            geom_bar() + 
-            coord_flip() + 
-            labs(x = "root", y = "count")
-BT_IV
+#plot most important predictors in the model
+EN_plot <- vip::vip(x_EN, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+EN_plot
+
+#save manually as "EN-plot.png" -- ggsave throws errors
+
+#find predictions and intervals
+EN_resid <- best_EN_fit %>%
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "EN Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+EN_pred_act
 
 #save file
-BT_IV_file = here("results","BT-IV-outlier.png")
-ggsave(filename = BT_IV_file, plot = BT_IV)
-
-#find predictions and intervals
-BT_resid <- best_BT_fit %>%
-                  broom.mixed::augment(new_data = req_train_data) %>%
-                  dplyr::select(.pred, ReqAmt) %>%
-                  dplyr::mutate(.resid = .pred - ReqAmt)
-
-#plot model predictions from tuned model versus actual outcomes
-#geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = ReqAmt)) +
-                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Requested Funds",
-                              x = "Predicted Requested Funds ($)",
-                              y = "Actual Requested Funds ($)")
-BT_pred_act
-
-#save file
-BT_pred_act_file = here("results","BT-pred-act.png")
-ggsave(filename = BT_pred_act_file, plot = BT_pred_act)
+EN_pred_act_file = here("results","EN-pred-act.png")
+ggsave(filename = EN_pred_act_file, plot = EN_pred_act)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
+EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
                             geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Fit: Residuals",
+                            geom_point(color = "#3b528b") +
+                            labs(title = "EN Fit: Residuals",
                                   x = "Observation Number",
                                   y = "Residual")
-BT_resids
+EN_resids
 
 #save file
-BT_resid_file = here("results","BT-resid.png")
-ggsave(filename = BT_resid_file, plot = BT_resids)
+EN_resid_file = here("results","EN-resid.png")
+ggsave(filename = EN_resid_file, plot = EN_resids)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
+EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Requested Funds",
-                                x = "Predicted Requested Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "EN Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Fitted Log(Requested Funds) ($)",
                                 y = "Residual")
-BT_pred_resid
+EN_pred_resid
 
 #save file
-BT_pred_resid_file = here("results","BT-pred-resid.png")
-ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)
+EN_pred_resid_file = here("results","EN-pred-resid.png")
+ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)
 
-#print model performance
-#print 10 best performing hyperparameter sets
-BT_res %>%
+#print the 10 best performing hyperparameter sets
+EN_res %>%
   tune::show_best(n = 10) %>%
-  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
-  dplyr::mutate(rmse = round(rmse, 3),
-                std_err = round(std_err, 4),
-                cost_complexity = scales::scientific(cost_complexity))
+  dplyr::select(rmse = mean, std_err)
 
-#print the best model performance
-BT_performance <- BT_res %>% tune::show_best(n = 1)
-print(BT_performance)
+#print the best model
+EN_performance <- EN_res %>% tune::show_best(n = 1)
+EN_performance
 
-#compare model performance to null model
-BT_RMSE <- BT_res %>%
+#compare model performance to null model (and other models)
+EN_RMSE <- EN_res %>%
               tune::show_best(n = 1) %>%
@@
                 SE = round(std_err, 4),
-                  model = "Bagged Tree") %>%
+                model = "EN") %>%
              dplyr::bind_rows(RF_RMSE)
-BT_RMSE
+EN_RMSE
+```
+The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors.
+The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors.
+The EN model has an improved RMSE, but the null model technically still has a lower RMSE.
+
+---
+
+## LASSO Model
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the LASSO model
+#mixture = 1 identifies the model to be a LASSO model
+lasso_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = 1) %>%
+                parsnip::set_engine("glmnet")
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define lasso workflow
+lasso_wflow <- workflows::workflow() %>%
+                 workflows::add_model(lasso_mod) %>%
+                 workflows::add_recipe(RA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+lasso_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
+
+#5 lowest penalty values
+lasso_grid %>%
+  dplyr::top_n(-5)
+
+#5 highest penalty values
+lasso_grid %>%
+  dplyr::top_n(5)
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+lasso_res <- lasso_wflow %>%
+              tune::tune_grid(resamples = req_folds_spatial,
+                              grid = lasso_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
+
+#look at top 5 EN models
+top_lasso_models <- lasso_res %>%
+                        tune::show_best("rmse", n = 5)
+top_lasso_models
+
+#default visualization
+lasso_auto_plot <- lasso_res %>% autoplot()
+lasso_auto_plot
+
+#save file
+lasso_auto_fig_file = here("results","lasso-auto-plot.png")
+ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the LASSO model with the lowest rmse
+lasso_lowest_rmse <- lasso_res %>%
+                        tune::select_best("rmse")
+
+#finalize the workflow by using the selected lasso model
+best_lasso_wflow <- lasso_wflow %>%
+                      tune::finalize_workflow(lasso_lowest_rmse)
+best_lasso_wflow
+
+#one last fit on the training data
+best_lasso_fit <- best_lasso_wflow %>%
+                    parsnip::fit(data = req_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_lasso <- best_lasso_fit$fit$fit$fit
+
+#plot most important predictors in the model
+lasso_plot <- vip::vip(x_lasso, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+lasso_plot
+
+#save manually as "LASSO-plot.png" -- ggsave throws errors
+
+#find predictions and intervals
+lasso_resid <- best_lasso_fit %>%
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+lasso_pred_act <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "LASSO Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+lasso_pred_act
+
+#save file
+lasso_pred_act_file = here("results","lasso-pred-act.png")
+ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+lasso_resids <- ggplot2::ggplot(lasso_resid, aes(x = as.numeric(row.names(lasso_resid)), y = .resid))+
+                                geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                geom_point(color = "#3b528b") +
+                                labs(title = "LASSO Fit: Residuals",
+                                      x = "Observation Number",
+                                      y = "Residual")
+lasso_resids
+
+#save file
+lasso_resid_file = here("results","lasso-resid.png")
+ggsave(filename = lasso_resid_file, plot = lasso_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+lasso_pred_resid <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = .resid))+
+                                    geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                    geom_point(color = "#3b528b") +
+                                    labs(title = "LASSO Fit: Residuals vs Fitted Log(Requested Funds)",
+                                          x = "Fitted Log(Requested Funds) ($)",
+                                          y = "Residual")
+lasso_pred_resid
+
+#save file
+lasso_pred_resid_file = here("results","lasso-pred-resid.png")
+ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+lasso_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+lasso_performance <- lasso_res %>% tune::show_best(n = 1)
+lasso_performance
+
+#compare model performance to null model (and other models)
+lasso_RMSE <- lasso_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "LASSO") %>%
+             dplyr::bind_rows(EN_RMSE)
+lasso_RMSE
 
@@
 model_comp_file = here("results", "ML-model-comp.Rds")
-saveRDS(BT_RMSE, file = model_comp_file)
+saveRDS(lasso_RMSE, file = model_comp_file)
 ```
+The LASSO model fit is almost identical to the Elastic Net model fit.
 
-In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.
-
---
+---
 
 ## Model Selection and Evaluation
-Using the bagged tree model
+Based on the results of the models above, the null model actually has the best fit, which warrants further investigation (Especially with the outliers).
+In the mean time, we'll fit the LASSO model as it has the best fit out of the models.
 ```{r}
 #fit model to training set but evaluate on the test data
-BT_fit_test <- best_BT_wflow %>%
+lasso_fit_test <- best_lasso_wflow %>%
                     tune::last_fit(split = req_data_split)
@@
 #compare test performance against training performance
-BT_rmse_test <- collect_metrics(BT_fit_test) %>%
+lasso_rmse_test <- collect_metrics(lasso_fit_test) %>%
                    dplyr::select(rmse = .estimate) %>%
@@
 
-BT_comp <- BT_RMSE %>%
-                  dplyr::filter(model == "Bagged Tree") %>%
+lasso_comp <- lasso_RMSE %>%
+                  dplyr::filter(model == "LASSO") %>%
                   dplyr::transmute(rmse, data = "train") %>%
-                  dplyr::bind_rows(BT_rmse_test) %>%
+                  dplyr::bind_rows(lasso_rmse_test) %>%
                   dplyr::slice(-3) #don't know why the third row shows up
-BT_comp
-#RMSEs are incredibly different...
+lasso_comp
+#RMSEs are different but not as much as previous iterations
+
+#save combined table
+test_comp_file = here("results", "test-train-model-comp.Rds")
+saveRDS(lasso_comp, file = test_comp_file)
 
 #find predictions and intervals
-BT_resid_fit <- BT_fit_test %>%
+lasso_resid_fit <- lasso_fit_test %>%
                       broom.mixed::augment() %>%
-                  dplyr::select(.pred, ReqAmt) %>%
-                  dplyr::mutate(.resid = .pred - ReqAmt)
+                      dplyr::select(.pred, log_req_amt) %>%
+                      dplyr::mutate(.resid = .pred - log_req_amt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = ReqAmt)) +
+lasso_pred_act_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = log_req_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Requested Funds",
-                              x = "Predicted Requested Funds ($)",
-                              y = "Actual Requested Funds ($)")
-BT_pred_act_test
+                        geom_point(color = "#3b528b") +
+                        labs(title = "LASSO Test Fit: Predicted vs. Actual Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+lasso_pred_act_test
 
 #save file
-BT_pred_act_test_file = here("results","BT-pred-act-test.png")
-ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)
+lasso_pred_act_test_file = here("results","lasso-pred-act-test.png")
+ggsave(filename = lasso_pred_act_test_file, plot = lasso_pred_act_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids_test <- ggplot2::ggplot(BT_resid_fit, aes(x = as.numeric(row.names(BT_resid_fit)), y = .resid))+
+lasso_resids_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = as.numeric(row.names(lasso_resid_fit)), y = .resid))+
                       geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Test Fit: Residuals",
+                      geom_point(color = "#3b528b") +
+                      labs(title = "LASSO Test Fit: Residuals",
                             x = "Observation Number",
                             y = "Residual")
-BT_resids_test
+lasso_resids_test
 
 #save file
-BT_resid_test_file = here("results","BT-resid-test.png")
-ggsave(filename = BT_resid_test_file, plot = BT_resids_test)
+lasso_resid_test_file = here("results","lasso-resid-test.png")
+ggsave(filename = lasso_resid_test_file, plot = lasso_resids_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
+lasso_pred_resid_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Requested Funds",
-                                x = "Predicted Requested Funds ($)",
-                                y = "Residual")
-BT_pred_resid_test
+                          geom_point(color = "#3b528b") +
+                          labs(title = "LASSO Test Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Predicted Log(Requested Funds) ($)",
+                                y = "Residuals")
+lasso_pred_resid_test
 
 #save file
-BT_pred_resid_test_file = here("results","BT-pred-resid-test.png")
-ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)
+lasso_pred_resid_test_file = here("results","lasso-pred-resid-test.png")
+ggsave(filename = lasso_pred_resid_test_file, plot = lasso_pred_resid_test)
 ```

### 2021-12-12 code/analysis_code/analysis-ML-RA.Rmd
ML update - new models, started outliers

@@
 * Decision Tree
+* Random Forest
+* Elastic Net
 * LASSO
-* Random Forest.
 
@@
 * ranger: for random forest models
+* baguette: for bagged decision tree
+* viridis: for a consistent color scheme in figures
 ```{r libraries, include = FALSE}
@@
 library(ranger) #for random forest models
+library(baguette) #for bagged decision tree
+library(viridis) #for a consistent color scheme in figures
 
@@
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","analysisdata.rds")
+data_location <- here::here("data","processed_data","analysisdata-RA.rds")
 
 #load data. 
-analysisdata <- readRDS(data_location)
+req_data <- readRDS(data_location)
 
 #summary of data using skimr package
-skimr::skim(analysisdata)
+skimr::skim(req_data)
 ```
@@
 
-* Drop the unused variables
-* Transform Incident Month into ordinal variable
+* Log transform the outcome
+* Drop `ReqAmt`, `disasterNumber`, `declarationTitle`
 * Set the random seed to 123
-* Split the dataset into 70% training, 30% testing, stratified on ReqAmt
-* 5-fold cross validation, 5 times repeated, stratified on ReqAmt for the CV folds
+* Split the dataset into 70% training, 30% testing
+* 5-fold cross validation for the spatial CV folds
 * Create a recipe for data and fitting that codes categorical variables as dummy variables
 ```{r data setup}
-#requested amount filter
-req_data <- analysisdata %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
+#log transform ReqAmt
+req_data$logReqAmt <- log(req_data$ReqAmt)
 
-#transform incident month into ordinal variable
-req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
-                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
+#drop unnecessary variables
+req_data <- req_data %>%
+              dplyr::select(-c(ReqAmt, disasterNumber, declarationTitle))
+
+#specify all character variables as factors and clean variable names
+req_data <- req_data %>% 
+              janitor::clean_names() %>%
+              dplyr::mutate_if(is.character, factor)
 
@@
 
-#split dataset into 70% training, 30% testing
-#use ReqAmt as stratification
-req_data_split <- rsample::initial_split(req_data, prop = 7/10,
-                                     strata = c(ReqAmt, state, incidentType))
-
-#optional tag row for later identification: 
-req_data$rowid<-1:nrow(req_data)
-dplyr_stratified <- function(df, percent, ...){
-  columns<-enquos(...)
-   #group then sample each group
-  out<-df %>% group_by(!!!columns)  %>% slice( sample(1:n(), percent*n())) 
-}
-
-train_group <- dplyr_stratified(req_data, 0.9, state, incidentType)
-train_group
-
-test_group <- dplyr_stratified(req_data, 0.5, state, incidentType)
-test_group
+#split dataset into 66.6% training, 33.3% testing
+#use incidentType as stratification
+req_data_split <- rsample::initial_split(req_data, prop = 2/3)
 
@@
 
-#training set proportions by ReqAmt
-req_train_data %>%
-      dplyr::count(ReqAmt) %>%
-      dplyr::mutate(prop = n / sum(n))
-
-#testing set proportions by ReqAmt
-req_test_data %>%
-      dplyr::count(ReqAmt) %>%
-      dplyr::mutate(prop = n / sum(n))
-
-#5-fold cross validation, 5 times repeated, stratified on `ReqAmt` for the CV folds
-req_folds <- rsample::vfold_cv(req_train_data,
-                               v = 5,
-                               repeats = 5,
-                               strata = ReqAmt)
+#5-fold spatial cross validation
+req_folds_spatial <- spatialsample::spatial_clustering_cv(req_train_data,
+                                                          coords = c(latitude, longitude),
+                                                          v = 5)
 
 #create recipe that codes categorical variables as dummy variables
-RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
+RA_rec <- recipes::recipe(log_req_amt ~ ., data = req_train_data) %>%
+          recipes::step_unknown(state, new_level = "unknown state") %>%
           recipes::step_dummy(all_nominal_predictors())
@@
 ```
+
+---
+
+## Decision Tree
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the tree model
+tree_mod <-
+  parsnip::decision_tree(cost_complexity = tune(),
+                         tree_depth = tune(),
+                         min_n = tune()) %>%
+  parsnip::set_engine("rpart") %>%
+  parsnip::set_mode("regression")
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define workflow for tree
+tree_wflow <- workflows::workflow() %>%
+               workflows::add_model(tree_mod) %>%
+               workflows::add_recipe(RA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+tree_grid <- dials::grid_regular(cost_complexity(),
+                                 tree_depth(),
+                                 min_n(),
+                                 levels = 5)
+
+#tree depth
+tree_grid %>%
+  dplyr::count(tree_depth)
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+tree_res <- tree_wflow %>%
+                tune::tune_grid(resamples = req_folds_spatial,
+                                grid = tree_grid,
+                                control = control_grid(verbose = TRUE),
+                                metrics = yardstick::metric_set(rmse))
+
+#collect metrics
+tree_res %>% workflowsets::collect_metrics()
+
+#default visualization
+DT_auto_plot <- tree_res %>% autoplot()
+DT_auto_plot
+
+#save file
+DT_auto_fig_file = here("results","DT-auto-plot.png")
+ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)
+
+#more detailed plot
+DT_detail_plot <- tree_res %>%
+                    workflowsets::collect_metrics() %>%
+                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
+                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
+                             geom_line(size = 1.5, alpha = 0.6) +
+                             geom_point(size = 2) +
+                             facet_wrap(~ .metric, scales = "free") +
+                             scale_x_log10(labels = scales::label_number()) +
+                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0) +
+                             labs(x = "cost complexity",
+                                  y = "mean",
+                                  color = "Tree Depth")
+DT_detail_plot
+
+#save file
+DT_det_fig_file = here("results","DT-det-plot.png")
+ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)
+
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the tree model with the lowest rmse
+tree_lowest_rmse <- tree_res %>%
+                        tune::select_best("rmse")
+
+#finalize the workflow by using the selected DT model
+best_tree_wflow <- tree_wflow %>%
+                      tune::finalize_workflow(tree_lowest_rmse)
+best_tree_wflow
+
+#one last fit on the training data
+best_tree_fit <- best_tree_wflow %>%
+                    parsnip::fit(data = req_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#plot the tree
+rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
+                                                     roundint = F,
+                                                     type = 5,
+                                                     digits = 5,
+                                                     main = "Selected Decision Tree Model",
+                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
+                                                     shadow.col = "grey65",
+                                                     col = "grey99")
+#save manually as "DT-plot.png" -- ggsave throws an error
+
+#find predictions and intervals
+tree_resid <- best_tree_fit %>%
+                  broom.mixed::augment(new_data = req_train_data) %>%
+                  dplyr::select(.pred, log_req_amt) %>%
+                  dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline draws a 45 degree line, along which the results should fall
+DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "Decision Tree Fit: Predicted vs. Actual Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)") 
+DT_pred_act
+
+#save file
+DT_pred_act_file = here("results","DT-pred-act.png")
+ggsave(filename = DT_pred_act_file, plot = DT_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
+                     geom_hline(yintercept = 0, color = "red", lty = 2) +
+                     geom_point(color = "#3b528b") +
+                     labs(title = "Decision Tree Fit: Residuals",
+                          x = "Observation Number",
+                          y = "Residual")
+DT_resid
+
+#save file
+DT_resid_file = here("results","DT-resid.png")
+ggsave(filename = DT_resid_file, plot = DT_resid)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
+                          geom_hline(yintercept = 0, color = "red", lty = 2) +
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Decision Tree Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Predicted Log(Requested Funds) ($)",
+                                y = "Residual")
+DT_pred_resid
+
+#save file
+DT_pred_resid_file = here("results","DT-pred-resid.png")
+ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)
+
+#print model performance
+#print 10 best performing hyperparameter sets
+tree_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
+  dplyr::mutate(rmse = round(rmse, 3),
+                std_err = round(std_err, 4),
+                cost_complexity = scales::scientific(cost_complexity))
+
+#print the best model performance
+tree_performance <- tree_res %>% tune::show_best(n = 1)
+print(tree_performance)
+
+#compare model performance to null model
+tree_RMSE <- tree_res %>%
+                tune::show_best(n = 1) %>%
+                dplyr::transmute(
+                  rmse = round(mean, 3),
+                  SE = round(std_err, 4),
+                  model = "Tree") %>%
+               dplyr::bind_rows(req_null_RMSE_train)
+tree_RMSE
+```
+
+The identified best performing tree model has some interesting results.
+It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding.
+New York and November are particularly surprising key predictors.
+It is interesting that incident type or declaration type are not included as significant.
+The RMSE actually increases from the Null, which suggests this is not an appropriate model.
+Moving on.
+
+<br>
+
+---
+
+## Random Forest Model
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the RF model
+RF_mod <-
+  parsnip::rand_forest(mtry = tune(),
+                       min_n = tune(),
+                       trees = tune()) %>%
+  parsnip::set_engine("ranger",
+                      importance = "permutation") %>%
+  parsnip::set_mode("regression")
+
+#use the recipe specified earlier (line 133)
+
+#check to make sure identified parameters will be tuned
+RF_mod %>% tune::parameters()
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define workflow for RF regression
+RF_wflow <- workflows::workflow() %>%
+               workflows::add_model(RF_mod) %>%
+               workflows::add_recipe(RA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+RF_grid <- expand.grid(mtry = c(3, 4, 5, 6),
+                       min_n = c(40, 50, 60),
+                       trees = c(500,1000))
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+RF_res <- RF_wflow %>%
+              tune::tune_grid(resamples = req_folds_spatial,
+                              grid = RF_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
+
+#look at top 5 RF models
+top_RF_models <- RF_res %>%
+                    tune::show_best("rmse", n = 5)
+top_RF_models
+
+#default visualization
+RF_auto_plot <- RF_res %>% autoplot()
+RF_auto_plot
+
+#save file
+RF_auto_fig_file = here("results","RF-auto-plot.png")
+ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the RF model with the lowest rmse
+RF_lowest_rmse <- RF_res %>%
+                      tune::select_best("rmse")
+
+#finalize the workflow by using the selected RF model
+best_RF_wflow <- RF_wflow %>%
+                      tune::finalize_workflow(RF_lowest_rmse)
+best_RF_wflow
+
+#one last fit on the training data
+best_RF_fit <- best_RF_wflow %>%
+                    parsnip::fit(data = req_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_RF <- best_RF_fit$fit$fit$fit
+
+#plot most important predictors in the model
+RF_plot <- vip::vip(x_RF, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+RF_plot
+
+#save manually as "RF-plot.png" -- ggsave throws errors
+
+#find predictions and intervals
+RF_resid <- best_RF_fit %>%
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "RF Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+RF_pred_act
+
+#save file
+RF_pred_act_file = here("results","RF-pred-act.png")
+ggsave(filename = RF_pred_act_file, plot = RF_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
+                            geom_hline(yintercept = 0, color = "red", lty = 2) +
+                            geom_point(color = "#3b528b") +
+                            labs(title = "RF Fit: Residuals",
+                                  x = "Observation Number",
+                                  y = "Residual")
+RF_resids
+
+#save file
+RF_resid_file = here("results","RF-resid.png")
+ggsave(filename = RF_resid_file, plot = RF_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
+                          geom_hline(yintercept = 0, color = "red", lty = 2) +
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Random Forest Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Fitted Log(Requested Funds) ($)",
+                                y = "Residual")
+RF_pred_resid
+
+#save file
+RF_pred_resid_file = here("results","RF-pred-resid.png")
+ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+RF_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+RF_performance <- RF_res %>% tune::show_best(n = 1)
+RF_performance
+
+#compare model performance to null model (and other models)
+RF_RMSE <- RF_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "RF") %>%
+             dplyr::bind_rows(tree_RMSE)
+RF_RMSE
+```
+
+In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
+It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and whether or not IH was awarded to any part of the state as the five most important predictors.
+Let's try another model to see if we can find an improvement in the RMSE.
+
+---
+
+## Elastic Net Model
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the elastic net model
+elastic_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = tune()) %>%
+                parsnip::set_engine("glmnet")
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define elastic net workflow
+elastic_wflow <- workflows::workflow() %>%
+                 workflows::add_model(elastic_mod) %>%
+                 workflows::add_recipe(RA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+EN_grid <- dials::grid_regular(penalty(),
+                               mixture(),
+                               levels = 50)
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+EN_res <- elastic_wflow %>%
+              tune::tune_grid(resamples = req_folds_spatial,
+                              grid = EN_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
+
+#look at top 5 EN models
+top_EN_models <- EN_res %>%
+                    tune::show_best("rmse", n = 5)
+top_EN_models
+
+#default visualization
+EN_auto_plot <- EN_res %>% autoplot()
+EN_auto_plot
+
+#save file
+EN_auto_fig_file = here("results","EN-auto-plot.png")
+ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the EN model with the lowest rmse
+EN_lowest_rmse <- EN_res %>%
+                      tune::select_best("rmse")
+
+#finalize the workflow by using the selected RF model
+best_EN_wflow <- elastic_wflow %>%
+                      tune::finalize_workflow(EN_lowest_rmse)
+best_EN_wflow
+
+#one last fit on the training data
+best_EN_fit <- best_EN_wflow %>%
+                    parsnip::fit(data = req_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_EN <- best_EN_fit$fit$fit$fit
+
+#plot most important predictors in the model
+EN_plot <- vip::vip(x_EN, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+EN_plot
+
+#save manually as "EN-plot.png" -- ggsave throws errors
+
+#find predictions and intervals
+EN_resid <- best_EN_fit %>%
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "EN Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+EN_pred_act
+
+#save file
+EN_pred_act_file = here("results","EN-pred-act.png")
+ggsave(filename = EN_pred_act_file, plot = EN_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
+                            geom_hline(yintercept = 0, color = "red", lty = 2) +
+                            geom_point(color = "#3b528b") +
+                            labs(title = "EN Fit: Residuals",
+                                  x = "Observation Number",
+                                  y = "Residual")
+EN_resids
+
+#save file
+EN_resid_file = here("results","EN-resid.png")
+ggsave(filename = EN_resid_file, plot = EN_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
+                          geom_hline(yintercept = 0, color = "red", lty = 2) +
+                          geom_point(color = "#3b528b") +
+                          labs(title = "EN Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Fitted Log(Requested Funds) ($)",
+                                y = "Residual")
+EN_pred_resid
+
+#save file
+EN_pred_resid_file = here("results","EN-pred-resid.png")
+ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+EN_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+EN_performance <- EN_res %>% tune::show_best(n = 1)
+EN_performance
+
+#compare model performance to null model (and other models)
+EN_RMSE <- EN_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "EN") %>%
+             dplyr::bind_rows(RF_RMSE)
+EN_RMSE
+```
+The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors.
+The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors.
+The EN model has an improved RMSE, but the null model technically still has a lower RMSE.
 
@@
 ### 1. Model Specification
-```{r lasso model specification}
-#define the lasso model
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the LASSO model
 #mixture = 1 identifies the model to be a LASSO model
-lasso_mod <-
-  parsnip::linear_reg(mode = "regression",
+lasso_mod <- linear_reg(mode = "regression",
                            penalty = tune(),
@@
                 parsnip::set_engine("glmnet")
-
-#use the recipe specified earlier
 ```
 
-<br>
+---
 
 ### 2. Workflow Definition
-```{r lasso wflow def}
-#define workflow for lasso regression
+```{r}
+#define lasso workflow
 lasso_wflow <- workflows::workflow() %>%
@@
 
-<br>
+---
 
 ### 3. Tuning Grid Specification
-```{r lasso tune grid spec}
+```{r}
 #tuning grid specification
-lasso_grid <- grid_regular(penalty(), levels = 10)
+lasso_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
 
@@
 
-<br>
+---
 
 ### 4. Tuning Using Cross-Validation and the `tune_grid()` function
-```{r lasso tuning}
+```{r}
 #tune the model with previously specified cross-validation and RMSE as target metric
 lasso_res <- lasso_wflow %>%
-                tune::tune_grid(resamples = req_folds,
+              tune::tune_grid(resamples = req_folds_spatial,
                               grid = lasso_grid,
-                                control = control_grid(verbose = TRUE, 
-                                                       save_pred = TRUE),
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                               metrics = metric_set(rmse))
 
-#look at 15 models with lowest RMSEs
+#look at top 5 EN models
 top_lasso_models <- lasso_res %>%
-                      tune::show_best("rmse", n = 15) %>%
-                      dplyr::arrange(penalty)
+                        tune::show_best("rmse", n = 5)
 top_lasso_models
@@
 #default visualization
-lasso_res %>% autoplot()
+lasso_auto_plot <- lasso_res %>% autoplot()
+lasso_auto_plot
+
+#save file
+lasso_auto_fig_file = here("results","lasso-auto-plot.png")
+ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
 ```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the LASSO model with the lowest rmse
+lasso_lowest_rmse <- lasso_res %>%
+                        tune::select_best("rmse")
+
+#finalize the workflow by using the selected lasso model
+best_lasso_wflow <- lasso_wflow %>%
+                      tune::finalize_workflow(lasso_lowest_rmse)
+best_lasso_wflow
+
+#one last fit on the training data
+best_lasso_fit <- best_lasso_wflow %>%
+                    parsnip::fit(data = req_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_lasso <- best_lasso_fit$fit$fit$fit
+
+#plot most important predictors in the model
+lasso_plot <- vip::vip(x_lasso, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+lasso_plot
+
+#save manually as "LASSO-plot.png" -- ggsave throws errors
+
+#find predictions and intervals
+lasso_resid <- best_lasso_fit %>%
+                broom.mixed::augment(new_data = req_train_data) %>%
+                dplyr::select(.pred, log_req_amt) %>%
+                dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+lasso_pred_act <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "LASSO Fit: Actual vs. Predicted Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+lasso_pred_act
+
+#save file
+lasso_pred_act_file = here("results","lasso-pred-act.png")
+ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+lasso_resids <- ggplot2::ggplot(lasso_resid, aes(x = as.numeric(row.names(lasso_resid)), y = .resid))+
+                                geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                geom_point(color = "#3b528b") +
+                                labs(title = "LASSO Fit: Residuals",
+                                      x = "Observation Number",
+                                      y = "Residual")
+lasso_resids
+
+#save file
+lasso_resid_file = here("results","lasso-resid.png")
+ggsave(filename = lasso_resid_file, plot = lasso_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+lasso_pred_resid <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = .resid))+
+                                    geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                    geom_point(color = "#3b528b") +
+                                    labs(title = "LASSO Fit: Residuals vs Fitted Log(Requested Funds)",
+                                          x = "Fitted Log(Requested Funds) ($)",
+                                          y = "Residual")
+lasso_pred_resid
+
+#save file
+lasso_pred_resid_file = here("results","lasso-pred-resid.png")
+ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+lasso_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+lasso_performance <- lasso_res %>% tune::show_best(n = 1)
+lasso_performance
+
+#compare model performance to null model (and other models)
+lasso_RMSE <- lasso_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "LASSO") %>%
+             dplyr::bind_rows(EN_RMSE)
+lasso_RMSE
+
+#save combined table
+model_comp_file = here("results", "ML-model-comp.Rds")
+saveRDS(lasso_RMSE, file = model_comp_file)
+```
+The LASSO model fit is almost identical to the Elastic Net model fit.
+
+---
+
+## Model Selection and Evaluation
+Based on the results of the models above, the null model actually has the best fit, which warrants further investigation (Especially with the outliers).
+In the mean time, we'll fit the LASSO model as it has the best fit out of the models.
+```{r}
+#fit model to training set but evaluate on the test data
+lasso_fit_test <- best_lasso_wflow %>%
+                    tune::last_fit(split = req_data_split)
+
+#compare test performance against training performance
+lasso_rmse_test <- collect_metrics(lasso_fit_test) %>%
+                   dplyr::select(rmse = .estimate) %>%
+                   dplyr::mutate(data = "test")
+
+lasso_comp <- lasso_RMSE %>%
+                  dplyr::filter(model == "LASSO") %>%
+                  dplyr::transmute(rmse, data = "train") %>%
+                  dplyr::bind_rows(lasso_rmse_test) %>%
+                  dplyr::slice(-3) #don't know why the third row shows up
+lasso_comp
+#RMSEs are different but not as much as previous iterations
+
+#save combined table
+test_comp_file = here("results", "test-train-model-comp.Rds")
+saveRDS(lasso_comp, file = test_comp_file)
+
+#find predictions and intervals
+lasso_resid_fit <- lasso_fit_test %>%
+                      broom.mixed::augment() %>%
+                      dplyr::select(.pred, log_req_amt) %>%
+                      dplyr::mutate(.resid = .pred - log_req_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline draws a 45 degree line, along which the results should fall
+lasso_pred_act_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = log_req_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
+                        geom_point(color = "#3b528b") +
+                        labs(title = "LASSO Test Fit: Predicted vs. Actual Log(Requested Funds)",
+                              x = "Predicted Log(Requested Funds) ($)",
+                              y = "Actual Log(Requested Funds) ($)")
+lasso_pred_act_test
+
+#save file
+lasso_pred_act_test_file = here("results","lasso-pred-act-test.png")
+ggsave(filename = lasso_pred_act_test_file, plot = lasso_pred_act_test)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+lasso_resids_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = as.numeric(row.names(lasso_resid_fit)), y = .resid))+
+                      geom_hline(yintercept = 0, color = "red", lty = 2) +
+                      geom_point(color = "#3b528b") +
+                      labs(title = "LASSO Test Fit: Residuals",
+                            x = "Observation Number",
+                            y = "Residual")
+lasso_resids_test
+
+#save file
+lasso_resid_test_file = here("results","lasso-resid-test.png")
+ggsave(filename = lasso_resid_test_file, plot = lasso_resids_test)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+lasso_pred_resid_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = .resid))+
+                          geom_hline(yintercept = 0, color = "red", lty = 2) +
+                          geom_point(color = "#3b528b") +
+                          labs(title = "LASSO Test Fit: Residuals vs Fitted Log(Requested Funds)",
+                                x = "Predicted Log(Requested Funds) ($)",
+                                y = "Residuals")
+lasso_pred_resid_test
+
+#save file
+lasso_pred_resid_test_file = here("results","lasso-pred-resid-test.png")
+ggsave(filename = lasso_pred_resid_test_file, plot = lasso_pred_resid_test)
+```

### 2021-12-12 code/analysis_code/analysis-eda.Rmd
ML update - new models, started outliers

@@
 
-# globally set theme for ggplots
+#set ggplot theme to classic
 ggplot2::theme_set(theme_classic())
+
+#center plot titles
+theme_update(plot.title = element_text(hjust = 0.5))
 ```
@@
 #define file path
-data_location <- here::here("data", "processed_data", "combinedprocessed.rds")
+data_location <- here::here("data", "processed_data", "analysisdata.rds")
 
@@
 summarytools::dfSummary(EDAdata)
+
+skimr::skim(EDAdata)
 ```
@@
 * There are 445 distinct disaster numbers across 2012 - 2021
-* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297
-* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486
-* There are 16 possible incident types
+* `OblAmt` ranges from -\$0 to \$4,408,424,297
+* `ReqAmt` ranges from -\$0 to \$4,388,540,486
+* There are 9 possible incident types
 * There is a clear effect of both incident month and incident year
@@
 #try a log transformation out of curiosity 
-#this does ignore all negative values
 ggplot2::ggplot(data = EDAdata, aes(x = log(ReqAmt))) +
@@
 We may need to remove outliers in the analysis, but for now, we'll leave them to capture the true nature of disaster declarations.
-The largest outlier is likely the COVID-19 Pandemic.
-The log transformation does a better job of showing the distribution of the data, so we may need to consider an exponential model in the analysis; however, log(negative number) is undefined, so we would be ignoring all reimbursement data.
+The largest outlier is likely Hurricane Maria.
+The log transformation does a better job of showing the distribution of the data, so we may need to consider an exponential model in the analysis.
 The transformed and untransformed figures suggest there are significant tails in the data.
@@
 It appears that obligated funds and requested funds have similar distributions and characteristics. The same commentary about the requested funds analysis applies here as well.
+
+Before moving on, I want to create a plot of Requested vs obligated Funds to understand the relationship between the two outcomes:
+```{r}
+funds_comp <- EDAdata %>%
+                ggplot2::ggplot(aes(x = log(ReqAmt), y = log(OblAmt))) +
+                         geom_point()
+
+funds_comp
+```
 

### 2021-12-12 code/analysis_code/analysis-ML-OA-outlier.Rmd
Last ML model update

--- 2021-11-12 code/analysis_code/analysis-ML-OA.Rmd
@@
 * Random Forest
-* Bagged Tree
+* Elastic Net
+* LASSO
 
-It compares the  models, and then finally fits the “best” model to the test data.
+It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis where the log(OblAmt) was less than zero.
 
@@
 ## Load Data
-Load the analysis data from the `processed_data` folder in the project file.
+Load the OblAmt analysis data from the `processed_data` folder in the project file.
 ```{r load data}
@@
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","analysisdata.rds")
+data_location <- here::here("data","processed_data","analysisdata-OA.rds")
 
@@
 
-* Drop the unused variables
-* Perform feature engineering
-* Transform Incident Month into ordinal variable
+* Log transform the outcome
+* Drop `OblAmt`, `disasterNumber`, `declarationTitle`
+* Remove the obvious outliers (-log(OblAmt))
 * Set the random seed to 123
-* Split the dataset into 70% training, 30% testing, stratified on incidentType
-* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
+* Split the dataset into 70% training, 30% testing
+* 5-fold cross validation for the spatial CV folds
 * Create a recipe for data and fitting that codes categorical variables as dummy variables
 ```{r data setup}
-#requested amount filter
-obl_data <- analysisdata %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)
+#log transform OblAmt
+analysisdata$logOblAmt <- log(analysisdata$OblAmt)
 
-#transform incident month into ordinal variable
-obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
-                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
+#drop unnecessary variables
+obl_data_2 <- analysisdata %>%
+              dplyr::select(-c(OblAmt, disasterNumber, declarationTitle))
 
-#not enough values per state, so create a variable that is FEMA Regions
-obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
-                                 obl_data$state == "ME" |
-                                 obl_data$state == "MA" |
-                                 obl_data$state == "NH" |
-                                 obl_data$state == "RI" |
-                                 obl_data$state == "VT"),
-                              "Region I",
-                              ifelse((obl_data$state == "NJ" |
-                                         obl_data$state == "NY" |
-                                         obl_data$state == "PR" |
-                                         obl_data$state == "VI"),
-                                      "Region II",
-                                       ifelse((obl_data$state == "DE" |
-                                                 obl_data$state == "MD" |
-                                                 obl_data$state == "PA" |
-                                                 obl_data$state == "VA" |
-                                                 obl_data$state == "DC" |
-                                                 obl_data$state == "WV"),
-                                              "Region III",
-                                               ifelse((obl_data$state == "AL" |
-                                                         obl_data$state == "FL" |
-                                                         obl_data$state == "GA" |
-                                                         obl_data$state == "KY" |
-                                                         obl_data$state == "MS" |
-                                                         obl_data$state == "NC" |
-                                                         obl_data$state == "SC" |
-                                                         obl_data$state == "TN"),
-                                                      "Region IV",
-                                                      ifelse((obl_data$state == "IL" |
-                                                               obl_data$state == "IN" |
-                                                               obl_data$state == "MI" |
-                                                               obl_data$state == "MN" |
-                                                               obl_data$state == "OH" |
-                                                               obl_data$state == "WI"),
-                                                            "Region V",
-                                                             ifelse((obl_data$state == "AR" |
-                                                                         obl_data$state == "LA" |
-                                                                         obl_data$state == "NM" |
-                                                                         obl_data$state == "OK" |
-                                                                         obl_data$state == "TX"),
-                                                                      "Region VI",
-                                                                       ifelse((obl_data$state == "IA" |
-                                                                                 obl_data$state == "KS" |
-                                                                                 obl_data$state == "MO" |
-                                                                                 obl_data$state == "NE"),
-                                                                              "Region VII",
-                                                                               ifelse((obl_data$state == "CO" |
-                                                                                         obl_data$state == "MT" |
-                                                                                         obl_data$state == "ND" |
-                                                                                         obl_data$state == "SD" |
-                                                                                         obl_data$state == "UT" |
-                                                                                         obl_data$state == "WY"),
-                                                                                      "Region VIII",
-                                                                                       ifelse((obl_data$state == "AZ" |
-                                                                                                 obl_data$state == "CA" |
-                                                                                                 obl_data$state == "HI" |
-                                                                                                 obl_data$state == "NV"),
-                                                                                              "Region IX",
-                                                                                               ifelse((obl_data$state == "AK" |
-                                                                                                         obl_data$state == "ID" |
-                                                                                                         obl_data$state == "OR" |
-                                                                                                         obl_data$state == "WA"),
-                                                                                                      "Region X",
-                                                                                                      NA
-                                                                                                      )
-                                                                                              )
-                                                                                      )
-                                                                              )
-                                                                      )
-                                                            )
-                                                      )
-                                              )
-                                     )
-                              )
+#specify all character variables as factors and clean variable names
+obl_data_1 <- obl_data_2 %>% 
+                janitor::clean_names() %>%
+                dplyr::mutate_if(is.character, factor)
 
-#drop state variable
-obl_data <- obl_data %>%
-              dplyr::select(-c(state))
-
-#add to "other" category for incident Types that are infrequent
-obl_data$incidentType[(obl_data$incidentType == "Chemical" |
-                        obl_data$incidentType == "Coastal Storm" |
-                        obl_data$incidentType == "Dam/Levee Break" |
-                        obl_data$incidentType == "Mud/Landslide" |
-                        obl_data$incidentType == "Volcano" |
-                        obl_data$incidentType == "Terrorist" |
-                        obl_data$incidentType == "Snow")] <- "Other"
+#remove outliers (rows that have a -log(oblamt))
+obl_data <- obl_data_1 %>%
+              dplyr::filter(log_obl_amt > 0)
 
@@
 #use incidentType as stratification
-obl_data_split <- rsample::initial_split(obl_data, prop = 2/3,
-                                     strata = incidentType)
+obl_data_split <- rsample::initial_split(obl_data, prop = 2/3)
 
@@
 
-#5-fold cross validation, repeated 5 times, stratified by incidentType
-obl_folds <- rsample::vfold_cv(obl_train_data,
-                               strata = incidentType,
-                               v = 5,
-                               repeats = 5)
+#5-fold spatial cross validation
+obl_folds_spatial <- spatialsample::spatial_clustering_cv(obl_train_data,
+                                                          coords = c(latitude, longitude),
+                                                          v = 5)
 
 #create recipe that codes categorical variables as dummy variables
-OA_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
+OA_rec <- recipes::recipe(log_obl_amt ~ ., data = obl_train_data) %>%
+          recipes::step_unknown(state, new_level = "unknown state") %>%
           recipes::step_dummy(all_nominal_predictors())
@@
 obl_null_RMSE_train <- tibble::tibble(
-                          rmse = rmse_vec(truth = obl_train_data$OblAmt,
-                                              estimate = rep(mean(obl_train_data$OblAmt), nrow(obl_train_data))),
+                          rmse = rmse_vec(truth = obl_train_data$log_obl_amt,
+                                              estimate = rep(mean(obl_train_data$log_obl_amt), nrow(obl_train_data))),
                           SE = 0,
@@
 obl_null_RMSE_test <- tibble::tibble(
-                          rmse = rmse_vec(truth = obl_test_data$OblAmt,
-                                              estimate = rep(mean(obl_test_data$OblAmt), nrow(obl_test_data))),
+                          rmse = rmse_vec(truth = obl_test_data$log_obl_amt,
+                                              estimate = rep(mean(obl_test_data$log_obl_amt), nrow(obl_test_data))),
                           SE = 0,
@@
   parsnip::set_mode("regression")
+```
 
-#use the recipe specified earlier (line 133)
-```
 ---
@@
 tree_res <- tree_wflow %>%
-                tune::tune_grid(resamples = obl_folds,
+                tune::tune_grid(resamples = obl_folds_spatial,
                                 grid = tree_grid,
@@
 #save file
-DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
+DT_auto_fig_file = here("results","DT-auto-plot-OA-outlier.png")
 ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)
@@
 #save file
-DT_detail_fig_file = here("results","DT-det-plot-obl.png")
-ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)
+DT_det_fig_file = here("results","DT-det-plot-OA-outlier.png")
+ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)
 
@@
                                                      digits = 5,
-                                         main = "Selected Decision Tree Model")
-#save file manually -- ggsave throws an error
+                                                     main = "Selected Decision Tree Model",
+                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
+                                                     shadow.col = "grey65",
+                                                     col = "grey99")
+#save manually as "DT-plot-OA-outlier.png" -- ggsave throws an error
 
@@
                   broom.mixed::augment(new_data = obl_train_data) %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                  dplyr::select(.pred, log_obl_amt) %>%
+                  dplyr::mutate(.resid = .pred - log_obl_amt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = OblAmt)) +
+DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Decision Tree Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "Decision Tree Fit: Predicted vs. Actual Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)") 
 DT_pred_act
@@
 #save file
-DT_pred_act_file = here("results","DT-pred-act-obl.png")
+DT_pred_act_file = here("results","DT-pred-act-OA-outlier.png")
 ggsave(filename = DT_pred_act_file, plot = DT_pred_act)
@@
 #save file
-DT_resid_file = here("results","DT-resid-obl.png")
+DT_resid_file = here("results","DT-resid-OA-outlier.png")
 ggsave(filename = DT_resid_file, plot = DT_resid)
@@
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Decision Tree Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Decision Tree Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Predicted Log(Obligated Funds) ($)",
                                 y = "Residual")
@@
 #save file
-DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
+DT_pred_resid_file = here("results","DT-pred-resid-OA-outlier.png")
 ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)
@@
 ```
+
+The identified best performing tree model has some interesting results.
+It identifies percent of state counties awarded IH program, total number of agencies involved, total number of counties per state, percent of counties awarded PA program, FEMA Region VI, incident month of October and March as important predictors.
+It is interesting that incident type or declaration type are not included as significant.
+The RMSE does decrease from the null, which suggests the outlier removal was a necessary step.
+Let's explore the other models.
 
@@
 RF_res <- RF_wflow %>%
-              tune::tune_grid(resamples = obl_folds,
+              tune::tune_grid(resamples = obl_folds_spatial,
                               grid = RF_grid,
@@
 #save file
-RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
+RF_auto_fig_file = here("results","RF-auto-plot-OA-outlier.png")
 ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
@@
 #plot most important predictors in the model
-RF_plot <- vip::vip(x_RF, num_features = 20)
+RF_plot <- vip::vip(x_RF, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
 RF_plot
 
-#save  manually - ggsave throws an error
+#save manually as "RF-plot-OA-outlier.png" -- ggsave throws errors
 
@@
                 broom.mixed::augment(new_data = obl_train_data) %>%
-                dplyr::select(.pred, OblAmt) %>%
-                dplyr::mutate(.resid = .pred - OblAmt)
+                dplyr::select(.pred, log_obl_amt) %>%
+                dplyr::mutate(.resid = .pred - log_obl_amt)
 
@@
 #geom_abline is a 45 degree line, along which the results should fall
-RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
+RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
-                        geom_point() +
-                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "RF Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+RF_pred_act
 
 #save file
-RF_pred_act_file = here("results","RF-pred-act-obl.png")
+RF_pred_act_file = here("results","RF-pred-act-OA-outlier.png")
 ggsave(filename = RF_pred_act_file, plot = RF_pred_act)
@@
 #save file
-RF_resid_file = here("results","RF-resid-obl.png")
+RF_resid_file = here("results","RF-resid-OA-outlier.png")
 ggsave(filename = RF_resid_file, plot = RF_resids)
@@
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Random Forest Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Fitted Requested Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "Random Forest Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Fitted Log(Obligated Funds) ($)",
                                 y = "Residual")
@@
 #save file
-RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
+RF_pred_resid_file = here("results","RF-pred-resid-OA-outlier.png")
 ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)
@@
 
+In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
+It identifies percent of counties awarded IH program, total number of federal agencies involved, percent of counties awarded HM program, average FEMA cost share, and whether or not the IH program was awarded to any county in the state as the five most important predictors.
+Let's try another model to see if we can find an improvement in the RMSE.
+
 ---
 
-## Bagged Decision Tree Model
+## Elastic Net Model
 
@@
 
-#start over with data split to allow for bootstrapping
-#set random seed
-set.seed(20140102)
+#define the elastic net model
+elastic_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = tune()) %>%
+                parsnip::set_engine("glmnet")
+```
 
-#use bootstrap instead of CV
-obl_data_bootsplit <- obl_train_data %>%
-                        rsample::bootstraps(times = 10,
-                                            strata = "incidentType")
-
-#define model
-bag_tree_mod <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-                parsnip::set_engine("rpart",
-                                    times = 10) %>%
-                parsnip::set_mode("regression")
-
-#create recipe that codes categorical variables as dummy variables
-BT_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
-          step_string2factor(all_nominal())
-```
 ---
@@
 ```{r}
-#define workflow for tree
-BT_wflow <- workflows::workflow() %>%
-               workflows::add_model(bag_tree_mod) %>%
-               workflows::add_recipe(BT_rec)
+#define elastic net workflow
+elastic_wflow <- workflows::workflow() %>%
+                 workflows::add_model(elastic_mod) %>%
+                 workflows::add_recipe(OA_rec)
 ```
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-          parsnip::set_engine("rpart",
-                              times = 10) %>%
-          parsnip::set_mode("regression") %>%
-          tune::tune_grid(preprocessor = BT_rec,
-                          resamples = obl_data_bootsplit,
-                          grid = tree_grid,
+EN_res <- elastic_wflow %>%
+              tune::tune_grid(resamples = obl_folds_spatial,
+                              grid = EN_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                               metrics = metric_set(rmse))
 
-#collect metrics
-BT_res %>% workflowsets::collect_metrics()
+#look at top 5 EN models
+top_EN_models <- EN_res %>%
+                    tune::show_best("rmse", n = 5)
+top_EN_models
 
 #default visualization
-BT_auto_plot <- BT_res %>% autoplot()
-BT_auto_plot
+EN_auto_plot <- EN_res %>% autoplot()
+EN_auto_plot
 
 #save file
-BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
-ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)
-
-#more detailed plot
-BT_det_plot <- BT_res %>%
-                    workflowsets::collect_metrics() %>%
-                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
-                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
-                             geom_line(size = 1.5, alpha = 0.6) +
-                             geom_point(size = 2) +
-                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
-                             scale_x_log10(labels = scales::label_number()) +
-                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
-BT_det_plot
-
-#save file
-BT_det_fig_file = here("results","BT-det-plot-obl.png")
-ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
+EN_auto_fig_file = here("results","EN-auto-plot-OA-outlier.png")
+ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
 ```
@@
 ```{r}
-#select the BT model with the lowest rmse
-BT_lowest_rmse <- BT_res %>%
+#select the EN model with the lowest rmse
+EN_lowest_rmse <- EN_res %>%
                       tune::select_best("rmse")
 
-#finalize the workflow by using the selected BT model
-best_BT_wflow <- BT_wflow %>%
-                      tune::finalize_workflow(BT_lowest_rmse)
-best_BT_wflow
+#finalize the workflow by using the selected RF model
+best_EN_wflow <- elastic_wflow %>%
+                      tune::finalize_workflow(EN_lowest_rmse)
+best_EN_wflow
 
 #one last fit on the training data
-best_BT_fit <- best_BT_wflow %>%
+best_EN_fit <- best_EN_wflow %>%
                     parsnip::fit(data = obl_train_data)
@@
 #extract model from final fit
-x_BT <- best_BT_fit$fit$fit$fit
+x_EN <- best_EN_fit$fit$fit$fit
 
-#vip package doesn't support BT
+#plot most important predictors in the model
+EN_plot <- vip::vip(x_EN, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+EN_plot
+
+#save manually as "EN-plot-OA-outlier.png" -- ggsave throws errors
 
 #find predictions and intervals
-BT_resid <- best_BT_fit %>%
+EN_resid <- best_EN_fit %>%
                 broom.mixed::augment(new_data = obl_train_data) %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                dplyr::select(.pred, log_obl_amt) %>%
+                dplyr::mutate(.resid = .pred - log_obl_amt)
 
 #plot model predictions from tuned model versus actual outcomes
-#geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
+#geom_abline is a 45 degree line, along which the results should fall
+EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
-                        geom_point() +
-                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
-BT_pred_act
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "EN Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+EN_pred_act
 
 #save file
-BT_pred_act_file = here("results","BT-pred-act-obl.png")
-ggsave(filename = BT_pred_act_file, plot = BT_pred_act)
+EN_pred_act_file = here("results","EN-pred-act-OA-outlier.png")
+ggsave(filename = EN_pred_act_file, plot = EN_pred_act)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
+EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
                             geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Fit: Residuals",
+                            geom_point(color = "#3b528b") +
+                            labs(title = "EN Fit: Residuals",
                                   x = "Observation Number",
                                   y = "Residual")
-BT_resids
+EN_resids
 
 #save file
-BT_resid_file = here("results","BT-resids-obl.png")
-ggsave(filename = BT_resid_file, plot = BT_resids)
+EN_resid_file = here("results","EN-resid-OA-outlier.png")
+ggsave(filename = EN_resid_file, plot = EN_resids)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
+EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "EN Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Fitted Log(Obligated Funds) ($)",
                                 y = "Residual")
-BT_pred_resid
+EN_pred_resid
 
 #save file
-BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
-ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)
+EN_pred_resid_file = here("results","EN-pred-resid-OA-outlier.png")
+ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)
 
-#print model performance
-#print 10 best performing hyperparameter sets
-BT_res %>%
+#print the 10 best performing hyperparameter sets
+EN_res %>%
   tune::show_best(n = 10) %>%
-  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
-  dplyr::mutate(rmse = round(rmse, 3),
-                std_err = round(std_err, 4),
-                cost_complexity = scales::scientific(cost_complexity))
+  dplyr::select(rmse = mean, std_err)
 
-#print the best model performance
-BT_performance <- BT_res %>% tune::show_best(n = 1)
-print(BT_performance)
+#print the best model
+EN_performance <- EN_res %>% tune::show_best(n = 1)
+EN_performance
 
-#compare model performance to null model
-BT_RMSE <- BT_res %>%
+#compare model performance to null model (and other models)
+EN_RMSE <- EN_res %>%
               tune::show_best(n = 1) %>%
@@
                 SE = round(std_err, 4),
-                  model = "Bagged Tree") %>%
+                model = "EN") %>%
              dplyr::bind_rows(RF_RMSE)
-BT_RMSE
+EN_RMSE
+```
+The Elastic Net model places great emphasis on Maine being the state receving the declaration.
+The next closest predictors include average FEMA cost share, Vermont, Montana, and Idaho.
+The EN model has an improved RMSE from the DT model but not the RF model.
+Let's try LASSO just to round out the analysis.
+
+---
+
+## LASSO Model
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the LASSO model
+#mixture = 1 identifies the model to be a LASSO model
+lasso_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = 1) %>%
+                parsnip::set_engine("glmnet")
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define lasso workflow
+lasso_wflow <- workflows::workflow() %>%
+                 workflows::add_model(lasso_mod) %>%
+                 workflows::add_recipe(OA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+lasso_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
+
+#5 lowest penalty values
+lasso_grid %>%
+  dplyr::top_n(-5)
+
+#5 highest penalty values
+lasso_grid %>%
+  dplyr::top_n(5)
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+lasso_res <- lasso_wflow %>%
+              tune::tune_grid(resamples = obl_folds_spatial,
+                              grid = lasso_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
+
+#look at top 5 EN models
+top_lasso_models <- lasso_res %>%
+                        tune::show_best("rmse", n = 5)
+top_lasso_models
+
+#default visualization
+lasso_auto_plot <- lasso_res %>% autoplot()
+lasso_auto_plot
+
+#save file
+lasso_auto_fig_file = here("results","lasso-auto-plot-OA-outlier.png")
+ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the LASSO model with the lowest rmse
+lasso_lowest_rmse <- lasso_res %>%
+                        tune::select_best("rmse")
+
+#finalize the workflow by using the selected lasso model
+best_lasso_wflow <- lasso_wflow %>%
+                      tune::finalize_workflow(lasso_lowest_rmse)
+best_lasso_wflow
+
+#one last fit on the training data
+best_lasso_fit <- best_lasso_wflow %>%
+                    parsnip::fit(data = obl_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_lasso <- best_lasso_fit$fit$fit$fit
+
+#plot most important predictors in the model
+lasso_plot <- vip::vip(x_lasso, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+lasso_plot
+
+#save manually as "LASSO-plot-OA-outlier.png" -- ggsave throws errors
+
+#find predictions and intervals
+lasso_resid <- best_lasso_fit %>%
+                broom.mixed::augment(new_data = obl_train_data) %>%
+                dplyr::select(.pred, log_obl_amt) %>%
+                dplyr::mutate(.resid = .pred - log_obl_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+lasso_pred_act <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = log_obl_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "LASSO Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+lasso_pred_act
+
+#save file
+lasso_pred_act_file = here("results","lasso-pred-act-OA-outlier.png")
+ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+lasso_resids <- ggplot2::ggplot(lasso_resid, aes(x = as.numeric(row.names(lasso_resid)), y = .resid))+
+                                geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                geom_point(color = "#3b528b") +
+                                labs(title = "LASSO Fit: Residuals",
+                                      x = "Observation Number",
+                                      y = "Residual")
+lasso_resids
+
+#save file
+lasso_resid_file = here("results","lasso-resid-OA-outlier.png")
+ggsave(filename = lasso_resid_file, plot = lasso_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+lasso_pred_resid <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = .resid))+
+                                    geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                    geom_point(color = "#3b528b") +
+                                    labs(title = "LASSO Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                          x = "Fitted Log(Obligated Funds) ($)",
+                                          y = "Residual")
+lasso_pred_resid
+
+#save file
+lasso_pred_resid_file = here("results","lasso-pred-resid-OA-outlier.png")
+ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+lasso_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+lasso_performance <- lasso_res %>% tune::show_best(n = 1)
+lasso_performance
+
+#compare model performance to null model (and other models)
+lasso_RMSE <- lasso_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "LASSO") %>%
+             dplyr::bind_rows(EN_RMSE)
+lasso_RMSE
 
 #save combined table
-model_comp_file = here("results", "ML-model-comp-obl.Rds")
-saveRDS(BT_RMSE, file = model_comp_file)
+model_comp_file = here("results", "ML-model-comp-OA-outlier.Rds")
+saveRDS(lasso_RMSE, file = model_comp_file)
 ```
+The LASSO model fit is almost identical to the Elastic Net model fit.
 
-In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.
-
---
+---
 
 ## Model Selection and Evaluation
-Using the bagged tree model
+Based on the results of the model comparison table, all of the models have a similar fit. However, RF has the lowest RMSE, so we will fit that to the test data.
 ```{r}
 #fit model to training set but evaluate on the test data
-BT_fit_test <- best_BT_wflow %>%
+RF_fit_test <- best_RF_wflow %>%
                     tune::last_fit(split = obl_data_split)
@@
 #compare test performance against training performance
-BT_rmse_test <- collect_metrics(BT_fit_test) %>%
+RF_rmse_test <- collect_metrics(RF_fit_test) %>%
                    dplyr::select(rmse = .estimate) %>%
@@
 
-BT_comp <- BT_RMSE %>%
-                  dplyr::filter(model == "Bagged Tree") %>%
+test_comp <- RF_RMSE %>%
+                  dplyr::filter(model == "Random Forest") %>%
                   dplyr::transmute(rmse, data = "train") %>%
-                  dplyr::bind_rows(BT_rmse_test) %>%
+                  dplyr::bind_rows(RF_rmse_test) %>%
                   dplyr::slice(-3) #don't know why the third row shows up
-BT_comp
-#RMSEs are incredibly different...
+test_comp
+#RMSEs are very different (but lower in test data?)
+
+#save combined table
+test_comp_file = here("results", "test-train-model-comp-OA-outlier.Rds")
+saveRDS(test_comp, file = test_comp_file)
 
 #find predictions and intervals
-BT_resid_fit <- BT_fit_test %>%
+RF_resid_fit <- RF_fit_test %>%
                       broom.mixed::augment() %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                      dplyr::select(.pred, log_obl_amt) %>%
+                      dplyr::mutate(.resid = .pred - log_obl_amt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = OblAmt)) +
+RF_pred_act_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
-BT_pred_act_test
+                        geom_point(color = "#3b528b") +
+                        labs(title = "RF Test Fit: Predicted vs. Actual Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+RF_pred_act_test
 
 #save file
-BT_pred_act_test_file = here("results","BT-pred-act-test-obl.png")
-ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)
+RF_pred_act_test_file = here("results","RF-pred-act-test-OA-outlier.png")
+ggsave(filename = RF_pred_act_test_file, plot = RF_pred_act_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids_test <- ggplot2::ggplot(BT_resid_fit, aes(x = as.numeric(row.names(BT_resid_fit)), y = .resid))+
+RF_resids_test <- ggplot2::ggplot(RF_resid_fit, aes(x = as.numeric(row.names(RF_resid_fit)), y = .resid))+
                       geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Test Fit: Residuals",
+                      geom_point(color = "#3b528b") +
+                      labs(title = "RF Test Fit: Residuals",
                             x = "Observation Number",
                             y = "Residual")
-BT_resids_test
+RF_resids_test
 
 #save file
-BT_resid_test_file = here("results","BT-resid-test-obl.png")
-ggsave(filename = BT_resid_test_file, plot = BT_resids_test)
+RF_resid_test_file = here("results","RF-resid-test-OA-outlier.png")
+ggsave(filename = RF_resid_test_file, plot = RF_resids_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
+RF_pred_resid_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
-                                y = "Residual")
-BT_pred_resid_test
+                          geom_point(color = "#3b528b") +
+                          labs(title = "RF Test Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Predicted Log(Requested Funds) ($)",
+                                y = "Residuals")
+RF_pred_resid_test
 
 #save file
-BT_pred_resid_test_file = here("results","BT-pred-resid-test-obl.png")
-ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)
+RF_pred_resid_test_file = here("results","RF-pred-resid-test-OA-outlier.png")
+ggsave(filename = RF_pred_resid_test_file, plot = RF_pred_resid_test)
 ```

### 2021-12-12 code/analysis_code/analysis-ML-OA.Rmd
Last ML model update

@@
 * Random Forest
-* Bagged Tree
+* Elastic Net
+* LASSO
 
@@
 * ranger: for random forest models
-* baguette: for bagged decision tree
+* viridis: for a consistent color scheme in figures
 ```{r libraries, include = FALSE}
@@
 library(ranger) #for random forest models
-library(baguette) #for bagged decision tree
+library(viridis) #for a consistent color scheme in figures
 
@@
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","analysisdata.rds")
+data_location <- here::here("data","processed_data","analysisdata-OA.rds")
 
 #load data. 
-analysisdata <- readRDS(data_location)
+obl_data <- readRDS(data_location)
 
 #summary of data using skimr package
-skimr::skim(analysisdata)
+skimr::skim(obl_data)
 ```
@@
 
-* Drop the unused variables
-* Perform feature engineering
-* Transform Incident Month into ordinal variable
+* Log transform the outcome
+* Drop `ReqAmt`, `disasterNumber`, `declarationTitle`
 * Set the random seed to 123
-* Split the dataset into 70% training, 30% testing, stratified on incidentType
-* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
+* Split the dataset into 70% training, 30% testing
+* 5-fold cross validation for the spatial CV folds
 * Create a recipe for data and fitting that codes categorical variables as dummy variables
 ```{r data setup}
-#requested amount filter
-obl_data <- analysisdata %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)
+#log transform ReqAmt
+obl_data$logOblAmt <- log(obl_data$OblAmt)
 
-#transform incident month into ordinal variable
-obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
-                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
+#drop unnecessary variables
+obl_data <- obl_data %>%
+              dplyr::select(-c(OblAmt, disasterNumber, declarationTitle))
 
-#not enough values per state, so create a variable that is FEMA Regions
-obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
-                                 obl_data$state == "ME" |
-                                 obl_data$state == "MA" |
-                                 obl_data$state == "NH" |
-                                 obl_data$state == "RI" |
-                                 obl_data$state == "VT"),
-                              "Region I",
-                              ifelse((obl_data$state == "NJ" |
-                                         obl_data$state == "NY" |
-                                         obl_data$state == "PR" |
-                                         obl_data$state == "VI"),
-                                      "Region II",
-                                       ifelse((obl_data$state == "DE" |
-                                                 obl_data$state == "MD" |
-                                                 obl_data$state == "PA" |
-                                                 obl_data$state == "VA" |
-                                                 obl_data$state == "DC" |
-                                                 obl_data$state == "WV"),
-                                              "Region III",
-                                               ifelse((obl_data$state == "AL" |
-                                                         obl_data$state == "FL" |
-                                                         obl_data$state == "GA" |
-                                                         obl_data$state == "KY" |
-                                                         obl_data$state == "MS" |
-                                                         obl_data$state == "NC" |
-                                                         obl_data$state == "SC" |
-                                                         obl_data$state == "TN"),
-                                                      "Region IV",
-                                                      ifelse((obl_data$state == "IL" |
-                                                               obl_data$state == "IN" |
-                                                               obl_data$state == "MI" |
-                                                               obl_data$state == "MN" |
-                                                               obl_data$state == "OH" |
-                                                               obl_data$state == "WI"),
-                                                            "Region V",
-                                                             ifelse((obl_data$state == "AR" |
-                                                                         obl_data$state == "LA" |
-                                                                         obl_data$state == "NM" |
-                                                                         obl_data$state == "OK" |
-                                                                         obl_data$state == "TX"),
-                                                                      "Region VI",
-                                                                       ifelse((obl_data$state == "IA" |
-                                                                                 obl_data$state == "KS" |
-                                                                                 obl_data$state == "MO" |
-                                                                                 obl_data$state == "NE"),
-                                                                              "Region VII",
-                                                                               ifelse((obl_data$state == "CO" |
-                                                                                         obl_data$state == "MT" |
-                                                                                         obl_data$state == "ND" |
-                                                                                         obl_data$state == "SD" |
-                                                                                         obl_data$state == "UT" |
-                                                                                         obl_data$state == "WY"),
-                                                                                      "Region VIII",
-                                                                                       ifelse((obl_data$state == "AZ" |
-                                                                                                 obl_data$state == "CA" |
-                                                                                                 obl_data$state == "HI" |
-                                                                                                 obl_data$state == "NV"),
-                                                                                              "Region IX",
-                                                                                               ifelse((obl_data$state == "AK" |
-                                                                                                         obl_data$state == "ID" |
-                                                                                                         obl_data$state == "OR" |
-                                                                                                         obl_data$state == "WA"),
-                                                                                                      "Region X",
-                                                                                                      NA
-                                                                                                      )
-                                                                                              )
-                                                                                      )
-                                                                              )
-                                                                      )
-                                                            )
-                                                      )
-                                              )
-                                     )
-                              )
-
-#drop state variable
+#specify all character variables as factors and clean variable names
 obl_data <- obl_data %>% 
-              dplyr::select(-c(state))
-
-#add to "other" category for incident Types that are infrequent
-obl_data$incidentType[(obl_data$incidentType == "Chemical" |
-                        obl_data$incidentType == "Coastal Storm" |
-                        obl_data$incidentType == "Dam/Levee Break" |
-                        obl_data$incidentType == "Mud/Landslide" |
-                        obl_data$incidentType == "Volcano" |
-                        obl_data$incidentType == "Terrorist" |
-                        obl_data$incidentType == "Snow")] <- "Other"
+              janitor::clean_names() %>%
+              dplyr::mutate_if(is.character, factor)
 
@@
 #save file
-DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
+DT_auto_fig_file = here("results","DT-auto-plot-OA.png")
 ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)
@@
 #save file
-DT_detail_fig_file = here("results","DT-det-plot-obl.png")
-ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)
+DT_det_fig_file = here("results","DT-det-plot-OA.png")
+ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)
 
@@
                                                      digits = 5,
-                                         main = "Selected Decision Tree Model")
-#save file manually -- ggsave throws an error
+                                                     main = "Selected Decision Tree Model",
+                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
+                                                     shadow.col = "grey65",
+                                                     col = "grey99")
+#save manually as "DT-plot-OA.png" -- ggsave throws an error
 
@@
 #save file
-DT_pred_act_file = here("results","DT-pred-act-obl.png")
+DT_pred_act_file = here("results","DT-pred-act-OA.png")
 ggsave(filename = DT_pred_act_file, plot = DT_pred_act)
@@
 #save file
-DT_resid_file = here("results","DT-resid-obl.png")
+DT_resid_file = here("results","DT-resid-OA.png")
 ggsave(filename = DT_resid_file, plot = DT_resid)
@@
 #save file
-DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
+DT_pred_resid_file = here("results","DT-pred-resid-OA.png")
 ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)
@@
 ```
+
+The identified best performing tree model has some interesting results.
+It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding.
+New York and November are particularly surprising key predictors.
+It is interesting that incident type or declaration type are not included as significant.
+The RMSE actually increases from the Null, which suggests this is not an appropriate model.
+Moving on.
 
@@
 #save file
-RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
+RF_auto_fig_file = here("results","RF-auto-plot-OA.png")
 ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
@@
 #plot most important predictors in the model
-RF_plot <- vip::vip(x_RF, num_features = 20)
+RF_plot <- vip::vip(x_RF, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
 RF_plot
 
-#save  manually - ggsave throws an error
+#save manually as "RF-plot-OA.png" -- ggsave throws errors
 
@@
 #geom_abline is a 45 degree line, along which the results should fall
-RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
+RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
-                        geom_point() +
-                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "RF Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+RF_pred_act
 
 #save file
-RF_pred_act_file = here("results","RF-pred-act-obl.png")
+RF_pred_act_file = here("results","RF-pred-act-OA.png")
 ggsave(filename = RF_pred_act_file, plot = RF_pred_act)
@@
 #save file
-RF_resid_file = here("results","RF-resid-obl.png")
+RF_resid_file = here("results","RF-resid-OA.png")
 ggsave(filename = RF_resid_file, plot = RF_resids)
@@
 #save file
-RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
+RF_pred_resid_file = here("results","RF-pred-resid-OA.png")
 ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)
@@
 
+In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
+It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and incident duration as the five most important predictors.
+Let's try another model to see if we can find an improvement in the RMSE.
+
 ---
 
-## Bagged Decision Tree Model
+## Elastic Net Model
 
@@
 #tune the model with previously specified cross-validation and RMSE as target metric
-BT_res <- baguette::bag_tree(cost_complexity = tune(),
-                                   tree_depth = tune(),
-                                   min_n = tune()) %>%
-          parsnip::set_engine("rpart",
-                              times = 10) %>%
-          parsnip::set_mode("regression") %>%
-          tune::tune_grid(preprocessor = BT_rec,
-                          resamples = obl_data_bootsplit,
-                          grid = tree_grid,
+EN_res <- elastic_wflow %>%
+              tune::tune_grid(resamples = obl_folds_spatial,
+                              grid = EN_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                               metrics = metric_set(rmse))
 
-#collect metrics
-BT_res %>% workflowsets::collect_metrics()
+#look at top 5 EN models
+top_EN_models <- EN_res %>%
+                    tune::show_best("rmse", n = 5)
+top_EN_models
 
 #default visualization
-BT_auto_plot <- BT_res %>% autoplot()
-BT_auto_plot
+EN_auto_plot <- EN_res %>% autoplot()
+EN_auto_plot
 
 #save file
-BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
-ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)
-
-#more detailed plot
-BT_det_plot <- BT_res %>%
-                    workflowsets::collect_metrics() %>%
-                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
-                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
-                             geom_line(size = 1.5, alpha = 0.6) +
-                             geom_point(size = 2) +
-                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
-                             scale_x_log10(labels = scales::label_number()) +
-                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
-BT_det_plot
-
-#save file
-BT_det_fig_file = here("results","BT-det-plot-obl.png")
-ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
+EN_auto_fig_file = here("results","EN-auto-plot-OA.png")
+ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
 ```
@@
 #extract model from final fit
-x_BT <- best_BT_fit$fit$fit$fit
+x_EN <- best_EN_fit$fit$fit$fit
 
-#vip package doesn't support BT
+#plot most important predictors in the model
+EN_plot <- vip::vip(x_EN, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+EN_plot
+
+#save manually as "EN-plot-OA.png" -- ggsave throws errors
 
 #find predictions and intervals
-BT_resid <- best_BT_fit %>%
+EN_resid <- best_EN_fit %>%
                 broom.mixed::augment(new_data = obl_train_data) %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                dplyr::select(.pred, log_obl_amt) %>%
+                dplyr::mutate(.resid = .pred - log_obl_amt)
 
 #plot model predictions from tuned model versus actual outcomes
-#geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
+#geom_abline is a 45 degree line, along which the results should fall
+EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
-                        geom_point() +
-                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
-BT_pred_act
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "EN Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+EN_pred_act
 
 #save file
-BT_pred_act_file = here("results","BT-pred-act-obl.png")
-ggsave(filename = BT_pred_act_file, plot = BT_pred_act)
+EN_pred_act_file = here("results","EN-pred-act-OA.png")
+ggsave(filename = EN_pred_act_file, plot = EN_pred_act)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
+EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
                             geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Fit: Residuals",
+                            geom_point(color = "#3b528b") +
+                            labs(title = "EN Fit: Residuals",
                                   x = "Observation Number",
                                   y = "Residual")
-BT_resids
+EN_resids
 
 #save file
-BT_resid_file = here("results","BT-resids-obl.png")
-ggsave(filename = BT_resid_file, plot = BT_resids)
+EN_resid_file = here("results","EN-resid-OA.png")
+ggsave(filename = EN_resid_file, plot = EN_resids)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
+EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
+                          geom_point(color = "#3b528b") +
+                          labs(title = "EN Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Fitted Log(Obligated Funds) ($)",
                                 y = "Residual")
-BT_pred_resid
+EN_pred_resid
 
 #save file
-BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
-ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)
+EN_pred_resid_file = here("results","EN-pred-resid-OA.png")
+ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)
 
-#print model performance
-#print 10 best performing hyperparameter sets
-BT_res %>%
+#print the 10 best performing hyperparameter sets
+EN_res %>%
   tune::show_best(n = 10) %>%
-  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
-  dplyr::mutate(rmse = round(rmse, 3),
-                std_err = round(std_err, 4),
-                cost_complexity = scales::scientific(cost_complexity))
+  dplyr::select(rmse = mean, std_err)
 
-#print the best model performance
-BT_performance <- BT_res %>% tune::show_best(n = 1)
-print(BT_performance)
+#print the best model
+EN_performance <- EN_res %>% tune::show_best(n = 1)
+EN_performance
 
-#compare model performance to null model
-BT_RMSE <- BT_res %>%
+#compare model performance to null model (and other models)
+EN_RMSE <- EN_res %>%
               tune::show_best(n = 1) %>%
@@
                 SE = round(std_err, 4),
-                  model = "Bagged Tree") %>%
+                model = "EN") %>%
              dplyr::bind_rows(RF_RMSE)
-BT_RMSE
+EN_RMSE
+```
+The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors.
+The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors.
+The EN model has an improved RMSE, but the null model technically still has a lower RMSE.
+
+---
+
+## LASSO Model
+
+<br>
+
+### 1. Model Specification
+```{r}
+#run parallels to determine number of cores
+cores <- parallel::detectCores() - 1
+cores
+
+cl <- makeCluster(cores)
+
+registerDoParallel(cl)
+
+#define the LASSO model
+#mixture = 1 identifies the model to be a LASSO model
+lasso_mod <- linear_reg(mode = "regression",
+                           penalty = tune(),
+                           mixture = 1) %>%
+                parsnip::set_engine("glmnet")
+```
+
+---
+
+### 2. Workflow Definition
+```{r}
+#define lasso workflow
+lasso_wflow <- workflows::workflow() %>%
+                 workflows::add_model(lasso_mod) %>%
+                 workflows::add_recipe(OA_rec)
+```
+
+---
+
+### 3. Tuning Grid Specification
+```{r}
+#tuning grid specification
+lasso_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
+
+#5 lowest penalty values
+lasso_grid %>%
+  dplyr::top_n(-5)
+
+#5 highest penalty values
+lasso_grid %>%
+  dplyr::top_n(5)
+```
+
+---
+
+### 4. Tuning Using Cross-Validation and the `tune_grid()` function
+```{r}
+#tune the model with previously specified cross-validation and RMSE as target metric
+lasso_res <- lasso_wflow %>%
+              tune::tune_grid(resamples = obl_folds_spatial,
+                              grid = lasso_grid,
+                              control = control_grid(verbose = TRUE, save_pred = TRUE),
+                              metrics = metric_set(rmse))
+
+#look at top 5 EN models
+top_lasso_models <- lasso_res %>%
+                        tune::show_best("rmse", n = 5)
+top_lasso_models
+
+#default visualization
+lasso_auto_plot <- lasso_res %>% autoplot()
+lasso_auto_plot
+
+#save file
+lasso_auto_fig_file = here("results","lasso-auto-plot-OA.png")
+ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
+```
+
+---
+
+### 5. Identify Best Model
+```{r}
+#select the LASSO model with the lowest rmse
+lasso_lowest_rmse <- lasso_res %>%
+                        tune::select_best("rmse")
+
+#finalize the workflow by using the selected lasso model
+best_lasso_wflow <- lasso_wflow %>%
+                      tune::finalize_workflow(lasso_lowest_rmse)
+best_lasso_wflow
+
+#one last fit on the training data
+best_lasso_fit <- best_lasso_wflow %>%
+                    parsnip::fit(data = obl_train_data)
+```
+
+---
+
+### 6. Model evaluation
+```{r}
+#extract model from final fit
+x_lasso <- best_lasso_fit$fit$fit$fit
+
+#plot most important predictors in the model
+lasso_plot <- vip::vip(x_lasso, 
+                    num_features = 20,
+                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
+lasso_plot
+
+#save manually as "LASSO-plot-OA.png" -- ggsave throws errors
+
+#find predictions and intervals
+lasso_resid <- best_lasso_fit %>%
+                broom.mixed::augment(new_data = obl_train_data) %>%
+                dplyr::select(.pred, log_obl_amt) %>%
+                dplyr::mutate(.resid = .pred - log_obl_amt)
+
+#plot model predictions from tuned model versus actual outcomes
+#geom_abline is a 45 degree line, along which the results should fall
+lasso_pred_act <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = log_obl_amt)) +
+                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
+                        geom_point(color = "#3b528b") +
+                        scale_y_continuous(expand = c(0,0),
+                                           limits = c(-30, 25)) +
+                        scale_x_continuous(limits = c(0, 25)) +
+                        labs(title = "LASSO Fit: Actual vs. Predicted Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+lasso_pred_act
+
+#save file
+lasso_pred_act_file = here("results","lasso-pred-act-OA.png")
+ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)
+
+#plot model with residuals
+#the geom_hline plots a straight horizontal line along which the results should fall
+lasso_resids <- ggplot2::ggplot(lasso_resid, aes(x = as.numeric(row.names(lasso_resid)), y = .resid))+
+                                geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                geom_point(color = "#3b528b") +
+                                labs(title = "LASSO Fit: Residuals",
+                                      x = "Observation Number",
+                                      y = "Residual")
+lasso_resids
+
+#save file
+lasso_resid_file = here("results","lasso-resid-OA.png")
+ggsave(filename = lasso_resid_file, plot = lasso_resids)
+
+#plot model fit vs residuals
+#the geom_hline plots a straight horizontal line along which the results fall
+lasso_pred_resid <- ggplot2::ggplot(lasso_resid, aes(x = .pred, y = .resid))+
+                                    geom_hline(yintercept = 0, color = "red", lty = 2) +
+                                    geom_point(color = "#3b528b") +
+                                    labs(title = "LASSO Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                          x = "Fitted Log(Obligated Funds) ($)",
+                                          y = "Residual")
+lasso_pred_resid
+
+#save file
+lasso_pred_resid_file = here("results","lasso-pred-resid-OA.png")
+ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)
+
+#print the 10 best performing hyperparameter sets
+lasso_res %>%
+  tune::show_best(n = 10) %>%
+  dplyr::select(rmse = mean, std_err)
+
+#print the best model
+lasso_performance <- lasso_res %>% tune::show_best(n = 1)
+lasso_performance
+
+#compare model performance to null model (and other models)
+lasso_RMSE <- lasso_res %>%
+              tune::show_best(n = 1) %>%
+              dplyr::transmute(
+                rmse = round(mean, 3),
+                SE = round(std_err, 4),
+                model = "LASSO") %>%
+             dplyr::bind_rows(EN_RMSE)
+lasso_RMSE
 
 #save combined table
-model_comp_file = here("results", "ML-model-comp-obl.Rds")
-saveRDS(BT_RMSE, file = model_comp_file)
+model_comp_file = here("results", "ML-model-comp-OA.Rds")
+saveRDS(lasso_RMSE, file = model_comp_file)
 ```
+The LASSO model fit is almost identical to the Elastic Net model fit.
 
-In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.
-
---
+---
 
 ## Model Selection and Evaluation
-Using the bagged tree model
+Based on the results of the models above, the null model actually has the best fit, which warrants further investigation (Especially with the outliers).
+In the mean time, we'll fit the LASSO model as it has the best fit out of the models.
 ```{r}
 #fit model to training set but evaluate on the test data
-BT_fit_test <- best_BT_wflow %>%
+lasso_fit_test <- best_lasso_wflow %>%
                     tune::last_fit(split = obl_data_split)
@@
 
-BT_comp <- BT_RMSE %>%
-                  dplyr::filter(model == "Bagged Tree") %>%
+lasso_comp <- lasso_RMSE %>%
+                  dplyr::filter(model == "LASSO") %>%
                   dplyr::transmute(rmse, data = "train") %>%
-                  dplyr::bind_rows(BT_rmse_test) %>%
+                  dplyr::bind_rows(lasso_rmse_test) %>%
                   dplyr::slice(-3) #don't know why the third row shows up
-BT_comp
-#RMSEs are incredibly different...
+lasso_comp
+#RMSEs are different but not as much as previous iterations
+
+#save combined table
+test_comp_file = here("results", "test-train-model-comp-OA.Rds")
+saveRDS(lasso_comp, file = test_comp_file)
 
 #find predictions and intervals
-BT_resid_fit <- BT_fit_test %>%
+lasso_resid_fit <- lasso_fit_test %>%
                       broom.mixed::augment() %>%
-                  dplyr::select(.pred, OblAmt) %>%
-                  dplyr::mutate(.resid = .pred - OblAmt)
+                      dplyr::select(.pred, log_obl_amt) %>%
+                      dplyr::mutate(.resid = .pred - log_obl_amt)
 
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = OblAmt)) +
+lasso_pred_act_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = log_obl_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
-                        geom_point() +
-                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Obligated Funds",
-                              x = "Predicted Obligated Funds ($)",
-                              y = "Actual Obligated Funds ($)")
-BT_pred_act_test
+                        geom_point(color = "#3b528b") +
+                        labs(title = "LASSO Test Fit: Predicted vs. Actual Log(Obligated Funds)",
+                              x = "Predicted Log(Obligated Funds) ($)",
+                              y = "Actual Log(Obligated Funds) ($)")
+lasso_pred_act_test
 
 #save file
-BT_pred_act_test_file = here("results","BT-pred-act-test-obl.png")
-ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)
+lasso_pred_act_test_file = here("results","lasso-pred-act-test-OA.png")
+ggsave(filename = lasso_pred_act_test_file, plot = lasso_pred_act_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-BT_resids_test <- ggplot2::ggplot(BT_resid_fit, aes(x = as.numeric(row.names(BT_resid_fit)), y = .resid))+
+lasso_resids_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = as.numeric(row.names(lasso_resid_fit)), y = .resid))+
                       geom_hline(yintercept = 0, color = "red", lty = 2) +
-                      geom_point() +
-                      labs(title = "Bagged Tree Test Fit: Residuals",
+                      geom_point(color = "#3b528b") +
+                      labs(title = "LASSO Test Fit: Residuals",
                             x = "Observation Number",
                             y = "Residual")
-BT_resids_test
+lasso_resids_test
 
 #save file
-BT_resid_test_file = here("results","BT-resid-test-obl.png")
-ggsave(filename = BT_resid_test_file, plot = BT_resids_test)
+lasso_resid_test_file = here("results","lasso-resid-test-OA.png")
+ggsave(filename = lasso_resid_test_file, plot = lasso_resids_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
+lasso_pred_resid_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
-                          geom_point() +
-                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Obligated Funds",
-                                x = "Predicted Obligated Funds ($)",
-                                y = "Residual")
-BT_pred_resid_test
+                          geom_point(color = "#3b528b") +
+                          labs(title = "LASSO Test Fit: Residuals vs Fitted Log(Obligated Funds)",
+                                x = "Predicted Log(Obligated Funds) ($)",
+                                y = "Residuals")
+lasso_pred_resid_test
 
 #save file
-BT_pred_resid_test_file = here("results","BT-pred-resid-test-obl.png")
-ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)
+lasso_pred_resid_test_file = here("results","lasso-pred-resid-test-OA.png")
+ggsave(filename = lasso_pred_resid_test_file, plot = lasso_pred_resid_test)
 ```

### 2021-12-12 code/analysis_code/analysis-ML-RA-outlier.Rmd
Last ML model update

@@
 
-It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis.
+It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis where the log(ReqAmt) was less than zero.
 
@@
 #save file
-DT_auto_fig_file = here("results","DT-auto-plot.png")
+DT_auto_fig_file = here("results","DT-auto-plot-outlier.png")
 ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)
@@
 #save file
-DT_det_fig_file = here("results","DT-det-plot.png")
+DT_det_fig_file = here("results","DT-det-plot-outlier.png")
 ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)
@@
                                                      col = "grey99")
-#save manually as "DT-plot.png" -- ggsave throws an error
+#save manually as "DT-plot-outlier.png" -- ggsave throws an error
 
@@
 #save file
-DT_pred_act_file = here("results","DT-pred-act.png")
+DT_pred_act_file = here("results","DT-pred-act-outlier.png")
 ggsave(filename = DT_pred_act_file, plot = DT_pred_act)
@@
 #save file
-DT_resid_file = here("results","DT-resid.png")
+DT_resid_file = here("results","DT-resid-outlier.png")
 ggsave(filename = DT_resid_file, plot = DT_resid)
@@
 #save file
-DT_pred_resid_file = here("results","DT-pred-resid.png")
+DT_pred_resid_file = here("results","DT-pred-resid-outlier.png")
 ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)
@@
 The identified best performing tree model has some interesting results.
-It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding.
-New York and November are particularly surprising key predictors.
+It identifies percent of state counties awarded IH program, total number of agencies involved, total number of counties per state, percent of counties awarded PA program, FEMA Region VI, incident month of October and March as important predictors.
 It is interesting that incident type or declaration type are not included as significant.
-The RMSE actually increases from the Null, which suggests this is not an appropriate model.
-Moving on.
+The RMSE does decrease from the null, which suggests the outlier removal was a necessary step.
+Let's explore the other models.
 
@@
 #save file
-RF_auto_fig_file = here("results","RF-auto-plot.png")
+RF_auto_fig_file = here("results","RF-auto-plot-outlier.png")
 ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
@@
 
-#save manually as "RF-plot.png" -- ggsave throws errors
+#save manually as "RF-plot-outlier.png" -- ggsave throws errors
 
@@
 #save file
-RF_pred_act_file = here("results","RF-pred-act.png")
+RF_pred_act_file = here("results","RF-pred-act-outlier.png")
 ggsave(filename = RF_pred_act_file, plot = RF_pred_act)
@@
 #save file
-RF_resid_file = here("results","RF-resid.png")
+RF_resid_file = here("results","RF-resid-outlier.png")
 ggsave(filename = RF_resid_file, plot = RF_resids)
@@
 #save file
-RF_pred_resid_file = here("results","RF-pred-resid.png")
+RF_pred_resid_file = here("results","RF-pred-resid-outlier.png")
 ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)
@@
 In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree.
-It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and whether or not IH was awarded to any part of the state as the five most important predictors.
+It identifies percent of counties awarded IH program, total number of federal agencies involved, percent of counties awarded HM program, average FEMA cost share, and whether or not the IH program was awarded to any county in the state as the five most important predictors.
 Let's try another model to see if we can find an improvement in the RMSE.
@@
 #save file
-EN_auto_fig_file = here("results","EN-auto-plot.png")
+EN_auto_fig_file = here("results","EN-auto-plot-outlier.png")
 ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
@@
 
-#save manually as "EN-plot.png" -- ggsave throws errors
+#save manually as "EN-plot-outlier.png" -- ggsave throws errors
 
@@
 #save file
-EN_pred_act_file = here("results","EN-pred-act.png")
+EN_pred_act_file = here("results","EN-pred-act-outlier.png")
 ggsave(filename = EN_pred_act_file, plot = EN_pred_act)
@@
 #save file
-EN_resid_file = here("results","EN-resid.png")
+EN_resid_file = here("results","EN-resid-outlier.png")
 ggsave(filename = EN_resid_file, plot = EN_resids)
@@
 #save file
-EN_pred_resid_file = here("results","EN-pred-resid.png")
+EN_pred_resid_file = here("results","EN-pred-resid-outlier.png")
 ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)
@@
 ```
-The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors.
-The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors.
-The EN model has an improved RMSE, but the null model technically still has a lower RMSE.
+The Elastic Net model places great emphasis on Maine being the state receving the declaration.
+The next closest predictors include average FEMA cost share, Vermont, Montana, and Idaho.
+The EN model has an improved RMSE from the DT model but not the RF model.
+Let's try LASSO just to round out the analysis.
 
@@
 #save file
-lasso_auto_fig_file = here("results","lasso-auto-plot.png")
+lasso_auto_fig_file = here("results","lasso-auto-plot-outlier.png")
 ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
@@
 
-#save manually as "LASSO-plot.png" -- ggsave throws errors
+#save manually as "LASSO-plot-outlier.png" -- ggsave throws errors
 
@@
 #save file
-lasso_pred_act_file = here("results","lasso-pred-act.png")
+lasso_pred_act_file = here("results","lasso-pred-act-outlier.png")
 ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)
@@
 #save file
-lasso_resid_file = here("results","lasso-resid.png")
+lasso_resid_file = here("results","lasso-resid-outlier.png")
 ggsave(filename = lasso_resid_file, plot = lasso_resids)
@@
 #save file
-lasso_pred_resid_file = here("results","lasso-pred-resid.png")
+lasso_pred_resid_file = here("results","lasso-pred-resid-outlier.png")
 ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)
@@
 ## Model Selection and Evaluation
-Based on the results of the models above, the null model actually has the best fit, which warrants further investigation (Especially with the outliers).
-In the mean time, we'll fit the LASSO model as it has the best fit out of the models.
+Based on the results of the model comparison table, all of the models have a similar fit. However, RF has the lowest RMSE, so we will fit that to the test data.
 ```{r}
 #fit model to training set but evaluate on the test data
-lasso_fit_test <- best_lasso_wflow %>%
+RF_fit_test <- best_RF_wflow %>%
                     tune::last_fit(split = req_data_split)
@@
 #compare test performance against training performance
-lasso_rmse_test <- collect_metrics(lasso_fit_test) %>%
+RF_rmse_test <- collect_metrics(RF_fit_test) %>%
                    dplyr::select(rmse = .estimate) %>%
@@
 
-lasso_comp <- lasso_RMSE %>%
-                  dplyr::filter(model == "LASSO") %>%
+test_comp <- RF_RMSE %>%
+                  dplyr::filter(model == "Random Forest") %>%
                   dplyr::transmute(rmse, data = "train") %>%
-                  dplyr::bind_rows(lasso_rmse_test) %>%
+                  dplyr::bind_rows(RF_rmse_test) %>%
                   dplyr::slice(-3) #don't know why the third row shows up
-lasso_comp
-#RMSEs are different but not as much as previous iterations
+test_comp
+#RMSEs are very different (but lower in test data?)
 
 #save combined table
-test_comp_file = here("results", "test-train-model-comp.Rds")
-saveRDS(lasso_comp, file = test_comp_file)
+test_comp_file = here("results", "test-train-model-comp-outlier.Rds")
+saveRDS(test_comp, file = test_comp_file)
 
 #find predictions and intervals
-lasso_resid_fit <- lasso_fit_test %>%
+RF_resid_fit <- RF_fit_test %>%
                       broom.mixed::augment() %>%
@@
 #geom_abline draws a 45 degree line, along which the results should fall
-lasso_pred_act_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = log_req_amt)) +
+RF_pred_act_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = log_req_amt)) +
                         geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                         geom_point(color = "#3b528b") +
-                        labs(title = "LASSO Test Fit: Predicted vs. Actual Log(Requested Funds)",
+                        labs(title = "RF Test Fit: Predicted vs. Actual Log(Requested Funds)",
                               x = "Predicted Log(Requested Funds) ($)",
                               y = "Actual Log(Requested Funds) ($)")
-lasso_pred_act_test
-
-#save file
-lasso_pred_act_test_file = here("results","lasso-pred-act-test.png")
-ggsave(filename = lasso_pred_act_test_file, plot = lasso_pred_act_test)
+RF_pred_act_test
+
+#save file
+RF_pred_act_test_file = here("results","RF-pred-act-test-outlier.png")
+ggsave(filename = RF_pred_act_test_file, plot = RF_pred_act_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results should fall
-lasso_resids_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = as.numeric(row.names(lasso_resid_fit)), y = .resid))+
+RF_resids_test <- ggplot2::ggplot(RF_resid_fit, aes(x = as.numeric(row.names(RF_resid_fit)), y = .resid))+
                       geom_hline(yintercept = 0, color = "red", lty = 2) +
                       geom_point(color = "#3b528b") +
-                      labs(title = "LASSO Test Fit: Residuals",
+                      labs(title = "RF Test Fit: Residuals",
                             x = "Observation Number",
                             y = "Residual")
-lasso_resids_test
-
-#save file
-lasso_resid_test_file = here("results","lasso-resid-test.png")
-ggsave(filename = lasso_resid_test_file, plot = lasso_resids_test)
+RF_resids_test
+
+#save file
+RF_resid_test_file = here("results","RF-resid-test-outlier.png")
+ggsave(filename = RF_resid_test_file, plot = RF_resids_test)
 
@@
 #the geom_hline plots a straight horizontal line along which the results fall
-lasso_pred_resid_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = .resid))+
+RF_pred_resid_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = .resid))+
                           geom_hline(yintercept = 0, color = "red", lty = 2) +
                           geom_point(color = "#3b528b") +
-                          labs(title = "LASSO Test Fit: Residuals vs Fitted Log(Requested Funds)",
+                          labs(title = "RF Test Fit: Residuals vs Fitted Log(Requested Funds)",
                                 x = "Predicted Log(Requested Funds) ($)",
                                 y = "Residuals")
-lasso_pred_resid_test
-
-#save file
-lasso_pred_resid_test_file = here("results","lasso-pred-resid-test.png")
-ggsave(filename = lasso_pred_resid_test_file, plot = lasso_pred_resid_test)
-```
+RF_pred_resid_test
+
+#save file
+RF_pred_resid_test_file = here("results","RF-pred-resid-test-outlier.png")
+ggsave(filename = RF_pred_resid_test_file, plot = RF_pred_resid_test)
+```

### 2021-12-13 code/analysis_code/analysis-eda.Rmd
Analysis Complete

@@
 * table1: for making summary tables
+* maps: for making maps
 
@@
 library(table1) #for making summary tables
+library(maps) #for making maps
 
@@
 
+Before moving through the entire EDA, I want to create a secondary dataset with the outliers removed.
+This is the data actually used in the ML analysis, so this is the dataset that should be used for the EDA.
+```{r outliers}
+EDAdata_outliers <- EDAdata %>%
+                      dplyr::filter(log(ReqAmt) > 0)
+```
+
 Let's dive a little deeper into each variable.
@@
 #summary statistics
-base::summary(EDAdata$ReqAmt)
+base::summary(EDAdata_outliers$ReqAmt)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = ReqAmt)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = ReqAmt)) +
   geom_density()
@@
 #try a log transformation out of curiosity 
-ggplot2::ggplot(data = EDAdata, aes(x = log(ReqAmt))) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = log(ReqAmt))) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = ReqAmt)) +
   geom_boxplot()
@@
 #crop to $1M to get a better idea
-ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
-  geom_boxplot() +
-  ylim(0, 1000000)
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = ReqAmt)) +
+  geom_boxplot()
 
 #log transformed boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = log(ReqAmt))) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = log(ReqAmt))) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$ReqAmt)
+car::qqPlot(EDAdata_outliers$ReqAmt)
 ```
@@
 #summary statistics
-base::summary(EDAdata$OblAmt)
+base::summary(EDAdata_outliers$OblAmt)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = OblAmt)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = OblAmt)) +
   geom_density()
@@
 #try a log transformation out of curiosity 
-ggplot2::ggplot(data = EDAdata, aes(x = log(OblAmt))) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = log(OblAmt))) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = OblAmt)) +
   geom_boxplot()
@@
 #crop to $1M to get a better idea
-ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
-  geom_boxplot() +
-  ylim(0, 1000000)
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = OblAmt)) +
+  geom_boxplot()
 
 #log transformed boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = log(OblAmt))) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = log(OblAmt))) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$OblAmt)
+car::qqPlot(EDAdata_outliers$OblAmt)
 ```
@@
 ```{r}
-funds_comp <- EDAdata %>%
+#including all points
+funds_comp <- EDAdata_outliers %>%
                 ggplot2::ggplot(aes(x = log(ReqAmt), y = log(OblAmt))) +
-                         geom_point()
+                         geom_smooth(color = "#440154",
+                                     size = 2) +
+                         geom_point(color = "#1f9e89",
+                                    alpha = 0.5) +
+                         scale_y_continuous(expand = c(0,0),
+                                            limits = c(0, NA)) +
+                         scale_x_continuous(expand = c(0,0),
+                                            limits = c(0, NA)) +
+                         labs(x = "log(Requested Funding)",
+                              y = "log(Obligated Funding)")
 
 funds_comp
+
+#save file
+funds_comp_fig_file = here("results","funds_comp.png")
+ggsave(filename = funds_comp_fig_file, plot = funds_comp)
+```
+
+I also want a density plot of both log-transformed outcomes on one plot, so that can go here as well.
+```{r}
+#convert to long format in order to be able to plot both outcomes with same variable
+data_long <- gather(EDAdata_outliers, FundType, funds, ReqAmt:OblAmt, factor_key=TRUE)
+data_long
+
+#change labels
+levels(data_long$FundType) <- c("Requested Amount", "Obligated Amount")
+
+#now usual ggplot2 density plot
+funds_density <- ggplot2::ggplot(data = data_long, aes(x = log(funds), color = FundType, fill = FundType)) +
+                          geom_density(alpha = 0.4) +
+                          facet_wrap(~ FundType) +
+                          scale_fill_manual(values = c("#6ece58", "#26828e")) +
+                          scale_color_manual(values = c("#6ece58", "#26828e")) +
+                          labs(x = "Log(Funding)",
+                               y = "Density") +
+                          theme(legend.position="none")
+
+funds_density
+
+#save file
+funds_density_fig_file = here("results","funds_density.png")
+ggsave(filename = funds_density_fig_file, plot = funds_density)
+
 ```
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$state, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$state, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_state <- EDAdata %>%
+obl_fig_state <- EDAdata_outliers %>%
                     dplyr::add_count(state) %>%
-                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
+                    dplyr::filter(dplyr::dense_rank(-n) < 5) %>%
                     ggplot2::ggplot(aes(x = state, y = OblAmt)) +
-                      geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_boxplot(color = "#440154") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
                       labs(x = "State or Territory",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated to five most disaster-prone states")
+                           y = "Obligated FEMA Funds")
 obl_fig_state
 
-#save file
-obl_state_fig_file = here("results","obl-state.png")
-ggsave(filename = obl_state_fig_file, plot = obl_fig_state)
-
 #repeat for requested funds
-req_fig_state <- EDAdata %>%
+req_fig_state <- EDAdata_outliers %>%
                     dplyr::add_count(state) %>%
@@
                     ggplot2::ggplot(aes(x = state, y = ReqAmt)) +
-                      geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_boxplot(color = "#440154") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) +
-                      labs(x = "State or Territory",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested by five most disaster-prone states")
+                      labs(x = " " ,
+                           y = "Requested FEMA Funds")
 req_fig_state
 
-#save file
-req_state_fig_file = here("results","req-state.png")
-ggsave(filename = req_state_fig_file, plot = req_fig_state)
+#combine into one plot
+plot_states <- cowplot::plot_grid(req_fig_state, 
+                                     obl_fig_state,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_states
+
+title_state <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds For Five Most Disaster Prone States",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_states_funds <- cowplot::plot_grid(
+                              title_state, plot_states,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_states_funds
+
+#save file
+states_fig_file = here("results","states-funds.png")
+ggsave(filename = states_fig_file, plot = plot_states_funds)
 
 #create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
-decs_state_5 <- EDAdata %>%
+decs_state_5 <- EDAdata_outliers %>%
                     dplyr::add_count(state) %>%
@@
 #summary statistics since it's continuous
-base::summary(EDAdata$IncidentYear)
+base::summary(EDAdata_outliers$IncidentYear)
 
@@
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = IncidentYear)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = IncidentYear)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = IncidentYear)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = IncidentYear)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$IncidentYear)
+car::qqPlot(EDAdata_outliers$IncidentYear)
 
 #histogram distribution
-year_hist <- EDAdata %>% 
+year_hist <- EDAdata_outliers %>% 
                ggplot2::ggplot(aes(x=IncidentYear)) + 
-                        geom_bar(fill="steelblue4") + 
+                        geom_bar(fill="#482878") + 
                         scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
@@
 #look at total cost per incident year
-req_fig_year <- EDAdata %>%
+req_fig_year <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IncidentYear)) +
@@
                                stat = "identity",
-                               fill = "tomato") +
+                               fill = "#6ece58") +
                       scale_y_continuous(expand = c(0, 0), 
@@
 #repeat for obligated funds
-obl_fig_year <- EDAdata %>%
+obl_fig_year <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IncidentYear)) +
@@
                                stat = "identity",
-                               fill = "turquoise4") +
+                               fill = "#26828e") +
                       scale_y_continuous(expand = c(0, 0),
@@
                     draw_label(
-                      "Requested vs obligated FEMA funds per year",
+                      "Requested vs Obligated FEMA Funds Per Year",
                       fontface = 'bold',
@@
 #create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
-decs_year_3 <- EDAdata %>%
+decs_year_3 <- EDAdata_outliers %>%
                     dplyr::add_count(IncidentYear) %>%
@@
 ```{r}
-#summary statistics since it's continuous
-base::summary(EDAdata$IncidentMonth)
+#frequency table since it's categorical
+#hide NAs because there are no NAs
+summarytools::freq(EDAdata_outliers$IncidentMonth, report.nas = FALSE)
 
 #examine graphical relationship with outcomes
-
-#density plot
-ggplot2::ggplot(data = EDAdata, aes(x = IncidentMonth)) +
-  geom_density()
-
-#normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = IncidentMonth)) +
-  geom_boxplot()
-
-##QQ plot
-car::qqPlot(EDAdata$IncidentMonth)
-
-#histogram distribution
-month_hist <- EDAdata %>% 
-               ggplot2::ggplot(aes(x=IncidentMonth)) + 
-                        geom_bar(fill="darkseagreen4") + 
-                        scale_x_continuous(breaks = seq(1, 12, by = 1),
-                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
-                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
-                        scale_y_continuous(expand = c(0, 0), limits = c(0, 110)) +
-                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
-                        labs(x = "Declaration Month",
-                             y = "Number of Declarations",
-                             title = "Disaster Declarations per Month (2012 - 2021)")
-month_hist
-
-#save file
-month_fig_file = here("results","incidentmonth.png")
-ggsave(filename = month_fig_file, plot = month_hist)
-
-#examine graphical relationship with outcomes of interest
-#look at total cost per incident month
-req_fig_month <- EDAdata %>%
-                    ggplot2::ggplot(aes(x = IncidentMonth)) +
-                      geom_bar(aes(y = ReqAmt),
-                               stat = "identity",
-                               fill = "tomato") +
-                      scale_y_continuous(expand = c(0, 0), 
-                                         labels = scales::dollar_format()) +
-                      scale_x_continuous(breaks = seq(1, 12, by = 1),
-                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
-                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
-                      labs(x = "Declaration Month",
+#include a jitter function to have a better idea of number of measurements and distribution
+#filter top 5 states
+#start with obligated amount
+obl_fig_month <- EDAdata_outliers %>%
+                    dplyr::add_count(state) %>%
+                    dplyr::filter(dplyr::dense_rank(-n) < 5) %>%
+                    ggplot2::ggplot(aes(x = IncidentMonth, y = OblAmt)) +
+                      geom_boxplot(color = "#440154") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      labs(x = "Incident Month",
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
+obl_fig_month
+
+#repeat for requested funds
+req_fig_month <- EDAdata_outliers %>%
+                    dplyr::add_count(state) %>%
+                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
+                    ggplot2::ggplot(aes(x = IncidentMonth, y = ReqAmt)) +
+                      geom_boxplot(color = "#440154") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
+                      scale_y_continuous(labels = scales::dollar_format()) +
+                      labs(x = " " ,
                            y = "Funds ($)",
@@
 
-#repeat for obligated funds
-obl_fig_month <- EDAdata %>%
-                    ggplot2::ggplot(aes(x = IncidentMonth)) +
-                      geom_bar(aes(y = OblAmt),
-                               stat = "identity",
-                               fill = "turquoise4") +
-                      scale_y_continuous(expand = c(0, 0),
-                                         labels = scales::dollar_format()) +
-                      scale_x_continuous(breaks = seq(1, 12, by = 1),
-                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
-                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
-                      labs(x = "Declaration Month",
-                           y = "Funds ($)",
-                           subtitle = "Obligated")
-obl_fig_month
-
 #combine into one plot
-plot_col1 <- cowplot::plot_grid(req_fig_month, 
+plot_month <- cowplot::plot_grid(req_fig_month, 
                                      obl_fig_month,
@@
                                      rel_heights = c(0.5,0.5))
-plot_col1
-
-title2 <- cowplot::ggdraw() + 
+plot_month
+
+title_month <- cowplot::ggdraw() + 
                     draw_label(
-                      "Requested vs obligated FEMA funds by declaration month",
+                      "Requested vs Obligated FEMA Funds By Month",
                       fontface = 'bold',
@@
                       hjust = 0) +
-                    theme( plot.margin = margin(0, 0, 0, 7))
-
-funds_fig_month <- cowplot::plot_grid(
-                              title2, plot_col1,
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_month_funds <- cowplot::plot_grid(
+                              title_month, plot_month,
                               ncol = 1,
@@
                               rel_heights = c(0.1, 1))
-funds_fig_month
-
-#save file
-funds_month_fig_file = here("results","funds-month.png")
-ggsave(filename = funds_month_fig_file, plot = funds_fig_month)
-
-#create a table where columns represent requested and obligated funds and rows are 3 most distaster prone-months
-decs_month_3 <- EDAdata %>%
+plot_month_funds
+
+#save file
+month_fig_file = here("results","month-funds.png")
+ggsave(filename = month_fig_file, plot = plot_month_funds)
+
+#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
+decs_month_5 <- EDAdata_outliers %>%
                     dplyr::add_count(IncidentMonth) %>%
-                    dplyr::filter(dense_rank(-n) < 4) %>%
+                    dplyr::filter(dense_rank(-n) < 6) %>%
                     gtsummary::select(IncidentMonth, ReqAmt, OblAmt) %>%
@@
                       by = IncidentMonth,
-                      label = list(IncidentMonth ~ "Month",
+                      label = list(IncidentMonth ~ "Incident Month",
                                    ReqAmt ~ "Requested Funds",
@@
                     gtsummary::modify_header(label ~ "**Variable**") %>% 
-                    gtsummary::modify_header(update = list(
-                                               stat_1 ~ "**Jan**",
-                                               stat_2 ~ "**Sept**",
-                                               stat_3 ~ "**Oct**"
-                                             )) %>% 
-                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Month**") %>%
-                    gtsummary::modify_caption("**Table 3. FEMA Funding for the 3 Costliest Months (2012 - 2021)**") %>%
+                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**Incident Month**") %>%
+                    gtsummary::modify_caption("**Table 4. FEMA Funding by Month 2012 - 2021**") %>%
                     gtsummary::bold_labels()
-decs_month_3
-
-#save file
-decs_month_file = here("results", "decs-month-3.Rds")
-saveRDS(decs_month_3, file = decs_month_file)
+decs_month_5
+
+#save file
+decs_month_file = here("results", "decs-month-5.Rds")
+saveRDS(decs_month_5, file = decs_month_file)
 ```
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$declarationType, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$declarationType, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_DT <- EDAdata %>%
+obl_fig_DT <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = declarationType, y = OblAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
@@
                       labs(x = "Declaration Type",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated for declaration types")
+                           y = "Funds ($)",
+                           subtitle = "Obligated")
 obl_fig_DT
 
-#save file
-obl_DT_fig_file = here("results","obl-DT.png")
-ggsave(filename = obl_DT_fig_file, plot = obl_fig_DT)
-
 #repeat for requested funds
-req_fig_DT <- EDAdata %>%
+req_fig_DT <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = declarationType, y = ReqAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) +
                       scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
-                      labs(x = "Declaration Type",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested for declaration types")
+                      labs(x = " ",
+                           y = "Funds ($)",
+                           subtitle = "Requested")
 req_fig_DT
 
-#save file
-req_DT_fig_file = here("results","req-DT.png")
-ggsave(filename = req_DT_fig_file, plot = req_fig_DT)
+#combine into one plot
+plot_DT <- cowplot::plot_grid(req_fig_DT, 
+                                     obl_fig_DT,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_DT
+
+title_DT <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds By Declaration Type",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_DT_funds <- cowplot::plot_grid(
+                              title_DT, plot_DT,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_DT_funds
+
+#save file
+DT_fig_file = here("results","DT-funds.png")
+ggsave(filename = DT_fig_file, plot = plot_DT_funds)
 
 #create a table where rows represent requested and obligated funds and columns are declaration type
-decs_dtype <- EDAdata %>%
+decs_dtype <- EDAdata_outliers %>%
                     gtsummary::select(declarationType, ReqAmt, OblAmt) %>%
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$incidentType, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$incidentType, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_IT <- EDAdata %>%
+obl_fig_IT <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = incidentType, y = OblAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
                       labs(x = "Incident Type",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated for incident types") +
+                           y = "Funds ($)",
+                           subtitle = "Obligated") +
                       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
@@
 
-#save file
-obl_IT_fig_file = here("results","obl-IT.png")
-ggsave(filename = obl_IT_fig_file, plot = obl_fig_IT)
-
 #repeat for requested funds
-req_fig_IT <- EDAdata %>%
+req_fig_IT <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = incidentType, y = ReqAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) + 
-                      labs(x = "Incident Type",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested for declaration types") +
+                      labs(x = " ",
+                           y = "Funds ($)",
+                           subtitle = "Requested") +
                       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
@@
 
-#save file
-req_IT_fig_file = here("results","req-IT.png")
-ggsave(filename = req_IT_fig_file, plot = req_fig_IT)
+#combine into one plot
+plot_IT <- cowplot::plot_grid(req_fig_IT, 
+                                     obl_fig_IT,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_IT
+
+title_IT <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds By Incident Type",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_IT_funds <- cowplot::plot_grid(
+                              title_IT, plot_IT,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_IT_funds
+
+#save file
+IT_fig_file = here("results","IT-funds.png")
+ggsave(filename = IT_fig_file, plot = plot_IT_funds)
 
@@
 #convert class to numeric
-EDAdata$IncidentDuration <- as.numeric(EDAdata$IncidentDuration, units="days")
+EDAdata_outliers$IncidentDuration <- as.numeric(EDAdata_outliers$IncidentDuration, units="days")
 
 #summary statistics since it's continuous
-base::summary(EDAdata$IncidentDuration)
+base::summary(EDAdata_outliers$IncidentDuration)
 #NAs are ongoing events
@@
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = IncidentDuration)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = IncidentDuration)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = IncidentDuration)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = IncidentDuration)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$IncidentDuration)
+car::qqPlot(EDAdata_outliers$IncidentDuration)
 
@@
 #look at total cost per incident month
-req_fig_dur <- EDAdata %>%
+req_fig_dur <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IncidentDuration)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
 #repeat for obligated funds
-obl_fig_dur <- EDAdata %>%
+obl_fig_dur <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IncidentDuration)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
                     draw_label(
-                      "Requested vs obligated FEMA funds by Incident Duration (Under $1B)",
+                      "Requested vs Obligated FEMA Funds by Incident Duration (Under $1B)",
                       fontface = 'bold',
@@
 #convert class to numeric
-EDAdata$Population <- as.numeric(gsub(",", "", EDAdata$Population))
+EDAdata_outliers$Population <- as.numeric(gsub(",", "", EDAdata_outliers$Population))
 
 #summary statistics since it's continuous
-base::summary(EDAdata$Population)
+base::summary(EDAdata_outliers$Population)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = Population)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = Population)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = Population)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = Population)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$Population)
+car::qqPlot(EDAdata_outliers$Population)
 
@@
 #look at total cost per pop
-req_fig_pop <- EDAdata %>%
+req_fig_pop <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = Population)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
 #repeat for obligated funds
-obl_fig_pop <- EDAdata %>%
+obl_fig_pop <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = Population)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
 #summary statistics since it's continuous
-base::summary(EDAdata$TotalAgencies)
+base::summary(EDAdata_outliers$TotalAgencies)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = TotalAgencies)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = TotalAgencies)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = TotalAgencies)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = TotalAgencies)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$TotalAgencies)
+car::qqPlot(EDAdata_outliers$TotalAgencies)
 
@@
 #look at total cost per incident month
-req_fig_ag <- EDAdata %>%
+req_fig_ag <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = TotalAgencies)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
 #repeat for obligated funds
-obl_fig_ag <- EDAdata %>%
+obl_fig_ag <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = TotalAgencies)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
 #summary statistics since it's continuous
-base::summary(EDAdata$Counties)
+base::summary(EDAdata_outliers$Counties)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = Counties)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = Counties)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = Counties)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = Counties)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$Counties)
+car::qqPlot(EDAdata_outliers$Counties)
 
@@
 #look at total cost per incident month
-req_fig_count <- EDAdata %>%
+req_fig_count <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = Counties)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
 #repeat for obligated funds
-obl_fig_count <- EDAdata %>%
+obl_fig_count <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = Counties)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
                               ncol = 1,
-                              # rel_heights values control vertical title margins
                               rel_heights = c(0.1, 1))
@@
 #summary statistics since it's continuous
-base::summary(EDAdata$FEMACSAvg)
+base::summary(EDAdata_outliers$FEMACSAvg)
 
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = FEMACSAvg)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = FEMACSAvg)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = FEMACSAvg)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = FEMACSAvg)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$FEMACSAvg)
+car::qqPlot(EDAdata_outliers$FEMACSAvg)
 
@@
 #look at total cost per incident month
-req_fig_cs <- EDAdata %>%
+req_fig_cs <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = FEMACSAvg)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
 #repeat for obligated funds
-obl_fig_cs <- EDAdata %>%
+obl_fig_cs <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = FEMACSAvg)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
 #convert to factor
-EDAdata$IH <- as.factor(EDAdata$IH)
+EDAdata_outliers$IH <- as.factor(EDAdata_outliers$IH)
 
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$IH, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$IH, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_IH <- EDAdata %>%
+obl_fig_IH <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IH, y = OblAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
@@
                       labs(x = "IH Program",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated when IH program is awarded") +
-                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+                           y = "Funds ($)",
+                           subtitle = "Obligated") 
 obl_fig_IH
 
-#save file
-obl_IH_fig_file = here("results","obl-IH.png")
-ggsave(filename = obl_IH_fig_file, plot = obl_fig_IH)
-
 #repeat for requested funds
-req_fig_IH <- EDAdata %>%
+req_fig_IH <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = IH, y = ReqAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) + 
                       scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
-                      labs(x = "IH Program",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested when IH program is awarded") +
-                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+                      labs(x = " ",
+                           y = "Funds ($)",
+                           subtitle = "Requested") 
 req_fig_IH
 
-#save file
-req_IH_fig_file = here("results","req-IH.png")
-ggsave(filename = req_IH_fig_file, plot = req_fig_IH)
+#combine into one plot
+plot_IH <- cowplot::plot_grid(req_fig_IH, 
+                                     obl_fig_IH,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_IH
+
+title_IH <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds By IH Program Awardance",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_IH_funds <- cowplot::plot_grid(
+                              title_IH, plot_IH,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_IH_funds
+
+#save file
+IH_fig_file = here("results","IH-funds.png")
+ggsave(filename = IH_fig_file, plot = plot_IH_funds)
 
 #create a table where rows represent requested and obligated funds and columns are declaration type
-decs_ih <- EDAdata %>%
+decs_ih <- EDAdata_outliers %>%
                     gtsummary::select(IH, ReqAmt, OblAmt) %>%
@@
 #convert to factor
-EDAdata$PA <- as.factor(EDAdata$PA)
+EDAdata_outliers$PA <- as.factor(EDAdata_outliers$PA)
 
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$PA, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$PA, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_PA <- EDAdata %>%
+obl_fig_PA <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = PA, y = OblAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
@@
                       labs(x = "PA Program",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated when PA program is awarded") +
-                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+                           y = "Funds ($)",
+                           subtitle = "FEMA funds obligated when PA program is awarded") 
 obl_fig_PA
 
-#save file
-obl_PA_fig_file = here("results","obl-PA.png")
-ggsave(filename = obl_PA_fig_file, plot = obl_fig_PA)
-
 #repeat for requested funds
-req_fig_PA <- EDAdata %>%
+req_fig_PA <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = PA, y = ReqAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) + 
                       scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
-                      labs(x = "PA Program",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested when PA program is awarded") +
+                      labs(x = " ",
+                           y = "Funds ($)",
+                           title = "Requested") +
                       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
@@
 
-#save file
-req_PA_fig_file = here("results","req-PA.png")
-ggsave(filename = req_PA_fig_file, plot = req_fig_PA)
+#combine into one plot
+plot_PA <- cowplot::plot_grid(req_fig_PA, 
+                                     obl_fig_PA,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_PA
+
+title_PA <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds By PA Program Awardance",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_PA_funds <- cowplot::plot_grid(
+                              title_PA, plot_PA,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_PA_funds
+
+#save file
+PA_fig_file = here("results","PA-funds.png")
+ggsave(filename = PA_fig_file, plot = plot_PA_funds)
 
 #create a table where rows represent requested and obligated funds and columns are declaration type
-decs_pa <- EDAdata %>%
+decs_pa <- EDAdata_outliers %>%
                     gtsummary::select(PA, ReqAmt, OblAmt) %>%
@@
 #convert to factor
-EDAdata$HM <- as.factor(EDAdata$HM)
+EDAdata_outliers$HM <- as.factor(EDAdata_outliers$HM)
 
@@
 #hide NAs because there are no NAs
-summarytools::freq(EDAdata$HM, report.nas = FALSE)
+summarytools::freq(EDAdata_outliers$HM, report.nas = FALSE)
 
@@
 #start with obligated amount
-obl_fig_HM <- EDAdata %>%
+obl_fig_HM <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = HM, y = OblAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                       scale_y_continuous(labels = scales::dollar_format()) +
@@
                       labs(x = "HM Program",
-                           y = "Obligated FEMA Funds",
-                           title = "FEMA funds obligated when HM program is awarded") +
-                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+                           y = "Funds ($)",
+                           subtitle = "Obligated") 
 obl_fig_HM
 
-#save file
-obl_HM_fig_file = here("results","obl-HM.png")
-ggsave(filename = obl_HM_fig_file, plot = obl_fig_HM)
-
 #repeat for requested funds
-req_fig_HM <- EDAdata %>%
+req_fig_HM <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = HM, y = ReqAmt)) +
                       geom_boxplot() +
-                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
+                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                       scale_y_continuous(labels = scales::dollar_format()) + 
                       scale_x_discrete(labels = c("Not Awarded", "Awarded")) +
-                      labs(x = "HM Program",
-                           y = "Requested FEMA Funds",
-                           title = "FEMA funds requested when HM program is awarded") +
-                      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+                      labs(x = " ",
+                           y = "Funds ($)",
+                           subtitle = "Requested") 
 req_fig_HM
 
-#save file
-req_HM_fig_file = here("results","req-HM.png")
-ggsave(filename = req_HM_fig_file, plot = req_fig_HM)
+#combine into one plot
+plot_HM <- cowplot::plot_grid(req_fig_HM, 
+                                     obl_fig_HM,
+                                     ncol = 1,
+                                     align = "v",
+                                     rel_heights = c(0.5,0.5))
+plot_HM
+
+title_HM <- cowplot::ggdraw() + 
+                    draw_label(
+                      "Requested vs Obligated FEMA Funds By HM Program Awardance",
+                      fontface = 'bold',
+                      x = 0,
+                      hjust = 0) +
+                    theme(plot.margin = margin(0, 0, 0, 7))
+
+plot_HM_funds <- cowplot::plot_grid(
+                              title_HM, plot_HM,
+                              ncol = 1,
+                              # rel_heights values control vertical title margins
+                              rel_heights = c(0.1, 1))
+plot_HM_funds
+
+#save file
+PA_fig_file = here("results","PA-funds.png")
+ggsave(filename = PA_fig_file, plot = plot_PA_funds)
 
@@
 #summary statistics since it's continuous
-base::summary(EDAdata$ResponseDays)
+base::summary(EDAdata_outliers$ResponseDays)
 #NAs are ongoing events
@@
 #density plot
-ggplot2::ggplot(data = EDAdata, aes(x = ResponseDays)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(x = ResponseDays)) +
   geom_density()
@@
 #normal boxplot
-ggplot2::ggplot(data = EDAdata, aes(y = ResponseDays)) +
+ggplot2::ggplot(data = EDAdata_outliers, aes(y = ResponseDays)) +
   geom_boxplot()
@@
 ##QQ plot
-car::qqPlot(EDAdata$ResponseDays)
+car::qqPlot(EDAdata_outliers$ResponseDays)
 
 #examine graphical relationship with outcomes of interest
-req_fig_resp <- EDAdata %>%
+req_fig_resp <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = ResponseDays)) +
                       geom_point(aes(y = ReqAmt),
-                                 color = "tomato",
+                                 color = "#6ece58",
                                  show.legend = FALSE) +
@@
                                          labels = scales::dollar_format()) +
-                      labs(x = "Response Duration (Days)",
+                      labs(x = " ",
                            y = "Funds ($)",
@@
 #repeat for obligated funds
-obl_fig_resp <- EDAdata %>%
+obl_fig_resp <- EDAdata_outliers %>%
                     ggplot2::ggplot(aes(x = ResponseDays)) +
                       geom_point(aes(y = OblAmt),
-                                 color = "turquoise4",
+                                 color = "#26828e",
                                  show.legend = FALSE) +
@@
 * ResponseDays
-* IH
-* PA
-* HM
+* IH Percent
+* PA Percent
+* HM Percent
 * Population
@@
 ```{r}
-#label factors
-EDAdata$IH <- factor(EDAdata$IH, 
-                     levels = c(0, 1),
-                     labels = c("Not Awarded", "Awarded"))
-EDAdata$PA <- factor(EDAdata$PA, 
-                     levels = c(0, 1),
-                     labels = c("Not Awarded", "Awarded"))
-EDAdata$HM <- factor(EDAdata$HM, 
-                     levels = c(0, 1),
-                     labels = c("Not Awarded", "Awarded"))
-
 #specify units
@@
 #create table
-table1 <- table1::table1(~ ReqAmt + OblAmt + state + IncidentYear + IncidentMonth + declarationType + incidentType +
-                           IncidentDuration + ResponseDays + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
-                          data = EDAdata)
+table1 <- table1::table1(~ ReqAmt + OblAmt + FEMACSAvg + TotalAgencies + IH_pct + PA_pct + HM_pct + ResponseDays +
+                           IncidentDuration + IncidentMonth + IncidentYear + incidentType + declarationType +
+                           FEMARegion + Counties + Population, data = EDAdata)
+table1
 
@@
 
-## Save Data
-Since we've made some changes to the format of the processed data, let's save it for the analysis.
-```{r}
-#location to save processed file
-analysis_data_location <- here::here("data","processed_data","analysisdata.rds")
-
-#save as an RDS
-saveRDS(EDAdata, file = analysis_data_location)
-```
+# Map Figure
+Make the following four maps:
+
+1. FEMA Regions
+2. Density by total number of declarations
+3. Density by total requested funds
+4. Density by obligated funds
+
+```{r}
+#pull geographic boundaries from maps package
+MainStates <- map_data("state")
+
+#fix capitalization issue in map data
+MainStates$name <- str_to_title(MainStates$region)
+
+#create summary df by state for EDAdata_analysis
+#count number of declarations per state
+EDA_sum_1 <- EDAdata_outliers %>%
+                dplyr::group_by(name,
+                                state,
+                                latitude,
+                                longitude,
+                                Population,
+                                Counties,
+                                FEMARegion) %>%
+                dplyr::count(state)
+
+#find total of requested and obligated funds
+EDA_sum_2 <- EDAdata_outliers %>%
+                dplyr::group_by(name) %>%
+                dplyr::summarize(RequestedAmt = sum(ReqAmt),
+                                 ObligatedAmt = sum(OblAmt))
+
+#combine the two
+EDA_sums <- inner_join(EDA_sum_1, EDA_sum_2, by = "name")
+
+#merge MainStates and EDAdata_outliers
+MergedStates <- inner_join(MainStates, EDA_sums, by = "name")
+
+#rename count variable
+MapsData <- MergedStates %>%
+                  dplyr::mutate(NumDecs = n) %>%
+                  dplyr::select(-n)
+
+#create map with FEMA Regions
+map1 <- ggplot(MapsData, aes(x = long, 
+                             y = lat, 
+                             group = group)) +
+      geom_polygon(aes(fill = FEMARegion), colour = alpha("white", 1 / 2), size = 0.2) +
+      geom_polygon(data = MergedStates, colour = "white", fill = NA) +
+      coord_fixed() +
+      scale_fill_manual(values = c("#fde725", "#b5de2b", "#6ece58", "#35b779", "#1f9e89", "#26828e", "#31688e",
+                                   "#3e4989", "#482878", "#440154")) +
+      theme(axis.line = element_blank(), axis.text = element_blank(),
+            axis.ticks = element_blank(), axis.title = element_blank())
+map1
+
+#save file
+map1_fig_file = here("results","map1.png")
+ggsave(filename = map1_fig_file, plot = map1)
+
+#create map with number of declarations
+map2 <- ggplot(MapsData, aes(x = long, 
+                             y = lat, 
+                             group = group)) +
+      geom_polygon(aes(fill = NumDecs), colour = alpha("white", 1 / 2), size = 0.2) +
+      geom_polygon(data = MergedStates, colour = "white", fill = NA) +
+      coord_fixed() +
+      scale_fill_viridis_c(option = "D",
+                           name = "Number of Declarations") +
+      theme(axis.line = element_blank(), axis.text = element_blank(),
+            axis.ticks = element_blank(), axis.title = element_blank())
+map2
+
+#save file
+map2_fig_file = here("results","map2.png")
+ggsave(filename = map2_fig_file, plot = map2)
+
+#create map with requested amount
+map3 <- ggplot(MapsData, aes(x = long, 
+                             y = lat, 
+                             group = group)) +
+      geom_polygon(aes(fill = RequestedAmt), colour = alpha("white", 1 / 2), size = 0.2) +
+      geom_polygon(data = MergedStates, colour = "white", fill = NA) +
+      coord_fixed() +
+      scale_fill_viridis_c(option = "D",
+                           labels=scales::dollar_format(),
+                           name = "Requested Amount") +
+      theme(axis.line = element_blank(), axis.text = element_blank(),
+            axis.ticks = element_blank(), axis.title = element_blank())
+map3
+
+#save file
+map3_fig_file = here("results","map3.png")
+ggsave(filename = map3_fig_file, plot = map3)
+
+#create map with obligated amount
+map4 <- ggplot(MapsData, aes(x = long, 
+                             y = lat, 
+                             group = group)) +
+      geom_polygon(aes(fill = ObligatedAmt), colour = alpha("white", 1 / 2), size = 0.2) +
+      geom_polygon(data = MergedStates, colour = "white", fill = NA) +
+      coord_fixed() +
+      scale_fill_viridis_c(option = "D",
+                           labels=scales::dollar_format(),
+                           name = "Obligated Amount") +
+      theme(axis.line = element_blank(), axis.text = element_blank(),
+            axis.ticks = element_blank(), axis.title = element_blank())
+map4
+
+#save file
+map4_fig_file = here("results","map4.png")
+ggsave(filename = map4_fig_file, plot = map4)
+```

### 2021-12-13 code/analysis_code/analysis-linearmodels.Rmd
Analysis Complete

@@
 * Whichever outcome is not currently fitted to the model will *not* be considered a predictor of interest
+* The outcomes of interest are log-transformed
 
@@
 ```{r load data}
-#define file path
-data_location <- here::here("data", "processed_data", "analysisdata.rds")
+#define file path to req data
+data_location_req <- here::here("data", "processed_data", "analysisdata-RA.rds")
 
-#load data
-analysis_data <- readRDS(data_location)
+#load req data
+analysis_data_req <- readRDS(data_location_req)
+
+#define file path to obl data
+data_location_obl <- here::here("data", "processed_data", "analysisdata-OA.rds")
+
+#load obl data
+analysis_data_obl <- readRDS(data_location_obl)
 ```
@@
 ```{r data overview}
-skimr::skim(analysis_data)
+skimr::skim(analysis_data_req)
+
+skimr::skim(analysis_data_obl)
 ```
@@
 ```{r data subsets}
-#obligated amount
-obl_data <- analysis_data %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)
+#log transform ReqAmt
+analysis_data_req$logReqAmt <- log(analysis_data_req$ReqAmt)
 
-#requested amount
-req_data <- analysis_data %>%
-              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
-                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
+#drop unnecessary variables
+req_data <- analysis_data_req %>%
+              dplyr::select(-c(ReqAmt, disasterNumber, declarationTitle))
+
+#log transform OblAmt
+analysis_data_obl$logOblAmt <- log(analysis_data_obl$OblAmt)
+
+#drop unnecessary variables
+obl_data <- analysis_data_obl %>%
+              dplyr::select(-c(OblAmt, disasterNumber, declarationTitle))
 ```
@@
 #define recipe
-req_SLR_rec <- recipes::recipe(ReqAmt ~ IncidentDuration, data = req_train_data)
+req_SLR_rec <- recipes::recipe(logReqAmt ~ IncidentDuration, data = req_train_data)
 
@@
 #using the prediction capacity
-req_SLR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
+req_SLR_train %>% yardstick::metrics(truth = logReqAmt, estimate = .pred)
 ```
-This is a terrible model. Time to try adding more predictors.
+This isn't a great model. Time to try adding more predictors.
 
@@
 #define recipe
-req_MVR_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data)
+req_MVR_rec <- recipes::recipe(logReqAmt ~ ., data = req_train_data)
 
@@
 #using the prediction capacity
-req_MVR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
+req_MVR_train %>% yardstick::metrics(truth = logReqAmt, estimate = .pred)
 ```
@@
                       dplyr::group_by(data, model) %>%
-                      yardstick::metrics(truth = ReqAmt, estimate = .pred)
+                      yardstick::metrics(truth = logReqAmt, estimate = .pred)
 
@@
 #define recipe
-obl_SLR_rec <- recipes::recipe(OblAmt ~ IncidentDuration, data = obl_train_data)
+obl_SLR_rec <- recipes::recipe(logOblAmt ~ IncidentDuration, data = obl_train_data)
 
@@
 #using the prediction capacity
-obl_SLR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
+obl_SLR_train %>% yardstick::metrics(truth = logOblAmt, estimate = .pred)
 ```
-This is a terrible model. Time to try adding more predictors.
+This isn't a great model. Time to try adding more predictors.
 
@@
 #define recipe
-obl_MVR_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data)
+obl_MVR_rec <- recipes::recipe(logOblAmt ~ ., data = obl_train_data)
 
@@
 #using the prediction capacity
-obl_MVR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
+obl_MVR_train %>% yardstick::metrics(truth = logOblAmt, estimate = .pred)
 ```
@@
                       dplyr::group_by(data, model) %>%
-                      yardstick::metrics(truth = OblAmt, estimate = .pred)
+                      yardstick::metrics(truth = logOblAmt, estimate = .pred)
 

### 2021-12-13 products/manuscript/Manuscript_Disasters.Rmd
Manuscript Updates - Part 1

@@
 As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments.
-While this has improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.
 
-It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices.
-Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster.
-Modeling approaches that include machine learning algorithms have been demonstrated to have extensive prediction capability for large datasets.
-As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster.
-In particular, a decision tree, bagged tree, and random forest model will be used to predict requested and obligated FEMA funds.
+While these acts have improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.
+Due to the federal government legislative process, these datasets have been publicly available for less than two years, during which the preponderance of the emergency management community was dedicated to managing the COVID-19 pandemic.
+To date, there is one published paper using predictive modeling to examine disaster declarations; however, it is only published in Spanish, and the authors do not have a reliable means of translation [@Perez].
+Based on an approximate translation, the authors used a logistic regression to predict whether or not a presidential declaration would be awarded in the state of Maryland, utilizing Maryland data 2003 - 2017 [@Perez].
+Given this newly available data from the federal government, there are numerous potential predictive models to be considered.
 
@@
 
-## Machine Learning Models
-There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on three different types of tree models.
-This type of model was chosen as the primary users of the prediction algorithms are non-statisticians, and tree models are among the easiest machine learning models to interpret and understand [@james_introduction_2021].
-
-* Decision Tree (DT): Also known as a classification and regression tree (CART), the DT examines each predictor and splits the outcome at a value of the predictor that leads to the best performance increase in the model.
-The tree development continues until a certain threshold criteria is met, such as number of observations in each leaf of the tree.
-A DT model is often the most intuitive machine learning algorithm, but it also typically has a reduced performance compared to other models [@boehmke_hands-machine_2019].
-
-* Bagged Tree (BT): A slightly more sophisticated model, a BT aims to reduce the variance by re-sampling the existing data (otherwise known as bootstrapping) and builds a tree for each sample.
-The final model then averages all of the individually sampled trees.
-However, the reducing the variance in the model also reduces the ease of interpretation of the model [@dietterich_experimental_nodate].
-
-* Random Forest (RF): A RF model follows similar principles to the BT, but instead of considering all possible predictors at each decision tree split, the split occurs based on the best predictors of a random sample of all predictors.
-This method allows for de-correlation in the trees by avoiding the default nature of decision trees that aim to include as many predictors as possible.
-Each individual tree is then averaged to find the final model.
-RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].
+It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices.
+Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster.
+Modeling approaches that include machine learning algorithms have been demonstrated to have extensive prediction capability for large datasets.
+As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster.
+In particular, a decision tree, bagged tree, and random forest model will be used to predict requested and obligated FEMA funds.
 
@@
 ## Model Development
-Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g.
+Four models were considered for the prediction analysis, and each were fit for both outcomes of interest (e.g.
 requested and obligated funds).
-The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type.
-Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type.
-The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type.
-Incident type was chosen as the stratification variable to address the fact that the frequency of incident types varies across FEMA regions.
+The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model.
+Spatial cross-validation was used for all models, using a five-fold resampling structure.
+The tidymodels framework utilized in the analysis does not support a repreated spatial cross-validation resampling structure, as is standard in typical cross-validation.
+
+## Model Definition
+There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on four different types of models.
+The tree-based models were chosen as the primary users of the prediction algorithms are non-statisticians, and such models are among the easiest machine learning models to interpret and understand [@james_introduction_2021].
+However, to robustly consider the machine learning methodology, two regularization-based methods were also included in the analysis.
+
+* Decision Tree (DT): Also known as a classification and regression tree (CART), the DT examines each predictor and splits the outcome at a value of the predictor that leads to the best performance increase in the model.
+The tree development continues until a certain threshold criteria is met, such as number of observations in each leaf of the tree.
+A DT model is often the most intuitive machine learning algorithm, but it also typically has a reduced performance compared to other models [@boehmke_hands-machine_2019].
+
+* Random Forest (RF): A slightly more sophisticated model, a RF model aims to reduce the variance by building a tree for each re-sampling of the existing data.
+However, instead of considering all possible predictors at each decision tree split, the split occurs based on the best predictors of a random sample of all predictors.
+This method allows for de-correlation in the trees by avoiding the default nature of decision trees that aim to include as many predictors as possible.
+Each individual tree is then averaged to find the final model.
+RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].
+
+* Elastic Net (EN):
+
+* LASSO:
 

### 2021-12-13 code/analysis_code/analysis-ML-RA-outlier.Rmd
Bookdown Issues - Permission Denied?

@@
 #save combined table
-model_comp_file = here("results", "ML-model-comp.Rds")
+model_comp_file = here("results", "ML-model-comp-outlier.Rds")
 saveRDS(lasso_RMSE, file = model_comp_file)

### 2021-12-13 products/manuscript/Manuscript_Disasters.Rmd
Bookdown Issues - Permission Denied?

@@
 On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T.
-Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government [@moss_stafford_2009].
+Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government [@mossStaffordActPriorities2009].
 Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government.
-That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither [@mccarthy_femas_2009].
+That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither [@mccarthyFEMADisasterDeclaration2009].
 FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type.
@@
 However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request.
-The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019 [@fema_guide_2021].
-These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration.
-As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments.
-
-While these acts have improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.
-Due to the federal government legislative process, these datasets have been publicly available for less than two years, during which the preponderance of the emergency management community was dedicated to managing the COVID-19 pandemic.
-To date, there is one published paper using predictive modeling to examine disaster declarations; however, it is only published in Spanish, and the authors do not have a reliable means of translation [@Perez].
-Based on an approximate translation, the authors used a logistic regression to predict whether or not a presidential declaration would be awarded in the state of Maryland, utilizing Maryland data 2003 - 2017 [@Perez].
-Given this newly available data from the federal government, there are numerous potential predictive models to be considered.
+This process, explained in greater detail below, involves numerous parameters that are not always obvious to non-FEMA representatives.
+As such, this analysis endeavors to discern the key predictors for receiving federal funding relief under a disaster declaration.
 
@@
 The local officials must have enacted their own emergency operations plans and demonstrate the need for more resources than the local entities can provide.
-Once this threshold is crossed, the municipal or county leader submits a formal request to the state governor for assistance by requesting a state disaster declaration [@mccarthy_femas_2009].
+Once this threshold is crossed, the municipal or county leader submits a formal request to the state governor for assistance by requesting a state disaster declaration [@mccarthyFEMADisasterDeclaration2009].
 If the governor approves such a request, state resources are mobilized to assist the affected area, but the local government is still the primary leading entity for the emergency management response.
 If the state has fully enacted  their emergency response plans and further assistance is still required, the governor requests the state emergency management agency to conduct a state-level Preliminary Damage Assessment (PDA).
-These assessments aim to estimate the impact of the disaster on individuals, infrastructure, and government agencies [@moss_stafford_2009].
+These assessments aim to estimate the impact of the disaster on individuals, infrastructure, and government agencies [@mossStaffordActPriorities2009].
 As federal resources are extraordinarily expensive, state governors utilize PDAs to ensure federal assistance is only requested when absolutely necessary.
 If the governor does decide that federal resources are required, a formal request is submitted by the governor's office through the regional FEMA office.
-This request includes the PDA, barring a few exceptional circumstances for catastrophic events, and a list of requested federal resources, including the anticipated duration and cost of each resource [@moss_stafford_2009].
-The regional FEMA office ensures the request is complete before sending it to a special declaration processing unit in FEMA headquarters, which further add to the request by incorporating detailed statistics about the region and their own assessment of needs and costs [@mccarthy_femas_2009].
+This request includes the PDA, barring a few exceptional circumstances for catastrophic events, and a list of requested federal resources, including the anticipated duration and cost of each resource [@mossStaffordActPriorities2009].
+The regional FEMA office ensures the request is complete before sending it to a special declaration processing unit in FEMA headquarters, which further add to the request by incorporating detailed statistics about the region and their own assessment of needs and costs [@mccarthyFEMADisasterDeclaration2009].
 The declaration processing unit then submits the request to the FEMA Administrator, who must make a recommendation for the President.
 The request and the recommendation are then submitted to the President, who ultimately decides whether or not to grant the declaration request.
-The President is not bound by the FEMA Administrator's recommendation, and, in the event the President denies the request, the governor is permitted to submit an appeal based on further damage assessments [@moss_stafford_2009].
+The President is not bound by the FEMA Administrator's recommendation, and, in the event the President denies the request, the governor is permitted to submit an appeal based on further damage assessments [@mossStaffordActPriorities2009].
 If the request is approved, FEMA becomes the primary agency to coordinate resource allocation and support the local response.
-The approval from the President also specifies which programs are authorized as well as the federal/state cost-sharing ratio [@fema_guide_2021].
+The approval from the President also specifies which programs are authorized as well as the federal/state cost-sharing ratio [@femaGuideDisasterDeclaration2021].
+
+## Newly Available Data
+The federal government funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019 [@femaGuideDisasterDeclaration2021].
+These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration.
+As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments.
+While these acts have improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.
+Due to the federal government legislative process, these datasets have been publicly available for less than two years, during which the preponderance of the emergency management community was dedicated to managing the COVID-19 pandemic.
+To date, there is one published paper using predictive modeling to examine disaster declarations; however, it is only published in Spanish, and the authors do not have a reliable means of translation [@araujoperezPredictiveModelsDisaster2019].
+Based on an approximate translation, the authors used a logistic regression to predict whether or not a presidential declaration would be awarded in the state of Maryland, utilizing Maryland data 2003 - 2017 [@araujoperezPredictiveModelsDisaster2019].
+Given this newly available data from the federal government, there are numerous potential predictive models to be considered.
 
@@
 ## Data and Processing
-The data were obtained from the OpenFEMA Data Sets website [@noauthor_openfema_nodate] as well as the US Department of Agriculture Economic Research Service website [@noauthor_usda_nodate].
+The data were obtained from the OpenFEMA Data Sets website [@OpenFEMADataSets] as well as the US Department of Agriculture Economic Research Service website [@USDAERSCountylevel].
 The mission assignments data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory.
-Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.
+Summary variables were created for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.
 Similarly, the disaster declaration summaries were also summarized to represent the unique disaster declarations for each state, bounded to the same time frame as the mission assignments data.
@@
 The mission assignment data and declaration data were then joined to create a comprehensive dataset that was used for analysis.
+At this point in the analysis, Google developer data [@googledeveloperStatesCsvDataset] was added to specify coordinates of each date, and the outcomes of interest were subset to include only circumstances in which the federal government provided funding (e.g.
+only positive requested or obligated amounts).
 
-Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed.
-Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category.
-Additionally, the relief provided to Puerto Rico in the aftermath of Hurricane Maria in 2017 was removed as an outlier, as it was more than $2 billion more than the next largest allocation of funding.
+Prior to fitting the models, the data were subset based on outcome of interest, and the least frequent disaster types were added to the "Other" category, due to the limited values in certain categories.
+The outcomes of interest were also log transformed to better represent the distribution of disaster relief funding.
+At this point, three outliers were removed from the analysis, as each had a negative value of the log transform of the outcomes of interest.
+
+## Parameters Considered
+Ultimately, the following parameters were considered in the analysis: state, longitude/latitude of state, state population, state number of counties, the FEMA Region to which the state belonged, declaration type, month and year of the first date of the event, duration of the incident, duration of the response efforts, when the Individual/Household, Public Assistance, and Hazard Mitigation Programs were awarded to any county in the state as well as the percent of counties receiving each program, number of federal agencies involved in the response, and the average FEMA cost share for the resources provided to the state under the declaration.
 
@@
 ## Model Definition
-There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on four different types of models.
-The tree-based models were chosen as the primary users of the prediction algorithms are non-statisticians, and such models are among the easiest machine learning models to interpret and understand [@james_introduction_2021].
+There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on two different types of models: tree-based and regularization-based.
+The tree-based models were chosen as the primary users of the prediction algorithms are non-statisticians, and such models are among the easiest machine learning models to interpret and understand [@jamesIntroductionStatisticalLearning2021].
 However, to robustly consider the machine learning methodology, two regularization-based methods were also included in the analysis.
+Broadly, regularization allows variables to be included in a model with less weight than other variables by reducing the value of the variable's coefficient [@hastieElementsStatisticalLearning2016].
 
@@
 The tree development continues until a certain threshold criteria is met, such as number of observations in each leaf of the tree.
-A DT model is often the most intuitive machine learning algorithm, but it also typically has a reduced performance compared to other models [@boehmke_hands-machine_2019].
+A DT model is often the most intuitive machine learning algorithm, but it also typically has a reduced performance compared to other models [@boehmkeHandsOnMachineLearning2019].
 
@@
 Each individual tree is then averaged to find the final model.
-RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].
+RF models are also more difficult to interpret than standard decision tree models [@breimanRandomForests2001].
 
-* Elastic Net (EN):
+* Least Absolute Shrinkage and Selection Operator (LASSO): LASSO balances the goodness of fit with a penalty for the coefficients, calculated by the respective absolute value.
+This methodology allows coefficients to go to zero and thus be dropped from the model.
+The disadvantage of the LASSO model comes predominantly in the form of selection of just one variable from highly correlated variables, ignoring the rest entirely [@rothGeneralizedLASSO2004].
 
-* LASSO:
+* Elastic Net (EN): Of particular use in circumstances with correlations between predictor variables, an EN model further adds to a LASSO model by incorporating a mixture parameter in addition to the overall weight given of the penalty.
+The mixture parameter determines the distribution of the penalty, which is the sum of the LASSO regularization component as well as a new regularization component, which is calculated by squaring the magnitude of the coefficient, and then all coefficients are shrunk by the same factor [@zouRegularizationVariableSelection2005].
 
@@
 ## Software
-This analysis was conducted in Rstudio 4.1 on a Windows Server operating system [@r_core_team_r_2021].
-The following R packages were utilized: here, tidyverse, skimr, lubridate, tidymodels, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.
+This analysis was conducted in R 1.4, using Rstudio 4.1 on a Windows Server operating system [@rcoreteamLanguageEnvironmentStatistical2021].
+The following R packages were utilized: here [@here], tidyverse [@tidyverse], tidymodels [@tidymodels], skimr [@skimr], spatialsample [@spatialsample], broom.mixed [@broommixed], rpart.plot [@rpartplot], vip [@vip], doParallel [@doParallel], ranger [@ranger], viridis [@viridis], maps [@maps], table1 [@table1], cowplot [@cowplot], scales [@scales], gtsummary [@gtsummary], car [@car], ggplot2 [@ggplot2], and summarytools [@summarytools].
 All processing and analysis code can be found in the supplementary materials.
@@
 # Results
-A total of 446 disaster declarations were included in this analysis, ranging from February 2010 to November 2021.
-On average, \$29,700,000 (SD = \$219,000,000) were requested and obligated in FEMA funding, with requests ranging from \-$855,000 to $4,390,000,000.
-All states and territories were represented in the analysis, and on average, the FEMA cost-share was 80.6% (SD = 35.4%).
-The year 2020 had more than double the average number of disaster declarations, which likely can be attributed to the COVID-19 pandemic.
-Further information about the distribution of the data can be found in the supplementary documentation.
 
-Three models were fit to predict requested and obligated FEMA funding in the event of a disaster declaration: decision tree, bagged tree, and random forest.
+## Outcomes of Interest
+A total 3266 disaster declarations were included in this analysis, ranging from February 2012 to November 2021.
+On average, \$40,600,000 (SD = \$255,000,000) were requested in FEMA funding, and \$40,700,000 (SD = \$256,000,000) were obligated in FEMA funding.
+A total 51 states and territories were represented in the analysis, and on average, the FEMA cost-share was 96.4% (SD = 5.73%).
+Figure \@ref(fig:funds-density) shows the almost identical density distribution of the two outcomes of interest, which is further confirmed by the correlation plot between the two shown in figure \@ref(fig:funds-comp).
+
+```{r funds-density,  fig.cap='Density distribution of outcomes of interest', echo=FALSE}
+knitr::include_graphics("../../results/funds_density.png")
+```
+
+```{r funds-comp,  fig.cap='Requested vs obligated FEMA funds provided to states or territories 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/funds_comp.png")
+```
+
+Figures \@ref(fig:map3) and \@ref(fig:map4) display the total amount of FEMA funding requested and obligated respectively to each state during the study period.
+Figure \@ref(fig:map2) further elucidates the role of states, as it shows the total number of declarations each state received during the study period.
+
+```{r map3,  fig.cap='FEMA funds requested by mainland US states 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/map3.png")
+```
+
+```{r map4,  fig.cap='FEMA funds obligated to mainland US states 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/map4.png")
+```
+
+```{r map2,  fig.cap='Total number of federal disaster declarations to mainland US states 2012 - 2021', echo=FALSE}
+knitr::include_graphics("../../results/map2.png")
+```
+
+Further information about the distribution of the covariates considered in this analysis can be found in the supplementary documentation.
+
+## Machine Learning Models
+
+Four models were fit to predict requested and obligated FEMA funding in the event of a disaster declaration: decision tree, random forest, LASSO, and elastic net.
+
 Table \@ref(tab:ML-model-comp) shows the model comparison for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding.
-In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.
+In the requested funding outcome, the random forest model had the lowest RMSE, while the obligated funding outcome model identified the decision tree as the model with the lowest RMSE.
+However, as the difference between the two metrics was <0.05, for the purposes of interpretation, we will choose the random forest model as the most appropriate for both outcomes.
 
 ```{r ML-model-comp,  echo=FALSE, message = FALSE}
-resulttable=readRDS("../../results/ML-model-comp.rds")
+resulttable=readRDS("../../results/ML-model-comp-outlier.rds")
 knitr::kable(resulttable, caption = 'ML model comparison for requested FEMA funding')
@@
 ```{r ML-model-comp-obl,  echo=FALSE}
-resulttable1=readRDS("../../results/ML-model-comp-obl.rds")
+resulttable1=readRDS("../../results/ML-model-comp-OA-outlier.rds")
 knitr::kable(resulttable1, caption = 'ML model comparison for obligated FEMA funding')
@@
 
-The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample.
-Figure \@ref(fig:BT-plot-outlier) shows the identified most important variables in the bagged tree model for requested FEMA funds, and Figure \@ref(fig:BT-plot-obl-outlier) illustrates the identical principle but for obligated FEMA funds.
-In both models, total number of federal agencies involved, incident type, incident duration, state population, awardance of the individual / household assistance program, and FEMA region were identified as the most important predictors of FEMA funding.
+Figure \@ref(fig:RF-plot-outlier) shows the identified most important variables in the random forest model for requested FEMA funds, and Figure \@ref(fig:RF-plot-OA-outlier) illustrates the identical principle but for obligated FEMA funds.
+In both models, percent of state counties awarded the Individual/Households program, total number of federal agencies involved in the response, percent of state counties awarded the Hazard Mitigation program, the FEMA cost share average for federal resources provided to the state, and whether or not the Individual/Households program was awarded to any part of the state were identified as the five most important predictors for requesting or receving FEMA funding in the aftermath of a disaster declaration.
 
-```{r BT-plot-outlier,  fig.cap='Most important variables to predict requested FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-IV-outlier.png")
+```{r RF-plot-outlier,  fig.cap='Most important variables to predict requested FEMA funds for the random forest fit', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot-outlier.png")
 ```
 
-```{r BT-plot-obl-outlier,  fig.cap='Most important variables to predict obligated FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-IV-obl-outlier.png")
+```{r RF-plot-OA-outlier,  fig.cap='Most important variables to predict obligated FEMA funds for the random forest fit', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot-OA-outlier.png")
 ```
 
-The RMSEs for the testing and training datasets were substantially different, with an RMSE of 45003235 for the training data and 69378994 for the testing data.
-Figure \@ref(fig:BT_pred_act_test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT_pred_act_test-obl) does the same for obligated FEMA funds.
-The supplementary information contains the code for the models as well as all diagnostic plots.
-```{r BT_pred_act_test,  fig.cap='Predicted vs actual requested FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-pred-act-test.png")
+To examine the fit of the model to the training data, the following diagnostic plots were generated.
+Figure \@ref(fig:RF-pred-act-outlier) displays the actual versus predicted values of the log(Requested FEMA Funds), Figure \@ref(fig:RF-resid-outlier) depicts the residuals, and Figure \@ref(fig:RF-pred-resid-outlier) illustrates residuals in comparison to the fitted values of the outcome of interest in the model.
+
+```{r RF-pred-act-outlier,  fig.cap='Actual vs predicted log(requested FEMA funds) for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-outlier.png")
 ```
 
-```{r BT_pred_act_test-obl,  fig.cap='Predicted vs actual obligated FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-pred-act-test-obl.png")
+```{r RF-resid-outlier,  fig.cap='Requested FEMA funds residuals for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-resid-outlier.png")
+```
+
+```{r RF-pred-resid-outlier,  fig.cap='Residuals versus fitted values of the log(requested FEMA funds) for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-resid-outlier.png")
+```
+
+Figures \@ref(fig:RF-pred-act-OA-outlier), \@ref(fig:RF-resid-OA-outlier), and \@ref(fig:RF-pred-resid-OA-outlier) generate the same plots as described above, but for obligated FEMA funds.
+
+```{r RF-pred-act-OA-outlier,  fig.cap='Actual vs predicted log(obligated FEMA funds) for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-OA-outlier.png")
+```
+
+```{r RF-resid-OA-outlier,  fig.cap='Obligated FEMA funds residuals for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-resid-OA-outlier.png")
+```
+
+```{r RF-pred-resid-OA-outlier,  fig.cap='Residuals versus fitted values of the log(obligated FEMA funds) for the random forest model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-resid-OA-outlier.png")
+```
+
+Following selection of the random forest model, it was fitted to the testing data sample.
+The RMSEs for the testing and training datasets were different for both outcomes, as shown by Tables \@ref(tab:test-train-model-comp-outlier) and \@ref(tab:test-train-model-comp-OA-outlier).
+In both models, the RMSE for the testing data was lower than for the training data.
+
+```{r TTmodelcomp,  echo=FALSE, message = FALSE}
+readRDS("../../results/test-train-model-comp-outlier.rds")
+```
+
+```{r TTmodelcompOA,  echo=FALSE, message = FALSE}
+readRDS("../../results/test-train-model-comp-OA-outlier.Rds")
+```
+
+```{r test-train-model-comp-outlier,  echo=FALSE, message = FALSE}
+resulttable=readRDS("../../results/test-train-model-comp-outlier.rds")
+knitr::kable(resulttable, caption = 'Random Forest model comparison for test/train data for requested FEMA funding')
+```
+
+```{r test-train-model-comp-OA-outlier,  echo=FALSE}
+resulttable1=readRDS("../../results/test-train-model-comp-OA-outlier.rds")
+knitr::kable(resulttable1, caption = 'Random Forest model comparison for test/train data for obligated FEMA funding')
+```
+
+Figure \@ref(fig:RF_pred_act_test_outlier) shows the predicted and actual requested funds values, and Figure \@ref(fig:RF_pred_act_test-OA-outlier) does the same for obligated FEMA funds.
+The supplementary information contains the code for the models as well as further diagnostic plots.
+```{r RF_pred_act_test_outlier,  fig.cap='Predicted vs actual requested FEMA funds for the random forest fit', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-test-outlier.png")
+```
+
+```{r RF_pred_act_test-OA-outlier,  fig.cap='Predicted vs actual obligated FEMA funds for the random forest fit', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-test-OA-outlier.png")
 ```
@@
 # Discussion
-In order to identify the most important predictors of requested and obligated FEMA funding in the aftermath of a disaster, three machine learning algorithms were fit to FEMA's disaster declaration data: decision tree, bagged tree, and random forest.
-Of the three models, the bagged tree model demonstrated the best performance with the lowest RMSE, the identified metric of evaluation.
-The bagged tree model identified total number of federal agencies involved, incident type, incident duration, state population, awardance of individual / household assistance program, and FEMA region.
+In order to identify the most important predictors of requested and obligated FEMA funding in the aftermath of a disaster, four machine learning algorithms were fit to FEMA's disaster declaration data: decision tree, random forest, LASSO, and elastic net.
+Of the four models, the random forest model demonstrated the best performance with the lowest RMSE, the identified metric of evaluation.
+The random forest model identified percent of state counties awarded the Individual/Households program, total number of federal agencies involved in the response, percent of state counties awarded the Hazard Mitigation program, the FEMA cost share average for federal resources provided to the state, and whether or not the Individual/Households program was awarded to any part of the state were identified as the five most important predictors.
 These variables are relatively expected, as the each of them plays a critical role in FEMA's decision of how much funding to award a state.
-However, it is interesting the state population is identified as such a critical predictor; it suggests that the number of individuals affected by a disaster plays a crucial role in the President's decision to provide assistance.
+However, it is interesting that the incident month and incident year were identified as more important that other factors, such as average FEMA cost share, incident type, or declaration type.
+It is possible that the year effect may be due to inflation, while the month may be skewed by hazards such as the COVID-19 pandemic or major hurricanes.
+
+The diagnostic plots generated for both the training and testing data fits suggest further attenuation of the model is required to accurately describe funding for presidential declarations.
+The residual plots do suggest a relatively even distribution, but Figures \@ref(fig:RF-pred-resid-outlier) and \@ref(fig:RF-pred-resid-OA-outlier) in particular suggest the random forest model does not comprehensively describe the data.
+There appears to be some underlying factor that is affecting the distribution of residuals against the fitted values of the model.
+As discussed below, further research is warranted to understand this behavior.
 
@@
 
-Ultimately, this analysis is the first of its kind to apply prediction algorithms to federal funding for disaster declarations in the United States.
+Ultimately, this analysis is the first of its kind to apply such prediction algorithms to federal funding for disaster declarations in the United States.
 It demonstrates the clear potential of such modeling capabilities not only to allow local and state emergency managers to predict which aspects of a PDA are critical in obtaining financial federal relief, but also to allow the regional FEMA offices to identify such predictors while developing the formal requests.

### 2021-12-13 products/manuscript/Manuscript_Supplemental.Rmd
Bookdown Issues - Permission Denied?

@@
 This document represents additional information about the data and results.
+The full code and detailed explanation of the figures can be found at the [GitHub repository](https://github.com/morganataylor/MorganTaylor-MADA-project).
 
 ## Exploratory analysis
-The processed data were then explored to characterize and summarize the relevant features in RStudio 4.1 with the following packages: here, tidyverse, skimr, summarytools, ggplot2, car, gtsummary, scales, cowplot, and table1.
-The supplementary files contain the code that generates the following results.
+The processed data were explored to characterize and summarize the relevant features, which can be found below.
 
@@
 readRDS("../../results/table1.rds")
+```
+
+Figure \@ref(fig:map1) shows the FEMA Regions by state.
+```{r map1,  fig.cap='FEMA Regions', echo=FALSE, message = FALSE}
+knitr::include_graphics("../../results/map1.png")
 ```
@@
 ```{r ML-model-comp-obl,  echo=FALSE}
-resulttable1=readRDS("../../results/ML-model-comp-obl.rds")
+resulttable1=readRDS("../../results/ML-model-comp-OA.rds")
 knitr::kable(resulttable1, caption = 'ML model comparison for obligated FEMA funding')

### 2021-12-13 products/manuscript/Manuscript_Disasters.Rmd
Fixed Bookdown Issue - Windows Server OS Issue

@@
 
-Figure \@ref(fig:RF_pred_act_test_outlier) shows the predicted and actual requested funds values, and Figure \@ref(fig:RF_pred_act_test-OA-outlier) does the same for obligated FEMA funds.
+Figure \@ref(fig:RF-pred-act-test-outlier) shows the predicted and actual requested funds values, and Figure \@ref(fig:RF-pred-act-test-OA-outlier) does the same for obligated FEMA funds.
 The supplementary information contains the code for the models as well as further diagnostic plots.
-```{r RF_pred_act_test_outlier,  fig.cap='Predicted vs actual requested FEMA funds for the random forest fit', echo=FALSE}
+```{r RF-pred-act-test-outlier,  fig.cap='Predicted vs actual requested FEMA funds for the random forest fit', echo=FALSE}
 knitr::include_graphics("../../results/RF-pred-act-test-outlier.png")
@@
 
-```{r RF_pred_act_test-OA-outlier,  fig.cap='Predicted vs actual obligated FEMA funds for the random forest fit', echo=FALSE}
+```{r RF-pred-act-test-OA-outlier,  fig.cap='Predicted vs actual obligated FEMA funds for the random forest fit', echo=FALSE}
 knitr::include_graphics("../../results/RF-pred-act-test-OA-outlier.png")

### 2021-12-13 products/manuscript/Manuscript_Supplemental.Rmd
Fixed Bookdown Issue - Windows Server OS Issue

@@
 
+## Modeling Analysis
+
 Prior to fitting the data to machine learning models, standard simple and multivariable regression models were fitted using the tidymodels framework in RStudio 4.1.
@@
 Figure \@ref(fig:req_MVR_bp) and \@ref(fig:obl_MVR_bp) show the significant variables fitted in the multivariable regression models for requested and obligated FEMA funds, respectively.
+There are simply too many variables to be able to report all variables on one figure.
 
@@
 
-## Full analysis
+## Machine Learning Models
 
-Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g.
-requested and obligated funds).
-Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed.
-Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category.
-The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type.
-Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type.
-The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type.
-The models were then compared based on the root mean square error (RMSE), the standard metric for regression models.
-The analysis was completed in RStudio 4.1 with the following packages: here, tidyverse, tidymodels, skmir, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.
+The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot-outlier) and  \@ref(fig:DT-plot-OA-outlier).
 
-The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot) and  \@ref(fig:DT-plot-obl).
-Both outcomes identified the same decision point features: total number of federal agencies involved in the response, duration of the incident, state population, state counties, and whether the individual / housing assistance program was awarded.
-
-```{r DT-plot,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/DT-plot.png")
+```{r DT-plot-outlier,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/DT-plot-outlier.png")
 ```
 
-```{r DT-plot-obl,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/DT-plot-obl")
+```{r DT-plot-OA-outlier,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/DT-plot-OA-outlier.png")
 ```
 
-The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot) and  \@ref(fig:RF-plot-obl).
-Both outcomes identified the same important features: incident duration, total number of agencies involved, individual/housing assistance program awardance, and incident month.
-Interestingly, however, the obligated funds model includes days of the entire response as the fifth most important feature, whereas the requested funds model includes it as the eleventh most important feature.
-Unfortunately, the tidymodels framework does not support visualization of the bagged tree model.
+The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot-outlier) and  \@ref(fig:RF-plot-OA-outlier).
 
-```{r RF-plot,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/RF-plot.png")
+```{r RF-plot-outlier,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot-outlier.png")
 ```
 
-```{r RF-plot-obl,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
-knitr::include_graphics("../../results/RF-plot-obl")
+```{r RF-plot-OA-outlier,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/RF-plot-OA-outlier.png")
+```
+
+The most important variables in the identified elastic net model for each outcome is illustrated by Figures  \@ref(fig:EN-plot-outlier) and  \@ref(fig:EN-plot-OA-outlier).
+
+```{r EN-plot-outlier,  fig.cap='20 most important variables for the elastic net model for requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/EN-plot-outlier.png")
+```
+
+```{r EN-plot-OA-outlier,  fig.cap='20 most important variables for the elastic net model for obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/EN-plot-OA-outlier.png")
+```
+
+The most important variables in the identified LASSO model for each outcome is illustrated by Figures  \@ref(fig:LASSO-plot-outlier) and  \@ref(fig:LASSO-plot-OA-outlier).
+
+```{r LASSO-plot-outlier,  fig.cap='20 most important variables for the LASSO model for requested FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/LASSO-plot-outlier.png")
+```
+
+```{r LASSO-plot-OA-outlier,  fig.cap='20 most important variables for the LASSO model for obligated FEMA funds', echo=FALSE}
+knitr::include_graphics("../../results/LASSO-plot-OA-outlier.png")
 ```
@@
 Table \@ref(tab:ML-model-comp) shows the model comparisons for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding.
-In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.
 
@@
 
-The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample.
-The RMSEs for the testing and training datasets were substantially different.
-Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds.
-The supplementary information contains the code for the models as well as all diagnostic plots.
-```{r BT_pred_act_test,  fig.cap='Predicted vs actual requested FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-pred-act-test.png")
+Once the random forest was fit to the test data, the following diagnostic plots were created for each outcome.
+
+For Requested FEMA Funding:
+```{r RF-pred-act-test-outlier,  fig.cap='Actual vs predicted log(requested FEMA funds) for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-test-outlier.png")
 ```
 
-```{r BT_pred_act_test-obl,  fig.cap='Predicted vs actual obligated FEMA funds for the bagged tree fit', echo=FALSE}
-knitr::include_graphics("../../results/BT-pred-act-test-obl.png")
+```{r RF-resid-test-outlier,  fig.cap='Requested FEMA funds residuals for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-resid-test-outlier.png")
 ```
+
+```{r RF-pred-resid-test-outlier,  fig.cap='Residuals versus fitted values of the log(requested FEMA funds) for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-resid-test-outlier.png")
+```
+
+For Obligated FEMA Funding:
+
+```{r RF-pred-act-test-OA-outlier,  fig.cap='Actual vs predicted log(obligated FEMA funds) for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-act-test-outlier.png")
+```
+
+```{r RF-resid-test-OA-outlier,  fig.cap='Obligated FEMA funds residuals for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-resid-test-OA-outlier.png")
+```
+
+```{r RF-pred-resid-test-OA-outlier,  fig.cap='Residuals versus fitted values of the log(obligated FEMA funds) for the random forest test model', echo=FALSE}
+knitr::include_graphics("../../results/RF-pred-resid-test-OA-outlier.png")
+```
